### 2012-10-05 notes.tex
First draft of variational derivation of model with common M0 across sites.

\documentclass[10pt, letterpaper]{article}

\usepackage{amsmath} % just math
\usepackage{amssymb} % allow blackboard bold (aka N,R,Q sets)
\linespread{1.6}  % double spaces lines
\usepackage[left=1in,top=1in,right=1in,bottom=1in,nohead]{geometry}

\newcommand{\E}{\mathbb{E}}

\begin{document}

\linespread{1} % single spaces lines
\small \normalsize %% dumb, but have to do this for the prev to work
\begin{flushright}
RVD2 Project Notes \\
\today
\end{flushright}

Consider the following model
\begin{align}
\mu_j &\thicksim \text{Gaussian}(u_0,\sigma^2_0)\\
\theta_{kj} &\thicksim \text{Beta}(\mu_j,M_0)\\
r_{kj} &\thicksim \text{Bino}(\theta_{kj}, n_{kj})
\end{align}

The inferential object of interest is the joint posterior distribution over the latent variables, $p(\mu,\theta|r)$. From this we can compute the marginal posterior distribution over the site-specific error rate - the null hypothesis distribution.
\begin{equation}
p(\mu|r) = \int_\theta p(\mu,\theta|r)d\theta
\end{equation}

Writing out the joint posterior distribution gives
\begin{equation}
\begin{split}
\log p(\mu,\theta | r)& = \log \frac{p(\mu,\theta,r;u_o,\sigma^2_0,M_0)}{p(r | u_o,\sigma^2_0,M_0)}\\
& = \log p(\mu,\theta,r;u_o,\sigma^2_0,M_0) - \log p(r ; u_o, \sigma^2_0,M_0)\\
& = \log \prod_{j=1}^J \prod_{k=1}^K p(\mu_j,\theta_{kj}, r_{kj}) - \log \prod_{j=1}^J \prod_{k=1}^K \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j,\theta_{kj}, r_{kj})d\mu_jd\theta_{kj}
\end{split}
\end{equation}

\begin{multline}
\log p(\mu,\theta | r) = K\sum_{j=1}^J \log p(\mu_j;\mu_0,\sigma^2_0) + \sum_{j=1}^J \sum_{k=1}^K \log p(\theta_{kj} | \mu_j;M_0)\\ + \sum_{j=1}^J \sum_{k=1}^K\log p(r_{kj} | \theta_{kj}, n_{kj})
- \sum_{j=1}^J \sum_{k=1}^K \log \int_{\theta_{kj}} \int_{\mu_j} p(r_{kj};\mu_0,\sigma^2_0,M_0) \end{multline}

The computation of $p(r;u_0,\sigma_0^2,M_0)$ is intractable due to the coupling in the latent variables induced by the integral so we must use an approximation or sampling method.

{\bf Variational approximation}
\begin{align}
\log p(r_{kj}; u_0, \sigma_0^2, M_0) & = \log \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j, \theta_{kj}, r_{kj})d\mu_jd\theta_{kj} \\
& = \log \int_{\theta_{kj}} \int_{\mu_j} p(\mu_j, \theta_{kj}, r_{kj}) \frac{q(\mu_j, \theta_{kj})}{q(\mu_j, \theta_{kj})}  d\mu_jd\theta_{kj}\\
& \geq  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log p(\mu_j, \theta_{kj}, r_{kj}) -  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log q(\mu_j, \theta_{kj}) \\
& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L}
\end{align}

The bound on the log likelihood of the data, $\mathcal{L}$, is seen to be the sum of the expected complete data log-likelihood and the entropy of the variational approximating distribution.

We now specify an averaging(variational) distribution
\begin{align}
q(\mu_j | \gamma_{1j}, \gamma_{2j}) & \thicksim \text{Gaussian}(\gamma_{1j}, \gamma_{2j})\\
q(\theta_{kj} | \alpha, \beta ) & \thicksim \text{Beta}(\alpha, \beta)
\end{align}

We are now able to decompose $\mathcal{L}$ into its constituent parts
\begin{equation}\label{eqn:composite}
\mathcal{L} = \E_q\log p(\mu_j|u_0,\sigma_0^2) + \E_q \log p(\theta_{kj} | \mu_j; M_0) 
+ \E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) - \E_q \log q(\mu) - \E_q \log q(\theta_{kj})
\end{equation}

Taking each term in \eqref{eqn:composite} individually we can cast the bound on the log-likelihood in terms of original and variational parameters only.

\begin{align}
\E_q \log p(\mu_j ; u_0, \sigma_0^2) & = -\frac{1}{2}\log (2\pi\sigma_0^2) -\frac{1}{2}\left[ \gamma_{2} + (\gamma_1 - u_0)^2 \right]
\end{align}

\begin{align}
\E_q \log p(\theta | \mu ; M_0) & = \int_\mu q(\mu) \int_\theta q(\theta) \log p(\theta | \mu, M_0)\\
\begin{split}
& = \log \Gamma(M_0) - \E_q \log \Gamma(\mu M_0) - \E_q \log \Gamma((1-\mu)M_0) \\
& \qquad+(u_0M_0-1)\left( \psi(\alpha) - \psi(\alpha + \beta) \right) + ((1-u_0)M_0-1)\left( \psi(\beta) - \psi(\alpha + \beta) \right)
\end{split}
\end{align}
where $\E_q \log \Gamma(\mu M_0)$ is easily computed numerically. When we get to differentiating with respect to the parameters, we can move that inside the integral. The kernel changes, but it can still be computed numerically.

\begin{align}
\E_q \log p(r | \theta, n) & = \int_\theta \\
\begin{split}
& = \log \Gamma(n+1) - \log \Gamma(r+1) - \log \Gamma(n-r+1) \\
& \qquad +r(\psi(\alpha)-\psi(\alpha + \beta)) +(n-r)(\psi(\beta) - \psi(\alpha+\beta))
\end{split}
\end{align}

\begin{align}
\E_q \log p(r | \theta, n) & = \int_\theta \\
\begin{split}
& = \log \Gamma(n+1) - \log \Gamma(r+1) - \log \Gamma(n-r+1) \\
& \qquad +r(\psi(\alpha)-\psi(\alpha + \beta)) +(n-r)(\psi(\beta) - \psi(\alpha+\beta))
\end{split}
\end{align}

\begin{equation}
\E_q \log q(\mu)  = \text{Negative Gaussian Entropy}
\end{equation}

\begin{equation}
\E_q \log q(\theta)  = \text{Negative Beta Entropy}
\end{equation}

Now that \eqref{eqn:composite} has been written in terms of only the variational and model parameters, we can numerically optimize the system with respect to variational and the model parameters and do coordinate ascent. Maximizing with respect to the variational parameters tightens the bound on the data log-likelihood and maximizing with respect to the model parameters maximizes the likelihood.
\end{document}

### 2012-10-08 notes.tex
Partial code for LL bound for common M0.

@@
 & \geq  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log p(\mu_j, \theta_{kj}, r_{kj}) -  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log q(\mu_j, \theta_{kj}) \\
-& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L}
+& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L_{kj}}
 \end{align}
@@
 q(\mu_j | \gamma_{1j}, \gamma_{2j}) & \thicksim \text{Gaussian}(\gamma_{1j}, \gamma_{2j})\\
-q(\theta_{kj} | \alpha, \beta ) & \thicksim \text{Beta}(\alpha, \beta)
+q(\theta_{kj} | \alpha_{kj}, \beta_{kj} ) & \thicksim \text{Beta}(\alpha_{kj}, \beta_{kj})
 \end{align}
@@
 We are now able to decompose $\mathcal{L}$ into its constituent parts
-\begin{equation}\label{eqn:composite}
-\mathcal{L} = \E_q\log p(\mu_j|u_0,\sigma_0^2) + \E_q \log p(\theta_{kj} | \mu_j; M_0) 
-+ \E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) - \E_q \log q(\mu) - \E_q \log q(\theta_{kj})
-\end{equation}
+\begin{multline}\label{eqn:composite}
+\mathcal{L} = K\sum_{j=1}^J \E_q\log p(\mu_j|u_0,\sigma_0^2) + \sum_{j=1}^J\sum_{k=1}^K\E_q \log p(\theta_{kj} | \mu_j; M_0) 
++ \sum_{j=1}^J\sum_{k=1}^K\E_q \log p(r_{kj} | \theta_{kj}, n_{kj})\\
+- K\sum_{j=1}^J\E_q \log q(\mu) - \sum_{j=1}^J\sum_{k=1}^K \E_q \log q(\theta_{kj})
+\end{multline}
 
@@
 \begin{align}
-\E_q \log p(\mu_j ; u_0, \sigma_0^2) & = -\frac{1}{2}\log (2\pi\sigma_0^2) -\frac{1}{2}\left[ \gamma_{2} + (\gamma_1 - u_0)^2 \right]
+\E_q \log p(\mu_j ; u_0, \sigma_0^2) & = -\frac{1}{2}\log (2\pi\sigma_0^2) -\frac{1}{2\sigma_0^2}\left[ \gamma_{2j} + (\gamma_{1j} - u_0)^2 \right]
 \end{align}
@@
 \begin{align}
-\E_q \log p(\theta | \mu ; M_0) & = \int_\mu q(\mu) \int_\theta q(\theta) \log p(\theta | \mu, M_0)\\
+\E_q \log p(\theta_{kj} | \mu_j ; M_0) & = \int_{\mu_j} q(\mu_j) \int_{\theta_{kj}} q(\theta_{kj}) \log p(\theta_{kj} | \mu_j, M_0)\\
 \begin{split}
-& = \log \Gamma(M_0) - \E_q \log \Gamma(\mu M_0) - \E_q \log \Gamma((1-\mu)M_0) \\
-& \qquad+(u_0M_0-1)\left( \psi(\alpha) - \psi(\alpha + \beta) \right) + ((1-u_0)M_0-1)\left( \psi(\beta) - \psi(\alpha + \beta) \right)
+& = \log \Gamma(M_0) - \E_q \log \Gamma(\mu_j M_0) - \E_q \log \Gamma((1-\mu_j)M_0) \\
+& \qquad+(u_0M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) + ((1-u_0)M_0-1)\left( \psi(\beta_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right)
 \end{split}
 \end{align}
-where $\E_q \log \Gamma(\mu M_0)$ is easily computed numerically.
+where $\E_q \log \Gamma(\mu_j M_0)$ is easily computed numerically.
 When we get to differentiating with respect to the parameters, we can move that inside the integral.
@@
 \begin{align}
-\E_q \log p(r | \theta, n) & = \int_\theta \\
+\E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) & = \int_{\theta_{kj}} \\
 \begin{split}
-& = \log \Gamma(n+1) - \log \Gamma(r+1) - \log \Gamma(n-r+1) \\
-& \qquad +r(\psi(\alpha)-\psi(\alpha + \beta)) +(n-r)(\psi(\beta) - \psi(\alpha+\beta))
-\end{split}
-\end{align}
-
-\begin{align}
-\E_q \log p(r | \theta, n) & = \int_\theta \\
-\begin{split}
-& = \log \Gamma(n+1) - \log \Gamma(r+1) - \log \Gamma(n-r+1) \\
-& \qquad +r(\psi(\alpha)-\psi(\alpha + \beta)) +(n-r)(\psi(\beta) - \psi(\alpha+\beta))
+& = \log \Gamma(n_{kj}+1) - \log \Gamma(r_{kj}+1) - \log \Gamma(n_{kj}-r_{kj}+1) \\
+& \qquad +r(\psi(\alpha_{kj})-\psi(\alpha_{kj} + \beta_{kj})) +(n_{kj}-r_{kj})(\psi(\beta_{kj}) - \psi(\alpha_{kj}+\beta_{kj}))
 \end{split}
@@
 \begin{equation}
-\E_q \log q(\mu)  = \text{Negative Gaussian Entropy}
+\E_q \log q(\mu_j)  = -\frac{1}{2} \log(2\pi e \gamma_{2j})
 \end{equation}
 
-\begin{equation}
-\E_q \log q(\theta)  = \text{Negative Beta Entropy}
-\end{equation}
+\begin{multline}
+\E_q \log q(\theta_{kj})  = \log \Gamma(\alpha_{kj} + \beta_{kj}) - \log \Gamma(\alpha_{kj}) - \log \Gamma(\beta_{kj}) \\+(\alpha_{kj}-1)\psi(\alpha_{kj}) +(\beta_{kj} -1)\psi(\beta_{kj}) -(\alpha_{kj}+\beta_{kj}-2)\psi(\alpha_{kj}+\beta_{kj})
+\end{multline}
 
@@
 Maximizing with respect to the variational parameters tightens the bound on the data log-likelihood and maximizing with respect to the model parameters maximizes the likelihood.
+
+{\bf Update Equations}
+Isolating the terms of \eqref{eqn:composite} that involve $u_0$ gives
+\begin{multline}
+\mathcal{L}_{[u_0]} = K\sum_{j=1}^J\left(-\frac{1}{2}(\gamma_{1j}-u_0)^2\right) \\+ \sum_{j=1}^J \sum_{k=1}^Ku_0M_0\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj}+\beta_{kj}) \right) - \sum_{j=1}^J \sum_{k=1}^Ku_0M_0\left( \psi(\beta_{kj}) - \psi(\alpha_{kj}+\beta_{kj}) \right)
+\end{multline}
+
+Taking the derivative with respect to $u_0$ gives
+\begin{equation}
+\frac{\partial \mathcal{L}}{\partial u_0} = K\sum_{j=1}^J\gamma_{1j} - KJu_0 +M_0 \sum_{j=1}^J \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right)
+\end{equation}
+
+Solving for $u_0$ gives
+\begin{equation}
+u_0 \leftarrow \frac{1}{J}\sum_{j=1}^J \gamma_{1j} +\frac{M_0}{KJ}\sum_{j=1}^J \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right)
+\end{equation}
+
+Isolating the terms of \eqref{eqn:composite} that involve $\sigma^2_0$ gives
+\begin{equation}
+\mathcal{L}_{[\sigma^2_0]} = -\frac{J}{2}\log(2\pi\sigma^2_0) -\frac{1}{2\sigma_0^2}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2 \right)
+\end{equation}
+
+Taking the derivative with respect to $\sigma^2_0$ gives
+\begin{equation}
+\frac{\partial \mathcal{L}}{\partial \sigma^2_0} = -\frac{J}{2\sigma^2_0} +\frac{1}{\sigma^4_0}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2\right)
+\end{equation}
+
+Setting the derivative equal to zero and solving for $\sigma^2_0$ gives
+\begin{equation}
+\sigma^2_0 \leftarrow \frac{1}{J}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2\right)
+\end{equation}
+
 \end{document}

### 2012-10-08 src/dLdM0.m

@@ -1,0 +1,26 @@
+function [ f ] = dLdM0( M0, u0, alpha, beta, gam1, gam2 )
+%DLDM0 Derivative of the bound with respect to M0
+%   Detailed explanation goes here
+[K,J] = size(alpha);
+
+for j = 1:J
+    Eq1(j) = integral(@(x)Eq1kernel(x, M0, gam1(j), gam2(j)), 0, 1);
+end
+Eq1 = sum(Eq1);
+
+for j = 1:J
+    Eq2(j) = integral(@(x)Eq2kernel(x, M0, gam1(j), gam2(j)), 0, 1);
+end
+Eq2 = sum(Eq2);
+
+f = K*J*psi(M0) + u0*sum(sum(psi(alpha))) +(1-u0)*sum(sum(psi(beta)))...
+    -sum(sum(psi(alpha+beta))) -K*Eq1 - K*Eq2;
+
+    function [f] = Eq1kernel(x, M0, gam1, gam2)
+        f = normpdf(gam1,sqrt(gam2)).*x.*psi(x.*M0);
+    end
+    function [f] = Eq2kernel(x, M0, gam1, gam2)
+        f = normpdf(gam1,sqrt(gam2)).*(1-x).*psi((1-x).*M0);
+    end
+end
+

### 2012-10-08 src/ll_bound.m

@@ -1,0 +1,40 @@
+function [ ll ] = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta )
+%LL_BOUND Variational bound on data log-likelihood
+[K,J] = size(n);
+
+EqlogPmu = -0.5*log(2*pi*sigma20) -(1/sigma20)*(gam2 + (gam1 - u0).^2);
+EqlogPmu = repmat(EqlogPmu, [K 1]);
+
+for i = 1:length(gam1)
+    EqlogG1(i) = integral(@(x)logG1(x,gam1(i),gam2(i),M0),0,1);
+end
+for i = 1:length(gam1)
+    EqlogG2(i) = integral(@(x)logG2(x,gam1(i),gam2(i),M0),0,1);
+end
+
+EqlogPtheta = gammaln(M0) -repmat(EqlogG1-EqlogG2,[K 1]) ...
+    +(u0*M0-1)*(psi(alpha) - psi(alpha+beta)) ...
+    +((1-u0)*M0-1)*(psi(beta) - psi(alpha+beta));
+
+EqlogPr = gammaln(n+1) -gammaln(r+1) -gammaln(n-r+1) ...
+    +r.*(psi(alpha)-psi(alpha+beta)) + (n-r).*(psi(beta)-psi(alpha+beta));
+
+EqlogQmu = -0.5*log(2*pi*exp(1)*gam2);
+EqlogQmu = repmat(EqlogQmu, [K 1]);
+
+EqlogQtheta = gammaln(alpha+beta) - gammaln(alpha) - gammaln(beta) ...
+    +(alpha-1).*psi(alpha) +(beta-1).*psi(beta) ...
+    -(alpha+beta-2).*psi(alpha+beta);
+
+ll = EqlogPmu + EqlogPtheta + EqlogPr - EqlogQmu - EqlogQtheta;
+ll = sum(sum(ll));
+
+
+end
+
+function [f] = logG1(x, gam1, gam2, M0)
+f = normpdf(x,gam1,sqrt(gam2)).*gammaln(x*M0);
+end
+function [f] = logG2(x, gam1, gam2, M0)
+f = normpdf(x,gam1,sqrt(gam2)).*gammaln((1-x)*M0);
+end

### 2012-10-08 src/test_ll_bound.m

@@ -1,0 +1,48 @@
+rng(1000,'twister');
+
+K = 3; J = 1000;
+
+u0 = 0.5; sigma20 = 0.1^2;
+
+mu = normrnd(u0,sqrt(sigma20),[1 J]);
+
+M0 = 1000; theta = NaN(K,J);
+for k = 1:K
+    theta(k,:) = betarnd(mu.*M0,M0.*(1-mu)); 
+end
+
+n = repmat(1000,K,J);
+r = binornd(n,theta);
+
+gam1 = repmat(0.5, [1 J]); gam2 = repmat(0.5, [1 J]);
+alpha = repmat(3,[K J]); beta = repmat(2, [K J]);
+
+% Compute the log-likelihood bound
+[ ll ] = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta );
+
+% Solve for u0
+u0 = mean(gam1) + (M0./(K*J)).*sum(sum(psi(alpha) - psi(beta)));
+
+% Solve for sigma20
+
+% Solve for M0
+[ f ] = dLdM0( M0, u0, alpha, beta, gam1, gam2 );
+M0 = fsolve( @(x)dLdM0(x, u0, alpha, beta, gam1, gam2), M0, ...
+    'Display','iter' );
+
+% Plot the LL bound varying M0
+v = logspace(-1 , 3, 10);
+ll_M0 = arrayfun(@(x)ll_bound( r, n, u0, sigma20, x, gam1, gam2, alpha, beta ), v);
+plot(v,ll_M0) 
+
+% Plot the LL bound varying u0
+v = linspace(0 , 1, 100);
+ll_u0 = arrayfun(@(x)ll_bound( r, n, x, sigma20, M0, gam1, gam2, alpha, beta ), v);
+plot(v,ll_u0) 
+
+% Plot the LL bound varying sigma20
+v = logspace(-2 , 1, 100);
+ll_sigma20 = arrayfun(@(x)ll_bound( r, n, u0, x, M0, gam1, gam2, alpha, beta ), v);
+semilogx(v,ll_sigma20) 
+
+

### 2012-10-12 notes.tex
Fix some bugs in code and notes. Mainly testing. u0 update works, M0 update is questionable. Sigma20 update works. alpha and beta updates work. Testing M0.

@@
 & = \log \Gamma(M_0) - \E_q \log \Gamma(\mu_j M_0) - \E_q \log \Gamma((1-\mu_j)M_0) \\
-& \qquad+(u_0M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) + ((1-u_0)M_0-1)\left( \psi(\beta_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right)
+& \qquad+(\gamma_{1j}M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) + ((1-\gamma_{1j})M_0-1)\left( \psi(\beta_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right)
 \end{split}
@@
 Isolating the terms of \eqref{eqn:composite} that involve $u_0$ gives
-\begin{multline}
-\mathcal{L}_{[u_0]} = K\sum_{j=1}^J\left(-\frac{1}{2}(\gamma_{1j}-u_0)^2\right) \\+ \sum_{j=1}^J \sum_{k=1}^Ku_0M_0\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj}+\beta_{kj}) \right) - \sum_{j=1}^J \sum_{k=1}^Ku_0M_0\left( \psi(\beta_{kj}) - \psi(\alpha_{kj}+\beta_{kj}) \right)
-\end{multline}
+\begin{equation}
+\mathcal{L}_{[u_0]} = K\sum_{j=1}^J\left(-\frac{1}{2\sigma_0^2}(\gamma_{1j}-u_0)^2\right)
+\end{equation}
 
 Taking the derivative with respect to $u_0$ gives
-\begin{equation}
-\frac{\partial \mathcal{L}}{\partial u_0} = K\sum_{j=1}^J\gamma_{1j} - KJu_0 +M_0 \sum_{j=1}^J \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right)
-\end{equation}
 
@@
 \begin{equation}
-u_0 \leftarrow \frac{1}{J}\sum_{j=1}^J \gamma_{1j} +\frac{M_0}{KJ}\sum_{j=1}^J \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right)
+u_0 \leftarrow \frac{1}{J}\sum_{j=1}^J \gamma_{1j} 
 \end{equation}
@@
 
+Isolating terms that involve $\alpha_{kj}$ gives
+\begin{multline}
+\mathcal{L}_{[\alpha_{kj}]} = (u_0M_0 -1)(\psi(\alpha_{kj}) - \psi(\alpha_{kj}-\beta_{kj}) -((1-u_0)M_0-1)\psi(\alpha_{kj}+\beta_{kj})\\
+ +r_{kj}*\psi(\alpha_{kj})-n_{kj}\psi(\alpha_{kj}+\beta_{kj}) -\log \Gamma (\alpha+\beta) + \log \Gamma(\alpha) -(\alpha-1)\psi(\alpha)+(\alpha+\beta-2)\psi(\alpha+\beta)
+\end{multline}
+
+Setting the derivative equal to zero and solving for $\alpha_{kj}$ gives
+\begin{equation}
+\alpha_{kj} \leftarrow \psi_1(\alpha_{kj})(u_0M_0+r-\alpha_{kj}) - \psi_1(\alpha_{kj}+\beta_{kj})(M_0+n-(\alpha_{kj}+\beta_{kj}))
+\end{equation}
+
+A similar procedure holds for $\beta_{kj}$ and the resulting update equation bears symmetry to that for $\alpha_{kj}$
+\begin{equation}
+\beta_{kj} \leftarrow \psi_1(\beta_{kj})\left[(1-u_0)M_0-1+(n-r)-\beta_{kj}\right] - \psi_1(\alpha_{kj}+\beta_{kj})\left[M_0+n-(\alpha_{kj}+\beta_{kj})\right]
+\end{equation}
+
 \end{document}

### 2012-10-12 src/dLdM0.m

@@ -3,6 +3,10 @@ function [ f ] = dLdM0( M0, u0, alpha, beta, gam1, gam2 )
 %   Detailed explanation goes here
 [K,J] = size(alpha);
 
+if M0 <= 0 % if M0 is undefined, set it equal to the value of M0=0
+    M0 = eps;
+end
+
 for j = 1:J
     Eq1(j) = integral(@(x)Eq1kernel(x, M0, gam1(j), gam2(j)), 0, 1);
 end
@@ -13,8 +17,8 @@ for j = 1:J
 end
 Eq2 = sum(Eq2);
 
-f = K*J*psi(M0) + u0*sum(sum(psi(alpha))) +(1-u0)*sum(sum(psi(beta)))...
-    -sum(sum(psi(alpha+beta))) -K*Eq1 - K*Eq2;
+f = K*J*psi(M0) + sum(gam1.*sum(psi(alpha)-psi(beta),1)) ...
+    +sum(sum(psi(beta)-psi(alpha+beta))) -K*Eq1 - K*Eq2;
 
     function [f] = Eq1kernel(x, M0, gam1, gam2)
         f = normpdf(gam1,sqrt(gam2)).*x.*psi(x.*M0);

### 2012-10-12 src/dLdalpha.m

@@ -1,0 +1,10 @@
+function [ f ] = dLdalpha( alpha, beta, M0, u0, r, n )
+%DLDALPHA Derivative of the bound with respect to alpha
+%   Detailed explanation goes here
+
+if alpha <=0
+    alpha = eps;
+end
+
+f = psi(1,alpha)*(u0*M0 +r -alpha) -psi(1,alpha+beta)*(M0+n-(alpha+beta));
+end

### 2012-10-12 src/dLdgam1.m

@@ -1,0 +1,31 @@
+function [ f ] = dLdgam1( gam1, gam2, alpha, beta, M0, u0, sigma20)
+%UNTITLED2 Summary of this function goes here
+%   Detailed explanation goes here
+[K,J] = size(alpha);
+
+Eq1 = integral(@Eq1kernel,0,1);
+Eq2 = integral(@Eq2kernel,0,1);
+Eq3 = integral(@Eq3kernel,0,1);
+Eq4 = integral(@Eq4kernel,0,1);
+
+f = -(K/sigma20)*(gam1-u0) -(1/gam2)*Eq1 +(gam1/gam2)*Eq2 ...
+    - (1/gam2)*Eq3 + (gam1/gam2)*Eq4 + M0*sum(psi(alpha)-psi(beta),1);
+
+        
+    function [f] = Eq1kernel(x)
+        f = normpdf(x,gam1,sqrt(gam2)).*x.*gammaln(x*M0);
+    end
+
+    function [f] = Eq2kernel(x)
+        f = normpdf(x,gam1,sqrt(gam2)).*gammaln(x*M0);
+    end
+
+    function [f] = Eq3kernel(x)
+        f = normpdf(x,gam1,sqrt(gam2)).*x.*gammaln((1-x)*M0);
+    end
+
+    function [f] = Eq4kernel(x)
+        f = normpdf(x,gam1,sqrt(gam2)).*gammaln((1-x)*M0);
+    end
+end
+

### 2012-10-12 src/dLdgam2.m

@@ -1,0 +1,23 @@
+function [ f ] = dLdgam2( gam1, gam2, M0, sigma20, J )
+%DLDgam2 Derivative of the bound with respect to gam2
+%   Detailed explanation goes here
+
+if gam2 <0
+    gam2 = eps;
+end
+
+Eq1 = integral(@(x)Eq1kernel(x, M0, gam1, gam2), 0, 1);
+Eq2 = integral(@(x)Eq2kernel(x, M0, gam1, gam2), 0, 1);
+
+f = -J/(2*sigma20) - Eq1 -Eq2 + 1/(2*gam2);
+
+    function [f] = Eq1kernel(x, M0, gam1, gam2)
+        f = (2*gam2).^(-2).*normpdf(gam1,sqrt(gam2)).*(x-gam1).^2.*gammaln(x*M0) ...
+                -pi*(2*pi*gam2)^(-3/2)*normpdf(gam1,sqrt(gam2)).*gammaln(x*M0);
+    end
+    function [f] = Eq2kernel(x, M0, gam1, gam2)
+        f = (2*gam2).^(-2).*normpdf(gam1,sqrt(gam2)).*(x-gam1).^2.*gammaln((1-x)*M0) ...
+                -pi*(2*pi*gam2)^(-3/2)*normpdf(gam1,sqrt(gam2)).*gammaln((1-x)*M0);
+    end
+end
+

### 2012-10-12 src/test_ll_bound.m

@@ -1,12 +1,16 @@
+%% Problem Setup
+
 rng(1000,'twister');
 
-K = 3; J = 100;
+K = 3; J = 500;
 
-u0 = 0.5; sigma20 = 0.1^2;
+u0 = 0.2; sigma20 = 0.01^2;
 
 mu = normrnd(u0,sqrt(sigma20),[1 J]);
+sum(mu<0 |mu>1)
+mu(mu<0) = 0; mu(mu>1) = 1;
 
-M0 = 10; theta = NaN(K,J);
+M0 = 50; theta = NaN(K,J);
 for k = 1:K
     theta(k,:) = betarnd(mu.*M0,M0.*(1-mu)); 
 end
@@ -14,25 +18,82 @@ end
 n = repmat(100,K,J);
 r = binornd(n,theta);
 
-gam1 = repmat(u0, [1 J]); gam2 = repmat(sigma20, [1 J]);
-alpha = repmat(3,[K J]); beta = repmat(2, [K J]);
+gam1 = repmat(u0,[1 J]); gam2 = repmat(sigma20, [1 J]);
+alpha = rand([K J]); beta = rand([K J]);
+
+%% Solve for log-likelihood bound
 
 % Compute the log-likelihood bound
 [ ll ] = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta );
 
-% Plot the LL bound varying M0
+
+%% Plot the LL bound varying M0
 v = logspace(-3 , 2, 50);
 ll_M0 = arrayfun(@(x)ll_bound( r, n, u0, sigma20, x, gam1, gam2, alpha, beta ), v);
 semilogx(v,ll_M0) 
 
-% Plot the LL bound varying u0
+%% Plot the LL bound varying u0
 v = linspace(0 , 1, 50);
 ll_u0 = arrayfun(@(x)ll_bound( r, n, x, sigma20, M0, gam1, gam2, alpha, beta ), v);
 plot(v,ll_u0) 
 
-% Plot the LL bound varying sigma20
-v = logspace(-3 , 0, 50);
+%% Plot the LL bound varying sigma20
+v = logspace(-3 , 2, 50);
 ll_sigma20 = arrayfun(@(x)ll_bound( r, n, u0, x, M0, gam1, gam2, alpha, beta ), v);
 semilogx(v,ll_sigma20) 
 
+for i = 1:10
+%% Open pool
+% matlabpool open
 
+%% Solve for u0
[…]

### 2012-10-13 src/dLdM0.m
Fix bug where x was missing in normpdf(). Now able to run iterations and does not go out of bounds!

@@ -17,14 +17,15 @@ for j = 1:J
 end
 Eq2 = sum(Eq2);
 
-f = K*J*psi(M0) + sum(gam1.*sum(psi(alpha)-psi(beta),1)) ...
-    +sum(sum(psi(beta)-psi(alpha+beta))) -K*Eq1 - K*Eq2;
+gam1K = repmat(gam1,[K 1]);
+f = K*J*psi(M0) -K*Eq1 -K*Eq2 ...
+    +sum(sum(gam1K.*psi(alpha) +(1-gam1K).*psi(beta) -psi(alpha+beta)));
 
     function [f] = Eq1kernel(x, M0, gam1, gam2)
-        f = normpdf(gam1,sqrt(gam2)).*x.*psi(x.*M0);
+        f = normpdf(x, gam1,sqrt(gam2)).*x.*psi(x.*M0);
     end
     function [f] = Eq2kernel(x, M0, gam1, gam2)
-        f = normpdf(gam1,sqrt(gam2)).*(1-x).*psi((1-x).*M0);
+        f = normpdf(x, gam1,sqrt(gam2)).*(1-x).*psi((1-x).*M0);
     end
 end
 

### 2012-10-13 src/dLdgam2.m

@@ -9,15 +9,16 @@ end
 Eq1 = integral(@(x)Eq1kernel(x, M0, gam1, gam2), 0, 1);
 Eq2 = integral(@(x)Eq2kernel(x, M0, gam1, gam2), 0, 1);
 
-f = -J/(2*sigma20) - Eq1 -Eq2 + 1/(2*gam2);
+f = -1/(2*sigma20) -(0.5/gam2)*Eq1 -(0.5/gam2)*Eq2 + 1/(2*gam2);
 
     function [f] = Eq1kernel(x, M0, gam1, gam2)
-        f = (2*gam2).^(-2).*normpdf(gam1,sqrt(gam2)).*(x-gam1).^2.*gammaln(x*M0) ...
-                -pi*(2*pi*gam2)^(-3/2)*normpdf(gam1,sqrt(gam2)).*gammaln(x*M0);
+        f = (1/gam2).*normpdf(x,gam1, sqrt(gam2)).*gammaln(x*M0).*(x-gam1).^2 ...
+            -normpdf(x, gam1,sqrt(gam2)).*gammaln(x.*M0);
+            
     end
     function [f] = Eq2kernel(x, M0, gam1, gam2)
-        f = (2*gam2).^(-2).*normpdf(gam1,sqrt(gam2)).*(x-gam1).^2.*gammaln((1-x)*M0) ...
-                -pi*(2*pi*gam2)^(-3/2)*normpdf(gam1,sqrt(gam2)).*gammaln((1-x)*M0);
+        f = (1/gam2).*normpdf(x, gam1, sqrt(gam2)).*gammaln((1-x)*M0).*(x-gam1).^2 ...
+            -normpdf(x,gam1,sqrt(gam2)).*gammaln((1-x).*M0);    
     end
 end
 

### 2012-10-13 src/test_ll_bound.m

@@ -2,9 +2,9 @@
 
 rng(1000,'twister');
 
-K = 3; J = 500;
+K = 3; J = 100;
 
-u0 = 0.2; sigma20 = 0.01^2;
+u0 = 0.5; sigma20 = 0.1^2;
 
 mu = normrnd(u0,sqrt(sigma20),[1 J]);
 sum(mu<0 |mu>1)
@@ -47,7 +47,6 @@ for i = 1:10
 % matlabpool open
 
 %% Solve for u0
-% u0 = fminbnd(@(x)-ll_bound( r, n, x, sigma20, M0, gam1, gam2, alpha, beta),0,1);
 u0 = mean(gam1);
 
 %% Solve for sigma20
@@ -57,6 +56,7 @@ sigma20 = mean(gam2 + (gam1-u0).^2);
 options = optimset('Display','iter');
 % M0 = fminbnd(@(x)-ll_bound(r, n, u0, sigma20, x, gam1, gam2, alpha, beta),0,1e10)
 M0 = fzero(@(x)dLdM0( x, u0, alpha, beta, gam1, gam2 ), M0, options);
+
 
 %% Solve for alpha & beta
 options = optimset('Display','off');
@@ -69,31 +69,40 @@ end
 options = optimset('Display','off');
 for j = 1:J
     for k = 1:K
-         beta(k,j) = fsolve(@(x)dLdbeta(alpha(k,j), x, M0,u0,r(k,j),n(k,j)), beta(k,j), options);
+         beta(k,j) = fzero(@(x)dLdbeta(alpha(k,j), x, M0,u0,r(k,j),n(k,j)), beta(k,j), options);
     end
 end
 
 %% Solve for gamma1
 
-gam1 = u0 - (sigma20*M0).*mean(psi(alpha)-psi(beta));
-% options = optimset('Display','off');
-% for j = 1:J
-%     gam1(j)= fzero(@(x)dLdgam1( x, gam2(j), alpha(:,j), beta(:,j), M0, u0, sigma20), gam1(j),...
-%         options);
-% end
-% plot(gam1)
-
-%% Solve for gamma2
-% % options = optimset('Display','iter');
-% options = optimset('Display','off');
-% for j = 1:J
-%     gam2(j)= fzero(@(x)dLdgam2( gam1(j), x, M0, sigma20, J ), gam2(j),options);
-% end
-% plot(gam2)
-disp('Done.')
+% gam1 = u0 - (sigma20*M0).*mean(psi(alpha)-psi(beta));
[…]

### 2012-10-13 notes.tex
Fix bug where x was missing in normpdf(). Now able to run iterations and does not go out of bounds!
Add some derivations wrt gamma2. Fixed rvd2_est and test_ll_bound to work on cluster.

@@
 Taking each term in \eqref{eqn:composite} individually we can cast the bound on the log-likelihood in terms of original and variational parameters only.
-
 \begin{align}
@@
 & = \log \Gamma(M_0) - \E_q \log \Gamma(\mu_j M_0) - \E_q \log \Gamma((1-\mu_j)M_0) \\
-& \qquad+(\gamma_{1j}M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) + ((1-\gamma_{1j})M_0-1)\left( \psi(\beta_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right)
+& \qquad+(\gamma_{1j}M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) \\
+& \qquad + ((1-\gamma_{1j})M_0-1)\left( \psi(\beta_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right)
 \end{split}
@@
 where $\E_q \log \Gamma(\mu_j M_0)$ is easily computed numerically.
-When we get to differentiating with respect to the parameters, we can move that inside the integral.
-The kernel changes, but it can still be computed numerically.
+In general, we will be able to differentiate under the integral by Leibniz integral rule and the fact that the bounds are constants $\mu_j \in [0,1]$.
 
 \begin{align}
-\E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) & = \int_{\theta_{kj}} \\
+\E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) & = \int_{\theta_{kj}} q(\theta_{kj}) \log p(r_{kj} | \theta_{kj}, n_{kj})\\
 \begin{split}
@@
 
+Isolating terms that involve $M_0$ gives
+\begin{multline}
+\mathcal{L}_{[M_0]} = \log \Gamma(M_0) -\E_q\log \Gamma(\mu_jM_0) -\E_q\log \Gamma((1-\mu_j)M_0) +\\
++\gamma_{1j}M_0(\psi(\alpha_{kj})-\psi(\alpha_{kj}+\beta_{kj})) +(1-\gamma_{1j})M_0(\psi(\beta_{kj})-\psi(\alpha_{kj}+\beta_{kj}))
+\end{multline}
+
+{\bf Update Equation for $\gamma_{2j}$ }
+
+Isolating terms that involve $\gamma_{2j}$ gives
+\begin{equation}
+\mathcal{L}_{[\gamma_{2j}]} = -\frac{K}{2\sigma_0^2}\gamma_{2j} -\E_q\log\Gamma(\mu_jM_0) - \E_q\log\Gamma((1-\mu_j)M_0)+\frac{1}{2}\log(2\pi e \gamma_{2j})
+\end{equation}
+
 \end{document}

### 2012-10-13 src/rvd2_est.m

@@ -12,7 +12,7 @@ alpha = rand([K J]); beta = rand([K J]);
 [ llCurr ] = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta );
 llDelta = Inf; iterCount =0; llSave = NaN(1,MAXITER);
 
-while iterCount < MAXITER
+while iterCount < MAXITER & llDelta > LLTOL
     %% Solve for u0
     u0 = mean(gam1);
     
@@ -27,14 +27,14 @@ while iterCount < MAXITER
     % TODO: Do we want to loop over the variational parameters inside.
     %% Solve for alpha & beta
     options = optimset('Display','off');
-    parfor j = 1:J
+    for j = 1:J
         for k = 1:K
             alpha(k,j) = fzero(@(x)dLdalpha(x, beta(k,j),M0,u0,r(k,j),n(k,j)), alpha(k,j), options);
         end
     end
     
     options = optimset('Display','off');
-    parfor j = 1:J
+    for j = 1:J
         for k = 1:K
             beta(k,j) = fzero(@(x)dLdbeta(alpha(k,j), x, M0,u0,r(k,j),n(k,j)), beta(k,j), options);
         end
@@ -43,14 +43,14 @@ while iterCount < MAXITER
     %% Solve for gamma1 & gamma2
     % gam1 = u0 - (sigma20*M0).*mean(psi(alpha)-psi(beta));
     options = optimset('Display','off');
-    parfor j = 1:J
+    for j = 1:J
         gam1(j)= fzero(@(x)dLdgam1( x, gam2(j), alpha(:,j), beta(:,j), M0, u0, sigma20), gam1(j),...
             options);
     end
     
     % options = optimset('Display','iter');
     options = optimset('Display','off');
-    parfor j = 1:J
+    for j = 1:J
         gam2(j)= fzero(@(x)dLdgam2( gam1(j), x, M0, sigma20, J ), gam2(j),options);
     end
 
@@ -58,6 +58,6 @@ while iterCount < MAXITER
     iterCount = iterCount+1;
     llSave(iterCount) = llCurr;
     llCurr = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta );
-    llDelta = (llCurr - llSave(iterCount))./llSave(iterCount)
+    llDelta = (llCurr - llSave(iterCount))./abs(llSave(iterCount))
 end
 

### 2012-10-13 src/test_ll_bound.m

@@ -21,38 +21,35 @@ r = binornd(n,theta);
 %% Solve for log-likelihood bound
 
 % Compute the log-likelihood bound
-[ ll ] = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta );
+% [ ll ] = ll_bound( r, n, u0, sigma20, M0, gam1, gam2, alpha, beta );
 
 
 %% Plot the LL bound varying M0
-v = logspace(-3 , 2, 50);
-ll_M0 = arrayfun(@(x)ll_bound( r, n, u0, sigma20, x, gam1, gam2, alpha, beta ), v);
-semilogx(v,ll_M0) 
+% v = logspace(-3 , 2, 50);
+% ll_M0 = arrayfun(@(x)ll_bound( r, n, u0, sigma20, x, gam1, gam2, alpha, beta ), v);
+% semilogx(v,ll_M0) 
 
 %% Plot the LL bound varying u0
-v = linspace(0 , 1, 50);
-ll_u0 = arrayfun(@(x)ll_bound( r, n, x, sigma20, M0, gam1, gam2, alpha, beta ), v);
-plot(v,ll_u0) 
+% v = linspace(0 , 1, 50);
+% ll_u0 = arrayfun(@(x)ll_bound( r, n, x, sigma20, M0, gam1, gam2, alpha, beta ), v);
+% plot(v,ll_u0) 
 
 %% Plot the LL bound varying sigma20
-v = logspace(-3 , 2, 50);
-ll_sigma20 = arrayfun(@(x)ll_bound( r, n, u0, x, M0, gam1, gam2, alpha, beta ), v);
-semilogx(v,ll_sigma20) 
+% v = logspace(-3 , 2, 50);
+% ll_sigma20 = arrayfun(@(x)ll_bound( r, n, u0, x, M0, gam1, gam2, alpha, beta ), v);
+% semilogx(v,ll_sigma20) 
 
 %% Estimate Model Parameters
-matlabpool open
+% matlabpool open
 [ u0,sigma20,M0,alpha,beta,gam1,gam2 ] ...
     = rvd2_est( r, n, u0, sigma20, M0 );
-matlabpool close
+% matlabpool close
 %%
 
-post_mu = gam1;
-plot(theta(:,1:10)','.','MarkerSize',14)
-hold on
-plot(post_mu(1:10))
-post_theta = alpha./(alpha+beta);
[…]

### 2012-10-13 notes.tex
Fixed and added some update equations in the notes.

@@
 
-{\bf Update Equations}
+{\bf Update Equation for $u_0$ }
+
 Isolating the terms of \eqref{eqn:composite} that involve $u_0$ gives
@@
 
-Taking the derivative with respect to $u_0$ gives
-
-Solving for $u_0$ gives
+Taking the derivative and solving for $u_0$ gives
 \begin{equation}
@@
 \end{equation}
+
+{\bf Update Equation for $\sigma_0^2$ }
 
@@
 
+{\bf Update Equation for $\alpha_{kj}$ and $\beta_{kj}$ }
+
 Isolating terms that involve $\alpha_{kj}$ gives
 \begin{multline}
-\mathcal{L}_{[\alpha_{kj}]} = (u_0M_0 -1)(\psi(\alpha_{kj}) - \psi(\alpha_{kj}-\beta_{kj}) -((1-u_0)M_0-1)\psi(\alpha_{kj}+\beta_{kj})\\
- +r_{kj}*\psi(\alpha_{kj})-n_{kj}\psi(\alpha_{kj}+\beta_{kj}) -\log \Gamma (\alpha+\beta) + \log \Gamma(\alpha) -(\alpha-1)\psi(\alpha)+(\alpha+\beta-2)\psi(\alpha+\beta)
+\mathcal{L}_{[\alpha_{kj}]} = (u_0M_0 -1)(\psi(\alpha_{kj}) - \psi(\alpha_{kj}+\beta_{kj})) -((1-u_0)M_0-1)\psi(\alpha_{kj}+\beta_{kj}) \\
++ r_{kj}\psi(\alpha_{kj})-n_{kj}\psi(\alpha_{kj}+\beta_{kj}) \\
+ - \log \Gamma (\alpha_{kj}+\beta_{kj}) + \log \Gamma(\alpha_{kj}) -(\alpha_{kj}-1)\psi(\alpha_{kj})+(\alpha_{kj}+\beta_{kj}-2)\psi(\alpha_{kj}+\beta_{kj})
 \end{multline}
@@
 \begin{equation}
-\alpha_{kj} \leftarrow \psi_1(\alpha_{kj})(u_0M_0+r-\alpha_{kj}) - \psi_1(\alpha_{kj}+\beta_{kj})(M_0+n-(\alpha_{kj}+\beta_{kj}))
+ \frac{\partial \mathcal{L}}{\partial \alpha_{kj}} = \psi_1(\alpha_{kj})(u_0M_0+r-\alpha_{kj}) - \psi_1(\alpha_{kj}+\beta_{kj})(M_0+n-(\alpha_{kj}+\beta_{kj}))
 \end{equation}
@@
 \end{equation}
+
+{\bf Update Equation for $M_0$ }
 
@@
 
+Taking the derivative with respect to $\gamma_{2j}$ gives
+\begin{equation}\label{eqn:gam2_deriv}
+\frac{\partial \mathcal{L}}{\partial \gamma_{2j}} = -\frac{K}{2\sigma_0^2} -\frac{\partial}{\partial \gamma_{2j}}\E_q\log\Gamma(\mu_jM_0) -\frac{\partial}{\partial \gamma_{2j}}\E_q\log\Gamma((1-\mu_j)M_0) +\frac{1}{2\gamma_{2j}}
+\end{equation}
+
+The two partials are very similar, so we show only the derivation of the first and apply the solution to the second replacing $\mu_j$ with $(1-\mu_j)$ in the gamma function.
+\begin{align}\label{eqn:gam2_int1}
+\frac{\partial}{\partial \gamma_{2j}}\E_q\log\Gamma(\mu_jM_0) &= \frac{\partial}{\partial \gamma_{2j}} \int_0^1 (2\pi\gamma_{2j})^{-1/2}\exp\left(-\frac{1}{2\gamma_{j2}}(\mu_j-\gamma_{1j})^2\right) \log\Gamma(\mu_jM_0)d\mu_j \\
+& = \frac{1}{2\gamma_{2j}}\left(\frac{1}{\gamma_{2j}} \E_q \left[ (\mu_j-\gamma_{1j})^2 \log\Gamma(\mu_jM_0) \right] - \E_q \left[ \log\Gamma(\mu_jM_0) \right] \right)
+\end{align}
+
+Replacing \eqref{eqn:gam2_int1} in \eqref{eqn:gam2_deriv} gives
+\begin{multline}
+\frac{\partial \mathcal{L}}{\partial \gamma_{2j}} = -\frac{K}{2\sigma_0^2} 
+- \frac{1}{2\gamma_{2j}}\left(\frac{1}{\gamma_{2j}} \E_q \left[ (\mu_j-\gamma_{1j})^2 \log\Gamma(\mu_jM_0) \right] - \E_q \left[ \log\Gamma(\mu_jM_0) \right] \right) \\
+- \frac{1}{2\gamma_{2j}}\left(\frac{1}{\gamma_{2j}} \E_q \left[ (\mu_j-\gamma_{1j})^2 \log\Gamma((1-\mu_j)M_0) \right] - \E_q \left[ \log\Gamma((1-\mu_j)M_0) \right] \right) +\frac{1}{2\gamma_{2j}}
+\end{multline}
+Solving for the root gives the maximizing value of $\gamma_{2j}$ and the update.
+
 \end{document}

### 2012-10-15 notes.tex
Finished update equations in notes.

@@
 Isolating the terms of \eqref{eqn:composite} that involve $u_0$ gives
-\begin{equation}
+\begin{equation*}
 \mathcal{L}_{[u_0]} = K\sum_{j=1}^J\left(-\frac{1}{2\sigma_0^2}(\gamma_{1j}-u_0)^2\right)
-\end{equation}
+\end{equation*}
 
@@
 
+%%%			sigma2 update			%%%
 {\bf Update Equation for $\sigma_0^2$ }
@@
 Isolating the terms of \eqref{eqn:composite} that involve $\sigma^2_0$ gives
-\begin{equation}
+\begin{equation*}
 \mathcal{L}_{[\sigma^2_0]} = -\frac{J}{2}\log(2\pi\sigma^2_0) -\frac{1}{2\sigma_0^2}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2 \right)
-\end{equation}
+\end{equation*}
 
 Taking the derivative with respect to $\sigma^2_0$ gives
-\begin{equation}
+\begin{equation*}
 \frac{\partial \mathcal{L}}{\partial \sigma^2_0} = -\frac{J}{2\sigma^2_0} +\frac{1}{\sigma^4_0}\sum_{j=1}^J\left(\gamma_{2j} + (\gamma_{1j}-u_0)^2\right)
-\end{equation}
+\end{equation*}
 
@@
 
+%%% 		M0 update 		%%%
+{\bf Update Equation for $M_0$ }
+
+Isolating terms that involve $M_0$ gives
+\begin{multline*}
+\mathcal{L}_{[M_0]} = \log \Gamma(M_0) -\E_q\log \Gamma(\mu_jM_0) -\E_q\log \Gamma((1-\mu_j)M_0) +\\
++\gamma_{1j}M_0(\psi(\alpha_{kj})-\psi(\alpha_{kj}+\beta_{kj})) +(1-\gamma_{1j})M_0(\psi(\beta_{kj})-\psi(\alpha_{kj}+\beta_{kj}))
+\end{multline*}
+
+Taking the derivative with respect to $M_0$ gives
+\begin{multline*}
+\frac{\partial \mathcal{L}}{\partial M_0} = \psi(M_0) -\frac{\partial}{\partial M_0}\E_q\log \Gamma(\mu_jM_0)
+-\frac{\partial}{\partial M_0}\E_q\log \Gamma((1-\mu_j)M_0) \\
++\gamma_{1j}(\psi(\alpha_{kj})-\psi(\alpha_{kj}+\beta_{kj})) 
++(1-\gamma_{1j})(\psi(\beta_{kj})-\psi(\alpha_{kj}+\beta_{kj}))
+\end{multline*}
+
+The partial derivative can be taken inside of the integral which gives $\frac{\partial}{\partial M_0}\E_q\log \Gamma(\mu_jM_0) = \E_q \left[ \mu_j \psi(\mu_j M_0) \right]$ and a similar expression for the second term $\frac{\partial}{\partial M_0}\E_q\log \Gamma((1-\mu_j)M_0) = \E_q \left[(1- \mu_j) \psi((1-\mu_j) M_0) \right]$.
+
+Plugging these simplifications into the partial with respect to $M_0$ gives
+\begin{multline}
+\frac{\partial \mathcal{L}}{\partial M_0} = \psi(M_0) - \E_q \left[ \mu_j \psi(\mu_j M_0) \right]
+-\E_q \left[(1- \mu_j) \psi((1-\mu_j) M_0) \right] \\
++\gamma_{1j}(\psi(\alpha_{kj})-\psi(\alpha_{kj}+\beta_{kj})) 
++(1-\gamma_{1j})(\psi(\beta_{kj})-\psi(\alpha_{kj}+\beta_{kj}))
+\end{multline}
+
+%%%		alpha & beta update			%%%
 {\bf Update Equation for $\alpha_{kj}$ and $\beta_{kj}$ }
@@
 Isolating terms that involve $\alpha_{kj}$ gives
-\begin{multline}
+\begin{multline*}
 \mathcal{L}_{[\alpha_{kj}]} = (u_0M_0 -1)(\psi(\alpha_{kj}) - \psi(\alpha_{kj}+\beta_{kj})) -((1-u_0)M_0-1)\psi(\alpha_{kj}+\beta_{kj}) \\
@@
  - \log \Gamma (\alpha_{kj}+\beta_{kj}) + \log \Gamma(\alpha_{kj}) -(\alpha_{kj}-1)\psi(\alpha_{kj})+(\alpha_{kj}+\beta_{kj}-2)\psi(\alpha_{kj}+\beta_{kj})
-\end{multline}
+\end{multline*}
 
@@
 
-A similar procedure holds for $\beta_{kj}$ and the resulting update equation bears symmetry to that for $\alpha_{kj}$
+A similar procedure holds for $\beta_{kj}$ and the resulting update equation bears symmetry to that for $\alpha_{kj}$,
 \begin{equation}
-\beta_{kj} \leftarrow \psi_1(\beta_{kj})\left[(1-u_0)M_0-1+(n-r)-\beta_{kj}\right] - \psi_1(\alpha_{kj}+\beta_{kj})\left[M_0+n-(\alpha_{kj}+\beta_{kj})\right]
+ \frac{\partial \mathcal{L}}{\partial \beta_{kj}} = \psi_1(\beta_{kj}) \left[ (1-u_0)M_0-1+(n_{kj}-r_{kj})-\beta_{kj}\right] - \psi_1(\alpha_{kj}+\beta_{kj})(M_0+n-(\alpha_{kj}+\beta_{kj}))
 \end{equation}
 
-{\bf Update Equation for $M_0$ }
-
-Isolating terms that involve $M_0$ gives
-\begin{multline}
-\mathcal{L}_{[M_0]} = \log \Gamma(M_0) -\E_q\log \Gamma(\mu_jM_0) -\E_q\log \Gamma((1-\mu_j)M_0) +\\
-+\gamma_{1j}M_0(\psi(\alpha_{kj})-\psi(\alpha_{kj}+\beta_{kj})) +(1-\gamma_{1j})M_0(\psi(\beta_{kj})-\psi(\alpha_{kj}+\beta_{kj}))
-\end{multline}
-
-{\bf Update Equation for $\gamma_{2j}$ }
+%%% 		gamma1 update 		%%%
+{\bf Update equation for $\gamma_{1j}$}\label{sec:gamma1_update}
 
 Isolating terms that involve $\gamma_{2j}$ gives
+\begin{equation*}
+\mathcal{L}_{[\gamma_{1j}]} = -\frac{K}{2\sigma_0^2}(\gamma_{1j}-u_0)^2 
++ \sum_{k=1}^K \gamma_{1j} M_0 ( \psi(\alpha_kj) - \psi(\alpha_{kj} + \beta_{kj}) )
+- \sum_{k=1}^K \gamma_{1j} M_0 ( \psi(\beta_kj) - \psi(\alpha_{kj} + \beta_{kj}) )
+\end{equation*}
+
+Taking the derivative with respect to $\gamma_{1j}$, setting it equal to zero and solving for $\gamma_{1j}$ gives
 \begin{equation}
+\gamma_{1j} \leftarrow u_0 + M_0 \sigma_0^2 \frac{1}{K} \sum_{k=1}^K \psi(\alpha_{kj}) - \psi(\beta_{kj})
+\end{equation}
+
+%%% 		gamma2 update 		%%%
+{\bf Update Equation for $\gamma_{2j}$ }\label{sec:gamma2_update}
+
+Isolating terms that involve $\gamma_{2j}$ gives
+\begin{equation*}
 \mathcal{L}_{[\gamma_{2j}]} = -\frac{K}{2\sigma_0^2}\gamma_{2j} -\E_q\log\Gamma(\mu_jM_0) - \E_q\log\Gamma((1-\mu_j)M_0)+\frac{1}{2}\log(2\pi e \gamma_{2j})
-\end{equation}
+\end{equation*}
 

### 2012-10-15 notes.tex
Partially correct gamma1 update.

@@
 
+%%%			log-likelihood bound			%%%
 We are now able to decompose $\mathcal{L}$ into its constituent parts
@@
 \begin{split}
-& = \log \Gamma(M_0) - \E_q \log \Gamma(\mu_j M_0) - \E_q \log \Gamma((1-\mu_j)M_0) \\
+& = \log \Gamma(M_0) -\E_q \log \Gamma(\mu_j M_0) -\E_q \log \Gamma((1-\mu_j)M_0) \\
 & \qquad+(\gamma_{1j}M_0-1)\left( \psi(\alpha_{kj}) - \psi(\alpha_{kj} + \beta_{kj}) \right) \\
@@
 
+%%%			u0 update				%%%
 {\bf Update Equation for $u_0$ }
@@
 
-Isolating terms that involve $\gamma_{2j}$ gives
-\begin{equation*}
+Isolating terms that involve $\gamma_{1j}$ gives
+\begin{multline*}
 \mathcal{L}_{[\gamma_{1j}]} = -\frac{K}{2\sigma_0^2}(\gamma_{1j}-u_0)^2 
+-\E_q \log \Gamma(\mu_j M_0) -\E_q \log \Gamma((1-\mu_j)M_0) \\
 + \sum_{k=1}^K \gamma_{1j} M_0 ( \psi(\alpha_kj) - \psi(\alpha_{kj} + \beta_{kj}) )
 - \sum_{k=1}^K \gamma_{1j} M_0 ( \psi(\beta_kj) - \psi(\alpha_{kj} + \beta_{kj}) )
-\end{equation*}
+\end{multline*}
 
-Taking the derivative with respect to $\gamma_{1j}$, setting it equal to zero and solving for $\gamma_{1j}$ gives
+Taking the derivative with respect to $\gamma_{1j}$  gives
 \begin{equation}
-\gamma_{1j} \leftarrow u_0 + M_0 \sigma_0^2 \frac{1}{K} \sum_{k=1}^K \psi(\alpha_{kj}) - \psi(\beta_{kj})
+.
 \end{equation}

### 2012-10-24 notes.tex

@@
 \end{flushright}
+
+\section*{RVD2.1: Independent variational parameters}
 
@@
 \begin{multline}
-\log p(\mu,\theta | r) = K\sum_{j=1}^J \log p(\mu_j;\mu_0,\sigma^2_0) + \sum_{j=1}^J \sum_{k=1}^K \log p(\theta_{kj} | \mu_j;M_0)\\ + \sum_{j=1}^J \sum_{k=1}^K\log p(r_{kj} | \theta_{kj}, n_{kj})
-- \sum_{j=1}^J \sum_{k=1}^K \log \int_{\theta_{kj}} \int_{\mu_j} p(r_{kj};\mu_0,\sigma^2_0,M_0) \end{multline}
+\log p(\mu,\theta | r) = K\sum_{j=1}^J \log p(\mu_j;\mu_0,\sigma^2_0) 
+	+\sum_{j=1}^J \sum_{k=1}^K \log p(\theta_{kj} | \mu_j;M_0)\\ 
+	+\sum_{j=1}^J \sum_{k=1}^K\log p(r_{kj} | \theta_{kj}, n_{kj})
+	- \sum_{j=1}^J \sum_{k=1}^K \log \int_{\theta_{kj}} \int_{\mu_j} p(r_{kj};\mu_0,\sigma^2_0,M_0) 
+\end{multline}
 
@@
 & \geq  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log p(\mu_j, \theta_{kj}, r_{kj}) -  \int_{\theta_{kj}} \int_{\mu_j} q(\mu_j, \theta_{kj}) \log q(\mu_j, \theta_{kj}) \\
-& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L_{kj}}
+& = \E_q \left[ \log p(r_{kj}, \mu_j, \theta_{kj}) \right] - \E_q\left[ \log q(\mu_j, \theta_{kj}) \right] = \mathcal{L}_{kj}
 \end{align}
@@
 \begin{align}
-\E_q \log p(\theta_{kj} | \mu_j ; M_0) & = \int_{\mu_j} q(\mu_j) \int_{\theta_{kj}} q(\theta_{kj}) \log p(\theta_{kj} | \mu_j, M_0)\\
+\E_q \log p(\theta_{kj} | \mu_j ; M_0) & = \int_{\mu_j} q(\mu_j) \int_{\theta_{kj}} q(\theta_{kj}) \log p(\theta_{kj} | \mu_j, M_0) \nonumber\\
 \begin{split}
@@
 \begin{align}
-\E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) & = \int_{\theta_{kj}} q(\theta_{kj}) \log p(r_{kj} | \theta_{kj}, n_{kj})\\
+\E_q \log p(r_{kj} | \theta_{kj}, n_{kj}) & = \int_{\theta_{kj}} q(\theta_{kj}) \log p(r_{kj} | \theta_{kj}, n_{kj}) \nonumber \\
 \begin{split}
@@
 Taking the derivative with respect to $\gamma_{1j}$  gives
-\begin{equation}
-.
-\end{equation}
+\begin{multline}
+\frac{\partial \mathcal{L}}{\partial \gamma_{1j}} = -\frac{K}{\sigma_0^2}(\gamma_{1j} - u_0) 
+	+M_0 \sum_{k=1}^K \left( \psi(\alpha_{kj}) - \psi(\beta_{kj}) \right) \\
+	-K \E_q \left[ \frac{1}{\gamma_{2j}} \log \Gamma(\mu_jM_0) \right]
+	-K \E_q \left[ \frac{1}{\gamma_{2j}} \log \Gamma((1-\mu_j)M_0) \right]
+\end{multline}
 

### 2013-02-11 notes.tex
Added notes for rvd2.3.

@@
 \usepackage{amssymb} % allow blackboard bold (aka N,R,Q sets)
+
+\usepackage{algorithm2e}
+
 \linespread{1.6}  % double spaces lines
@@
 
+\section{RVD 2.3: Gamma Prior}
+Consider the following model
+\begin{align}
+\alpha_j &\thicksim \text{Gamma}(a, b)\\
+\theta_{ij} &\thicksim \text{Dirichlet}(\alpha_j)\\
+r_{ij} &\thicksim \text{Multinomial}(\theta_{ij}, n_{ij})
+\end{align}
+
+For each genomic location, $j=1,\ldots,M$, $K$ random variables are independently chosen from a Gamma distribution with shape parameter $a$ and scale parameter $b$.
+Within each location, $N$ Dirichlet random variables, $\theta_{ij}$ are chosen with parameter $\alpha_j$ for each replicate.
+Finally, a read count variable $r_{ij}$ is drawn from a multinomial distribution with probability parameter $\theta_{ij}$ and total count $n_{ij}$.
+The count of nucleotide $k\in\{A, C, T, G\}$ at position $j$ in replicate $i$ is then $r_{ij}^k$.
+
+The inferential object of interest here is the joint posterior distribution over the latent variables $p(\theta, \alpha | r)$.
+From the posterior distribution we can estimate the location-specific Dirichlet parameters $\hat{\alpha}_j$ and thus the location-specific nucleotide probabilities $\frac{ \hat{\alpha}^k_{j} } { \hat{\alpha}^0_{j} }$, where $\hat{\alpha}^0_{j} = \sum_k \hat{\alpha}^k_j$.
+
+The joint posterior distribution is
+\begin{equation}
+p(\alpha, \theta | r; a, b) = \prod_{i=1}^N \prod_{j=1}^M \frac{ p(r_{ij} | \theta_{ij}) p(\theta_{ij} | \alpha_j) p(\alpha_j; a, b) } { p(r_{ij} ; a, b) }.
+\end{equation}
+But, normalization factor in the posterior distribution,
+\begin{equation}
+p(r_{ij}; a, b) = \int_{\theta_{ij}} \int_{\alpha_j} p(r_{ij} | \theta_{ij}) p(\theta_{ij} | \alpha_j) p(\alpha_j; a, b),
+\end{equation}
+is computationally intractable due to the coupling between $\theta_{ij}$ and $\alpha_j$.
+
+Instead of attempting to directly compute the posterior distribution, we instead sample it using  a Metropolis-Hastings within Gibbs scheme.
+
+\begin{algorithm}[H]
+\SetAlgoLined
+\KwData{Counts $r_{ij}$ for experimental replicate $i$ at genomic location $j$}
+\KwResult{MLE estimates for hyper-parameters $\hat{a}$ and $\hat{b}$.
+Samples from joint posterior distribution $p(\theta_{ij}, \alpha_j | r_{ij}; \hat{a}, \hat{b})$ for $i=1,\ldots,N$ and $j=1,\ldots,M$}
+Initialize $a$ and $b$\;
+\Repeat {convergence} {
+	Sample $\alpha_j$ from $p(\alpha_j | a, b, \theta_{1j}, \ldots, \theta_{Nj})$ using M-H\;
+	Sample $\theta_{ij}$ from $p(\theta_{ij} | \alpha_j, r_{ij})$ using Gibbs\;
+	Compute MLE $\hat{a}$ and $\hat{b}$
+}
+\caption{Metropolis-within-Gibbs Sampling for RVD2.3 Model}
+\end{algorithm}
+
+The maximum-likelihood estimates $\hat{a}$ and $\hat{b}$ are computed using generalized Newton-Raphson as outlined in http://research.microsoft.com/en-us/um/people/minka/papers/minka-gamma.pdf
+
+Samples from $p(\theta_{ij} | \alpha_j, r_{ij})$ can be obtained directly because the Dirichlet is conjugate to the Multinomial. The posterior sampling distribution is
+\begin{equation}
+	\theta_{ij} | \alpha_j, r_{ij} \thicksim \text{Dirichlet}(\alpha_j + r_{ij}).
+\end{equation}
+
+The Gamma distribution is not conjugate to the Dirichelt, so we draw samples from $p(\alpha_j | a, b, \theta_{1j}, \ldots, \theta_{Nj})$ by Metropolis-Hastings. We use a proposal distribution,
+\begin{equation}
+	\alpha_j^{k \text{(new)}} | \alpha_j^k \thicksim \text{Gaussian}(\alpha_j^k, 0.05),
+\end{equation}
+which we find mixes sufficiently fast to sample from the posterior after a burn-in of length X.
+
 \end{document}

### 2013-02-11 src/python/rvd23/rvd23.py

@@ -72,7 +72,7 @@ def log_cond_theta(phi, alpha, theta, r):
     return (logPr + logPtheta)
 
 
-def sample_alpha(alpha, theta, phi, nsample=1):
+def sample_alpha_mh(alpha, theta, phi, nsample=1):
     """Return a sample from the posterior marginal distribution for alpha"""
     
     K = np.shape(alpha)[0]
@@ -159,17 +159,11 @@ def metro_gibbs(r, phi, nsample=10000):
     alpha_s = np.zeros((npos, K, nsample))
     for s in xrange(0,nsample):
         for j in xrange(0,npos):
-            alpha[j,:] = sample_alpha(alpha[j,:], 
-                                      theta[:,j,:], 
-                                      phi, 
-                                      nsample=1)
+            alpha[j,:] = sample_alpha_mh(alpha[j,:], theta[:,j,:], phi)
             for i in xrange(0, nrep):
                 theta[i,j,:] = sample_theta_gibbs(alpha[j,:], r[i,j,:])
-                # theta[i,j,:] = sample_theta_mh(alpha[j,:], 
-                #                                theta[i,j,:], 
-                #                                r[i,j,:], 
-                #                                phi,
-                #                                nsample=1)
+                # theta[i,j,:] = sample_theta_mh(alpha[j,:], theta[i,j,:], r[i,j,:], phi)
+        # Store the sample and update Gamma parameters by MLE
         alpha_s[:,:,s] = np.copy(alpha)
         theta_s[:,:,:,s] = np.copy(theta)
         phi['a'], phi['b'] = gamma_mle(alpha_s[:,:,s])
@@ -186,8 +180,7 @@ def gamma_mle(x):
     a = 0.5 / ( np.log(np.mean(x)) - np.mean(np.log(x)) )
     
     
-    # Generalized Newton updates for a
-    # stop when change in a is small
+    # Generalized Newton updates for "a"
     while True:
         a_new = 1 / (1/a + (np.mean(np.log(x)) \
                         - np.log(np.mean(x)) \
@@ -213,9 +206,6 @@ if __name__ == '__main__':
     alpha_hat = alpha_s.mean(2)
     phi['a'], phi['b'] = gamma_mle(alpha_hat)
 
-    print np.shape(theta_hat)
-    # print "Estimated Alpha"
-    # print alpha_hat
     
     print "Estimated phi"
     print phi

### 2013-02-22 doc/notes.tex
version 1 of RVD2.6 is working now.

--- 2013-02-11 notes.tex
@@
 
-\usepackage{algorithm2e}
+\usepackage[boxed]{algorithm2e}
 
@@
 \Repeat {convergence} {
-	Sample $\alpha_j$ from $p(\alpha_j | a, b, \theta_{1j}, \ldots, \theta_{Nj})$ using M-H\;
-	Sample $\theta_{ij}$ from $p(\theta_{ij} | \alpha_j, r_{ij})$ using Gibbs\;
-	Compute MLE $\hat{a}$ and $\hat{b}$
+	Sample $\alpha_j$ from $p(\alpha_j | a, b, \theta_{1j}, \ldots, \theta_{Nj})$ using Metropolis-Hastings\;
+	Sample $\theta_{ij}$ from $p(\theta_{ij} | \alpha_j, r_{ij})$\;
+	Compute maximum-likelihood estimates $\hat{a}$ and $\hat{b}$
 }
@@
 
-The Gamma distribution is not conjugate to the Dirichelt, so we draw samples from $p(\alpha_j | a, b, \theta_{1j}, \ldots, \theta_{Nj})$ by Metropolis-Hastings. We use a proposal distribution,
+The Gamma distribution is not conjugate to the Dirichlet, so we draw samples from $p(\alpha_j | a, b, \theta_{1j}, \ldots, \theta_{Nj})$ by Metropolis-Hastings. We use a proposal distribution,
 \begin{equation}

### 2013-02-22 src/python/rvd26/rvd26.py

@@ -118,8 +118,9 @@ def gibbs_map(r, n, tol=1e-4):
         theta = ss.beta.rvs(alpha+1e-3, beta+1e-3)
         
         # Draw samples from p(mu | theta, mu0, M0) by Metropolis-Hastings
-        mu_s = sampleMuMH(theta, phi['mu0'], phi['M0'], M, mu=mu, nsample=1000)
+        mu_s = sampleMuMH(theta, phi['mu0'], phi['M0'], M, mu=mu, nsample=5000)
         mu = np.median(mu_s, 0)
+        muErr = np.abs(np.percentile(mu_s, (2.5, 97.5), 0) - mu)
         
         # Draw samples from p(M | a, b, theta, mu)
         M_s = sampleMMH(theta, mu, phi['a'], phi['b'], M=M, nsample=5000)
@@ -156,12 +157,11 @@ def gibbs_map(r, n, tol=1e-4):
         ax1.xaxis.grid(True)
         ax1.yaxis.grid(True)
 
-        
         plt.show()
         
         
-        plt.plot(range(0,J), mu, color='b')
-        # plt.plot(range(0,J), np.transpose(muErr), color='b', linestyle='--')
+        # plt.plot(range(0,J), mu, color='b')
+        plt.errorbar(range(0,J), mu, yerr=muErr, color='b', linestyle='--')
         plt.plot(range(0,J), np.transpose(theta), linestyle='None', marker='+', markersize=10)
         plt.plot(range(0,J), np.transpose(r/n), linestyle='None', marker='o', markersize=6)
         plt.plot(range(0,J), np.tile(phi['mu0'], J), linestyle='--', color='r')
@@ -221,7 +221,7 @@ def sampleMMH(theta, mu, a, b, M=ss.gamma.rvs(1, 1), burnin=0.2, nsample=5000, t
     N,J = np.shape(theta)    
     
     accP = np.zeros(J) # track the acceptance probability for each position
-    Qsd = 1000*np.ones(J) # keep the std, dev of the proposal
+    Qsd = 10*np.ones(J) # keep the std, dev of the proposal
     
     M_s = np.zeros( (nsample, J) ) 
     for ns in xrange(0, nsample):
@@ -254,8 +254,6 @@ def sampleMMH(theta, mu, a, b, M=ss.gamma.rvs(1, 1), burnin=0.2, nsample=5000, t
         #     for j in xrange(0, J):
[…]

### 2013-08-22 doc/bioinformatics2013/manuscript-formatted.tex
added latex skeleton

\documentclass[11pt]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{algorithm, algorithmic}
\usepackage{graphicx}

\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcommand{\RR}{I\!\!R} %real numbers
\DeclareMathOperator{\diag}{diag}

\title{A Mixed-Membership Model for Heterogenous Tumor Subtype Classification}
\author{}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle
\section{Introduction}

Genomic analyses of many solid cancers have shown that there is extensive genetic heterogeneity between individual tumors \cite{Parsons:2008fu, Bonavia:2011hf} as well as within individual tumors \cite{Heppner:1984ve, Dexter:1978ux}. Microdissection and DNA sequencing has recently validated the existence of intratumor heterogeneity at high resolution \cite{Gerlinger:2012fs}. Characterizing intratumor heterogeneity is important because it may be a contributing factor to mono therapy treatment failure \cite{} and accurately estimating the subtypes present in an individual tumor may lead to improved combinatorial therapies.

Many cancers have been classified into distinct genetic subtypes that develop by means of activation or repression of different driver pathways. These tumor subtypes are commonly identified and characterized by clustering the genomic data from hundreds of samples \cite{Eisen:1998ue}. In an effort to disambiguate driver from passenger mutations, the genomic signatures associated with each subtype are sparse and comprised only of those aberrations that are thought to be involved in oncogenesis. New tumors are then classified based on their similarity to the centroids or signatures of those subtypes. However, this classification approach makes an all-or-none assumption about the primary tumor that is not the case for heterogeneous tumors.

Mixed-membership models have emerged in recent years as a modeling tool for data where the all-or-none clustering assumption doesn't make sense. The topic-modeling framework \cite{Blei:2003tn} has proven useful for modeling the structure in large corpri of text documents where each document may contain a mixture of topics. Mixed-membership models have also been used in population genetics \cite{Pritchard:2000uv},  and social network analysis \cite{Airoldi:2008wi}.

We show that using a novel mixed-membership model for genomic classification of primary tumors samples, we achieve a more accurate and complete representation of primary, heterogeneous tumors. Our model achieves the dual purpose of (1) representing each sample as a mixture of genomic subtypes and (2) representing each subtype signature as a sparse set of genomic features that delineate driving oncogenic pathways. We demonstrate our model on RNA expression data from primary gioblastoma samples with thousands of genomic features and hundreds of samples. We show that we recover known subtypes with a sparse set of driving aberrations and that many of the primary samples are in fact mixed.

\section{Model Structure}

Given a data matrix $y \in \RR^{M \times N}$ where the element $y_{ji}$ is an observation of feature $j$ in sample $i$, we would like to find a representation of $y$ such that each sample vector is $y_i = x\theta_i$ where $x \in \RR^{M \times K}$ is a matrix of $K$ cluster centroids and $\theta_i \in \RR^K$ is a distribution over clusters for sample $i$. Furthermore, we would like $x$ to be sparse for purposes of cluster interpretability and generalizability to test cases. In the specific case of cancer subtyping, $y_{ji}$ is a relative gene expression measurement for gene $j$ in patient $i$.

\subsection{Generative Process}\label{subset:gen_process}
The generative process for the data set under the Gaussian Laplace Dirichlet (GLaD) model is:
%maybe mention what GLaD stands for (Gaussian Laplace Dirichlet) ? 
\begin{enumerate}
\item Choose K cluster centroids from $x_k \thicksim \text{Laplace}(\lambda)$
\item For each sample $y_i$:
	\begin{enumerate}
	\item Choose a distribution over clusters $\theta_i \thicksim \text{Dirichlet}(\alpha)$
	\item Sample the expression vector $y_i \thicksim \text{Normal}(x\theta_i, \Sigma)$
	\end{enumerate}
\end{enumerate}

{\bf This model has two levels of sampling.} First, the cluster centroids are chosen once and for all for the entire data set. Then, the expected value of each gene expression vector is a mixture of cluster centroids.  The hierarchical model can be represented as a graphical model (Figure \ref{fig:graphical_model}). The Laplace distribution over $x_{jk}$ enforces sparsity in the subtype matrix \cite{Kabn 2007} and the Dirichlet distribution over $\theta_i$ places a distribution over the $K$ clusters for each sample. For simplicity, we assume that $\Sigma = \sigma^2 I_J$ for the remainder, but this assumption can easily be relaxed for a structured or full covariance matrix.

\begin{figure}[h]
\begin{center}
\includegraphics[width=60mm]{pdf_figs/graphical_model.pdf}
\caption{GLaD Graphical Model}
\label{fig:graphical_model}
\end{center}
\end{figure}

Given this model structure, the joint distribution over the latent and observed variables for a sample given the parameters is
\begin{equation}\label{eqn:jointpdf}
\begin{split}
p\left( x,  \theta_i, y_i; \alpha, \lambda, \sigma^2 \right) &= p\left( x ; \lambda \right) p\left( \theta_i ; \alpha \right) p\left( y_i | x, \theta_i ; \sigma^2 \right)\\
&=\left[ \prod_{j=1}^M \prod_{k=1}^K \frac{1}{2 \lambda} \exp \left( -\frac{|x_{jk}|}{\lambda} \right) \right]\left[ \frac{\Gamma\left( \sum_{k=1}^K \alpha_k \right)} { \prod_{k=1}^K \Gamma \left( \alpha_k \right)}
\prod_{k=1}^K \theta_{ki}^{\alpha_k - 1 } \right]\\
&\quad\ \cdot \left[ \prod_{j=1}^M \left( 2\pi \sigma^2 \right)^{ -\frac{1}{2} } \exp \left( -\frac{1}{2 \sigma^2 }\left( y_{ij} - x_j\theta_i \right)^2 \right)\right]
\end{split}
\end{equation}

Integrating over the latent variables $x$ and $\theta_i$ yields the marginal distribution of a sample expression vector,
\begin{equation}
p\left( y_i ; \alpha, \lambda, \Sigma \right) = \int_x p\left( x ; \lambda \right) \int_{\theta_i} p\left( \theta_i ; \alpha \right) p\left( y_i | x, \theta_i; \Sigma \right) d\theta_i dx
\end{equation}

Finally, assuming independent samples, the probability of the dataset is
% we may only need to assume exchangeable samples
\begin{equation}
p\left( y ; \alpha, \lambda, \Sigma \right) = \prod_{i=1}^N \int_x p\left( x ; \lambda \right) \int_{\theta_i} p\left( \theta_i ; \alpha \right) p\left( y_i | x, \theta_i; \Sigma \right) d\theta_i dx
\end{equation}

\subsection{Comparison to other clustering and topic models}
The GLaD model is distinguished from other clustering and topic models. A classical clustering model would sample a Dirichlet once for the entire data set. Then, each expression vector is allocated to one of the clusters. This is the assumption of the PAM50 method \cite{} and many other all-or-none approaches \cite{}.

A topic model choses from the Dirichlet distribution for sample as we have done here. The conditional distribution (a multinomial in LDA and a Gaussian for Latent Process Decomposition) is then sampled repeatedly. Mixing of subtypes in a given sample is accomplished  combining multiple Gaussian kernels. Instead, in our model, the mixing of clusters happens exclusively in the mean value parameter of the Gaussian distribution resulting in a single Gaussian whose expected value is a mixture of the cluster centroids.

\subsection {Interpretation as  Matrix Factorization}\label{subsec:matfac_interp}
The hierarchical model \eqref{eqn:jointpdf} can be interpreted as a matrix factorization problem with particular constraints. The negative complete data log-likelihood, retaining only terms involving $x$ and $\theta$, is

\begin{equation}\label{eqn:loglike1}
\begin{split}
-LL\left( x, \theta | y \right) = &\sum_{j=1}^M \sum_{i=1}^N \frac{1}{2 \sigma^2} \left( y_{ij} -x_j\theta_i \right)^2  + N\sum_{j=1}^M \sum_{k=1}^K \frac{|x_{jk}|}{\lambda}  \\
& - \sum_{k=1}^K \left( \alpha_k - 1 \right) \sum_{i=1}^N \log \theta_{ki},
\end{split}
\end{equation}
with the constraints $\sum_{k=1}^K \theta_{ki} = 1, \ \forall i$ and $\theta_{ki} \geq 0, \ \forall i,k$. We  aim to minimize \eqref{eqn:loglike1} in order to maximize the complete data log-likelihood. Setting $\sigma^2 = 1$, $\alpha_k=1$ and $\lambda=1$ to eliminate parameters of the statistical model from the more general matrix factorization problem leaves the objective function

\begin{equation}\label{eqn:loglike2}
f\left( x, \theta | y\right) = \sum_{i=1}^N \sum{j=1}^M \frac{1}{2} \left( y_{ij} -x_j\theta_i \right)^2 + N\sum_{j=1}^M \sum_{k=1}^K |x_{jk}|.
\end{equation}

The first term in \eqref{eqn:loglike2} can be viewed as a loss function for the difference between the predicted measurement $\hat{y_i} = x\theta_i$ and the actual data, $y_i$. The second term is a sparsity penalty function where setting $x_{jk}$ to a non-zero value has a cost associated with it proportional to the magnitude of $x_{jk}$ and inversely proportional to $\lambda$. Viewing \eqref{eqn:loglike2} as a Lagrangian function where only the terms involving $x$ and $\theta$ have been retained, the matrix factorization problem can be written as

\begin{equation}\label{eqn:matfac1}
\begin{split}
\underset{x,\theta}{\text{minimize}} & \quad\ \| y - x\theta \|_2^2\\
\text{subject to} & \quad\ \|x\|_1 \leq b\\
& \quad\  \theta^T \mathbf{1}_K=\mathbf{1}_N\\
& \quad\  \theta \geq 0.
\end{split}
\end{equation}
where the inequality on $\theta$ is component-wise for the matrix. The optimization problem \eqref{eqn:matfac1} is biconvex in $x$ and $\theta$.

% [talk about relation to MMF Mackey at el]

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width = 0.4\linewidth]{eps_figs/model1.eps}
%\caption{Graphical Model}
%\label{fig:graph1}
%\end{center}
%\end{figure}

\section{Inference and Parameter Estimation}

Statistical inference for this model focuses on the  posterior distribution over latent variables
\begin{equation}
p(\theta, x | y; \alpha, \lambda, \Sigma) = \frac{ p(\theta, x, y; \alpha, \lambda, \Sigma) } { p(y; \alpha, \lambda, \Sigma) }.
\end{equation}
Since exact inference is intractable for this model, we have developed approximate inference algorithms using iterated conditional modes, Gibbs sampling and a variational approximation. We explore the advantages and disadvantages of each approach after presenting the algorithms.

\subsection{Iterated Conditional Modes (ICM)}

Iterated Conditional Modes (ICM) (Besag 1986) is a a greedy ascent approximation algorithm that proceeds in two steps: (1) assign $\hat{x}^{(p+1)}$ and $\hat{\theta}^{(p+1)}$ by maximizing the complete data likelihood conditional on the current parameter estimates $\hat{\phi}^{(p)}$ and (2) update the parameter estimates $\hat{\phi}^{(p+1 )}$ by maximizing the complete data likelihood conditional on the current latent variable estimates $\hat{x}^{(p)}$ and $\hat{\theta}^{(p)}$.
\begin{algorithm}
\caption{Iterated Conditional Modes Algorithm}
\label{alg:assign_update}
\begin{algorithmic}[1]

\REPEAT

\STATE $\hat{x}^{(p+1)} \leftarrow \arg \max\limits_x \log p\left( x, \hat{\theta}^{(p)}, y ; \hat{\phi}^{(p)} \right)$

\STATE $\hat{\theta}^{(p+1)} \leftarrow \arg \max\limits_\theta \log p\left( \hat{x}^{(p+1)}, \theta, y ; \hat{\phi}^{(p)} \right)$

\STATE $\hat{\phi}^{(p+1)} \leftarrow \arg \max\limits_\phi \log p\left( \hat{x}^{(p+1)}, \hat{\theta}^{(p+1)}, y ; \phi \right)$

\UNTIL {complete data log-likelihood convergence}
\end{algorithmic}
\end{algorithm}

\subsection{Gibbs Sampling}
A Markov chain Monte Carlo (MCMC) sampling approach approximates the posterior distribution $p \left( x, \theta | y; \phi \right)$ by sampling. We use the Metropolis--Hastings (M--H) algorithm to obtain samples from the conditional distributions of each latent variable in turn and then optimize the joint distribution over the model parameters.

\begin{algorithm}[ht]
\caption{Metropolis within Gibbs Algorithm}
\label{alg:metro_gibbs}
\begin{algorithmic}[1]

\REPEAT
\FOR {j = 1 to M}
	\STATE Draw T samples from $p \left( x_{j} | \theta^{(p)}, y; \hat{\phi}^{(p)} \right)$ using M--H
	\STATE Set $x_j^{(p+1)}$ to the sample mean
\ENDFOR

\FOR {i = 1 to N}
	\STATE Draw T samples from $p \left( \theta_i | x^{(p+1)}, y ; \hat{\phi}^{(p)} \right)$ using M--H
	\STATE Set $\theta_i^{(p+1)}$ to the sample mean
\ENDFOR

\STATE Set $\hat{\phi}^{(p+1)} \leftarrow \arg \max\limits_{\phi} p \left( y , x^{(p+1)}, \theta^{(p+1)}; \hat{\phi}^{(p)} \right)$

\UNTIL {sample size is sufficient}
\end{algorithmic}
\end{algorithm}

%TODO: Comment on difference between mean and mode for laplace
The sample mean is only one of many functions that may be applied to the MCMC samples to obtain a point estimate from the posterior distribution. The posterior mode or median may perform better than the mean in certain applications (Kyung et al. 2010). Convergence may be monitored in many ways as well and is an important consideration when running time is a limitation (Cowles \& Carlin 1996). After estimating the parameters, samples can be obtained from the joint posterior distribution $p(x,\theta | y, \hat{\phi}^{(p)})$ using the same procedure except keeping $\hat{\phi}^{(p)}$ fixed.

\subsection{Variational Inference}

We derive a fast non-conjugate variational inference algorithm to estimate the posterior distribution $p(\theta, x | y)$. The algorithm alternates between optimizing the approximate posterior distribution $q(x, \theta)$ and updating the parameters $\phi=\{\alpha, \lambda, \sigma^2\}$. The algorithm performs conjugate gradient ascent on the evidence lower bound (ELBO) $\mathcal{L}(q, p)$ where $q$ is the approximate posterior distribution and $p$ is the true posterior distribution. We describe the algorithm briefly here and give an algorithm in Algorithm~\ref{alg:lapvar} and provide details in Appendix~\ref{sec:varlap}.

The variational algorithm performs coordinate ascent on the auxiliary function by alternating between updating variational distribution and updating the parameters of the model.
\begin{description}
	\item[(E step)] Maximize $\mathcal{L}(q, p)$ with respect to $q$
	\item[(M step)] Maximize $\mathcal{L}(q, p)$ with respect to $\phi$
\end{description}
The variational E-step optimizes the tightness of the approximation to the true posterior distribution and the M-step maximizes the log-likelihood of the data with the misisng data  completed by the variational approximation. In practice, we loop over the variational E-step until the change in the evidence lower bound is small and then move to the E-step \cite{Blei}.

Since we do not have a natural conjugate pair for $q(x)$ and $y$ we cannot derive an analytical update for $q(x)$. Instead, we use the Laplace approximation and numerically approximate $q(x)$ with a Gaussian. We use a variational dirichlet to approximate $q(\theta)$ as in latent Dirichlet allocation \cite{Blei}.

We chose to factorize the variational distribution as
\begin{equation}
	q(\theta,x) = \prod_{j=1}^J q(x_j) \prod_{i=1}^N q(\theta_i).
\end{equation}
This factorization for $q(x)$ is different than the mean-field factoriztion $q(x)=\prod_{k,j} q(x_{jk})$. We find that preserving some of the dependencies between random variables in the variational approximation yields an algorithm that performs much better and avoids trival solutions. More details are provided in Appendix~\ref{sec:varlap}.

\begin{algorithm}[ht]
\caption{Laplace Variational Approximation}
\label{alg:lapvar}
\begin{algorithmic}[1]

\REPEAT
\FOR {j = 1 to J}
	\STATE Set $x_j^{(p+1)} \leftarrow \arg \max\limits_{x_j} f(x_j)$
	\STATE Approximate $q( x_j ) \approx \mathcal{N} \left( x_j^{(p+1)}, -\triangledown^2 f( x_j^{(p+1)} ) \right)$
	\STATE Compute sufficient statistics $E_{q(x_j)} \left[ x_j \right]$, $E_{q(x_j)} \left[ x_j^T x_j \right]$ and $E_{q(x_j)} \left[ | x_j | \right]$
\ENDFOR

\FOR {i = 1 to N}
	\STATE Compute maximum likelihood estimate for $\gamma_i$ where $q(\theta_i; \gamma_i) \thicksim \text{Dir} (\gamma_i)$
	\STATE Compute sufficient statistics $E_{q(\theta_i)} \left[ \theta_i \right]$, $E_{q(\theta_i)} \left[ \theta_i \theta_i^T \right]$ and $E_{q(\theta_i)} [ \log \theta_i ]$
\ENDFOR

\STATE Set $\hat{\phi}^{(p+1)} \leftarrow \arg \max\limits_{\phi} \mathcal{L}(q,p)$

\UNTIL {change in evidence lower bound is small}
\end{algorithmic}
\end{algorithm}

\subsection{Estimating the number of clusters (K)}

The problem of estimating the number of clusters for any clustering method is challenging. Methods for choosing K include the silhouette measure \cite{}, information theoretic measures \cite{}, cross-validation \cite{} and non-parametric Bayes methods \cite{}. Prediction accuracy on held-out data (i.e. prediction -perplexity) has been used in the context of topic-model for documents \cite{}. Here, we use a simple Bayesian Information criterion approach to roughly estimate $K$,

\begin{equation}
\text{BIC}(q,\phi) = -2 \cdot \mathcal{L}(q, p) + \left( K \cdot (J + N + 1) +2 \right) \log(N).
\end{equation} 
We count a parameter for each element of $x$ and $\theta$ as well as for $\lambda$, $\sigma^2$ and $\alpha$.

We do not expect that this method will achieve perfect accuracy, only that it will guide selection towards a reasonable value. We avoid computing the true likelihood and instead use the ELBO to avoid performing the intervals over $x$ and $\theta$. As such, this version is only an approximation. We show results of this metric on tissue mixture data later.

\section{Results on Simulation Data}

We used a simulated data set with known parameter values to assess the performance of the three inference approaches and to assess the model in comparison to similar methods for clustering gene expression data.

Two data sets with 500 features and either 40 or 60 samples were generated from a random Gaussian distribution. We refer to these data sets as "pure" and "mixed". For both, the expected value for the first 20 features is +2 for subtype A and -2 for subtype B. The remaining features have an expected value of zero. The standard deviation is set to 1. For the pure sample data set, 20 samples were generated from subtype A and 20 from subtype B for a total of 40. For the mixed sample data set, the first 40 samples are identical to the pure data set and an additional 20 samples were generated as a 50:50 mixture of A and B for a total of 60 samples.

\subsection{Speed and Accuracy of Inference Algorithms}
We developed three different inference algorithms: ICM, MCMC and variational, to obtain at least point estimates for the parameters and latent variables of the model. We compare those algorithms on the same data set and measure the time to arrive at solutions and the accuracy of those solutions. All three algorithms were initialized with the same parameter values. We terminated ICM and MCMC optimization when the change in complete data log-likelihood was less than 0.1\% and we terminated the variational algorithm when the change in the ELBO \eqref{eqn:elbo} was less than 0.1\%. We ran the MCMC algorithm as described in \ref{} collecting $T=8000$ samples of the posterior distribution and took a 3000 sample burn-in and thinned every 2 samples. All three algorithms are run serially for these tests, but may be made much faster using standard multi-threading or map-reduce methods.

To measure the performance of these algorithms, we compared the true value, $x$, with the point estimate, $\hat{x}$, by the total $L_1$ distance $\sum_j \sum_k | x_{jk} - \hat{x}_{jk} |$  and the true value $\theta$ with the point estimate $\hat{\theta}$ using the classification error rate (CER),
\begin{equation}\label{eqn:cer}
\text{CER}(\theta, \hat{\theta}) = \frac{1}{N} \sum_{i=1}^N \sum_{k=1}^K | \theta_{ki} - \hat{\theta}_{ki} |.
\end{equation}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=5in]{pdf_figs/algorithm_time_compare.pdf}
%the caption mentions that the variational algorithm is returning results faster than both ICM and MCMC but it's not the case. ICM is much faster than both of them.
\caption{Timing comparison for ICM, MCMC and variational algorithms for GLaD model}
\label{fig:alg_compare}
\end{center}
\end{figure}

The ICM algorithm runs very quickly and completes in 42 sec. While the accuracy of the ICM estimate for $x$ is as good as MCMC and the variation algorithm, the CER for $\theta$ is much worse. The MCMC algorithm takes more time at 62,120 sec to complete, but arrives at a similar solution as the ICM algorithm. Assessing convergence for the MCMC algorithm is difficult, though there are several methods to do so \cite{}. Generally, these convergence assessment methods require more sample collection and thus result in slower algorithms. Finally, the variational approach yields good CER and total $L_1$ distance for $x$ in a reasonable amount of time (8,626 sec). We use the variational algorithm for the following experiments.

\subsection{Comparison to LPD and Sparse K-means}

To assess the performance of GLaD with other mixed-membership and sparse clustering algorithms, we compared the CER to sparse K-means and latent process decomposition on pure and mixed data sets. All algorithms were initialized with random starting points. Figure~\ref{fig:model_compare} shows the sample fraction estimates under two simulated data scenarios for Latent Process Decomposition (LPD), Sparse K-means and GLaD. The pure data scenario is shown in the left column and mixed scenario is shown in the right column. All models are estimated with $K=2$ subtypes in both scenarios and the true mixture fractions for each sample are shown in the top row of the figures. Black and white bars denote the two subtype fractions for each sample.

LPD is able to accurately estimate the sample fractions under both the pure and mixed scenarios. The pure data is very accurate with a classification error rate (CER) of 0.116. The mixed CER is also good at 0.202. Sparse K-means is very accurate for the pure scenario data (CER=0.0), but incorrectly classifies all of the mixed samples (CER = 0.33). This is expected since the algorithm was not intended to classify mixed samples. GLaD classifies the pure data with a CER of 0.317 and the mixed data with a CER of 0.196.

\begin{figure}[htbp]
\begin{center}
\includegraphics{pdf_figs/model_compare.pdf}
\caption{Comparison of sample classification by Latent Process Decomposition, Sparse K-means and GLaD }
\label{fig:model_compare}
\end{center}
\end{figure}

It is evident in the figure that LPD has a higher variance in the classification proportion for the mixed samples while GLaD has less variability. However, GLaD errs on the side of more mixing for the pure samples in both the pure and mixed data scenarios.

\section{Results on In-vitro Mixture Data}

While simulation data is useful to obtain quantitative performance metrics with a known data set under ideal conditions, it does not necessarily represent the variability in real gene expression data. Shen-Or et al performed an in-vitro mixture experiment that yields a good data set to test our algorithm while still providing a true positive control. They independently isolated RNA from rat lung, liver and brain tissue. Then the mixed those samples at varying fractions and measured the levels on microarrays with replicates. This data set gives us known mixture fractions and a known number of subtypes with real gene expression microarray data variability.

The initial data set contains 42 samples and 31,099 features. The measurements are the RMA normalized microarray intensities on a $\log_2$ scale. We preprocessed the probes to select only those that have a coefficient of variation greater than 20\% yielding a data matrix that is $1198 \times 42$ in size. We set $K=3$ and estimated the model and variational parameters for the GLaD model on this in-vitro mixture experiment.

Since the true number of mixing components is known to be three for this data set, we used the BIC approach outlined above to estimate $K$. Figure~\ref{fig:rat_bic_K} shows that the BIC method optimally chooses $K=5$, there is a clear minimum at 4. We use this criterion as a rough guide to estimate $K$ in general.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/rat_bic_K.pdf}
\caption{Bayesian Information Criterion metric to estimate K}
\label{fig:rat_bic_K}
\end{center}
\end{figure}

Figure~\ref{fig:rat_mixture_theta} shows true values of $\theta$ and point estimates, $\hat{\theta}$ for three subtypes which we have labeled by their associated tissue types. Clearly, the GLaD model identifies both the pure and mixed samples in the data set.

\begin{figure}[htbp]
\begin{center}
\includegraphics{pdf_figs/rat_mixture_estimate.pdf}
\caption{Posterior estimates for mixture fractions for in-vitro mixed tissue samples.}
\label{fig:rat_mixture_theta}
\end{center}
\end{figure}

The data set is biased towards brain tissue with more samples containing higher fractions of that tissue. The empirical Bayes estimates of $\alpha$ reflect this bias in the data set and thus bias the subtype fraction estimates for the pure samples. In applications where one would prefer not to bias the subtype predictions a-priori, uniform values for $\alpha$ can be set and fixed. However, the estimated $\alpha$ values do provide some information about the entire data set if that is of interest.

The genes within each subtype signature are clearly associated with the corresponding tissue type. Table~\ref{tbl:rat_signature} shows the top 10 genes ranked by absolute value of $x_{j}$ for the three subtypes. All of the top 10 genes are over expressed in the subtype indicated. We have annotated each subtype by the most closely matching tissue type. In the liver signature, there are genes associated with blood components (ITIH4, FGA, MUG1, FGB, PZP) and genes that are associated with liver regeneration (A1BG) and molecule uptake (FABP1). In the brain subtype, the top component is MBP - a myelin sheath component. The signature also includes VSNL1 - a neuronal calcium sensor, SNAP25 - a synaptic vesicle dicing protein and PLP1 - a transmembrane myelin protein. Finally, the lung signature contains SLC24A2 which if defective causes pulmonary alveolar microlithiasis and SFTPD which plays a role in defense response in the lung.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=60mm]{tables/rat_sig.pdf}
\caption{Tissue mixture subtype signatures}
\label{tbl:rat_signature}
\end{center}
\end{figure}

\section{Results on Glioblastoma Tumor Classification}

We applied the GLaD model to microarray measurements of RNA expression levels in glioblastoma (GBM) tumors obtained as part of The Cancer Genome Atlas (TCGA) project \cite{McLendon:2008gr, Verhaak:2010bj}. We used the unified filtered matrix provided in the supplementary materials \cite{Verhaak:2010bj}. This data has been normalized and filtered for the most differentially expressed genes leaving a data matrix with $1740$ features and $202$ primary tumor samples.

We set $K=4$ for this data because the TCGA group showed four canonical subtypes: Classical (CL), Proneural (PN), Neural(NL) and Mesenchymal(MES) \cite{Verhaak:2010bj}. Since these are biopsy samples, we expect that there is some normal cell contamination in the samples. Indeed, normal contamination was also observed by Verhaak et al and clustered within the Neural subtypes. As such, we expect that mixture fractions of the Neural subtype may represent normal cell contamination in our model.

Figure~\ref{fig:gbm_distribution} shows the distribution over subtypes for each sample in the glioblastoma data set. We have labeled the subtypes based on their similarity to the reported subtypes in Verhaak et al.  As expected there is considerable mixing between the classical subtype and other subtypes possibly due to normal cell contamination. We also see significant mixing among the other subtypes indicating that the samples are indeed generally not pure. There are several mesenchymal samples that show little to no mixing with the other subtypes (bottom-left) and these represent more examples of the subtype while there are very few pure pro-neural. This may be due to a distinct oncogenic mechanism at work in the mesenchymal subtype compared to some pathway overlap for the pro neural subtype.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=60mm]{pdf_figs/gbm_theta.pdf}
\caption{Subtype distributions for each of 202 GBM samples}
\label{fig:gbm_distribution}
\end{center}
\end{figure}

Table~\ref{tbl:gbm_signature} shows the genes involved in the four subtypes identified by the GLaD model. Like topic models, each subtype has coefficient values for all features. However, the Laplace distribution prior has the effect of regularizing uninformative coefficients towards zero yielding a sparse signature. We report only the top 15 candidates for the signature here and the full table is available in supplementary information. Genes in green are unregulated in the subtype and genes in blue are down regulated. Genes in bold were also identified by Verhaak et al. for that subtype except UGT8 which was associated with neural-like while we associated it with proneural.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=120mm]{tables/gbm_sig.pdf}
\caption{Sparse subtype signatures for GBM}
\label{tbl:gbm_signature}
\end{center}
\end{figure}

Upregulation of DKK1 and POSTN is associated with the mesenchymal subtype in our analysis. DKK1 plays a role in inhibition of the WNT signaling pathway and the presence of bone lesions in multiple myeloma patents \cite{Tian:2003ht}. Although the entire function of perstonin (POSTN) is not known, it is frequently unregulated in cancers and has been associated with adhesion and differentiation of osteoblasts \cite{Kudo:2007we}. These markers in combination with collagen-specific genes COL1A1 and COL1A2 point to the role of these genes in the precursor to bone and cartilage development this subtype's mesenchymal-like genomic character.

The classical subtype has only one highly unregulated gene - EGFR. Verhaak et al observed this association at both the mRNA and DNA levels in 97\% of classical-like cells indicating its strong association\cite{Verhaak:2010bj}. The proneural actin and microtubule related genes TMSL8 and DCX are over expressed in the proneural subtype \cite{Brown:2003hc}. Finally, the neural-like subtype contains the central-nervous system water channel AQP4 and other genes associated with normal neural cells.

\appendix
\section{Derivation of Variational Inference Algorithm}\label{sec:varlap}

We propose the following factorized variational distribution for the approximate posterior distribution over latent variables
\begin{equation}
	q(x, \theta) = q(x)q(\theta) = \prod_{j=1}^J q(x_{j}) \prod_{i=1}^N q(\theta_i).
\end{equation}

%The optimal variational distributions, given by \cite{Bishop} are of the form
%\begin{eqnarray}
%	q(\theta)  \propto \exp \left\{ E_{q(x)} \left[ \log p(y|x, \theta) p(\theta) \right] \right\} \\
%	q(x) \propto \exp \left\{ E_{q(\theta)} \left[ \log p(y|x, \theta) p(x) \right] \right\}.
%\end{eqnarray}
%
%We see the expected complete log-likelihood appears in the exponents  $f( \theta_i ) = E_{q(x)} \left[ \log p(y|x, \theta_i) p(\theta_i) \right]$ and $f( x_j ) = E_{q(\theta)} \left[\log p(y|x_j, \theta) p(x_j) \right]$. In Laplace variational inference \cite{Wang 2013} we take the Taylor approximation around the maximum of $q(\theta_i)$ and $q(x_j)$ which naturally leads to the following multivariate Gaussian approximations: 
%\begin{eqnarray}
%	q( \hat{ \theta }_i)  \approx \mathcal{N} \left( \hat{\theta}_i, -\triangledown^2 f( \hat{ \theta }_i ) \right)\\
%	q( \hat{ x }_j) \approx \mathcal{N} \left( \hat{x}_j, -\triangledown^2 f( \hat{ x }_j ) \right)
%\end{eqnarray}
%for each sample $i$ and feature $j$.

The log-likelihood of the data can be bound using the usual variational approximation
\begin{equation}
\begin{split}
\log p(y ; \phi)   &= \log \int_x \int_\theta p(y,x,\theta) dx d\theta \\
&= \log \int_x \int_\theta p(y,x,\theta) \frac{ q(x,\theta) } { q(x,\theta) } dx d\theta \\
&\geq \int_x \int_\theta q(x, \theta) \log \frac{ p(y,x,\theta) } { q(x,\theta) } dx d\theta \\
&= E_q \left[ \log p(y, x, \theta) \right] - E_q \left[ \log q(x, \theta) \right] \\
&\triangleq \mathcal{L}(q, p)
\end{split}
\end{equation}

The \emph{evidence lower bound} (ELBO),  $\mathcal{L}(q, p)$, is composed of the expected complete log-likelihood and the entropy of the variational distribution. Maximizing the auxiliary function suggests a tradeoff between opposing the objectives of maximizing the likelihood of the data under the variational distribution and maximizing the entropy of the variational distribution. The auxiliary function can also be seen as minimizing the KL-divergence between the variational distribution and the data distribution, $D_{KL}(q\|p)$.

Writing out the ELBO gives
\begin{equation}\label{eqn:elbo}
\begin{split}
	\mathcal{L}(q,p) &= E_q \left[ \log p(y|x,\theta; \sigma^2) \right] + E_q \left[ \log p(x ; \lambda) \right] + E_q \left[ \log p(\theta; \alpha) \right] \\
	& \quad- E_q \left[ \log q(x) \right] - E_q \left[ \log q(\theta) \right]
\end{split}
\end{equation}

The individual terms of \eqref{eqn:elbo} are shown in \eqref{eqn:llbnd1} -- \eqref{eqn:llbnd5}:
\begin{equation}\label{eqn:llbnd1}
\begin{split}
	E_q \left[ \log p(y|x,\theta; \sigma^2) \right] &= \sum_{i=1}^N \sum_{j=1}^J E_q \left[ \log p(y_{ji} | x_j, \theta_i) \right] \\
	&= \sum_{i=1}^N \sum_{j=1}^J \left( -\frac{1}{2} \log (2 \pi \sigma^2 ) - \frac{1}{2 \sigma^2} E_q \left[ (y_{ji} - x_j \theta_i)^2 \right] \right) \\
	&=   -\frac{NJ}{2} \log (2 \pi \sigma^2 ) - \frac{1}{2 \sigma^2} \sum_{i=1}^N \sum_{j=1}^J \left( E_q [ y_{ji}^2 ]  - 2 y_{ji} E_q [ x_j ] E_q[ \theta_i ] + E_q \left[ (x_j \theta_i)^2 \right] \right) \\
	&= -\frac{NJ}{2} \log (2 \pi \sigma^2 ) - \frac{1}{2 \sigma^2} \sum_{i=1}^N \sum_{j=1}^J \left( E_q [ y_{ji}^2 ]  - 2 y_{ji} E_q [ x_j ] E_q[ \theta_i ] + \text{Tr} ( E_q [  \theta_i \theta_i^T ] E_q [ x_j^T x_j ] ) \right)
\end{split}
\end{equation}
%The last term of \eqref{eqn:llbnd1} is simplified using the trace trick and the linearity of both the trace and expectation operator.
%\begin{eqnarray}
%	E_q \left[ (x_j \theta_i)^2 \right] &= E_q \left[ \text{Tr} (\theta_i^T x_j^T x_j \theta_i) \right] \\
%	& = E_q \left[ \text{Tr} ( \theta_i \theta_i^T x_j^T x_j) \right]  \\
%	& = \text{Tr} ( E_q \left[  \theta_i \theta_i^T x_j^T x_j \right] ) \\
%	& = \text{Tr} ( E_q [  \theta_i \theta_i^T ] E_q [ x_j^T x_j ] ).
%\end{eqnarray}

\begin{equation}\label{eqn:llbnd2}
	E_q \left[ \log p(x; \lambda) \right] = - K J \log(2 \lambda) -  \sum_{j=1}^J \sum_{k=1}^K \frac{ E_q[ | x_{jk} | ] }{ \lambda }
\end{equation}

\begin{equation}\label{eqn:llbnd3}
	E_q \left[ \log p(\theta; \alpha) \right] = N \log \Gamma \left( \sum_k \alpha_k \right) - N \sum_{k=1}^K \log \Gamma(\alpha_k) + \sum_{n=1}^N \sum_{k=1}^K (\alpha_k - 1) E_q[ \log \theta_{kn} ]
\end{equation}

\begin{equation}\label{eqn:llbnd4}
	E_q \left[ \log q(x) \right] = \sum_{j=1}^J \sum_{k=1}^K E_q \left[ \log q(x_{jk}) \right]
\end{equation}

\begin{equation}\label{eqn:llbnd5}
	E_q \left[ \log q(\theta; \gamma) \right] = \sum_{n=1}^N E_q \left[ \log q(\theta_i) \right]
\end{equation}

In order to compute the ELBO, we require these expected values computed with respect to the variational distribution: $E_q(x_j)$, $E_q(x_j^Tx_j)$, $E_q |x_{jk} |$, $E_q(\theta_i)$, $E_q(\theta_i \theta_i^T)$, $E_q(\log \theta_i)$. We aim to find a variational distribution that makes the computation of these expectations simple, so we choose the variational distribution,
\begin{equation}
	q(x, \theta) =\prod_{j=1}^J q(x_j) \prod_{i=1}^N q(\theta_i),
\end{equation}
where $q(x_j) \thicksim \mathcal{N} \left( \hat{x}_{j}, {-\triangledown f( \hat{x}_j ) }^{-1} \right)$ and $q(\theta_i) \thicksim \text{Dir}(\gamma_i)$. We use the Laplace approximation for $q(x_j)$ and the Dirichlet distribution for $q(\theta_i)$ to allow for the analytical computation of the required expected values.

\subsection{Variational Distribution $q(\theta_i)$}

The probability density function for $q(\theta_i) \thicksim \text{Dir}(\gamma_i)$ is 
\begin{equation}
	q(\theta_i; \gamma_i) = \log \Gamma \left( \sum_k \gamma_{ki} \right) - \sum_k \log \Gamma (\gamma_{ki}) + \sum_k (\gamma_{ki} - 1) \left( \psi(\gamma_{ki}) - \psi ( \gamma_{0i} ) \right)
\end{equation}
where we define $\gamma_{0i} = \sum_k \gamma_{ki}$.

The expected values needed for the ELBO can then be easily found to be for the Dirichlet
\begin{eqnarray}
	E_q[ \log \theta_i ] =& \psi(\gamma_{ki}) - \psi ( \gamma_{0i} ) \\
	E_q[\theta_i] =& \frac{ \gamma_{ki} } { \gamma_{0i} } \\
	E_q[\theta_i \theta_i^T] =& \frac{ \gamma_i \gamma_i^T + \text{diag}(\gamma_i) } { \gamma_{0i} ( 1 + \gamma_{0i} ) }
\end{eqnarray}

Note that it is difficult to use the Laplace approximation for $q(\theta_i)$ as we have done for $q(x_j)$.  The normal distribution has support over the real line and presents complications for the computation of $E_q[\log \theta_i]$.

\subsection{Variational Distribution $q( x_j )$}

We use the Laplace approximation for the variational distribution $q(x_j)$, 
\begin{equation}
	q(x_j) \approx \mathcal{N} \left( \hat{x}_j, -\triangledown^2 f(x_j)^{-1} \right)
\end{equation}
where $\hat{x}_j$ maximizes the ELBO \eqref{eqn:elbo} and $f(x_j)$ isolates the terms of the ELBO that depend on $x_j$. Under the Laplace approximation, the Hessian of the ELBO is used to estimate the covariance matrix.

Isolating the terms in ELBO that depend on $x_j$ gives
\begin{equation}\label{eqn:f_x}
	f( x_j ) = \frac{1}{\sigma^2}\sum_{i=1}^N y_{ij}E_{q} [\theta_i] x_j - \frac{1}{2 \sigma^2}\sum_{i=1}^N \text{Tr}(E_q[ \theta_i \theta_i^T ] x_j^T x_j) -\frac{1^T_K | x_j | } { \lambda }.
\end{equation}

The gradient of \eqref{eqn:f_x} is
\begin{equation}
	\triangledown f( x_j ) =  \frac{1} { \sigma^2 } \sum_{i=1}^N y_{ij}E_{q} [\theta_i] - \frac{1}{\sigma^2}\sum_{i=1}^N E_q[ \theta_i \theta_i^T ] x_j^T -\frac{ \text{sgn} (x_j) } { \lambda }.
\end{equation}

The Hessian of \eqref{eqn:f_x} is
\begin{equation}
	\triangledown^2 f( x_j )=  - \frac{1}{\sigma^2}\sum_{i=1}^N E_q[ \theta_i \theta_i^T ].
\end{equation}
We have used the fact that $\frac{d\text{sgn} (x) } { dx } = \delta(x)$ where $\delta(x)$ is the Kronecker delta function. Since this function is zero everywhere except for at the point $x=0$ we approximate it to be zero everywhere for the purpose of estimating the Hessian.

 The expected values needed for the ELBO are
\begin{eqnarray}
	E_q[ |x_j| ] =& \sqrt{ \frac{ 2 \Sigma(\hat{x}_j)_{kk} } {\pi} } \exp \left( -\frac{ \hat{x}_{jk}^2 } { 2 \Sigma(\hat{x}_j)_{kk} } \right) + \hat{x}_{jk} \left( 1 - 2 \Phi \left( - \frac{ \hat{x}_{jk} } { \sqrt{ \Sigma(\hat{x}_j)_{kk} } } \right) \right)\\
	E_q[\theta_i] =& \frac{ \gamma_{ki} } { \gamma_{0i} } \\
	E_q[\theta_i \theta_i^T] \approx& \frac{ \gamma_i \gamma_i^T + \text{diag}(\gamma_i) } { \gamma_{0i} ( 1 + \gamma_{0i} ) }
\end{eqnarray}
where $\Phi(x)$ is the standard normal cumulative distribution function.

We see that the normal approximation for $q(x_j)$ allows for the analytical computation of $E_q(|x_j|)$ because $|x_j|$ has a folded normal distribution. In our notation, we define $\Sigma(\hat{x}_j)_{kk}$ to be the element in the k-th row and column of the covariance matrix estimate $\Sigma(\hat{x}_j)  \triangleq -\triangledown^2 f( \hat{x}_j )^{-1}$.

\subsection{Optimizing Model Parameters $\phi=\{\alpha, \lambda, \sigma^2\}$}

The M-step in the variational EM algorithm requires us to maximize the ELBO with respect to each model parameter.

\subsubsection{Optimizing $\lambda$} The ELBO with respect to $\lambda$ is
\begin{equation}
\mathcal{L}_{[\lambda]} = - K J \log(2 \lambda) -  \sum_{j=1}^J \sum_{k=1}^K \frac{ E_q[ | x_{jk} | ] }{ \lambda }.
\end{equation}

Taking the derivative with respect to $\lambda$ and setting it equal to zero gives an analytical update,
\begin{equation}
	\lambda \leftarrow  \frac{1}{KJ} \sum_{j=1}^J \sum_{k=1}^K E_q[ | x_{jk} | ],
\end{equation}
where $E_q[ | x_{jk} | ]$ is obtained from the folded Normal distribution of the Laplace approximation.

\subsubsection{Optimizing $\sigma^2$} The ELBO with respect to $\sigma^2$ is
\begin{equation}
\mathcal{L}_{[\sigma^2]} = \sum_{i=1}^N \sum_{j=1}^J \left( -\frac{1}{2} \log (2 \pi \sigma^2 ) - \frac{1}{2 \sigma^2} E_q \left[ (y_{ji} - x_j \theta_i)^2 \right] \right).
\end{equation}

Taking the derivative with respect to $\sigma^2$ and setting it equal to zero gives an analytical update,
\begin{equation}
		\sigma^2 \leftarrow  \frac{1}{NJ} \sum_{i=1}^N \sum_{j=1}^J  E_q \left[ (y_{ji} - x_j \theta_i)^2 \right].
\end{equation}

Expanding the expected sum-of-squares error term gives
\begin{equation}
	\sigma^2 \leftarrow  \frac{1}{NJ} \sum_{i=1}^N \sum_{j=1}^J  \left( E_q [ y_{ji}^2 ]  - 2 y_{ji} E_q [ x_j ] E_q[ \theta_i ] + \text{Tr} ( E_q [  \theta_i \theta_i^T ] E_q [ x_j^T x_j ] ) \right)
\end{equation}

\subsubsection{Optimizing $\alpha$} 
Only one term in the ELBO involves $\alpha$ and we optimize the ELBO numerically. We simply maximize $E_q \left[ \log p(\theta; \alpha) \right]$ with respect to $\alpha$ using standard constrained numerical optimization methods. While it is possible to use the structure of the Hessian of the ELBO with respect to $\alpha$ to derive a fast Newton-Raphson update step \cite{Blei}, we opt to use packaged optimization software here.

\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}

### 2013-08-27 doc/bioinformatics2013/RVD2.tex
Paper draft for RVD2. Up to Bayesian Hypothesis Testing part.

\documentclass[11pt,reqno]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{algorithm, algorithmic}
\usepackage{graphicx}

\usepackage{amssymb}
\usepackage{amsmath}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcommand{\RR}{I\!\!R} %real numbers
\DeclareMathOperator{\diag}{diag}

\title{RVD2}
\author{}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\maketitle

\section{Introduction}

\section{Materials and Methods}

\subsection{Materials}
\subsection{Model structure}\

Consider the following hierarchical model as the generative process for RVD2 model

\begin{align}
\mu_j|\mu_0,M_0 &\thicksim \text{Beta}(a, b)\\
\theta_{ij}|\mu_j,M_j &\thicksim \text{Beta}(\mu_j,M_j)\\
r_{ij}|\theta_{ij},n_{ij} &\thicksim \text{Binomial}(\theta_{ij}, n_{ij})
\end{align}

For each genomic location,  $j=1,\ldots,J$, and each repilcate  $i=1,\ldots,N$, $n_{ij}$ is the total counts, while $r_{ij}$ stands for the non-reference read counts. Variable $r_{ij}$ is drawn from a binomial distribution with probability parameter $\theta_{ij}$ and total count $n_{ij}$.  The error probability $\theta_{ij}$ has a prior beta distribution with position -specific rate parameter $mu_j$ and precision parameter $M_j$.

The position error rate $\mu_j$ has a beta distribution as hypoprior distribution with parameter $\mu_0$ and $M_0$. This is to ensure that the error rate is between 0 and 1. The prior is useful for situations when there is a significant minor allele.

This model only addresses error/reference positions  and does not model individual nucleotide frequencies.

Given this model, the inferential object of interest here is the joint posterior distribution over the latent variable $p(\theta,\mu|r,n)$.

The joint posterior distribution is
\begin{equation}
\begin{split}
 p(\theta, \mu | r,n; \mu_0,M_0,M) &= \frac{p(r,\theta,\mu;\mu_0,M_0,M)}{p(r;\mu_0,M_0,M)}\\
 & = \prod_{i=1}^N\prod_{j=1}^J\frac{p(r_{ij}|\theta_{ij},n_{ij})p(\theta_{ij}|\mu_j,M_j)p(\mu_j|\mu_0,M_0)}{p(r_{ij};n_{ij},M,\mu_0,M_0)}
\end{split} 
\end{equation}

However, the normalization factor in the posterior distribution,
\begin{equation}
 {p(r_{ij};n_{ij},M,\mu_0,M_0)}=\int_{\theta_{ij}}\int_{\mu_j}p(r_{ij}\theta_{ij},n_{ij})p(\theta_{ij}|\mu_j,M_j)p(\mu_j|\mu_0,M_0)d\theta_{ij}d\mu_j
\end{equation}

is computationally intractable due to the coupling between $\theta_{ij}$ and $\mu_j$.

Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as discribed in Metropolis-within-Gibbs section.

\section{Inference algorithm and variant calling}
\subsection{Model parameter estimation}\

Method of moments is used to find the optimal model parameters and hyperparameters. From the feature of beta distribution
The estimate of

The estimated error probability is calculated from the observed sample
\begin{equation}
 \theta_{ij}=\frac{r_{ij}}{n_{ij}}
\end{equation}

The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment esitmates for the variance of $\theta_{ij}$.
\begin{equation}
 \mu_j=E(\theta_{ij}|\mu_j,M_j)
\end{equation}
\begin{equation}
 M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
\end{equation}

Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
\begin{equation}
 \mu_0=E(\mu_j|\mu_0,M_0)
\end{equation}
\begin{equation}
 M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
\end{equation}

\subsection{Metropolis-within-Gibbs sampling}\

Because of the computation complexition of the nomalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approch is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$. Specifically Metropo-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.

\begin{algorithm}[ht]
\caption{Metropolis within Gibbs Algorithm}
\label{alg:metro_gibbs}
\begin{algorithmic}[1]
\REPEAT
\FOR {j = 1 to J}
  \FOR {i =1 to N}
	\STATE Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs sampling
  \ENDFOR
\ENDFOR

\FOR {j = 1 to J}
	\STATE Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
	\STATE Set $mu_j$ to the sample median
\ENDFOR
\UNTIL {sample size is sufficient}

\end{algorithmic}
\end{algorithm}

\subsection{Bayesian Hypothesis Testing}

\section{Result}
Figure 1 Overall of the process
\subsection{Simulation Results}
\subsection{Comparison Results on Synthetic DNA}
Figure 3 ROC comparison with other methods: varscan/samtools/rvd
\subsection{Empirical Results on Multiple Sclerosis Data}
We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that

Table: Table of Variants
\subsection{Performance with read depth}
Figure 4 ROC by read depth
picard to thin segment data, showing performance as read depth decreasing

\bibliographystyle{apalike}
\bibliography{bioinfo}
\end{document}

### 2013-08-28 doc/bioinformatics2013/RVD2.tex
paper draft with update on statistic inference.

@@
 \usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{algorithm, algorithmic}
+%\usepackage{algorithm, algorithmic}
+
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+
 \usepackage{graphicx}
 
+\usepackage{verbatim}
 \usepackage{amssymb}
@@
 
-For each genomic location,  $j=1,\ldots,J$, and each repilcate  $i=1,\ldots,N$, $n_{ij}$ is the total counts, while $r_{ij}$ stands for the non-reference read counts.
+For each genomic location,  $j=1,\ldots,J$, and each replicate  $i=1,\ldots,N$, $n_{ij}$ is the total counts, while $r_{ij}$ stands for the non-reference read counts.
 Variable $r_{ij}$ is drawn from a binomial distribution with probability parameter $\theta_{ij}$ and total count $n_{ij}$.
@@
 
-The position error rate $\mu_j$ has a beta distribution as hypoprior distribution with parameter $\mu_0$ and $M_0$.
+The average positional error rate $\mu_j$ across replicates has a beta distribution as hyperprior distribution with parameter $\mu_0$ and $M_0$.
 This is to ensure that the error rate is between 0 and 1.
@@
 
-Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as discribed in Metropolis-within-Gibbs section.
+Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as described in Metropolis-within-Gibbs section.
 
@@
 
-The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment esitmates for the variance of $\theta_{ij}$.
+The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
 \begin{equation}
@@
 
-Because of the computation complexition of the nomalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approch is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$.
-Specifically Metropo-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.
+Because of the computation complexities of the normalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approach is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$.
+Specifically Metropolis-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.
 
@@
 \begin{algorithmic}[1]
-\REPEAT
-\FOR {j = 1 to J}
-  \FOR {i =1 to N}
-	\STATE Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs sampling
-  \ENDFOR
-\ENDFOR
+\Repeat
+\For {j = 1 to J}
+  \For {i =1 to N}
+	\State Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs; sampling
+  \EndFor
+\EndFor
 
-\FOR {j = 1 to J}
-	\STATE Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
-	\STATE Set $mu_j$ to the sample median
-\ENDFOR
-\UNTIL {sample size is sufficient}
+\For {j = 1 to J}
+	\State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H;
+	\State Set $mu_j$ to the sample median;
+\EndFor
+\Until {sample size is sufficient}
 
@@
 
-\subsection{Bayesian Hypothesis Testing}
+\subsection{Statistic Inference}\
+
+Full Bayesian Hypothesis Testing is done to test rare variants in the model.
+
+\subsubsection{Posterior distribution difference sampling}\
+
+Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample.
+In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done
+\begin{algorithm}
+\caption{Posterior distribution difference sampling}
+\begin{algorithmic}[1]
+\Procedure{}{}
+\Repeat
+\State  Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
+\State  Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
+\State Subtract the former from the later to get one sample for the posterior difference distribution 
+\Until {Sample size is sufficient}
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.
+
+$p \left( \mu_{testing}-\mu_{ref} |r,n; \mu_0,M_0\right)$
+
+\subsubsection{Bayesian Hypothesis Testing}
+To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
+\begin{align}
+ H_1: \mu_{testing}-\mu_{ref}\geq\tau \\
+ H_2: \mu_{testing}-\mu_{ref}<\tau
+\end{align}
+
+We can calculate the posterior probability of the hypothesis $H_1$ being true by integrating the posterior density over the correct region:
+
+\begin{equation}
+ p(H1:\mu_{testing}-\mu_{ref}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
+\end{equation}
+
+We accept the hypothesis $H_1$ if this posterior probability is no lower than 95\%, which means we are 95\% sure that there is a mutation in this position.
+
+ROC curve is plotted to evaluate the performance of the detection algorithm with different threshold for different mutation rate.
+
+\subsubsection{Uniformity Testing}
 

### 2013-08-28 doc/bioinformatics2013/RVD2.tex
Edited model structure section of manuscript.

@@
 
-\title{RVD2}
+\title{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
 \author{}
@@
 
-\section{Materials and Methods}
+\section{Model structure}\
 
-\subsection{Materials}
-\subsection{Model structure}\
+Given a data matrix $r \in \RR^{J \times N}$ where the element $r_{ji}$ is the count of the reads that contain a non-reference base at position $j$ in experimental replicate $i$.
+The total count of reads at position $i$ in replicate $i$ is $n_{ji}$.
+A typical data set is comprised of case and control data sets, $\{ r^{\text{case}} , n^{\text{case}} \}$ and  $\{ r^{\text{control}} , n^{\text{control}} \}$.
 
-Consider the following hierarchical model as the generative process for RVD2 model
+We have developed a hierarchical Bayesian model for this data set because of the nested structure of noise in the data.
+There is sampling error due to the fact that only a finite number of reads aligned and mapped for each location.
+There is inter-replicate errors due to experimental protocol and reproducibility.
+Finally there is sequence context error due to the fidelity of the DNA polymerase enzymes in synthesizing a base.
+These errors have been shown to depend heavily on the neighboring DNA bases and whether the base is at the end of a molecule or at the center of a molecule.
 
-\begin{align}
-\mu_j|\mu_0,M_0 &\thicksim \text{Beta}(a, b)\\
-\theta_{ij}|\mu_j,M_j &\thicksim \text{Beta}(\mu_j,M_j)\\
-r_{ij}|\theta_{ij},n_{ij} &\thicksim \text{Binomial}(\theta_{ij}, n_{ij})
-\end{align}
+\subsection{Generative Process} We have codified the statistical structure of these errors in the following generative process (we have used the mean-precision parameterized of the beta distribution):
 
-For each genomic location,  $j=1,\ldots,J$, and each replicate  $i=1,\ldots,N$, $n_{ij}$ is the total counts, while $r_{ij}$ stands for the non-reference read counts.
-Variable $r_{ij}$ is drawn from a binomial distribution with probability parameter $\theta_{ij}$ and total count $n_{ij}$.
-The error probability $\theta_{ij}$ has a prior beta distribution with position -specific rate parameter $mu_j$ and precision parameter $M_j$.
+\begin{enumerate}
+	\item Choose a baseline error rate parameter $\mu_0$ and a inter-location precision parameter $M_0$.
+	\item For each location $j$:
+	\begin{enumerate}
+		\item Sample a location specific error rate parameter $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
+		\item Choose an inter-replicate precision parameter $M_j$
+		\item For each replicate $i$:
+		\begin{enumerate}
+			\item Sample an error rate $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ for location $j$ in sequencing replicate $i$
+			\item Sample the non-reference read count $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
+		\end{enumerate}
+	\end{enumerate}
+\end{enumerate}
 
-The average positional error rate $\mu_j$ across replicates has a beta distribution as hyperprior distribution with parameter $\mu_0$ and $M_0$.
-This is to ensure that the error rate is between 0 and 1.
-The prior is useful for situations when there is a significant minor allele.
+This model has three levels of sampling.
+First the baseline error rate across all locations and all replicates is chosen once and for all for the entire data set.
+Then, based on that baseline error rate, location-specific error rates are sampled.
+Based on the location specific error rate, replicate-specific error rates are selected.
+Finally, the observable data, the count of the non-reference reads is sampled.
 
-This model only addresses error/reference positions  and does not model individual nucleotide frequencies.
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
+\caption{RVD2 Graphical Model}
+\label{fig:graphical_model}
+\end{center}
+\end{figure}
 
-Given this model, the inferential object of interest here is the joint posterior distribution over the latent variable $p(\theta,\mu|r,n)$.
+% Describe graphical model framework here
 
-The joint posterior distribution is
-\begin{equation}
-\begin{split}
- p(\theta, \mu | r,n; \mu_0,M_0,M) &= \frac{p(r,\theta,\mu;\mu_0,M_0,M)}{p(r;\mu_0,M_0,M)}\\
- & = \prod_{i=1}^N\prod_{j=1}^J\frac{p(r_{ij}|\theta_{ij},n_{ij})p(\theta_{ij}|\mu_j,M_j)p(\mu_j|\mu_0,M_0)}{p(r_{ij};n_{ij},M,\mu_0,M_0)}
-\end{split} 
+Given this model structure, the joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
+
+\begin{equation}\label{eqn:jointpdf}
+p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right)
 \end{equation}
 
-However, the normalization factor in the posterior distribution,
+\begin{gather}
+= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1} \\
+\cdot \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1} \\
+\cdot \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}
+\end{gather}
+
+Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data at location $j$ in replicate $i$, 
 \begin{equation}
- {p(r_{ij};n_{ij},M,\mu_0,M_0)}=\int_{\theta_{ij}}\int_{\mu_j}p(r_{ij}\theta_{ij},n_{ij})p(\theta_{ij}|\mu_j,M_j)p(\mu_j|\mu_0,M_0)d\theta_{ij}d\mu_j
+p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j
 \end{equation}
 
-is computationally intractable due to the coupling between $\theta_{ij}$ and $\mu_j$.
+Finally, assuming the conditional independence structure in the graphical model in Figure~\ref{fig:graphical_model}, the likelihood of the data set is
+
+\begin{equation}
+p \left( r ; \mu_0, M_0, M \right) = \prod_{j=1}^J p\left( \mu_j; \mu_0, M_0 \right) \prod_{i=1}^N p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) 
+\end{equation}
 
@@
 
-\section{Inference algorithm and variant calling}
+\section{Inference and Hypthesis Testing}
 \subsection{Model parameter estimation}\

### 2013-08-31 doc/bioinformatics2013/RVD2.tex
Paper draft. Major update is on the goodness of fit test.

@@
 In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done
-\begin{algorithm}
-\caption{Posterior distribution difference sampling}
-\begin{algorithmic}[1]
-\Procedure{}{}
-\Repeat
-\State  Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
-\State  Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
-\State Subtract the former from the later to get one sample for the posterior difference distribution 
-\Until {Sample size is sufficient}
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
+
+\begin{enumerate}
+ \item Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
+ \item Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
+ \item Subtract the former from the later to get one sample for the posterior difference distribution 
+\end{enumerate}
+Repeat the above three steps until sample size is sufficient.
 
 With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.
-
-$p \left( \mu_{testing}-\mu_{ref} |r,n; \mu_0,M_0\right)$
 
@@
 \begin{align}
- H_1: \mu_{testing}-\mu_{ref}\geq\tau \\
- H_2: \mu_{testing}-\mu_{ref}<\tau
+ H_1: \mu_{case}-\mu_{control}\geq\tau \\
+ H_2: \mu_{case}-\mu_{control}<\tau
 \end{align}
@@
 \begin{equation}
- p(H1:\mu_{testing}-\mu_{ref}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
+ p(H1:\mu_{case}-\mu_{control}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
 \end{equation}
@@
 
-\subsubsection{Uniformity Testing}
+\subsubsection{Goodness of fit Testing for Uniformity}\
+
+We assume that at one position in one experimental replicate, the non-reference reads follows multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.
+
+The non-reference reads can be because of either mutation or various errors including sampling error, inter-replicated error, $etc$.
+In theory, the non-reference read counts caused by non-oriented errors should be uniformly distributed among three non-reference categories, namely $p_1=p_2=p_3=1/3$.
+However, the distribution of non-reference read counts among three non-reference categories caused by mutation theoretically is biased.
+
+The null hypothesis can be denoted as
+\begin{align}
+ H_0: p=p_0
+\end{align}
+where
+\begin{align} 
+ p_0=(p_1,p_2,p_3), p_1=p_2=p_3=1/3
+\end{align}
+
+The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).
+Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is
+\begin{equation}
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
+\end{equation}
+
+Person's $X^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family.
+Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.
+
+Particulaly in our case, Let $(r_1,r_2,r_3)$ denotes the non-reference base frequencies and $\sum_{i=1}^3 r_i = r$.
+The expected frquencies are $(rp_1, rp_2, rp_3)$.
+Then the power-divergence statistics can be written as
+
+\begin{equation}
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^3 r_i \left[\left(\frac{r_i}{rp_i}\right)^\lambda-1\right];\lambda \in R
+\end{equation}
+
+We use $\lambda =2/3$ for the goodness of fit testing.
+
+For each position we have six power-divergence statistics from 6 experimental replicates.
+We use Fisher's method to combines the six statistics into a single one for inference.
+Bonferroni correction is applied to maintaining the familywise error rate (FWER).
+The significance level is chosen as 0.05.
+
+Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.
 

### 2013-09-02 doc/bioinformatics2013/RVD2.tex
edited methods.

@@
 
-\title{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
+\algnewcommand{\Inputs}[1]{%
+  \State \textbf{Inputs:}
+  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
+}
+\algnewcommand{\Initialize}[1]{%
+  \State \textbf{Initialize:}
+  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
+}
+
+\title[RVD2]{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
 \author{}
@@
 
-Instead of attempting to directly compute the posterior distribution, we instead sample it using a Metropolis-Hastings within Gibbs scheme as described in Metropolis-within-Gibbs section.
+Instead of attempting to directly compute the posterior distribution, we sample it using a Metropolis-Hastings within Gibbs scheme as described in Section~\ref{sec:algorithm}.
 
-\section{Inference and Hypthesis Testing}
-\subsection{Model parameter estimation}\
+\section{Inference and Hypothesis Testing}
 
-Method of moments is used to find the optimal model parameters and hyperparameters. From the feature of beta distribution
-The estimate of
+The primary object of inference in this model the joint posterior distribution function over the latent variables,
+\begin{equation}
+	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
+\end{equation}
+where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.
+
+We use a Metropolis-within-Gibbs sampling approach shown in Algorithm~\ref{alg:metro_gibbs} to approximate the posterior distribution.
+The model parameters are initialized using method-of-moments.
+Given those parameter estimates we first sample from the marginal posterior distribution for $\mu_j$ given it's Markov blanket using Metropolis Hasting.
+Then, we sample from the marginal posterior distribution for $\theta_{ji}$ given it's Markov blanket.
+
+\begin{algorithm}[ht]
+\caption{Metropolis-within-Gibbs Algorithm}
+\label{alg:metro_gibbs}
+\begin{algorithmic}[1]
+
+\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
+\Repeat
+\For {each position j} \Comment{Sample $\mu_j$}
+  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
+  \State Set $\mu_j$ to the sample median for the T samples
+
+  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
+	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
+  \EndFor
+
+\EndFor
+\Until {convergence and sample size sufficient}
+\end{algorithmic}
+\end{algorithm}
+
+Once we have obtained an estimate of the posterior distribution, we test the hypothesis that there is a sub-population of DNA molecules in the pool with a non-reference base at position $j, \forall j \in 1, \ldots, J$.
+First, we test the hypothesis that the error rate in the case data, $\mu_j^{\text{case}}$ is greater than the error rate in the control data, $\mu_j^{\text{control}}$.
+If so, then we test the hypothesis that the high error rate is due to excess reads with a particular non-reference base rather than a uniform distribution over all possible non-reference bases.
+
+\subsection{Initialization}
+
+The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure.
+Essentially, we set the parameter value (population moment) equal to the sample moment.
+In some cases, the method-of-moments estimate is outside the parameter space.
+Then, we XXX.
 
@@
 
-\subsection{Metropolis-within-Gibbs sampling}\
+\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
 
-Because of the computation complexities of the normalization factor in the posterior probability, A Markov Chain Monte Carlo (MCMC) sampling approach is employed to approximate the posterior distribution $p(\theta, \mu | r,n; \mu_0,M_0,M)$.
-Specifically Metropolis-within-Gibbs algorithm is used to obtain the posterior distribution of each latent variable.
+We are able to draw samples from $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
 
-\begin{algorithm}[ht]
-\caption{Metropolis within Gibbs Algorithm}
-\label{alg:metro_gibbs}
-\begin{algorithmic}[1]
-\Repeat
-\For {j = 1 to J}
-  \For {i =1 to N}
-	\State Draw 1 samples from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ using Gibbs; sampling
-  \EndFor
-\EndFor
+\subsection{Sampling from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$}
 
-\For {j = 1 to J}
-	\State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H;
-	\State Set $mu_j$ to the sample median;
-\EndFor
-\Until {sample size is sufficient}
+\subsection{Bayesian hypothesis test for high error rate}
 
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Statistic Inference}\
-
-Full Bayesian Hypothesis Testing is done to test rare variants in the model.
-
-\subsubsection{Posterior distribution difference sampling}\
-
-Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample.
-In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done
-
-\begin{enumerate}
- \item Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
- \item Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
- \item Subtract the former from the later to get one sample for the posterior difference distribution 
-\end{enumerate}
-Repeat the above three steps until sample size is sufficient.
-
-With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.
-
-\subsubsection{Bayesian Hypothesis Testing}
 To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
@@
 
-\subsubsection{Goodness of fit Testing for Uniformity}\
+\subsection{$\chi^2$ test for non-uniform base distribution}
 
@@
 
-\section{Result}
+\subsubsection{Posterior distribution difference sampling}\
+
+Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample.
+In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done
+
+\begin{enumerate}
+ \item Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
+ \item Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
+ \item Subtract the former from the later to get one sample for the posterior difference distribution 
+\end{enumerate}
+Repeat the above three steps until sample size is sufficient.
+
+With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.
+
+\section{Results}
 Figure 1 Overall of the process

### 2013-09-03 doc/bioinformatics2013/RVD2.tex
Paper draft update on Moments derivation

@@
 \subsection{Initialization}
-
 The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure.
@@
 \end{equation}
-
 The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
@@
 
+\section{Appendix}
+\subsection{Moments derivation}
+The probability density function of beta distribution, for $x\in[0,1]$, and shape real parameter $\alpha>1$ and $\beta>1$ is 
+\begin{equation}
+\begin{split}
+  f(x;\alpha,\beta) &= \frac{x^{\alpha-1}(1-x)^{\beta-1}}{\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx}\\ &=\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}x^{\alpha-1}(1-x)^{\beta-1}\\ &=\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}
+\end{split}
+\end{equation}
+where $\Gamma(z)$ is the Gammma function, $B(\alpha,\beta)$ is the beta function, and
+\begin{equation}
+\begin{split}
+ B(\alpha,\beta) &=\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx\\  &= \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}
+\end{split}
+\end{equation}
+
+The expected value (mean) of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
+\begin{equation}
+\begin{split}
+  E[X] &= \int_0^1xf(x;\alpha,\beta)dx\\ &= \int_0^1x\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx\\ &= \frac{\int_0^1x^{\alpha}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}\\ &= \frac{B(\alpha+1,\beta)}{B(\alpha,\beta)}\\ &= \frac{\Gamma(\alpha+1)\Gamma(\beta)}{\Gamma(\alpha+\beta+1)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha}{\alpha+\beta}
+\end{split}
+\end{equation}
+Note that Gamma function satisfies the functional equation
+\begin{equation}
+ \Gamma(z) = z\Gamma(z)
+\end{equation}
+
+The variance of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
+\begin{equation}
+\begin{split}
+  var(X) &= E[(X-E(X))^2]\\ &= E(X^2)-[E(x)]^2\\ &=\int_0^1x^2\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\int_0^1x^{\alpha+1}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{B(\alpha+2,\beta)}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\Gamma(\alpha+2)\Gamma(\beta)}{\Gamma(\alpha+\beta+2)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta+1)(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha}{(\alpha+\beta+1)(\alpha+\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\alpha\beta}{(\alpha+\beta+1)(\alpha+\beta)^2}
+\end{split}
+\end{equation}
+
+Instead of the standard scale parameter $\alpha$ and $\beta$, we use the mean $\mu$ and sample-size paramenter $M$ for the Beta distribution in our hierarchical model.
+\begin{equation}
+\begin{split}
+  f(x;\mu,M) &=\frac{\Gamma(M)}{\Gamma(M\mu)\Gamma(M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}\\ &= \frac{1}{B(M\mu,M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}
+\end{split}
+\end{equation}
+where 
+\begin{equation}
+ \mu = \frac{\alpha}{\alpha+\beta}
+\end{equation}
+\begin{equation}
+ M = \alpha+\beta
+\end{equation}
+From Equation 24 and Equation 26 it can be seen that
+\begin{equation}
+ E(x|\mu,M) = \mu
+\end{equation}
+\begin{equation}
+ Var(x|\mu,M) = \frac{\mu(1-\mu)}{M+1}
+\end{equation}
+
 \bibliographystyle{apalike}

### 2013-09-05 doc/bioinformatics2013/RVD2.tex
Paper draft update---modified derivation of parameter initialization using MOM

--- 2013-09-02 doc/bioinformatics2013/RVD2.tex
@@
 \begin{equation}
- \theta_{ij}=\frac{r_{ij}}{n_{ij}}
+ \tilde{\theta}_{ij}=\frac{r_{ij}}{n_{ij}}
 \end{equation}
-
 The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
@@
 \begin{equation}
+ \tilde{\mu}_j =\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{equation}
+
+\begin{equation}
  M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
 \end{equation}
-
+\begin{equation}
+ \tilde{M}_j = \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
+\end{equation}
 Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
@@
 \begin{equation}
+ \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
+\end{equation}
+
+\begin{equation}
  M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
+\end{equation}
+\begin{equation}
+ \tilde{M}_0 = \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
 \end{equation}
@@
 
+\section{Appendix}
+\subsection{Parameter initialization derivation}
+The model parameters and latent variables are initialized using method of moments by equating population moment to sample moment.
+
+The estimated error probability is calculated from the observed sample
+\subsubsection{$\theta_{ji}$}\
+\begin{equation}
+ \tilde{\theta}_{ji}=\frac{r_{ji}}{n_{ji}}
+\end{equation}
+
+\subsubsection{$\mu_j$ and $M_j$}\
+\begin{align}
+ \theta_{ji}|\mu_j,M_j\thicksim \text{Beta}(\mu_j,M_j)
+\end{align}
+
+The probability density function of beta distribution parameterized with the mean $\mu_j$ and sample-size paramenter $M_j$ is
+\begin{equation}
+\begin{split}
+  f(\theta_{ji};\mu_j,M_j) = \frac{1}{B(M_j\mu_j,M_j(1-\mu_j))}\theta_{ji}^{M_j\mu_j-1}(1-\theta_{ji})^{M_j(1-\mu_j)-1}.
+\end{split}
+\end{equation}
+The first polulation moment (mean) is
+\begin{equation}
+ \mu_j=\text{E}[\theta_{ji}],
+\end{equation}
+while the first sample moment
+\begin{equation}
+\begin{split}
+ \text{E}[\tilde{\theta}_{ji}] &= \frac{1}{N-1}\sum_{i=1}^N\tilde{\theta}_{ji}\\ &=\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{split}
+\end{equation}
+Equat the population mean to sample mean we have
+\begin{equation}
+ \tilde{\mu}_j =\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{equation}
+
+The second population moment centered around mean (variance) is
+\begin{equation}
+ Var(\theta_{ji}) = \frac{\mu_j(1-\mu_j)}{M_j+1}
+\end{equation}
+while the sample variance is
+\begin{equation}
+ \begin{split}
+  Var(\tilde{\theta}_{ji}) &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\text{E}[\tilde{\theta}_{ji}])^2\\ &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2
+ \end{split}
+\end{equation}
+By definition of Method of moments,
+\begin{equation}
+  \frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\tilde{M}_j+1} = \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2,
+\end{equation}
+so,
+\begin{equation}
+\begin{split}
+ \tilde{M}_j &= (N-1)\cdot\frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2}-1\\ &= (N-1)\cdot\frac{\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}\\ &= \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
+\end{split}
+\end{equation}
+
+\subsubsection{$\mu_0$ and $M_0$}
+\begin{align}
+ \mu_j|\mu_0,M_0\thicksim \text{Beta}(\mu_0,M_0)
+\end{align}
+Likewise,
+\begin{equation}
+ \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
+\end{equation}
+
+The expected value ($\mu_0$) is
+\begin{equation}
+ \mu_0=\text{E}[\mu_j],
+\end{equation}
+while the first sample moment
+\begin{equation}
+\begin{split}
+\text{E}[\tilde{\mu}_j] &= \frac{1}{J-1}\sum_{i=1}^J\tilde{\mu}_j\\ &=  \frac{1}{J-1}\sum_{i=1}^J\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\\ &=\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{split}
+\end{equation}
+
+Using Methond of Moment
+\begin{equation}
+  \tilde{\mu}_0 = \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{equation}
+
+The second population moment centered around mean (variance) is
+\begin{equation}
+ Var(\mu_j) = \frac{\mu_0(1-\mu_0)}{M_0+1}
+\end{equation}
+while the sample variance is
+\begin{equation}
+ \begin{split}
+  Var(\tilde{\mu}_j) &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\text{E}[\tilde{\mu}_j])^2\\ &= \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2
+ \end{split}
+\end{equation}
+By definition of Method of moments,
+\begin{equation}
+  \frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\tilde{M}_0+1} = \frac{1}{J-1}\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2,
+\end{equation}
+so,
+\begin{equation}
+\begin{split}
+ \tilde{M}_0 &= (J-1)\cdot\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2}-1\\ &= (J-1)\cdot\frac{\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1\\ &= \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
+\end{split}
+\end{equation}
 \bibliographystyle{apalike}

### 2013-09-05 doc/bioinformatics2013/RVD2.tex
fixed appendix title
Refined the Initialization section and its derivation.

@@
 \begin{center}
-\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
+%\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
 \caption{RVD2 Graphical Model}
@@
 \begin{equation}
- \tilde{\theta}_{ij}=\frac{r_{ij}}{n_{ij}}
+ \tilde{\theta}_{ji}=\frac{r_{ji}}{n_{ji}}
 \end{equation}
-The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
+
+In our model, $\theta_{ji}$ is follows beta distribution with parameters $\mu_j$ and $M_j$
+\begin{align}
+ \theta_{ji}|\mu_j,M_j\thicksim \text{Beta}(\mu_j,M_j)
+\end{align}
+
+The first polulation moment (mean) is 
 \begin{equation}
- \mu_j=E(\theta_{ij}|\mu_j,M_j)
+ \mu_j=E(\theta_{ij})
+\end{equation}
+By definition of Method of Moments, 
+\begin{equation}
+ \tilde{\mu}_j = E(\tilde{\theta}_{ij}) = \frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{equation}
+
+The second population moment centered around mean (variance) is
+\begin{equation}
+ Var(\theta_{ji}) = \frac{\mu_j(1-\mu_j)}{M_j+1}
+\end{equation}
+while the sample variance is
+\begin{equation}
+ \begin{split}
+  Var(\tilde{\theta}_{ji}) &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\text{E}[\tilde{\theta}_{ji}])^2\\ &= \frac{1}{N-1}\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2
+ \end{split}
+\end{equation}
+Equating polulation variance to sample variance we have
+\begin{equation}
+ \begin{split}
+  \tilde{M}_j &=\frac{\mu_j(1-\mu_j)}{\text{Var}(\tilde{\theta}_{ji})}-1\\ &= \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
+ \end{split}
+\end{equation}
+
+$\mu_j$ is modeled as a Beta distribution random variable with paprameters $\mu_o$ and $M_0$. Likewise,
+\begin{equation}
+\begin{split}
+ \tilde{\mu}_0 &= E(\mu_j)\\ &= \frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j\\ &= \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+\end{split}
 \end{equation}
 \begin{equation}
- \tilde{\mu}_j =\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
-\end{equation}
-
-\begin{equation}
- M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
-\end{equation}
-\begin{equation}
- \tilde{M}_j = \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
-\end{equation}
-Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
-\begin{equation}
- \mu_0=E(\mu_j|\mu_0,M_0)
-\end{equation}
-\begin{equation}
- \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
-\end{equation}
-
-\begin{equation}
- M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
-\end{equation}
-\begin{equation}
- \tilde{M}_0 = \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
+\begin{split}
+ \tilde{M}_0 &=\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{Var(\tilde{\mu}_j)}-1\\ &= \frac{\left(\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
+\end{split}
 \end{equation}
@@
 
-The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).
-Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is
+The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984). By definition, the power-divergence family of statistcs parameterized with $\lambda$ is 
 \begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R,
 \end{equation}
+
+where $\{r_i; i=1,...,k\}$ denote observed frequencies and $\{E_i; i=1,...,k\}$ denote the expected frequencies.
 
@@
 \section{Appendix}
-\subsection{Parameter initialization derivation}
-The model parameters and latent variables are initialized using method of moments by equating population moment to sample moment.
+\subsection{Parameter initialization derivation}\
+
+The model parameters and latent variables are initialized using Method of Moments.
+\subsubsection{$\theta_{ji}$}\
 
 The estimated error probability is calculated from the observed sample
-\subsubsection{$\theta_{ji}$}\
 \begin{equation}
@@
 \subsubsection{$\mu_j$ and $M_j$}\
+
+$\theta_{ji}$ is modeled as beta distribution random variable with parameter $\mu_j$ and $M_j$
 \begin{align}
@@
 \begin{split}
- \tilde{M}_j &= (N-1)\cdot\frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2}-1\\ &= (N-1)\cdot\frac{\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}\\ &= \frac{\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
+ \tilde{M}_j &= (N-1)\cdot\frac{\tilde{\mu}_j(1-\tilde{\mu}_j)}{\sum_{i=1}^N(\tilde{\theta}_{ji}-\tilde{\mu}_j)^2}-1\\ &= (N-1)\cdot\frac{\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}\\ &= \frac{\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^N\left(\frac{r_{ji}}{n_{ji}}-\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}
 \end{split}
@@
 
-\subsubsection{$\mu_0$ and $M_0$}
+\subsubsection{$\mu_0$ and $M_0$}\
+
+We set a hyperprior distribution $Beta(\mu_0,M_0)$ on $\mu_j$ based on our belief that $\mu_j\in[0,1]$
 \begin{align}
- \mu_j|\mu_0,M_0\thicksim \text{Beta}(\mu_0,M_0)
+ \mu_j|\mu_0,M_0\thicksim \text{Beta}(\mu_0,M_0).
 \end{align}
-Likewise,
-\begin{equation}
- \tilde{\mu}_0 =\frac{1}{N-1}\sum_{i=1}^N\tilde{\mu}_j
-\end{equation}
 
@@
 \begin{equation}
-  \tilde{\mu}_0 = \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
+  \tilde{\mu}_0 = \text{E}[\tilde{\mu}_j] = \frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}
 \end{equation}
@@
 \begin{split}
- \tilde{M}_0 &= (J-1)\cdot\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2}-1\\ &= (J-1)\cdot\frac{\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1\\ &= \frac{\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
+ \tilde{M}_0 &= (J-1)\cdot\frac{\tilde{\mu}_0(1-\tilde{\mu}_0)}{\sum_{i=1}^J(\tilde{\mu}_j-\tilde{\mu}_0)^2}-1\\ &= (J-1)\cdot\frac{\left(\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\frac{1}{N-1}\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1\\ &= \frac{\left(\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)\cdot\left(1-\frac{1}{(J-1)(N-1)}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)}{\sum_{i=1}^J\left(\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}-\frac{1}{J-1}\sum_{i=1}^J\sum_{i=1}^N\frac{r_{ji}}{n_{ji}}\right)^2}-1
 \end{split}

### 2013-09-06 doc/bioinformatics2013/RVD2.tex
Finished parameter initialization section and appendix.

--- 2013-09-03 doc/bioinformatics2013/RVD2.tex
@@
 
-Once we have obtained an estimate of the posterior distribution, we test the hypothesis that there is a sub-population of DNA molecules in the pool with a non-reference base at position $j, \forall j \in 1, \ldots, J$.
-First, we test the hypothesis that the error rate in the case data, $\mu_j^{\text{case}}$ is greater than the error rate in the control data, $\mu_j^{\text{control}}$.
-If so, then we test the hypothesis that the high error rate is due to excess reads with a particular non-reference base rather than a uniform distribution over all possible non-reference bases.
+Once we have obtained an estimate of the posterior distribution for the case and control data, we use a two-stage test for variant at each position.
+First, we test whether the error rate in the case sample is significantly greater than the error rate in the control data for each position.
+Then, we test the hypothesis that the error rate is high is due to excess reads with a particular non-reference base.
+We accomplish the first test with a Bayesian posterior density test and the second with a $\chi^2$ goodness-of-fit test to a uniform distribution.
+In the remainder of this section, we describe the initialization, MCMC sampling approach and hypothesis testing procedure.
 
+%%%%%%%%%%%%%
+% Initialization
+%%%%%%%%%%%%%
 \subsection{Initialization}
-The initial values of the model parameters and latent variables is obtained by a method-of-moments procedure.
-Essentially, we set the parameter value (population moment) equal to the sample moment.
-In some cases, the method-of-moments estimate is outside the parameter space.
-Then, we XXX.
+The initial values for the model parameters and latent variables is obtained by a method-of-moments(MoM) procedure.
+MoM works by setting the population moment equal to the sample moment.
+A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
+We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.
 
-The estimated error probability is calculated from the observed sample
-\begin{equation}
- \theta_{ij}=\frac{r_{ij}}{n_{ij}}
-\end{equation}
-The estimate of parameter $\mu_j$ is obtained using the expected value of $\theta_{ij}$, while $M_j$ is estimated using the moment estimates for the variance of $\theta_{ij}$.
-\begin{equation}
- \mu_j=E(\theta_{ij}|\mu_j,M_j)
-\end{equation}
-\begin{equation}
- M_j=\frac{\mu_j(1-\mu_j)}{Var(\theta_{ij}|\mu_j,M_j)}-1
-\end{equation}
+We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}.
+The MoM estimate for replicate-level parameters are $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
+The estimates for the position-level parameters are $\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ and $\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
+The estimates for the genome-level parameters are $\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ and $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
 
-Likewise,the estimate of hyperparameters $\mu_o$ and $M_0$ are
-\begin{equation}
- \mu_0=E(\mu_j|\mu_0,M_0)
-\end{equation}
-\begin{equation}
- M_0=\frac{\mu_0(1-\mu_0)}{Var(\mu_j|\mu_0,M_0)}-1
-\end{equation}
+The method-of-moments is known to have the potential to give estimates that are outside the true parameter bounds.
+For example, the MoM estimate $\tilde{M}_j$ may be negative while $M_j$ must be non-negative.
+We have not encounter such pathological behavior of the estimates in this application.
 
@@
 
-We are able to draw samples from $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
+Samples from the posterior distribution $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ are drawn analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
+The posterior distribution is
 
-\subsection{Sampling from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$}
+\begin{equation}
+	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right),
+\end{equation}
+where we have used the mean-precision parameterization of the beta distribution.
 
+\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
+
+The posterior distribution for $\mu_j$ is
+\begin{equation}
+	p \left( \mu_j |\theta_{ij},\mu_0,M_0\right) = 
+\end{equation}
 \subsection{Bayesian hypothesis test for high error rate}
@@
 
-\section{Appendix}
-\subsection{Moments derivation}
-The probability density function of beta distribution, for $x\in[0,1]$, and shape real parameter $\alpha>1$ and $\beta>1$ is 
+\appendix
+\section{Parameter Initialization}\label{sec:appendix_mom}
+Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
+Therefore the MoM estimator is
 \begin{equation}
-\begin{split}
-  f(x;\alpha,\beta) &= \frac{x^{\alpha-1}(1-x)^{\beta-1}}{\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx}\\ &=\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}x^{\alpha-1}(1-x)^{\beta-1}\\ &=\frac{1}{B(\alpha,\beta)}x^{\alpha-1}(1-x)^{\beta-1}
-\end{split}
-\end{equation}
-where $\Gamma(z)$ is the Gammma function, $B(\alpha,\beta)$ is the beta function, and
-\begin{equation}
-\begin{split}
- B(\alpha,\beta) &=\int_0^1x^{\alpha-1}(1-x)^{\beta-1}dx\\  &= \frac{\Gamma(\alpha)\Gamma(\beta)}{\Gamma(\alpha+\beta)}
-\end{split}
+	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
 \end{equation}
 
-The expected value (mean) of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
-\begin{equation}
-\begin{split}
-  E[X] &= \int_0^1xf(x;\alpha,\beta)dx\\ &= \int_0^1x\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx\\ &= \frac{\int_0^1x^{\alpha}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}\\ &= \frac{B(\alpha+1,\beta)}{B(\alpha,\beta)}\\ &= \frac{\Gamma(\alpha+1)\Gamma(\beta)}{\Gamma(\alpha+\beta+1)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}\\ &= \frac{\alpha}{\alpha+\beta}
-\end{split}
-\end{equation}
-Note that Gamma function satisfies the functional equation
-\begin{equation}
- \Gamma(z) = z\Gamma(z)
-\end{equation}
+We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
+The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
+The first and second population moments are
+\begin{eqnarray}
+	E[\theta_{ji}] =& \mu_j,\\
+	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
+\end{eqnarray}
+The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
+Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
+\begin{eqnarray}
+	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
+	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
+\end{eqnarray}
 
-The variance of a Beta distribution random variable X with parameter $\alpha$ and $\beta$ is 
-\begin{equation}
-\begin{split}
-  var(X) &= E[(X-E(X))^2]\\ &= E(X^2)-[E(x)]^2\\ &=\int_0^1x^2\frac{x^{\alpha-1}(1-x)^{\beta-1}}{B(\alpha,\beta)}dx-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\int_0^1x^{\alpha+1}(1-x)^{\beta-1}dx}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{B(\alpha+2,\beta)}{B(\alpha,\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\Gamma(\alpha+2)\Gamma(\beta)}{\Gamma(\alpha+\beta+2)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha\Gamma(\alpha)\Gamma(\beta)}{(\alpha+\beta+1)(\alpha+\beta)\Gamma(\alpha+\beta)}\cdot\frac{\Gamma(\alpha+\beta)}{\Gamma(\alpha)\Gamma(\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{(\alpha+1)\alpha}{(\alpha+\beta+1)(\alpha+\beta)}-\left(\frac{\alpha}{\alpha+\beta}\right)^2\\ &= \frac{\alpha\beta}{(\alpha+\beta+1)(\alpha+\beta)^2}
-\end{split}
-\end{equation}
-
-Instead of the standard scale parameter $\alpha$ and $\beta$, we use the mean $\mu$ and sample-size paramenter $M$ for the Beta distribution in our hierarchical model.
-\begin{equation}
-\begin{split}
-  f(x;\mu,M) &=\frac{\Gamma(M)}{\Gamma(M\mu)\Gamma(M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}\\ &= \frac{1}{B(M\mu,M(1-\mu))}x^{M\mu-1}(1-x)^{M(1-\mu)-1}
-\end{split}
-\end{equation}
-where 
-\begin{equation}
- \mu = \frac{\alpha}{\alpha+\beta}
-\end{equation}
-\begin{equation}
- M = \alpha+\beta
-\end{equation}
-From Equation 24 and Equation 26 it can be seen that
-\begin{equation}
- E(x|\mu,M) = \mu
-\end{equation}
-\begin{equation}
- Var(x|\mu,M) = \frac{\mu(1-\mu)}{M+1}
-\end{equation}
+Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
+\begin{eqnarray}
+	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
+	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
+\end{eqnarray}
 

### 2013-09-06 doc/bioinformatics2013/RVD2.tex
Finished sampling section.

@@
 
-This model has three levels of sampling.
-First the baseline error rate across all locations and all replicates is chosen once and for all for the entire data set.
-Then, based on that baseline error rate, location-specific error rates are sampled.
-Based on the location specific error rate, replicate-specific error rates are selected.
-Finally, the observable data, the count of the non-reference reads is sampled.
+This model has three levels of sampling that capture three scales of resolution in the data.
+First the genome-wide baseline error rate across all locations, $\mu_0$,  is chosen once for the entire data set.
+The precision parameter $M_0$ measures the variability of the location-specific error rates, $\mu_j$ across positions.
+Then, based on that genome-wide parameters, location-specific error rates, $\mu_j$ and precisions $M_j$ are sampled for each position.
+The precision parameter $M_j$ measures the variability between replicates at a particular position.
+Based on the location specific parameters, replicate-specific error rates, $\theta_{ji}$ are selected.
+Finally, the observable data, the count of the non-reference reads  $r_{ji}$ is sampled.
+
+It is important to note that this model partitions noise in the data hierarchically.
+The sampling error, represented by the binomial random variable, can be reduced by increasing the read depth $n_{ji}$.
+In contrast, increasing the read depth does not influence the parameter for the error rate variability between replicates $M_j$ and the error rate variability across positions $M_0$.
+The former parameter is influenced by the experimental protocol and the latter parameter is influenced by the DNA sequence itself and the fidelity of the base call based on it's sequence context.
 
@@
 
-% Describe graphical model framework here
+Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
+In this graphical model framework~\cite{} a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes.
+A rounded box or ``plate" represents replication of the nodes within the plate.
+This graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
 
-Given this model structure, the joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
+Given the model structure shown in Figure~\ref{fig:graphical_model}, the joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
 
@@
 
+%%%%%%%%%%%%%%%%%%
+% Sampling theta
+%%%%%%%%%%%%%%%%%%
 \subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
@@
 \end{equation}
-where we have used the mean-precision parameterization of the beta distribution.
+where we have used the mean-precision parameterization of the beta distribution. We note p-th sample as $\theta^{(p+1)}_{ji}$.
 
+%%%%%%%%%%%%%%%%%%
+% Sampling mu
+%%%%%%%%%%%%%%%%%%
 \subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
+The posterior distribution over $\mu_j$ given it's Markov blanket is 
+\begin{equation}
+	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j)
+\end{equation}
 
-The posterior distribution for $\mu_j$ is
+Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution.
+Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
+
+A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim $. The acceptance probability is then
 \begin{equation}
-	p \left( \mu_j |\theta_{ij},\mu_0,M_0\right) = 
+	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
 \end{equation}
+
+Since, each position $j$ is exchangeable given the genome-wide parameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors which increases the speed considerably.
+
+%%%%%%%%%%%%%%%%%%
+% Posterior Density Test
+%%%%%%%%%%%%%%%%%%
 \subsection{Bayesian hypothesis test for high error rate}

### 2013-09-06 doc/bioinformatics2013/RVD2.tex
Finished bayesian hyp test section.

@@
 Since, each position $j$ is exchangeable given the genome-wide parameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors which increases the speed considerably.
+We have found that the algorithm performance improves when we take the median of ten or more M-H samples as a single Gibbs step for each position.
+This computational effort only minimally increases run time when run on a parallel computing platform.
+In the experiments in the remainder, we use a XX sample burn-in to allow the Markov chain to approach its stationary distribution and we thin the chain by a factor of 2 to reduce autocorrelation among samples.
 
@@
 \subsection{Bayesian hypothesis test for high error rate}
+The Metropolis-within-Gibbs (M-w-G) sampling algorithm provides samples from the posterior distribution of $\mu_j$ for both the control and case data.
+For notational simplicity in this section, we define the random variables associated with these two distributions $\mu_j^{\text{case}}$ and $\mu_j^{\text{control}}$.
 
-To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
-\begin{align}
- H_1: \mu_{case}-\mu_{control}\geq\tau \\
- H_2: \mu_{case}-\mu_{control}<\tau
-\end{align}
+A variant is called if the data indicate that $\mu_j^{\text{case}} > \mu_j^{\text{control}}$ with high confidence,
+\begin{equation}
+	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} \geq \tau ) > 1-\alpha,
+\end{equation}
+where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
+While analytical forms for the posterior distributions are not available, hthe M-w-G algorithm does provide samples from those distributions.
+So, we sample from the posterior distribution $\mu_j^{\Delta} \triangleq \mu_j^{\text{case}} - \mu_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\mu_j^{\text{case}}$ and$\mu_j^{\text{control}}$ and calculating samples from $\mu_j^{\Delta}$ directly.
 
-We can calculate the posterior probability of the hypothesis $H_1$ being true by integrating the posterior density over the correct region:
-
-\begin{equation}
- p(H1:\mu_{case}-\mu_{control}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
-\end{equation}
-
-We accept the hypothesis $H_1$ if this posterior probability is no lower than 95\%, which means we are 95\% sure that there is a mutation in this position.
-
-ROC curve is plotted to evaluate the performance of the detection algorithm with different threshold for different mutation rate.
-
+%%%%%%%%%%%%%%%%%%
+% Chi^2 Test
+%%%%%%%%%%%%%%%%%%
 \subsection{$\chi^2$ test for non-uniform base distribution}

### 2013-09-07 doc/bioinformatics2013/RVD2.tex
Paper draft update on result section ROC curve

--- 2013-09-06 doc/bioinformatics2013/RVD2.tex
@@
 
-\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
+\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}\
 
@@
 \section{Results}
-Figure 1 Overall of the process
-\subsection{Simulation Results}
-\subsection{Comparison Results on Synthetic DNA}
+
+\subsection{Overall of the process}
+Figure 1
+
+\subsection{Performance with read depth}\
+Figure 2 ROC plot showing performance with different read depth
+
+picard to thin segment data, showing performance as read depth decreasing
+
+\begin{figure}[h]
+\begin{center}
+%\includegraphics[width=160mm]{eps_figs/ROC_without_chi2.eps}
+\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\label{fig:ROC_without_chi2}
+\end{center}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/ROC_without_chi2.pdf}
+\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\label{fig:ROC_without_chi2}
+\end{center}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/MuBarPlot.pdf}
+\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\label{fig:ROC_without_chi2}
+\end{center}
+\end{figure}
+
+We tested the mutation detection power of our graphical model on synthetic dataset with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard.The model had high detection power when the read depth was at $10^5$ level.
+At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%.
+We missed one mutation at position 205 and had one false positive at position 8.
+The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve.
+As expected, the detection power of the graphical model decreased when the read depth was reduced.
+
+\begin{figure}[h] %needs to go to supplementary document.
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
+\caption{ROC curve varying read depth showing method performance with Bayesian baHypothesis Testing and Chi-Square Test}
+\label{fig:R}
+\end{center}
+\end{figure}
+We also evaluateed the performance of the graphical model with Chi-square test included in the hypothesis testing.
+
+Low read depth limited the False positive rate. we are confident in the callings
+
+High read depth Not much influence. do not bring about bad influence
+
+\subsection{Comparison Results on Synthetic DNA}\
+
 Figure 3 ROC comparison with other methods: varscan/samtools/rvd
-\subsection{Empirical Results on Multiple Sclerosis Data}
+
+\subsection{Empirical Results on Multiple Sclerosis Data}\
 We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that
@@
 Table: Table of Variants
-\subsection{Performance with read depth}
-Figure 4 ROC by read depth
-picard to thin segment data, showing performance as read depth decreasing
 

### 2013-09-07 results/2013-08-15_Compute_ROC_Synthetic_avg_all/plotROCbyDilution.py

@@ -21,13 +21,13 @@ def main():
     folderList = ('2013-08-14_Compute_ROC_Synthetic_avg10',\
                   '2013-08-14_Compute_ROC_Synthetic_avg100',\
                   '2013-08-14_Compute_ROC_Synthetic_avg1000',\
-                  '2013-08-14_Compute_ROC_Synthetic_avg10000',\
-                  '2013-08-14_Compute_ROC_Synthetic_avg100000')
+                  '2013-08-14_Compute_ROC_Synthetic_avg10000')
     
     N=1000 # Z sampling size  
     fig=plt.figure(figsize=(10, 9))
     chi2=False
-    plt.suptitle('Comparing ROC plot for different read depth' )
+##    lstyle=('-','--','-.',':')
+    lcolor=('c','r','g','b')
     for d in dilutionList:
         logging.debug("Processing dilution: %0.1f%%" % d)
         ax = fig.add_subplot(2,2,dilutionList.index(d)+1)
@@ -37,27 +37,26 @@ def main():
             caseFile = "Case%s.hdf5" % str(d).replace(".","_")
             caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
              # ROC
-            
             [fpr,tpr,cov, T] = ROCpoints(controlFile,caseFile,P=0.95,chi2=chi2)
-            
-            ax.plot(fpr,tpr, label='%d' % cov)
-
+            ax.plot(fpr,tpr, color=lcolor[folderList.index(f)], label='%d' % cov)
+##            ax.plot(fpr,tpr,linestyle=lstyle[folderList.index(f)], color=lcolor[folderList.index(f)], label='%d' % cov)
+##            ax.plot(fpr[0],tpr[0],marker='o',markerfacecolor=lcolor[folderList.index(f)])
         l = ax.legend(loc=4,prop={'size':9},title='Read depth')
         l.get_title().set_fontsize(9)
         ax.plot([0,1],[0,1],color='k',linestyle='dashed')
-        ax.set_title('%0.1f%% Mutant Mixture' % d)
-        ax.set_xlim((-0.01,1.01))
-        ax.set_ylim((-0.01,1.01))
+        ax.set_title('%0.1f%% Mutant Mixture' % d, fontsize=10)
+        ax.set_xlim((-0.03,1.03))
+        ax.set_ylim((-0.03,1.03))
                  
         ax.set_xlabel('False Positive Rate')
[…]

### 2013-09-07 results/2013-09-03_optimal_threshold_confusion_matrix/ConfMatrix.py

@@ -23,14 +23,10 @@ def main():
                   '2013-08-14_Compute_ROC_Synthetic_avg1000',\
                   '2013-08-14_Compute_ROC_Synthetic_avg10000',\
                   '2013-08-14_Compute_ROC_Synthetic_avg100000')
-##    dilutionList = (1.0,)
-##
-##    folderList = ('2013-08-14_Compute_ROC_Synthetic_avg100000',)
-
     
     N=1000 # Z sampling size  
     fig=plt.figure(figsize=(10, 9))
-    chi2=True
+    chi2=False
     plt.suptitle('Comparing ROC plot for different read depth' )
     for d in dilutionList:
         logging.debug("Processing dilution: %0.1f%%" % d)
@@ -42,13 +38,13 @@ def main():
             caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
             
             [fpr,tpr, cov, optimalT, CalledCaseMu, CalledControlMu, CalledLoc, predList, trueList] = ROCpoints(controlFile,caseFile,P=0.95,chi2=chi2)
-            title='MuBarPlot_Dilution%(dilution)0.1f_Depth%(depth)d.png'%{'dilution':d,'depth':cov}
+            title='MuBarPlot_Dilution%(dilution)0.1f_Depth%(depth)d.pdf'%{'dilution':d,'depth':cov}
             
             #MuBarPlot
             MuBarPlot(CalledControlMu,CalledCaseMu,CalledLoc,optimalT,title)
             
             #Confusion Matrix
-            title='ConfusionMatrix_Dilution%(dilution)0.1f_Depth%(depth)d.png'%{'dilution':d,'depth':cov}
+            title='ConfusionMatrix_Dilution%(dilution)0.1f_Depth%(depth)d.pdf'%{'dilution':d,'depth':cov}
             Conf_Matrix(predList,trueList,title)
 
             # ROC curve
@@ -163,7 +159,7 @@ def MuBarPlot(CallControlMu,CallCaseMu,Loc,optimalT,title):
         ax.set_xticklabels( [str(x)[8:] for x in Loc] )
         ax.set_ylim
         ax.legend( (rects1[0], rects2[0]), ('Control', 'Case') )
-        fig.suptitle('Optimal Threshold = %0.5f'%optimalT )
+##        fig.suptitle('Optimal Threshold = %0.5f'%optimalT )
         plt.savefig(title.replace('.','_',1))
     else:
         print 'No variant is called'

### 2013-09-11 doc/bioinformatics2013/RVD2.tex
Finished result section on ROC curves. Update Notebook on ROC experiment.

@@
 
-The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).
-Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is
+The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).By definition, the power-divergence family of statistcs parameterized with $\lambda$ is 
 \begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R,
 \end{equation}
+where $\{r_i; i=1,...,k\}$ are the observed frequencies and $\{E_i; i=1,...,k\}$ are the  expected frequencies.
 
-Person's $X^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family.
+Person's $\chi^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family.
 Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.
@@
 
-Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.
-
-\subsubsection{Posterior distribution difference sampling}\
+\subsection{Posterior distribution difference sampling}\
 
@@
 
-picard to thin segment data, showing performance as read depth decreasing
-
-\begin{figure}[h]
-\begin{center}
-%\includegraphics[width=160mm]{eps_figs/ROC_without_chi2.eps}
-\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
-\label{fig:ROC_without_chi2}
-\end{center}
-\end{figure}
-
 \begin{figure}[h]
@@
 
-We tested the mutation detection power of our graphical model on synthetic dataset with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard.The model had high detection power when the read depth was at $10^5$ level.
+We tested the mutation detection power of our graphical model on synthetic dataset using a two-stage test scheme with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard.
+$\chi^2$ test, the second stage test can be optionally included based on average read depth.
+
+Excluding $\chi^2$ test, the ROC curves of our graphical model using Bayesian Hypothesis Testing with different threshold $\tau$ is shown in Figure??.
+The model had high detection power when the read depth was at $10^5$ level.
 At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%.
@@
 \includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
-\caption{ROC curve varying read depth showing method performance with Bayesian baHypothesis Testing and Chi-Square Test}
+\caption{ROC curve varying read depth showing method performance with Bayesian Hypothesis Testing and Chi-Square Testing}
 \label{fig:R}
@@
 \end{figure}
-We also evaluateed the performance of the graphical model with Chi-square test included in the hypothesis testing.
 
-Low read depth limited the False positive rate. we are confident in the callings
+Figure ??
+shows the ROC curves when both Bayesian Hypothesis Testing and $\chi^2$ testing are carried out for variant calling.
+Comparing with Figure ??, it can be seen that when read depth is low (from $10$ to $10^3$), the false positive rate is severely limited to zero, which means we are very confident in our variant callings.
+However, the low false positve rate can be achieved at a sacrifice of low true postive rate especially when the dilution rate and read depth are low.
+The $\chi^2$ test is a stringent test when the read depth is low.
 
-High read depth Not much influence. do not bring about bad influence
+However, if the read depth is high enough (up to $10^4$ or higher), the $\chi^2$ test can limit the false positive rate in a slight degree but no longer at a sacrifice of True Positive Rate.
+Actually the performance is not much influenced by $\chi_2$ testing with high read depth.
 

### 2013-09-11 doc/bioinformatics2013/RVD2.tex
Merged RVD2.tex

--- 2013-09-06 doc/bioinformatics2013/RVD2.tex
@@
 %%%%%%%%%%%%%%%%%%
+
 % Chi^2 Test
+
 %%%%%%%%%%%%%%%%%%
@@
 
-We assume that at one position in one experimental replicate, the non-reference reads follows multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.
+We assume that at one position in one experimental replicate, the non-reference read count follows a multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.
 
-The non-reference reads can be because of either mutation or various errors including sampling error, inter-replicated error, $etc$.
-In theory, the non-reference read counts caused by non-oriented errors should be uniformly distributed among three non-reference categories, namely $p_1=p_2=p_3=1/3$.
-However, the distribution of non-reference read counts among three non-reference categories caused by mutation theoretically is biased.
+The non-reference reads can be because of either mutation or various errors including sampling error or inter-replicated error.
+We assume, non-reference read counts caused by these random mechanisms should be uniformly distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$.
+In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.
 
-The null hypothesis can be denoted as
-\begin{align}
- H_0: p=p_0
-\end{align}
-where
-\begin{align} 
- p_0=(p_1,p_2,p_3), p_1=p_2=p_3=1/3
-\end{align}
+We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios.
+The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
+Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$.
+The test statistic is
+\begin{equation}
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
+\end{equation}
+where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
+Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.
 
-The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).
-Considering observed frequencies $\{r_i; i=1,...,k\}$ and expected frequencies $\{E_i; i=1,...,k\}$, by definition the power-divergence family of statistcs indexed by real parameter $\lambda$ is
+For each position, there are $N$ power-divergence statistics for each experimental replicate.
+We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position.
 \begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R
+	X^2 = -2 \sum_{i=1}^N \ln(p_i)
 \end{equation}
 
-Person's $X^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family.
-Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.
+To control for Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.
 
-Particulaly in our case, Let $(r_1,r_2,r_3)$ denotes the non-reference base frequencies and $\sum_{i=1}^3 r_i = r$.
-The expected frquencies are $(rp_1, rp_2, rp_3)$.
-Then the power-divergence statistics can be written as
-
-\begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^3 r_i \left[\left(\frac{r_i}{rp_i}\right)^\lambda-1\right];\lambda \in R
-\end{equation}
-
-We use $\lambda =2/3$ for the goodness of fit testing.
+Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.
 
@@
 The significance level is chosen as 0.05.
-
-Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.
 
@@
 \section{Results}
-Figure 1 Overall of the process
-\subsection{Simulation Results}
-\subsection{Comparison Results on Synthetic DNA}
+
+\subsection{Overall of the process}
+Figure 1
+
+\subsection{Performance with read depth}\
+Figure 2 ROC plot showing performance with different read depth
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/ROC_without_chi2.pdf}
+\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\label{fig:ROC_without_chi2}
+\end{center}
+\end{figure}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/MuBarPlot.pdf}
+\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\label{fig:ROC_without_chi2}
+\end{center}
+\end{figure}
+
+We tested the mutation detection power of our graphical model on synthetic dataset using a two-stage test scheme with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard.
+$\chi^2$ test, the second stage test can be optionally included based on average read depth.
+
+Excluding $\chi^2$ test, the ROC curves of our graphical model using Bayesian Hypothesis Testing with different threshold $\tau$ is shown in Figure??.
+The model had high detection power when the read depth was at $10^5$ level.
+At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%.
+We missed one mutation at position 205 and had one false positive at position 8.
+The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve.
+As expected, the detection power of the graphical model decreased when the read depth was reduced.
+
+\begin{figure}[h] %needs to go to supplementary document.
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
+\caption{ROC curve varying read depth showing method performance with Bayesian Hypothesis Testing and Chi-Square Testing}
+\label{fig:R}
+\end{center}
+\end{figure}
+
+Figure ??
+shows the ROC curves when both Bayesian Hypothesis Testing and $\chi^2$ testing are carried out for variant calling.
+Comparing with Figure ??, it can be seen that when read depth is low (from $10$ to $10^3$), the false positive rate is severely limited to zero, which means we are very confident in our variant callings.
+However, the low false positve rate can be achieved at a sacrifice of low true postive rate especially when the dilution rate and read depth are low.
+The $\chi^2$ test is a stringent test when the read depth is low.
+
+However, if the read depth is high enough (up to $10^4$ or higher), the $\chi^2$ test can limit the false positive rate in a slight degree but no longer at a sacrifice of True Positive Rate.
+Actually the performance is not much influenced by $\chi_2$ testing with high read depth.
+
+\subsection{Comparison Results on Synthetic DNA}\
+
 Figure 3 ROC comparison with other methods: varscan/samtools/rvd
-\subsection{Empirical Results on Multiple Sclerosis Data}
+
+\subsection{Empirical Results on Multiple Sclerosis Data}\
 We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that

### 2013-09-18 doc/bioinformatics2013/RVD2.tex
updates to hypothesis testing discussion in manuscript

@@
 where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
-While analytical forms for the posterior distributions are not available, hthe M-w-G algorithm does provide samples from those distributions.
-So, we sample from the posterior distribution $\mu_j^{\Delta} \triangleq \mu_j^{\text{case}} - \mu_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\mu_j^{\text{case}}$ and$\mu_j^{\text{control}}$ and calculating samples from $\mu_j^{\Delta}$ directly.
+We draw a sample from the posterior distribution $\mu_j^{\Delta} \triangleq \mu_j^{\text{case}} - \mu_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\mu_j^{\text{case}}$ and$\mu_j^{\text{control}}$ and calculating samples from $\mu_j^{\Delta}$ directly.
 
@@
 
-We assume that at one position in one experimental replicate, the non-reference read count follows a multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.
+We assume the counts for the non-reference bases at poisiuton $j$ in replicate $i$ follow a multinomial distribution with parameter  $p = (p_1, p_2, p_3)$ where $p_k \geq 0$ for $k =1, 2, 3$ and $\sum_{k=1}^3 p_k = 1$.
 
-The non-reference reads can be because of either mutation or various errors including sampling error or inter-replicated error.
-We assume, non-reference read counts caused by these random mechanisms should be uniformly distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$.
+The non-reference reads may be due to a true mutation or due to a random sequencing error.
+We assume, non-reference read counts caused by a random mechanism results in a uniform distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$.
 In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.
@@
 The test statistic is
+
 \begin{equation}
@@
 \end{equation}
+
 where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
@@
 For each position, there are $N$ power-divergence statistics for each experimental replicate.
-We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position.
+We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position,
+
 \begin{equation}
-	X^2 = -2 \sum_{i=1}^N \ln(p_i)
+	X^2 = -2 \sum_{i=1}^N \ln(p_i).
 \end{equation}
+This gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.
 
-To control for Bonferroni correction is applied to maintaining the familywise error rate (FWER). The significance level is chosen as 0.05.
-
-Finally, if read counts in one position passes the Bayesian Hypothesis Testing and fails the Goodness-of-fit Testing, the position will be called.
-
-For each position we have six power-divergence statistics from 6 experimental replicates.
-We use Fisher's method to combines the six statistics into a single one for inference.
-Bonferroni correction is applied to maintaining the familywise error rate (FWER).
-The significance level is chosen as 0.05.
+While Fisher's combined probability test collapses the p-values across replicates, we still have $J$ tests across positions.
+We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions.
 

### 2013-10-03 doc/bioinformatics2013/RVD2.tex
edits to manuscript.

--- 2013-09-11 doc/bioinformatics2013/RVD2.tex
@@
 \usepackage{amsmath}
+
+\usepackage{enumitem}
+
+\usepackage{setspace}
+\doublespacing
 %\usepackage{epstopdf}
@@
 
-\section{Model structure}\
+% DIscuss low-purity samples and somatic mutations
 
-Given a data matrix $r \in \RR^{J \times N}$ where the element $r_{ji}$ is the count of the reads that contain a non-reference base at position $j$ in experimental replicate $i$.
-The total count of reads at position $i$ in replicate $i$ is $n_{ji}$.
-A typical data set is comprised of case and control data sets, $\{ r^{\text{case}} , n^{\text{case}} \}$ and  $\{ r^{\text{control}} , n^{\text{control}} \}$.
+Next-generation sequencing (NGS) technology has enabled the genomic interrogation for a fraction of the cost of traditional assays~\cite{Koboldt2013Cell}.
+Protocol and platform engineering improvements have reduced the cost of the assay to the point where it is now possible to generate $1\times10^9$ bases of sequence data in under 40 hours for less than \$1000 \cite{}.
+As a result NGS is increasingly being used as a general platform for research assays for methylation state \cite{}, DNA mutations \cite{}, copy number variation \cite{}, promoter occupancy \cite{} and others \cite{}.
+Recently, NGS has been used as a clinical diagnostic tool for testing for mutations in non-invasive fetal DNA \cite{}, infectious disease samples \cite{}, cancerous tumors \cite{}, and human microbial communities \cite{}.
 
-We have developed a hierarchical Bayesian model for this data set because of the nested structure of noise in the data.
-There is sampling error due to the fact that only a finite number of reads aligned and mapped for each location.
-There is inter-replicate errors due to experimental protocol and reproducibility.
-Finally there is sequence context error due to the fidelity of the DNA polymerase enzymes in synthesizing a base.
-These errors have been shown to depend heavily on the neighboring DNA bases and whether the base is at the end of a molecule or at the center of a molecule.
+Clinical samples are often genetically heterogeneous.
+Non-invasive fetal DNA testing uses a sample of the maternal blood and filters for the minor fraction that is cell-free DNA from the fetus \cite{}.
+Infectious diseases such as HIV and influenza mutate rapidly and a given sample from an individual host may contain many sub-populations \cite{}.
+DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample.
 
-\subsection{Generative Process} We have codified the statistical structure of these errors in the following generative process (we have used the mean-precision parameterized of the beta distribution):
+However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
+Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
+Samtools/bcftools use a relatively straightforward XXX to call genotypes \cite{}.
+GATK uses a naive Bayesian decision rule to call a variant as one of two possible alleles \cite{McKenna2010}.
+VarScan2-mpileup uses information from multiple replicates to call homogeneous and heterogeneous genotypes.
 
-\begin{enumerate}
-	\item Choose a baseline error rate parameter $\mu_0$ and a inter-location precision parameter $M_0$.
+Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
+MuTect is able to handle low-purity samples by computing a log-odds score.
+If the likelihood of the data assuming the locus is mutated exceeds the likelihood of the data assuming the site is not mutated by a threshold then a variant is called.
+There is an additional log-odds check on the normal sample to filter false positives~\cite{Cibulskis2013}.
+Strelka models the joint distribution for a tumor and normal sample as a mixture model using continuous values allele frequencies \cite{}.
+VarScan2 has a subcommand to call somatic mutations \cite{}.
+Recently VarScan2 was shown to outperform many other algorithms for calling low-frequency (5\%) variants \cite{}.
+
+The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data.
+We present the hierarchical Bayesian model for the method in Section X and the Bayesian hypothesis test to detect mutations in Section X.
+In Section X we present sensitivity and specificity results of our method on pure DNA samples mixed at defined fractions.
+We compare our algorithm to the most accurate methods available to date in Section X.
+Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients.
+This strategy effectively reduces the cost of the assay by $n$-fold where $n$ is the pool size.
+
+\section{Materials and Methods}
+
+RVD uses a two-stage approach for detecting for rare variants.
+First, it estimates the parameters of two hierarchical Bayesian models; one using data from the sample of interest (case) and one using data from a known reference sample (control).
+Then, it tests for a significant difference between model parameters in the case and control samples and returns called variant positions.
+
+\subsection{Model Structure}
+
+For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
+We model the data generation process as a hierarchical Bayesian model; independent models are fit to the case and control data.
+The generative process is as follows:
+
+\begin{enumerate}[noitemsep]
 	\item For each location $j$: 
 	\begin{enumerate}
-		\item Sample a location specific error rate parameter $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
-		\item Choose an inter-replicate precision parameter $M_j$
+		\item Draw an error rate $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
 		\item For each replicate $i$:
 		\begin{enumerate}
-			\item Sample an error rate $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ for location $j$ in sequencing replicate $i$
-			\item Sample the non-reference read count $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
+			\item Draw $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$
+			\item Draw $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
 		\end{enumerate}
@@
 
-This model has three levels of sampling.
-First the baseline error rate across all locations and all replicates is chosen once and for all for the entire data set.
-Then, based on that baseline error rate, location-specific error rates are sampled.
-Based on the location specific error rate, replicate-specific error rates are selected.
-Finally, the observable data, the count of the non-reference reads is sampled.
+This process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
+The global error rate, $\mu_0$, estimates the expected error rate across all locations.
+The global precision, $M_0$, estimates the variation in the error rate  across locations.
+The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.
+
+RVD2 has three levels of sampling.
+First, a global error rate and global precision are chosen once for the entire data set.
+Then, at each location, a local precision is chosen and a local error rate is sampled from a beta distribution.
+Finally, the error rate for replicate $i$ at location $j$ is drawn from a beta distribution and the number of non-reference reads is drawn from a binomial.
+
+RVD2 hierarchically partitions sources of variation in the data.
+$r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
+$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility.
+The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
+Importantly, increasing the read depth $n_{ji}$ only reduces the sampling error, but does nothing to reduce experimental variation or variation due to sequence context.
 
@@
 
-% Describe graphical model framework here
+Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
+In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\cite{}.
+A rounded box or ``plate" represents replication of the nodes within the plate.
+This graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
 
-Given this model structure, the joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
+The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
 
@@
 
-Finally, assuming the conditional independence structure in the graphical model in Figure~\ref{fig:graphical_model}, the likelihood of the data set is
+Finally, the complete likelihood is
 
@@
 
-Instead of attempting to directly compute the posterior distribution, we sample it using a Metropolis-Hastings within Gibbs scheme as described in Section~\ref{sec:algorithm}.
-
 \section{Inference and Hypothesis Testing}
 
-The primary object of inference in this model the joint posterior distribution function over the latent variables,
+The primary object of inference in this model is the joint posterior distribution function over the latent variables,
 \begin{equation}
@@
 
-We use a Metropolis-within-Gibbs sampling approach shown in Algorithm~\ref{alg:metro_gibbs} to approximate the posterior distribution.
-The model parameters are initialized using method-of-moments.
-Given those parameter estimates we first sample from the marginal posterior distribution for $\mu_j$ given it's Markov blanket using Metropolis Hasting.
-Then, we sample from the marginal posterior distribution for $\theta_{ji}$ given it's Markov blanket.
+Exact inference in this model is intractable due to the coupling between the local and global beta distributions. [XXX complete this proof]
+
+Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
+First, the hyperparameters are initialized using method-of-moments.
+Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule.
+Finally, we sample from the marginal posterior distribution for $\theta_{ji}$ given its Markov blanket.
+Samples from $\theta_{ji}$ can be drawn from the posterior distribution directly  because the prior and likelihood form a conjugate pair.
+This sampling procedure is repeated until the chain converges to a stationary distribution then we draw samples from the posterior distribution over latent variables.
 
@@
 \Repeat
-\For {each position j} \Comment{Sample $\mu_j$}
+\For {each location j} \Comment{Sample $\mu_j$}
   \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
@@
 \EndFor
-\Until {convergence and sample size sufficient}
+\Until {sample size sufficient}
 \end{algorithmic}
 \end{algorithm}
-
-Once we have obtained an estimate of the posterior distribution for the case and control data, we use a two-stage test for variant at each position.
-First, we test whether the error rate in the case sample is significantly greater than the error rate in the control data for each position.
-Then, we test the hypothesis that the error rate is high is due to excess reads with a particular non-reference base.
-We accomplish the first test with a Bayesian posterior density test and the second with a $\chi^2$ goodness-of-fit test to a uniform distribution.
-In the remainder of this section, we describe the initialization, MCMC sampling approach and hypothesis testing procedure.
 
@@
 \subsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments(MoM) procedure.
+The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
 MoM works by setting the population moment equal to the sample moment.
@@
 The estimates for the genome-level parameters are $\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ and $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
+While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encounter such pathological behavior in this application.
 
-The method-of-moments is known to have the potential to give estimates that are outside the true parameter bounds.
-For example, the MoM estimate $\tilde{M}_j$ may be negative while $M_j$ must be non-negative.
-We have not encounter such pathological behavior of the estimates in this application.
-
-\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}\
+%%%%%%%%%%%%%%%%%%
+% Sampling theta
+%%%%%%%%%%%%%%%%%%
+\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
 
@@
 \begin{equation}
-	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right),
-\end{equation}
-where we have used the mean-precision parameterization of the beta distribution.
-
-\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
-
-The posterior distribution for $\mu_j$ is
-\begin{equation}
-	p \left( \mu_j |\theta_{ij},\mu_0,M_0\right) = 
-\end{equation}
-\subsection{Bayesian hypothesis test for high error rate}
-
-To detect a mutation in a specific position, a threshold $\tau$ is set for the posterior difference distribution testing.
-\begin{align}
- H_1: \mu_{case}-\mu_{control}\geq\tau \\
- H_2: \mu_{case}-\mu_{control}<\tau
-\end{align}
-
-We can calculate the posterior probability of the hypothesis $H_1$ being true by integrating the posterior density over the correct region:
-
-\begin{equation}
- p(H1:\mu_{case}-\mu_{control}\geq\tau)=\int_\tau^\infty p(\mu_{diff} |r,n; \mu_0,M_0)d\mu_{diff}
+	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
 \end{equation}
 
-We accept the hypothesis $H_1$ if this posterior probability is no lower than 95\%, which means we are 95\% sure that there is a mutation in this position.
-
-ROC curve is plotted to evaluate the performance of the detection algorithm with different threshold for different mutation rate.
-
-\subsection{$\chi^2$ test for non-uniform base distribution}
-
-We assume that at one position in one experimental replicate, the non-reference reads follows multinomial distribution of $k = 3$ categories, with probability vector $p = (p_1, p_2, p_3)$ where $p_i \geq 0$ for $i =1, 2, 3$ and $\sum_{i=1}^3 p_i = 1$.
-
-The non-reference reads can be because of either mutation or various errors including sampling error, inter-replicated error, $etc$.
-In theory, the non-reference read counts caused by non-oriented errors should be uniformly distributed among three non-reference categories, namely $p_1=p_2=p_3=1/3$.
-However, the distribution of non-reference read counts among three non-reference categories caused by mutation theoretically is biased.
-
-The null hypothesis can be denoted as
-\begin{align}
- H_0: p=p_0
-\end{align}
-where
-\begin{align} 
- p_0=(p_1,p_2,p_3), p_1=p_2=p_3=1/3
-\end{align}
-
-The null hypothesis is tested using power-divergence family of statistics proposed by Cressie and Read (1984).By definition, the power-divergence family of statistcs parameterized with $\lambda$ is 
+%%%%%%%%%%%%%%%%%%
+% Sampling mu
+%%%%%%%%%%%%%%%%%%
+\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
+The posterior distribution over $\mu_j$ given its Markov blanket is 
 \begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^k r_i \left[\left(\frac{r_i}{E_i}\right)^\lambda-1\right];\lambda \in R,
-\end{equation}
-where $\{r_i; i=1,...,k\}$ are the observed frequencies and $\{E_i; i=1,...,k\}$ are the  expected frequencies.
-
-Person's $\chi^2 (\lambda = 1)$ statistics, the log likelih ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$ are special cases of the family.
-Cressie and Read (1984) also proposed that when the $\lambda =2/3$, the new power-divergence statistics outcompete the traditional power-divergence statistics.
-
-Particulaly in our case, Let $(r_1,r_2,r_3)$ denotes the non-reference base frequencies and $\sum_{i=1}^3 r_i = r$.
-The expected frquencies are $(rp_1, rp_2, rp_3)$.
-Then the power-divergence statistics can be written as
-
-\begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{i=1}^3 r_i \left[\left(\frac{r_i}{rp_i}\right)^\lambda-1\right];\lambda \in R
+	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
 \end{equation}
 
-We use $\lambda =2/3$ for the goodness of fit testing.
+Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution.
+Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
 
-For each position we have six power-divergence statistics from 6 experimental replicates.
-We use Fisher's method to combines the six statistics into a single one for inference.
-Bonferroni correction is applied to maintaining the familywise error rate (FWER).
-The significance level is chosen as 0.05.
+A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim $. The acceptance probability is then
+\begin{equation}
+	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
+\end{equation}
 
-\subsection{Posterior distribution difference sampling}\
+Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors.
+We have found that the algorithm performance improves when we take the median of ten or more M-H samples as a single Gibbs step for each position.
+This computational effort only minimally increases run time when run on a parallel computing platform.
+In the experiments, we use a XX sample burn-in to allow the Markov chain to approach its stationary distribution and we thin the chain by a factor of 2 to reduce autocorrelation among samples.
 
-Using MCMC sampling, we can obtain the posterior distribution $p \left( \mu |\theta,\mu_0,M_0\right)$ for reference sample and testing sample.
-In order to compare the difference of position error rate $\mu$ between reference sample and testing sample , a random sampling of posterior distribution is done
+%%%%%%%%%%%%%%%%%%
+% Posterior Density Test
+%%%%%%%%%%%%%%%%%%
+\subsection{Hypothesis Test}
+Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data.
+For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
 
-\begin{enumerate}
- \item Randomly  draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0, M_0\right)$ from reference samples
- \item Randomly draw one Gibbs sample from $p \left( \mu |r, n, M, \mu_0,M_0\right) $ from testing samples
- \item Subtract the former from the later to get one sample for the posterior difference distribution 
-\end{enumerate}
-Repeat the above three steps until sample size is sufficient.
+A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
+\begin{equation}
+	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
+\end{equation}
+where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
+We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
 
-With the posterior difference distribution of position error rate $\mu$ across replicates obtained, Bayesian inference can be done to summarize the difference.
+\section{Comparison with other variant calling algorithms}
+
+GATK uses considerable preprocessing and a simple naive Bayes decision rule to call a variant.
+The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score \cite{McKenna2010}.
+Samtools
+
+%%%%%%%%%%%%%%%%%%
+
+% Chi^2 Test
+
+%%%%%%%%%%%%%%%%%%
+%\subsection{$\chi^2$ test for non-uniform base distribution}
+%
+%We assume the counts for the non-reference bases at poisiuton $j$ in replicate $i$ follow a multinomial distribution with parameter  $p = (p_1, p_2, p_3)$ where $p_k \geq 0$ for $k =1, 2, 3$ and $\sum_{k=1}^3 p_k = 1$.
+%
+%The non-reference reads may be due to a true mutation or due to a random sequencing error.
+We assume, non-reference read counts caused by a random mechanism results in a uniform distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$.
+In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.
+%
+%We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios.
+The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
+Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$.
+The test statistic is
+%
+%\begin{equation}
+% 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
+%\end{equation}
+%
+%where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
+Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.
+%
+%For each position, there are $N$ power-divergence statistics for each experimental replicate.
+We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position,
+%
+%\begin{equation}
+%	X^2 = -2 \sum_{i=1}^N \ln(p_i).
+%\end{equation}
+%This gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.
+%
+%While Fisher's combined probability test collapses the p-values across replicates, we still have $J$ tests across positions.
+We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions.
 
@@
 
-\subsection{Overall of the process}
-Figure 1
+\subsection{Performance with read depth}\
 
-\subsection{Performance with read depth}\
-Figure 2 ROC plot showing performance with different read depth
+We synthesized a 400bp control DNA sequence and another 400bp sequence that is identical to the control except at 14 loci with variant bases.
+We denote the indices of these true positive variant positions $j^*$.
+Pure case and control samples were mixed at defined fractions to yield minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+The sequencing protocol and data set are available from the original publication~\cite{}.
+
+BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ to simulate lower coverage data while retaining the error structure of real NGS data.
+The final data set contains pairs of reads for three replicates of each case and a pair of reads for the control sample giving a total of 36 BAM files for each downsampled data set.
+
+We ran the RVD2 algorithm on each downsampled data set sand varied the threshold $\tau$ with fixed $\alpha=0.05$ to generate receiver operating characteristic (ROC) curves in Figure~\ref{fig:ROC}.
+Figure~\ref{ig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
+At the lowest depth the sensitivity and specificity is no better than a random decision.
+We would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43.
+The performance improves monotonically and is much better at a depth of approximately 41,000.
+This depth exceed the inverse MAF by a $10\times$ which we find to be generally a good coverage threshold for detecting variants.
+We find a similar pattern in Figures~\ref{fig:ROC}B-C.
+Interestingly, the algorithm performance is much better than a random decision rule to detect a 10\% MAF with only a depth of 25.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=80mm]{pdf_figs/ROC_without_chi2.pdf}
+\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\label{fig:ROC}
+\end{center}
+\end{figure}
+
+Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 41,573$.
+The only false positive is at position 8, though posterior distributions for the case and control clearly show a difference at that position.
+The posterior mean estimates are shrunken towards the global error rate parameter $\mu_0 = 0.0023$ for the case and control due to the empirical Bayes procedure.
 
@@
 \begin{center}
-\includegraphics[width=160mm]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
-\label{fig:ROC_without_chi2}
+\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
+\caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
+\label{fig:MAF}
 \end{center}
@@
 
+\subsection{Comparison Results on Synthetic DNA}\
+
+Figure 3 ROC comparison with other methods: varscan2/samtools/rvd/strelka/virmid/muTect
+
+\subsection{Pooled Multiple-Sclerosis Samples}\
+
+We tested our algorithm on data from pooled samples from patients with multiple-sclerosis (MS).
+The MS patients are ``high-burden"; meaning that they are known to carry one or more of several genetic risk alleles.
+Our aim with studying these patients is to identify rare risk alleles that may not yet be known to be associated with MS.
+
+DNA was extracted from whole blood samples from 100 patients with multiple sclerosis and 100 unaffected, unrelated individuals.
+Each sample was subjected to targeted PCR amplification of IL2RA and interleukin 7 receptor (IL7R); genes with known risk alleles~\cite{HauserNEJM2007}.
+We targeted X eons covering XXbp in IL2RA and X exon covering XXbp in IL7R.
+The experiment was repeated on two independent cohorts to filter for spurious associations due to population sampling.
+
+We took 4000 samples from the Markov chain, removed the first 10\% for burn-in and thinned the chain by a factor of 2.
+We used 10 Metropolis samples for each Gibbs step.
+We called variants with $\tau = 0.005$ and $\alpha = 0.05$ because we expect that the minimum detectable genotype is a heterozygous mutation in 1 out of 100 samples.
+
+We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA.
+All position information is referenced to human genome reference hg19.
+One variant in IL7R is rs1494558 at chr5:35861068 T$>$C and has a MAF of 76\% and 71\% in the two cohorts.
+This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\cite{HahnCho2011}.
+The second IL7R variant is at chr5:35875593 A$>$T and has a MAF of 60\% and 55\% in the two cohorts.
+This variant has not yet been shown to be associated with any known disease.
+The IL2RA variant is at chr10:6061965 A$>$T and has a MAF  of 3\% and 5\% in the two cohorts.
+This variant has not yet been shown to be associated with any known disease.
+
 \begin{figure}[h]
 \begin{center}
-\includegraphics[width=160mm]{pdf_figs/MuBarPlot.pdf}
-\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
-\label{fig:ROC_without_chi2}
+%\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
+\caption{Estimated minor allele fractions for consensus variants in MS patient pools.}
+\label{fig:MS_MAF}
 \end{center}
@@
 
-We tested the mutation detection power of our graphical model on synthetic dataset using a two-stage test scheme with dilution rates $0.1\%$, $0.3\%$, $1.0\%$, $10.0\%$ and read depth varying from $10^2$ to $10^5$.Different read depth dataset was achieve by thining dataset with read depth up to $10^6$ using Picard.
-$\chi^2$ test, the second stage test can be optionally included based on average read depth.
-
-Excluding $\chi^2$ test, the ROC curves of our graphical model using Bayesian Hypothesis Testing with different threshold $\tau$ is shown in Figure??.
-The model had high detection power when the read depth was at $10^5$ level.
-At dilution rate at $0.1\%$, the model achieved performance with sensitivity at 99.74\% and specificity at 92.86\%.
-We missed one mutation at position 205 and had one false positive at position 8.
-The threshold was chosen at 0.022\% which achieved the optimal sensitivity and specificity in the ROC curve.
-As expected, the detection power of the graphical model decreased when the read depth was reduced.
-
-\begin{figure}[h] %needs to go to supplementary document.
-\begin{center}
-\includegraphics[width=160mm]{pdf_figs/ROC_subplots_with_chi2.pdf}
-\caption{ROC curve varying read depth showing method performance with Bayesian Hypothesis Testing and Chi-Square Testing}
-\label{fig:R}
-\end{center}
-\end{figure}
-
-Figure ??
-shows the ROC curves when both Bayesian Hypothesis Testing and $\chi^2$ testing are carried out for variant calling.
-Comparing with Figure ??, it can be seen that when read depth is low (from $10$ to $10^3$), the false positive rate is severely limited to zero, which means we are very confident in our variant callings.
-However, the low false positve rate can be achieved at a sacrifice of low true postive rate especially when the dilution rate and read depth are low.
-The $\chi^2$ test is a stringent test when the read depth is low.
-
-However, if the read depth is high enough (up to $10^4$ or higher), the $\chi^2$ test can limit the false positive rate in a slight degree but no longer at a sacrifice of True Positive Rate.
-Actually the performance is not much influenced by $\chi_2$ testing with high read depth.
-
-\subsection{Comparison Results on Synthetic DNA}\
-
-Figure 3 ROC comparison with other methods: varscan/samtools/rvd
-
-\subsection{Empirical Results on Multiple Sclerosis Data}\
-We tested our method on sequence data from clinical Multiple Sclerosis samples. We expect to see that
-
-Table: Table of Variants
+The validity of these loci are subject to confirmation on a larger independent sample.
+However, we note that most GWAS studies focus on common SNPs due to the cost-effective microarray technology used for those studies.
+We have instead used sequencing technology which yields and unbiased coverage of the targeted regions and has the potential to uncover rare variants that are not targeted by microarray technology.
+Indeed, the common SNP that was found is associated with an autoimmune disorder and the other two variants are not at dbSNP loci.
 

### 2013-10-03 doc/RECOMB2014/recomb2014.tex
create RECOMB2014 manuscript

--- 2013-10-03 doc/bioinformatics2013/RVD2.tex
@@
 \usepackage{setspace}
-\doublespacing
+%\doublespacing
 %\usepackage{epstopdf}
@@
 
-\title[RVD2]{RVD2: An ultra-sensitive variant detection model for heterogeneous population sequencing}
+\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
 \author{}
@@
 
-\section{Comparison with other variant calling algorithms}
+\subsection{Comparison to other algorithms}
 
+We compare the RVD2 model structure to other variant calling algorithms here.
 GATK uses considerable preprocessing and a simple naive Bayes decision rule to call a variant.
 The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score \cite{McKenna2010}.
-Samtools
+
+[XXX Finish this text]
 
@@
 \begin{center}
-\includegraphics[width=80mm]{pdf_figs/ROC_without_chi2.pdf}
+\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
 \caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
@@
 \begin{center}
-\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
+\includegraphics[width=120mm]{pdf_figs/MuBarPlot.pdf}
 \caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
@@
 
-\subsection{Comparison Results on Synthetic DNA}\
+\subsection{Empirical performance compared with other algorithms}
+
+We compare the empirical performance of RVD2 to other variant calling algorithms.
+We use the synthetic DNA data sets described in the previous section to explore performance across a dynamic range of read depths as well as minor-allele frequencies.
+We report the sensitivity and specificity of each algorithm to detect the 14 true positive variant positions out of 400 tested positions.
+
+GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
+Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
+Varscan-somatic has been shown to outperform XXX,XXX, and XXX to detect a 5\% MAF for read depths between 100 and 1000.
+We show its performance on much lower MAF variants and across a wider range of coverage depths.
+Because Varscan2-somatic does not accept replicate data for the ``normal" or ``tumor" bam files, we used a single bam file from each group.
+
+Table~\ref{t/l:comparison} shows the sensitivity and specificity for other variant calling algorithms. 
+[XXX Finish this text]
 
@@
 
-\subsection{Pooled Multiple-Sclerosis Samples}\
+\subsection{Pooled Multiple-Sclerosis Samples}
 

### 2013-10-03 doc/RECOMB2014/recomb2014.tex
added discussion of variant calling algorithm comparison.

@@
 \usepackage{setspace}
-%\doublespacing
+\doublespacing
+
+\usepackage{natbib}
+
 %\usepackage{epstopdf}
@@
 \section{Introduction}
-
+\citep{Ferretti:2013hp}
 % DIscuss low-purity samples and somatic mutations
 
-Next-generation sequencing (NGS) technology has enabled the genomic interrogation for a fraction of the cost of traditional assays~\cite{Koboldt2013Cell}.
-Protocol and platform engineering improvements have reduced the cost of the assay to the point where it is now possible to generate $1\times10^9$ bases of sequence data in under 40 hours for less than \$1000 \cite{}.
-As a result NGS is increasingly being used as a general platform for research assays for methylation state \cite{}, DNA mutations \cite{}, copy number variation \cite{}, promoter occupancy \cite{} and others \cite{}.
-Recently, NGS has been used as a clinical diagnostic tool for testing for mutations in non-invasive fetal DNA \cite{}, infectious disease samples \cite{}, cancerous tumors \cite{}, and human microbial communities \cite{}.
+Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}.
+Protocol and platform engineering improvements have reduced the cost of the assay to the point where it is now possible to generate $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}.
+As a result NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
+Recently, NGS has been used as a clinical diagnostic tool for testing for mutations in non-invasive fetal DNA~\citep{Kitzman:2012hea}, infectious disease samples~\citep{Capobianchi:2012em}, cancer~\citep{Navin:2010gu}, and human microbial communities~\citep{Consortium:2013iz}.
 
 Clinical samples are often genetically heterogeneous.
-Non-invasive fetal DNA testing uses a sample of the maternal blood and filters for the minor fraction that is cell-free DNA from the fetus \cite{}.
-Infectious diseases such as HIV and influenza mutate rapidly and a given sample from an individual host may contain many sub-populations \cite{}.
-DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample.
+For example, non-invasive fetal DNA testing uses a data from maternal blood sample and filters for the minor fraction of reads that originate from cell-free DNA from the fetus~\citep{Fan:2008di}.
+Infectious diseases such as HIV and influenza may contain many sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
+DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
 
@@
 
-The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data.
-We present the hierarchical Bayesian model for the method in Section X and the Bayesian hypothesis test to detect mutations in Section X.
-In Section X we present sensitivity and specificity results of our method on pure DNA samples mixed at defined fractions.
+The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology.
+We present the hierarchical Bayesian model in Section X and a Bayesian hypothesis test to detect mutations in Section X.
+In Section X we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
 We compare our algorithm to the most accurate methods available to date in Section X.
 Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients.
-This strategy effectively reduces the cost of the assay by $n$-fold where $n$ is the pool size.
 
@@
 
-RVD uses a two-stage approach for detecting for rare variants.
+RVD uses a two-stage approach for detecting for rare variants; the first stage is a model estimation and the second stage is a hypothesis test.
 First, it estimates the parameters of two hierarchical Bayesian models; one using data from the sample of interest (case) and one using data from a known reference sample (control).
@@
 Table~\ref{t/l:comparison} shows the sensitivity and specificity for other variant calling algorithms.
-[XXX Finish this text]
+Each cell contains the sensitivity and specificity of the method separated by a slash.
+The standard genotype variant calling algorithms: samtools, GATK, and VarScan2-mpileup are shown in the first columns.
+Samtools performs very well on 100\% MAF sample and has nearly perfect sensitivity and specificity for median coverage depths of 27 and 298.
 
-Figure 3 ROC comparison with other methods: varscan2/samtools/rvd/strelka/virmid/muTect
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=140mm]{pdf_figs/comparison_table.pdf}
+\caption{Comparison of RVD2 with other variant calling algorithms.}
+\label{fig:comparison}
+\end{center}
+\end{figure}
+
+VarScan2-somatic has perfect performance on a 100\% MAF samples at a low depth.
+As the depth increases, it calls more false positives and the specificity decreases monotonically from 1.0 to 0.31.
+For MAFs between 0.1\% and 10\% specificity generally decreases with read depth similarly.
+There appears to be an optimal read depth for sensitivity.
+The 100\% MAF sample shows optimal sensitivity at a depth of 27, the 10\% optimal depth is 260 and the 1\% optimal depth is 5584.
+One would expect that increasing the read depth would lead to better sensitivity, but contrary to expectation further increasing the depth decreases performance.
+This effect is not exclusive to VarScan2-somatic and is also observed in samtools and GATK.
+
+Strelka has better sensitivity and specificity than the previous algorithms for 100\% and 10\% MAF.
+However, it has zero sensitivity to detect variants at or below 1\% MAF.
+MuTect [XXX complete this section]
 

### 2013-10-04 doc/bioinformatics2013/RVD2.tex
edits to rvd2 manuscript. moved recomb back to bioinformatics.

@@
 \documentclass[11pt,reqno]{amsart}
-\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
+\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
 \geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
@@
 \usepackage{setspace}
-\doublespacing
+%\doublespacing
+
+\usepackage{natbib}
+
 %\usepackage{epstopdf}
@@
 \begin{document}
+
+\begin{abstract}
+Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
+Unlike most research samples, these clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
+However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
+We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and a Bayesian posterior hypothesis test to call variants in heterogeneous samples.
+We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depths and a minor allele frequencies.
+We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
+\end{abstract}
+
 \maketitle
@@
 
-% DIscuss low-purity samples and somatic mutations
-
-Next-generation sequencing (NGS) technology has enabled the genomic interrogation for a fraction of the cost of traditional assays~\cite{Koboldt2013Cell}.
-Protocol and platform engineering improvements have reduced the cost of the assay to the point where it is now possible to generate $1\times10^9$ bases of sequence data in under 40 hours for less than \$1000 \cite{}.
-As a result NGS is increasingly being used as a general platform for research assays for methylation state \cite{}, DNA mutations \cite{}, copy number variation \cite{}, promoter occupancy \cite{} and others \cite{}.
-Recently, NGS has been used as a clinical diagnostic tool for testing for mutations in non-invasive fetal DNA \cite{}, infectious disease samples \cite{}, cancerous tumors \cite{}, and human microbial communities \cite{}.
+Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}.
+Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}.
+As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
+NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.
 
 Clinical samples are often genetically heterogeneous.
-Non-invasive fetal DNA testing uses a sample of the maternal blood and filters for the minor fraction that is cell-free DNA from the fetus \cite{}.
-Infectious diseases such as HIV and influenza mutate rapidly and a given sample from an individual host may contain many sub-populations \cite{}.
-DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample.
+For example, non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
+Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
+DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
 
@@
 Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
-Samtools/bcftools use a relatively straightforward XXX to call genotypes \cite{}.
-GATK uses a naive Bayesian decision rule to call a variant as one of two possible alleles \cite{McKenna2010}.
-VarScan2-mpileup uses information from multiple replicates to call homogeneous and heterogeneous genotypes.
+Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{}.
+GATK involves more sophisticate pre and post-processing steps.
+The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
 
 Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
-MuTect is able to handle low-purity samples by computing a log-odds score.
-If the likelihood of the data assuming the locus is mutated exceeds the likelihood of the data assuming the site is not mutated by a threshold then a variant is called.
-There is an additional log-odds check on the normal sample to filter false positives~\cite{Cibulskis2013}.
-Strelka models the joint distribution for a tumor and normal sample as a mixture model using continuous values allele frequencies \cite{}.
-VarScan2 has a subcommand to call somatic mutations \cite{}.
-Recently VarScan2 was shown to outperform many other algorithms for calling low-frequency (5\%) variants \cite{}.
+VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
+Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders}.
+With the joint distribution available, one is able to identify locations with dissimilar allele frequencies.
+muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis}.
 
-The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data.
-We present the hierarchical Bayesian model for the method in Section X and the Bayesian hypothesis test to detect mutations in Section X.
-In Section X we present sensitivity and specificity results of our method on pure DNA samples mixed at defined fractions.
+The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology.
+We present the hierarchical Bayesian model in Section X and a Bayesian hypothesis test to detect mutations in Section X.
+In Section X we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
 We compare our algorithm to the most accurate methods available to date in Section X.
-Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients.
-This strategy effectively reduces the cost of the assay by $n$-fold where $n$ is the pool size.
+Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.
 
@@
 For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-We model the data generation process as a hierarchical Bayesian model; independent models are fit to the case and control data.
 The generative process is as follows:
@@
 
-\section{Comparison with other variant calling algorithms}
-
-GATK uses considerable preprocessing and a simple naive Bayes decision rule to call a variant.
-The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score \cite{McKenna2010}.
-Samtools
-
 %%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Performance with read depth}\
+\subsection{Performance with read depth}
 
@@
 
-\subsection{Comparison Results on Synthetic DNA}\
+\subsection{Empirical performance compared with other algorithms}
 
-Figure 3 ROC comparison with other methods: varscan2/samtools/rvd/strelka/virmid/muTect
+We compare the empirical performance of RVD2 to other variant calling algorithms.
+We use the synthetic DNA data sets described in the previous section to explore performance across a dynamic range of read depths as well as minor-allele frequencies.
+We report the sensitivity and specificity of each algorithm to detect the 14 true positive variant positions out of 400 tested positions.
 
-\subsection{Pooled Multiple-Sclerosis Samples}\
+GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
+Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
+\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
+We find its performance on much lower MAF variants and across a wider range of coverage depths.
+Varscan2-somatic, streak and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the median overall median depth of the replicates.
+
+Table~\ref{t/l:comparison} shows the sensitivity and specificity for other variant calling algorithms.
+We computed the Matthews Correlation Coefficient (MCC) as a combined performance measure and present those results in Appendix B~\cite{}.
+Each cell contains the sensitivity and specificity of the method separated by a slash.
+The standard genotype variant calling algorithms: samtools, GATK, and VarScan2-mpileup are shown in the first columns.
+Samtools performs very well on 100\% MAF sample and has nearly perfect sensitivity and specificity for median coverage depths of 27 and 298.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=140mm]{pdf_figs/comparison_table.pdf}
+\caption{Comparison of RVD2 with other variant calling algorithms (sensitivity / specificity).}
+\label{fig:comparison}
+\end{center}
+\end{figure}
+
+VarScan2-somatic has perfect performance on a 100\% MAF samples at a low depth.
+As the depth increases, it calls more false positives and the specificity decreases monotonically from 1.0 to 0.31.
+For MAFs between 0.1\% and 10\% specificity generally decreases with read depth similarly.
+There appears to be an optimal read depth for sensitivity.
+The 100\% MAF sample shows optimal sensitivity at a depth of 27, the 10\% optimal depth is 260 and the 1\% optimal depth is 5584.
+One would expect that increasing the read depth would lead to better sensitivity, but contrary to expectation further increasing the depth decreases performance.
+This effect is not exclusive to VarScan2-somatic and is also observed in samtools and GATK.
+
+Strelka has better sensitivity and specificity than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths and the 100\% and 10\% MAF.
+However, it has zero sensitivity to detect variants at or below 1\% MAF.
+
+MuTect has the best overall performance of the other algorithms we tested over a wide range of MAF and depths.
+Unlike several of the other algorithms muTect generally improves with a higher read depth or the specificity degradation due to false positives is modest.
+Specifically, muTect calls some of the true variant positions with a 0.3\% MAF and a median read depth of only 53 and 410.
+
+We tested our model RVD2 using two variations of the hypothesis test.
+We set $\tau$, the hypothesis test threshold to the naive value of zero and we optimized $\tau$ for the read depth and MAF ($\tau^*$).
+These optimal values were computed from this data set, but in general may be computed once and used as a lookup table for future experiments.
+With $\tau=0$, RVD2 is not able to call variants with a MAF of 0.1\% if the read depth is less than 4,129.
+This generally agrees with previous observations and intuition that more than 1,000 reads are required to call a 1:1000 MAF.
+The specificity degrades slightly at higher read depths with this fixed $\tau$ in line with muTect.
+At high read depths and MAFs classification is perfect.
+With $\tau$ set to the optimal value $\tau^*$, we find a significant improvement in both sensitivity and specificity.
+The algorithm is able to call a 0.1\% MAF with a median read depth of only 39.
+This performance is due to two properties: (1) the empirical Bayes structure of the model allows parameters $\mu_j$ to borrow strength from one another through $\mu_0$ and (2) the actual coverage at a particular location may be higher than the median coverage across all locations.
+
+\subsection{Pooled Multiple-Sclerosis Samples}
 
@@
 The MS patients are ``high-burden"; meaning that they are known to carry one or more of several genetic risk alleles.
-Our aim with studying these patients is to identify rare risk alleles that may not yet be known to be associated with MS.
+Our aim in studying this  patient population is to identify rare risk alleles that may not yet be known to be associated with MS.
 
@@
 Each sample was subjected to targeted PCR amplification of IL2RA and interleukin 7 receptor (IL7R); genes with known risk alleles~\cite{HauserNEJM2007}.
-We targeted X eons covering XXbp in IL2RA and X exon covering XXbp in IL7R.
+We targeted X exons covering XXbp in IL2RA and X exon covering XXbp in IL7R.
 The experiment was repeated on two independent cohorts to filter for spurious associations due to population sampling.
@@
 
+\section{Comparison to other algorithms}\label{sec:comparison}
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
+\caption{Comparison of RVD2 with other variant calling algorithms (MCC).}
+\label{fig:comparison_mcc}
+\end{center}
+\end{figure}
+
 \bibliographystyle{apalike}

### 2013-10-07 doc/bioinformatics2013/RVD2.tex
updates to manuscript

@@
 \usepackage{setspace}
-%\doublespacing
+\doublespacing
 
@@
 Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}.
-As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
+As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
 NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.
@@
 VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
-Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders}.
+Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}.
 With the joint distribution available, one is able to identify locations with dissimilar allele frequencies.
-muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis}.
+muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.
 
 The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology.
-We present the hierarchical Bayesian model in Section X and a Bayesian hypothesis test to detect mutations in Section X.
-In Section X we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
-We compare our algorithm to the most accurate methods available to date in Section X.
-Finally, in section X we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.
+We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}.
+In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
+We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
+Finally, in Section~\ref{sec:ms} we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.
 
@@
 
-\subsection{Model Structure}
+\subsection{Model Structure}\label{sec:model_structure}
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Hypothesis Test}
+\subsection{Hypothesis Test}\label{sec:hypothesis_test}
 Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data.
@@
 We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
+
+The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ maximizes the Matthews correlation coefficient.
+\begin{equation}
+	\tau^* = \arg\max_\tau \text{MCC}(\tilde{n}, \text{MAF})
+\end{equation}
 
@@
 
-\subsection{Performance with read depth}
+\subsection{Performance with read depth}\label{sec:read_depth}
 
@@
 
-\subsection{Empirical performance compared with other algorithms}
+\subsection{Empirical performance compared with other algorithms}\label{sec:comparison}
 
@@
 
-\subsection{Pooled Multiple-Sclerosis Samples}
+\subsection{Pooled Multiple-Sclerosis Samples}\label{sec:ms}
 
@@
 One variant in IL7R is rs1494558 at chr5:35861068 T$>$C and has a MAF of 76\% and 71\% in the two cohorts.
-This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\cite{HahnCho2011}.
+This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\citep{Hahn:2009bi}.
 The second IL7R variant is at chr5:35875593 A$>$T and has a MAF of 60\% and 55\% in the two cohorts.
@@
 
-\section{Comparison to other algorithms}\label{sec:comparison}
+\section{Comparison to other algorithms}\label{sec:app_comparison}
 \begin{figure}[h]

### 2013-11-04 doc/bioinformatics2013/RVD2.tex
Update on HCC1197 description

@@
 \begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table.pdf}
+%\includegraphics[width=140mm]{pdf_figs/comparison_table.pdf}
 \caption{Comparison of RVD2 with other variant calling algorithms (sensitivity / specificity).}
@@
 Indeed, the common SNP that was found is associated with an autoimmune disorder and the other two variants are not at dbSNP loci.
+\subsection{HCC1187 primary ductual carcinoma sample}
+We test our algorithm on normal-tumor paired data from sample HCC1187, a 41 years old white female with primary ductal carcinoma in TNM stage IIA, grade 3.
+The tumor DNA was extracted from HCC1187 breast ductal carcinoma cell line while the normal DNA from a matching lymphoblastoid cell line.
 
+500 ng of DNA were prepped using an early access version of Illumina’s TruSeq DNA PCR-Free kit.
+The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
+Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% of loci covered with 10 or more reads.
 \appendix
@@
 \begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
+%\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
 \caption{Comparison of RVD2 with other variant calling algorithms (MCC).}

### 2013-11-04 results/2013-10-30_HCC1187_TP53/bayesTest.py
Add somatic status classification

@@ -63,7 +63,7 @@ def main():
     f.create_dataset('chi2pvalue',data=chi2P)
     f.close()
 
-    idx = [i for i, j in enumerate(postP) if j>=0.99]
+    idx = [i for i, j in enumerate(postP) if j>=0.95]
 
     # Read in called position depth chart
     contdc=pd.read_table('depth_chart/HCC1187BL_S1.dc')
@@ -93,13 +93,13 @@ def main():
             print >> f, i+1,'\t', caseR[0,i,0],'\t',caseR[0,i,1],'\t',caseR[0,i,2]
         f.close()   
     with open('result.txt','w') as f:
-        print >> f, 'loc\tpostP\trefb\tA0\tC0\tG0\tT0\tA1\tC1\tG1\tT1'
-        for f1, f2,f3,f4,f5,f6,f7,f8,f9,f10,f11 in zip(loc[idx], postP[idx],refb[idx],
+        print >> f, 'count\tloc\tpostP\trefb\tA0\tC0\tG0\tT0\tA1\tC1\tG1\tT1'
+        for f0, f1, f2,f3,f4,f5,f6,f7,f8,f9,f10,f11 in zip(idx,loc[idx], postP[idx],refb[idx],
                                                        A0[idx],C0[idx],
                                                        G0[idx],T0[idx],
                                                        A1[idx],C1[idx],
                                                        G1[idx],T1[idx],):
-            print >> f, f1,'\t',f2,'\t',f3,'\t',f4,'\t',f5,'\t',f6,'\t',f7,'\t',f8,'\t',f9,'\t',f10,'\t',f11
+            print >> f, f0,'\t', f1,'\t',f2,'\t',f3,'\t',f4,'\t',f5,'\t',f6,'\t',f7,'\t',f8,'\t',f9,'\t',f10,'\t',f11
         f.close()
         
 if __name__ == "__main__":

### 2013-11-04 results/2013-10-30_HCC1187_PAXIP1/runall.py
Experiment on gene PAXIP1 on HCC1187 dataset.

@@ -1,0 +1,57 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../rvd2/src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=60)
+#pool = None
+
+gibbs_nsample = 8000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["depth_chart/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():    
+    # Estimate the model for the cases
+    logging.debug("Processing control data.")
+    controlList = ['HCC1187BL_S1.dc',]
+    fit_model(controlList, 'Control.hdf5')
+
+
+    # Estimate the model for the cases
+    logging.debug("Processing case data.")
+    caseList = ['HCC1187C_S1.dc',]
+    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2013-11-04 results/2013-11-03_HCC1187_TP53_somatic/bayesTest.py
Experiment on gene TP53 in HCC1187 dataset. Add somatic status classification.

@@ -1,0 +1,151 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import pandas as pd
+
+os.chdir('S:/yhe2/Research/rvd2/results/2013-11-03_HCC1187_TP53_somatic')
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../rvd2/src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    N=1000 # Z sampling size  
+            
+    # Load the control model samples
+    controlFile="../2013-10-30_HCC1187_TP53/Control.hdf5"
+    (controlPhi, controlTheta, controlMu, controlloc, controlR, controlN) = rvd27.load_model(controlFile)
+    
+    caseFile="../2013-10-30_HCC1187_TP53/Case.hdf5"
+    (casePhi, caseTheta, caseMu, caseLoc, caseR, caseN) = rvd27.load_model(caseFile)
+    
+    T = 0 # detection threshold
+    print("Detection Threshold = %f" % T)
+
+    # Extract the common locations in case and control
+    caseLocIdx = [i for i in xrange(len(caseLoc)) if caseLoc[i] in controlloc]
+    controllocIdx = [i for i in xrange(len(controlloc)) if controlloc[i] in caseLoc]
+
+    caseMu = caseMu[caseLocIdx,:]
+    controlMu = controlMu[controllocIdx,:]
+    caseR = caseR[:,caseLocIdx,:]
+    controlR = controlR[:,controllocIdx,:]
+    loc = caseLoc[caseLocIdx]
+    J = len(loc)
+    
+    # Sample from the posterior Z = muCase - muControl        
+    (Z, caseMuS, controlMuS) = rvd27.sample_post_diff(caseMu, controlMu, N)
+
+    postP = rvd27.bayes_test(Z, [(T, np.inf)]) # Posterior Prob that muCase is greater than muControl by T
+
+    def Mu_distr(Z, roi):
+        """ Return posterior probabilities in regions defined in list of tuples (roi)
+            from samples in columns of Z. """
+
+        (J,N)=np.shape(Z)
+        
[…]

### 2013-11-04 results/2013-11-03_HCC1187_TP53_somatic/runall.py

@@ -1,0 +1,57 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../rvd2/src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=48)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["depth_chart/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():    
+    # Estimate the model for the cases
+    logging.debug("Processing control data.")
+    controlList = ['HCC1187BL_S1.dc',]
+    fit_model(controlList, 'Control_trial.hdf5')
+
+
+    # Estimate the model for the cases
+    logging.debug("Processing case data.")
+    caseList = ['HCC1187C_S1.dc',]
+    fit_model(caseList, 'Case_trial.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2013-11-06 doc/bioinformatics2013/RVD2.tex

@@
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depths and a minor allele frequencies.
-We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
+%We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
 \end{abstract}
@@
 The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology.
+Typically such a variant calling algorithms are comprised of multiple pre- and post filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
+These steps may improve algorithm performance, but it is difficult to isolate what step is most responsible for the good performance.
+Instead, we focus exclusively on the model structure and statistical inference.
+Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study.
+We have isolated the statistical inference and hypothesis testing steps so that they may be used independently or as part of a larger variant calling pipeline.
+
 We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}.
@@
 We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
-Finally, in Section~\ref{sec:ms} we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.
+%Finally, in Section~\ref{sec:ms} we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.
 
@@
 First, a global error rate and global precision are chosen once for the entire data set.
-Then, at each location, a local precision is chosen and a local error rate is sampled from a beta distribution.
-Finally, the error rate for replicate $i$ at location $j$ is drawn from a beta distribution and the number of non-reference reads is drawn from a binomial.
+Then, at each location, a local precision is chosen and a local error rate is sampled from a Beta distribution.
+Finally, the error rate for replicate $i$ at location $j$ is drawn from a Beta distribution and the number of non-reference reads is drawn from a binomial.
 
@@
 
-\begin{gather}
-= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1} \\
-\cdot \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1} \\
-\cdot \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}
-\end{gather}
+The individual conditional distributions are:
+\begin{align}
+p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
+p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
+p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
+\end{align}
 
@@
 \begin{equation}
-p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j
+p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
 \end{equation}
 
-Finally, the complete likelihood is
+Finally, the complete log-likelihood for the data set is
 
 \begin{equation}
-p \left( r ; \mu_0, M_0, M \right) = \prod_{j=1}^J p\left( \mu_j; \mu_0, M_0 \right) \prod_{i=1}^N p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) 
+\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
 \end{equation}
@@
 
-Exact inference in this model is intractable due to the coupling between the local and global beta distributions. [XXX complete this proof]
+The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution.
+However, there is not a closed form for the product of a Beta distribution with another Beta distribution and exact inference is intractable.
+exact inference is intractable.
 
 Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
-First, the hyperparameters are initialized using method-of-moments.
+First, the hyperparameters are initialized using method-of-moments (MoM).
 Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule.
@@
 
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim $. The acceptance probability is then
+A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, 0.1 \cdot \mu_j^{(p)})$. The acceptance probability is then
 \begin{equation}
@@
 
-The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ maximizes the Matthews correlation coefficient.
+The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
+The optimal $\tau$ maximizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot.
 \begin{equation}
-	\tau^* = \arg\max_\tau \text{MCC}(\tilde{n}, \text{MAF})
+	\tau^* = \arg\max_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}
 \end{equation}
+
+While we are able to compute the optimal $\tau$ threshold for this test data set, in general we would not have access to $\tau^*$.
+With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest.
 
@@
 We use the synthetic DNA data sets described in the previous section to explore performance across a dynamic range of read depths as well as minor-allele frequencies.
-We report the sensitivity and specificity of each algorithm to detect the 14 true positive variant positions out of 400 tested positions.
+We report false discovery rate, sensitivity and specificity for each algorithm to detect the 14 true positive variant positions out of 400 tested positions.
+In a research application, the false discovery rate is more relevant because the aim is generally to identify interesting variants and not spend resources following up false positives.
+The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in calling all of the true variants and none of the true negatives.
 
@@
 
-Table~\ref{t/l:comparison} shows the sensitivity and specificity for other variant calling algorithms.
-We computed the Matthews Correlation Coefficient (MCC) as a combined performance measure and present those results in Appendix B~\cite{}.
-Each cell contains the sensitivity and specificity of the method separated by a slash.
+\subsubsection*{False Discovery Rate Comparison}
+Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 and other variant calling algorithms.
+We computed the Matthews Correlation Coefficient (MCC) and sensitivity and specificity as a combined performance measure and present those results in Appendix B~\cite{}.
+Blank cells indicate no positive calls were made.
 The standard genotype variant calling algorithms: samtools, GATK, and VarScan2-mpileup are shown in the first columns.
-Samtools performs very well on 100\% MAF sample and has nearly perfect sensitivity and specificity for median coverage depths of 27 and 298.
+Samtools performs very well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
+GATK performs well to detect both 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
+VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
 
@@
 \begin{center}
-%\includegraphics[width=140mm]{pdf_figs/comparison_table.pdf}
-\caption{Comparison of RVD2 with other variant calling algorithms (sensitivity / specificity).}
-\label{fig:comparison}
+\includegraphics[width=140mm]{pdf_figs/comparison_table_fdr.pdf}
+\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
+\label{tbl:comparison_fdr}
 \end{center}
@@
 
-VarScan2-somatic has perfect performance on a 100\% MAF samples at a low depth.
-As the depth increases, it calls more false positives and the specificity decreases monotonically from 1.0 to 0.31.
-For MAFs between 0.1\% and 10\% specificity generally decreases with read depth similarly.
-There appears to be an optimal read depth for sensitivity.
-The 100\% MAF sample shows optimal sensitivity at a depth of 27, the 10\% optimal depth is 260 and the 1\% optimal depth is 5584.
-One would expect that increasing the read depth would lead to better sensitivity, but contrary to expectation further increasing the depth decreases performance.
+VarScan2-somatic generally has a poor performance across all MAF and read depths.
+The FDR is close to 1 for the lower MAF experiments.
+The FDR drastically increases with read depth due many more false positive calls.
 This effect is not exclusive to VarScan2-somatic and is also observed in samtools and GATK.
-
-Strelka has better sensitivity and specificity than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths and the 100\% and 10\% MAF.
-However, it has zero sensitivity to detect variants at or below 1\% MAF.
-
-MuTect has the best overall performance of the other algorithms we tested over a wide range of MAF and depths.
-Unlike several of the other algorithms muTect generally improves with a higher read depth or the specificity degradation due to false positives is modest.
-Specifically, muTect calls some of the true variant positions with a 0.3\% MAF and a median read depth of only 53 and 410.
+Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths and the 100\% and 10\% MAF.
+However, it does not call any variants at or below 1\% MAF.
+MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
+The FDR is stable across a wide range of read depths.
+But the stable FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\%-100\% MAF.
+Indeed, muTect performs worse than samtools, GATK and varScan2 for high MAF experiments.
 
@@
 This generally agrees with previous observations and intuition that more than 1,000 reads are required to call a 1:1000 MAF.
-The specificity degrades slightly at higher read depths with this fixed $\tau$ in line with muTect.
-At high read depths and MAFs classification is perfect.
-With $\tau$ set to the optimal value $\tau^*$, we find a significant improvement in both sensitivity and specificity.
-The algorithm is able to call a 0.1\% MAF with a median read depth of only 39.
-This performance is due to two properties: (1) the empirical Bayes structure of the model allows parameters $\mu_j$ to borrow strength from one another through $\mu_0$ and (2) the actual coverage at a particular location may be higher than the median coverage across all locations.
+The FDR degrades at higher read depths with this fixed $\tau$ as more false positives are called.
+At high read depths and MAFs FDR is perfect.
+With $\tau$ set to the optimal value $\tau^*$, we find a significant improvement at all read depths and the FDR decreases with read depth.
+Comparing $\tau=0$ with $\tau^*$, we see that the $\tau=0$ parameter setting is more conservative and negatively affected by higher read depths.
 
-\subsection{Pooled Multiple-Sclerosis Samples}\label{sec:ms}
+\subsubsection*{Sensitivity/Specificity Comparison}
 
-We tested our algorithm on data from pooled samples from patients with multiple-sclerosis (MS).
+Figure~\ref{tbl:comparison_ss} shows the sensitivity and specificity for other variant calling algorithms. Each cell contains the sensitivity and specificity of the method separated by a slash.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=140mm]{pdf_figs/comparison_table_ss.pdf}
+\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
+\label{tbl:comparison_ss}
+\end{center}
+\end{figure}
+
+Samtools, GATK and VarScan2-pileup all have similar performance.
+They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
+VarScan2-somatic is able to call mixed samples as expected.
+However, as the read depth increases the specificity degrades.
+Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below.
+muTect has good performance across a wide range of MAF levels.
+But even at the highest depth only has around 0.5 sensitivity for low MAF levels.
+
+The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels.
+The sensitivity increases considerably with read depth at a slight expense to specificity.
+With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAF levels.
+However, the algorithm performance is still limited at low MAF levels and low read depth.
+\subsection{HCC1187 primary ductual carcinoma sample}
+We test our algorithm on normal-tumor paired data from sample HCC1187, a 41 years old white female with primary ductal carcinoma in TNM stage IIA, grade 3.
+The tumor DNA was extracted from HCC1187 breast ductal carcinoma cell line while the normal DNA from a matching lymphoblastoid cell line.
+
+%\subsection{Pooled Multiple-Sclerosis Samples}\label{sec:ms}
+%
+%We tested our algorithm on data from pooled samples from patients with multiple-sclerosis (MS).
 The MS patients are ``high-burden"; meaning that they are known to carry one or more of several genetic risk alleles.
 Our aim in studying this  patient population is to identify rare risk alleles that may not yet be known to be associated with MS.
-
-DNA was extracted from whole blood samples from 100 patients with multiple sclerosis and 100 unaffected, unrelated individuals.
+%
+%DNA was extracted from whole blood samples from 100 patients with multiple sclerosis and 100 unaffected, unrelated individuals.
 Each sample was subjected to targeted PCR amplification of IL2RA and interleukin 7 receptor (IL7R); genes with known risk alleles~\cite{HauserNEJM2007}.
@@
 The experiment was repeated on two independent cohorts to filter for spurious associations due to population sampling.
-
-We took 4000 samples from the Markov chain, removed the first 10\% for burn-in and thinned the chain by a factor of 2.
+%
+%We took 4000 samples from the Markov chain, removed the first 10\% for burn-in and thinned the chain by a factor of 2.
 We used 10 Metropolis samples for each Gibbs step.
 We called variants with $\tau = 0.005$ and $\alpha = 0.05$ because we expect that the minimum detectable genotype is a heterozygous mutation in 1 out of 100 samples.
-
-We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA.
+%
+%We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA.
 All position information is referenced to human genome reference hg19.
@@
 This variant has not yet been shown to be associated with any known disease.
-
-\begin{figure}[h]
-\begin{center}
-%\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
-\caption{Estimated minor allele fractions for consensus variants in MS patient pools.}
-\label{fig:MS_MAF}
-\end{center}
-\end{figure}
-
-The validity of these loci are subject to confirmation on a larger independent sample.
+%
+%\begin{figure}[h]
+%\begin{center}
+%%\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
+%\caption{Estimated minor allele fractions for consensus variants in MS patient pools.}
+%\label{fig:MS_MAF}
+%\end{center}
+%\end{figure}
+%
+%The validity of these loci are subject to confirmation on a larger independent sample.
 However, we note that most GWAS studies focus on common SNPs due to the cost-effective microarray technology used for those studies.
@@
 Indeed, the common SNP that was found is associated with an autoimmune disorder and the other two variants are not at dbSNP loci.
-\subsection{HCC1187 primary ductual carcinoma sample}
-We test our algorithm on normal-tumor paired data from sample HCC1187, a 41 years old white female with primary ductal carcinoma in TNM stage IIA, grade 3.
-The tumor DNA was extracted from HCC1187 breast ductal carcinoma cell line while the normal DNA from a matching lymphoblastoid cell line.
 
@@
 \appendix
+
+%%%%%%%%%%%%%%%%%%%
+% Appendix A: Parameter Initialization
+%%%%%%%%%%%%%%%%%%%
 \section{Parameter Initialization}\label{sec:appendix_mom}
@@
 
+%%%%%%%%%%%%%%%%%%%
+% Appendix B: Comparison Statistics
+%%%%%%%%%%%%%%%%%%%
 \section{Comparison to other algorithms}\label{sec:app_comparison}
+
 \begin{figure}[h]
@@
 
+%%%%%%%%%%%%%%%%%
+% Appendix C: Data Set Statistics
+%%%%%%%%%%%%%%%%%
+\section{Estimated Parameters for Synthetic DNA Data}\label{sec:synthetic_estimate}
+The RVD2 algorithm provides estimates of the model parameters given the data. We show several of these key parameters in Figure~\ref{fig:depthM}.
+
+The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
+For the 100\% mutant data set, the read depth drops at the mutant locations.
+This is due to the parameters imposed at the alignment stage.
+The reads are 36bp long and we required no more than 2 mismatches.
+Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
+\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
+\label{fig:depthM}
+\end{center}
+\end{figure}
+
+The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
+$M_j$ measures the variance between replicates at location $j$.
+There is little variability across positions indicating that the replication variance does not change greatly across position.
+Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
+
+The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
+We see that the variation between replicates is smaller than the variation between location.
+$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
+Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
+
 \bibliographystyle{apalike}

### 2013-11-20 results/2013-11-19_test_rvd29_synthetic_data_dc=1000/character.py
Test rvd29 on synthetic data (dc=1000)

@@ -1,0 +1,186 @@
+import sys
+import os
+import numpy as np
+
+import logging
+import pdb
+
+import vcf
+import xlwt
+
+# Insert the source directory at front of the path
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+import rvd29
+
+def main():
+    book=xlwt.Workbook(encoding="utf-8")
+    sheet1=book.add_sheet("TPR_TNR")
+    sheet1.write(0, 0, "MAF")
+    sheet1.write(0, 1, "Median Depth")
+
+    sheet2=book.add_sheet("Multi-measures")
+    sheet2.write(1, 0, "MAF")
+    sheet2.write(1, 1, "Median Depth")
+
+    sheet3=book.add_sheet("FDR")
+    sheet3.write(0, 0, "MAF")
+    sheet3.write(0, 1, "Median Depth")
+
+    sheet4=book.add_sheet("MCC")
+    sheet4.write(0, 0, "MAF")
+    sheet4.write(0, 1, "Median Depth")
+    
+    method = {'RVD2(T=0.00001)':'./vcf'}
+    
+    DilutionList = (0.1, 0.3, 1.0, 10.0,100.0)
+    
+##  DepthList = (10000, 1000, 100, 10)
+    DepthList = (100, 10)
+    i=0
+    
+    for k, v in method.iteritems():
+        i=i+1
+        print 'Method %(number)d: %(method)s' %{'number':i, 'method': k}
+        sheet1.write(0, i+1, k)
+        sheet2.write(0, 9*(i-1)+6, k)
+        sheet3.write(0, i+1, k)
+        sheet4.write(0, i+1, k)
+        character=('Sensitiviy', 'Specificity', 'FPR', 'FNR', 'PPV', 'NPV', 'FDR', 'ACC', 'MCC')
+        for j in xrange(9):
+            sheet2.write(1,9*(i-1)+j+2,character[j])
+        for d in DilutionList:
+            if i==1:
+                sheet1.write(DilutionList.index(d)*len(DepthList)+1,0,"%0.1f%%" %d)
+                sheet2.write(DilutionList.index(d)*len(DepthList)+2,0,"%0.1f%%" %d)
+                sheet3.write(DilutionList.index(d)*len(DepthList)+1,0,"%0.1f%%" %d)
+                sheet4.write(DilutionList.index(d)*len(DepthList)+1,0,"%0.1f%%" %d)
+                
+            for r in DepthList:
+                # read in the median coverage
[…]

### 2013-11-20 results/2013-11-19_test_rvd29_synthetic_data_dc=1000/runall.py

@@ -1,0 +1,67 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd29 directory at front of the path
+
+rvddir = os.path.join('./rvd29')
+sys.path.insert(0, rvddir)
+import rvd29
+
+##pool =None
+pool = mp.Pool(processes=4)
+tocfilename = "synthetic_toc_p1.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the control
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd29.load_depth(controlFileList)
+    phi, theta_s, mu_s, M_s = rvd29.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd29.save_model(h5FileName, phi, mu=mu_s, M=M_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2013-11-20 results/2013-11-19_test_rvd29_synthetic_data_dc=1000/rvd29.py

@@ -1,0 +1,789 @@
+#!/usr/bin/env python
+
+
+from __future__ import print_function
+from __future__ import division
+
+import numpy as np
+
+import scipy.stats as ss
+from scipy.special import gammaln, polygamma
+from scipy import special
+
+import logging
+import multiprocessing as mp
+from itertools import repeat
+import h5py
+import tempfile
+
+import os
+import subprocess
+from datetime import date
+
+import re
+import pdb
+import time
+
+
+"""rvd29.py: put Jeffrey's Prior on Mj."""
+
+
+
+def main():
+    import argparse
+
+    # Populate our options, -h/--help is already there.
+    description = """
+                    RVD is a hierarchical bayesian model for identifying
+                    rare variants from short-read sequence data. """
+                    
+    # create the top-level parser
+    argp = argparse.ArgumentParser(prog='rvd', description=description)
+    argp.add_argument('--version', action='version', version='%(prog)s 2.7')
+    argp.add_argument('-v', '--verbose', dest='verbose', action='count',
+                help="increase verbosity (specify multiple times for more)")
+    
+    # argp.add_argument('cmd', action='store', nargs='*',
+    #             choices=['gen', 'gibbs'])
+    subparsers = argp.add_subparsers(help='sub-command help')
+    
+    # create subparser for gibbs fitting
+    argpGibbs = subparsers.add_parser('gibbs', 
+                        help='fit the RVD model using Gibbs sampling')
+    argpGibbs.add_argument('dcfile', nargs='+',
+                        help='depth chart file name')
+    argpGibbs.add_argument('-o', dest='outputFile', 
+                default='output.hdf5',
+                help='output HDF5 file name')
+    argpGibbs.add_argument('-p', '--pool', type=int, default=None,
+                help='number of workers in multithread pool')
+    argpGibbs.set_defaults(func=gibbs)
+                
+                
+    # create subparser to compare two model files
+    argpTest = subparsers.add_parser('test', 
+                        help='test if case error rate is greater than control by T')
+    argpTest.add_argument('controlHDF5Name',
+                help='control model file (HDF5)')
[…]

### 2013-11-20 results/2013-11-19_test_rvd29_synthetic_data_dc=1000/test_rvd29.py

@@ -1,0 +1,56 @@
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the source directory at front of the path
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+import rvd29
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0,100.0)
+
+    folderList = ('./')
+
+    for d in dilutionList:
+        logging.debug("Processing dilution: %0.1f%%" % d)
+        for f in folderList:
+            path='vcf/%s' % str(10**(folderList.index(f)+1))
+            if not os.path.exists(path):
+                os.makedirs(path)
+            controlFile = "./%s/Control.hdf5" %f
+            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
+            caseFile = "./%(file)s" %{'folder':f,'file':caseFile}
+            
+            outputFile='%(path)s/vcf%(dilution)s' %{'path':path,'dilution':str(d).replace('.','_')}
+           
+            rvd29.test(controlFile,caseFile,T=0.00001,N=1000,outputFile=outputFile)
+if __name__ == '__main__':
+    main()
+
+
+"""import sys
+import os
+import rvd29
+#Insert the source directory at front of the path
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+
+ControlFile = './Control.hdf5'
+CaseFile = './Case0_1.hdf5'
+OutputFile = 'vcf0_1'
+
+assert os.path.isfile(ControlFile), "ControlFile file does not exist: %s" % ControlFile
+assert os.path.isfile(CaseFile), "CaseFile file does not exist: %s" % CaseFile
+
+rvd29.test(ControlFile, CaseFile, T=0, N=1000, outputFile=OutputFile)
+"""

### 2013-11-20 results/2013-11-15_Priors_Manuscript/Prior.tex
Manuscript draft for priors on Mj (need revising)

\documentclass[11pt,reqno]{amsart}
\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
%\usepackage{algorithm, algorithmic}

\usepackage{algorithm}
\usepackage{algpseudocode}

\usepackage{graphicx}

\usepackage{verbatim}
\usepackage{amssymb}
\usepackage{amsmath}

\usepackage{enumitem}

\usepackage{setspace}
\doublespacing

\usepackage{natbib}

%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newcommand{\RR}{I\!\!R} %real numbers
\DeclareMathOperator{\diag}{diag}

\algnewcommand{\Inputs}[1]{%
  \State \textbf{Inputs:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}
\algnewcommand{\Initialize}[1]{%
  \State \textbf{Initialize:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}

\title[Priors]{Priors for $M_{j}$}
\author{}
%\date{}                                           % Activate to display a given date or no date

\begin{document}

\maketitle

\section{Parametric Gamma Prior}
\subsection{Initialization}
The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure. MoM works by setting the population moment equal to the sample moment. We present the initial parameter estimates and provide the derivations here.

Since the distribution is $ M_{j} \thicksim \text{Gamma}(a, b)$. The first and second population moments are

\begin{eqnarray}
	E[M_{j}] =& a b,\\
	E[{M_{j}}^2] =& b^2 a (a+1).
\end{eqnarray}

The first and second sample moments are $m_1 = \frac{1}{J}\sum_{j=1}^J M_{j}$ and $m_2 = \frac{1}{J}\sum_{j=1}^J M_{j}^2$. Setting the population moments equal to the sample moments and we get

\begin{equation}
    a = \frac {{m_1}^2}{{m_2}-{m_1}^2}
\end{equation}

\begin{equation}
    b = \frac {{m_2}-{m_1}^2}{m_1}
\end{equation}

\subsection{Sampling from $p \left( M_{j} |a, b, \theta_{ji},\mu_j\right)$}
The posterior distribution over $M_{j}$ given its Markov blanket is

\begin{equation}
	p( M_{j} |a, b, \theta_{ji},\mu_j) \propto p(\theta_{ji} | \mu_j, M_j) p(M_{j} | a, b)
\end{equation}

We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{Gamma}(a, b)$. Assuming there is only one replicate,

\begin{equation}
	p( M_{j} |a, b, \theta_{ji},\mu_j) \propto \frac{\Gamma \left({\mu}_{j} {M}_{j} \right)}{\Gamma \left({\mu }_{j}\right)\Gamma \left({M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}\frac{{b}^{a}}{\Gamma \left({a}\right)}M_{j}^{a-1}{e}^{-bM_{j}}
\end{equation}

Since we cannot give an analytical form for this, so we sample from the posterior distribution using the Metropolis-Hastings algorithm.

\section{Jefferys Priors}\label{sec:model_structure}
Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
\begin{equation}\label{eqn:JefferyPrior}
\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
\end{equation}

Since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by

\begin{equation}\label{equ:JefferyInference}
I\left({M}_{j}\right)={E}_{{M}_{j}}\left[ -\frac{\delta ^{2}\log p\left(\theta _{j}|\mu_{j},M_{j}\right)}{\delta M^{2}_{j}}\right]
\end{equation}

Now to calculate the equations to acquire the Fisher information, we assume there is only one replicate,

\begin{equation}\label{eqn:Betapdf}
p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
\end{equation}

\begin{equation}\label{equ:JefferyInference1}
\begin{split}
\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
\end{split}
\end{equation}

\begin{equation}
\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
\end{equation}

\begin{equation}
\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
\end{equation}

Now we have the Jeffreys' prior for $M_{j}$:

\begin{equation}
\pi\left({M}_{j}\right) = [-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
\end{equation}

\section{Reference Prior}
We want to make the prior as less informative as possible, so to maximize a certain distance measure of prior and posterior is considered using Kullback-Leibler divergence.

\subsection{KL-Divergence}
For distributions $p$ and $q$, their KL-divergence is defined as $KL\left(p||\right) = E_p[\log\left(p/q\right)] = int_p\left(x\right)\log \frac{p\left(x\right)}{q\left(x\right)}dx$

\subsection{Definition for Reference Prior}
Suppose posterior is $p\left(\theta | x \right)$, depending on pror $p\left(\theta \right)$, marginal is $p\left(x \right)$, then $p\left(\theta \right)$ is called the reference pror if it maximizes $max_{p\left(\theta\right)}E_{p\left(x\right)}KL\left(p\left(\theta|x\right)||p\left(\theta\right)\right)$
The Jeffreys prior is proved to be reference prior in one dimension (by Michael I. Jordan).

\section{Log-normal Prior}
In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.

\subsection{Initialization}
The initial values is also obtained by a method-of-moments (MoM) procedure. We present the initial parameter estimates below.

Since the distribution is $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$. The first and second population moments are

\begin{eqnarray}
	E[M_{j}] = & e^{\mu+\frac{{\sigma}^2}{2}},\\
	E[{M_{j}}^2] = & \left(e^{{\sigma}^2}-1\right)[E[M_{j}]]^2.
\end{eqnarray}

We can acquire the sample moments, then parameters $\mu$ and $\sigma$ can be obtained

\begin{equation}
    \mu = \ln\left(E[M_{j}]\right)-\frac{{\sigma}^2}{2}
\end{equation}

\begin{equation}
    {\sigma}^2 = \ln\left(1+\frac{E[{M_{j}}^2]}{{E[M_{j}]}^2}\right)
\end{equation}

\subsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
The posterior distribution over $M_{j}$ given its Markov blanket is

\begin{equation}
	p( M_{j} |\mu, \sigma, \theta_{ji},\mu_j) \propto p(\theta_{ji} | \mu_j, M_j) p(M_{j} | \mu, \sigma)
\end{equation}

We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$. Since we cannot give an analytical form for this, so we sample from the posterior distribution using the Metropolis-Hastings algorithm.

\end{document}

### 2013-11-21 doc/bioinformatics2013/RVD2.tex

@@
 
-BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ to simulate lower coverage data while retaining the error structure of real NGS data.
+BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ using Picard Version 1.96 to simulate lower coverage data while retaining the error structure of real NGS data.
 The final data set contains pairs of reads for three replicates of each case and a pair of reads for the control sample giving a total of 36 BAM files for each downsampled data set.
@@
 
+\subsubsection*{Parameters setting for variant calling}\
+
+The variant calling algorithms we compare RVD2 to include samtools/bcftools, GATK VarScan2-mpileup, Varscan2-somatic, Strelka, and muTect.
+
+\paragraph{\textbf{Samtools}}We used samtools/bcftools (v0.1.19) for variant calling.
+mpileup was the function to call variants and bcftools was used to save the result in standard Variant Call Format (VCF) files.
+In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap.
+Our synthetic data was mapped using BWA Version 0.7.4, -C50 option was applied to minimize the effect from reads with overestimated mapping quality.
+mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling.
+Read depth for each sample was kept with -D enabled, and -u was turned on to make sure the output bcf files uncompressed.
+
+\paragraph{\textbf{GATK}}
+UnifiedGenotyper is the function we used in GATK (v2.1-8) to detect mutations.We followed the general GATK recomended workflow to detect variants.
+Due to some format  incompatibility, we applied Picard to format read group and GATK for realignment as preprocessing.
+When calling variants using UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches.
+All other parameters were set at default.
+Six replicates was passed to GATK for variant calling.
+
+\paragraph{\textbf{VarScan2-mpileup}}
+VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.
+We set -d at $10^6$ and enabled -50 for same concern previously stated.
+In afterwards mpileup2snp function, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.
+Six replicates were all introduced for variant calling.
+
+\paragraph{\textbf{VarScan2-somatic}}
+We tested VarScan2-somatic on our six replicates synthetic data set as well.
+Considering that RVD2 doesn't call somatic status, we combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown).
+--normal-purity was set at 1.00, while --tumor-purity was set as the dilution rate.
+--min-var-freq was set at $10^{-5}$.
+
+\paragraph{\textbf{Strelka}}
+Configuration and Analysis for Strelka is standard so we just installed the program and ran it on our data set.
+One thing to note is that since Strelka can not accept relicate data, we used a single replicate with a read depth that most closely matched the overall median depth of the replicates.
+
+\paragraph{\textbf{muTect}}
+Similar to Strelka, we ran muTect with standard settings and only passed one replicate of synthetic data.
+
 GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
@@
 We find its performance on much lower MAF variants and across a wider range of coverage depths.
-Varscan2-somatic, streak and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the median overall median depth of the replicates.
+Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
 

### 2013-11-21 doc/bioinformatics2013/RVD2.tex
updates to manuscript.

@@
 \begin{document}
+%TODO: Label all results/figures with comment indicating what experiment folder those results depend on and come from.
 
@@
 Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
-Unlike most research samples, these clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
+Clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
 However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
-We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and a Bayesian posterior hypothesis test to call variants in heterogeneous samples.
-We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depths and a minor allele frequencies.
+We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
+We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency.
+We identify XX mutations in the PAXXX gene in a matched clinical XXX tumor sample with 80\% tumor purity and XX median read depth.
 %We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
@@
 
-Clinical samples are often genetically heterogeneous.
-For example, non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
+Increasingly, NGS is being used to interrogate mutations in heterogeneous clinical samples.
+For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
 Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
@@
 
-The remainder of this paper presents a novel algorithm for calling rare variants in heterogeneous samples using NGS data that is an improvement upon currently available technology.
-Typically such a variant calling algorithms are comprised of multiple pre- and post filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
-These steps may improve algorithm performance, but it is difficult to isolate what step is most responsible for the good performance.
-Instead, we focus exclusively on the model structure and statistical inference.
+Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
+These steps may improve algorithm performance, but their presence makes it difficult to isolate what step is most responsible for the performance.
+Instead, we focus exclusively on the model structure and statistical inference steps.
 Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study.
@@
 We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
-%Finally, in Section~\ref{sec:ms} we present results on detecting mutations associated with multiple sclerosis from data obtained from pooling 100 affected patients~\citep{Ferretti:2013hp}.
+Finally, in Section~\ref{sec:brca} we present results on detecting mutations in the PAXIP1 gene from matched tumor-normal data from the HCC1187 cell line.
 
-\section{Materials and Methods}
+%%%%%%%%%%%%%%%%%%
+% Model Structure
+%%%%%%%%%%%%%%%%%%
+\section{Model Structure}\label{sec:model_structure}
 
@@
 Then, it tests for a significant difference between model parameters in the case and control samples and returns called variant positions.
-
-\subsection{Model Structure}\label{sec:model_structure}
 
@@
 
+%%%%%%%%%%%%%%%%%%
+% Inference & Hypothesis Testing
+%%%%%%%%%%%%%%%%%%
 \section{Inference and Hypothesis Testing}
@@
 
-We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}.
-The MoM estimate for replicate-level parameters are $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
-The estimates for the position-level parameters are $\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ and $\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
-The estimates for the genome-level parameters are $\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ and $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
+We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are 
+$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. 
+The estimates for the position-level parameters are 
+$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ 
+and 
+$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. 
+The estimates for the genome-level parameters are 
+$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ 
+and 
+$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. 
 While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encounter such pathological behavior in this application.
@@
 
-Samples from the posterior distribution $p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ are drawn analytically because of the Bayesian conjugacy between the prior $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ and the likelihood $p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
+Samples from the posterior distribution 
+$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ 
+are drawn analytically because of the Bayesian conjugacy between the prior 
+$p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ 
+and the likelihood 
+$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$. 
 The posterior distribution is 
-
 \begin{equation}
@@
 
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, 0.1 \cdot \mu_j^{(p)})$. The acceptance probability is then
+A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$. The acceptance probability is then
 \begin{equation}
@@
 
+We have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
+We set the proposal distribution variance $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise.
+We resample from the proposal if the sample is outside of the support of the posterior distribution.
+In the experiments, we discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
 Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors.
-We have found that the algorithm performance improves when we take the median of ten or more M-H samples as a single Gibbs step for each position.
-This computational effort only minimally increases run time when run on a parallel computing platform.
-In the experiments, we use a XX sample burn-in to allow the Markov chain to approach its stationary distribution and we thin the chain by a factor of 2 to reduce autocorrelation among samples.
 
@@
 A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
-\begin{equation}
+\begin{equation}\label{eqn:bayes_test}
 	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
@@
 where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
-We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by randomly sampling with replacement from the samples of $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
+We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
 
 The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ maximizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot.
+The optimal $\tau$ maximizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
 \begin{equation}
-	\tau^* = \arg\max_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}
+	\tau^* = \arg\max_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
 \end{equation}
@@
 With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest.
+Absent this information we set $\tau = 0$.
 
 %%%%%%%%%%%%%%%%%%
+% Chi^2 Test
+%%%%%%%%%%%%%%%%%%
+\subsection{$\chi^2$ test for non-uniform base distribution}
 
-% Chi^2 Test
+The non-reference bases may be due to a true mutation or due to a random sequencing error and we would like to differentiate these two scenarios.
+We assume non-reference read counts caused by a non-biological mechanism results in a uniform distributed among three non-reference categories.
+In contrast, the distribution of counts among three non-reference bases caused by biological mutation would be biased to one base.
+
+We use a $\chi^2$ goodness-of-fit test on a multinomial distribution over the non-reference bases to distinguish these two possible scenarios.
+The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
+Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
+The test statistic is
+
+\begin{equation}
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
+\end{equation}
+
+where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
+Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.
+
+At each position, we have one statistic for each replicate. We use Fisher's combined probability test to combine the $N$ p-values into a single p-value for each position,
+
+\begin{equation}\label{eqn:fisher_combined}
+	X^2 = -2 \sum_{i=1}^N \ln(p_i).
+\end{equation}
+
+Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.
+
+Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test}.
 
 %%%%%%%%%%%%%%%%%%
-%\subsection{$\chi^2$ test for non-uniform base distribution}
-%
-%We assume the counts for the non-reference bases at poisiuton $j$ in replicate $i$ follow a multinomial distribution with parameter  $p = (p_1, p_2, p_3)$ where $p_k \geq 0$ for $k =1, 2, 3$ and $\sum_{k=1}^3 p_k = 1$.
-%
-%The non-reference reads may be due to a true mutation or due to a random sequencing error.
-We assume, non-reference read counts caused by a random mechanism results in a uniform distributed among three non-reference categories, giving $p_1=p_2=p_3=1/3$.
-In contrast, the distribution of non-reference read counts among three non-reference categories caused by mutation would be biased to one base.
-%
-%We use a $\chi^2$ goodness-of-fit test to distinguish these two possible scenarios.
-The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
-Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistics, the log likelihood ratio statistics $(\lambda = 0)$, the Freeman-Tukey statistics $(\lambda = -1/2)$ the Neyman modified $X^2 (\lambda = -2)$.
-The test statistic is
-%
-%\begin{equation}
-% 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
-%\end{equation}
-%
-%where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.
-%
-%For each position, there are $N$ power-divergence statistics for each experimental replicate.
-We use Fisher's combined probability test to combine the p-values from each replicates into a single one for each position,
-%
-%\begin{equation}
-%	X^2 = -2 \sum_{i=1}^N \ln(p_i).
-%\end{equation}
-%This gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.
-%
-%While Fisher's combined probability test collapses the p-values across replicates, we still have $J$ tests across positions.
-We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions.
+% Data Sets
+%%%%%%%%%%%%%%%%%%
+\section{Data Sets}
+
+We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms.
+The synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
+The HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
+
+\subsection{Synthetic DNA Sequence Data}
+
+Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
+Sample of the case and control DNA were mixed at defined fractions to yield minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+More details of the experimental protocol are available from the original publication~\cite{}.
+We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
+
+To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
+The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
+
+\subsection{HCC1187 Sequence Data}
+
+The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
+A matched normal cells were derived from lymphoblastoid cells from whole blood.
+% TODO: check matched cells line origin
+% TODO: mention tumor purity here
+
+Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
+The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
+Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% coverage with 10 or more reads~\cite{}.
+% TODO: write this in own words.
+% TODO: add citation here.
 
@@
 
+We test RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
+The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
+We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
+
 \subsection{Performance with read depth}\label{sec:read_depth}
 
-We synthesized a 400bp control DNA sequence and another 400bp sequence that is identical to the control except at 14 loci with variant bases.
-We denote the indices of these true positive variant positions $j^*$.
-Pure case and control samples were mixed at defined fractions to yield minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-The sequencing protocol and data set are available from the original publication~\cite{}.
-
-BAM files for the synthetic DNA data were down sampled by $10\times$, $100\times$, $1,000\times$, $10,000\times$ using Picard Version 1.96 to simulate lower coverage data while retaining the error structure of real NGS data.
-The final data set contains pairs of reads for three replicates of each case and a pair of reads for the control sample giving a total of 36 BAM files for each downsampled data set.
-
-We ran the RVD2 algorithm on each downsampled data set sand varied the threshold $\tau$ with fixed $\alpha=0.05$ to generate receiver operating characteristic (ROC) curves in Figure~\ref{fig:ROC}.
-Figure~\ref{ig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
-At the lowest depth the sensitivity and specificity is no better than a random decision.
-We would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43.
-The performance improves monotonically and is much better at a depth of approximately 41,000.
-This depth exceed the inverse MAF by a $10\times$ which we find to be generally a good coverage threshold for detecting variants.
-We find a similar pattern in Figures~\ref{fig:ROC}B-C.
-Interestingly, the algorithm performance is much better than a random decision rule to detect a 10\% MAF with only a depth of 25.
+We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
+For these ROC curves, we only used the Bayesian test without the $\chi^2$ test.
+Adding the $\chi^2$ improves specificity at the expense of sensitivity.
+Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
+Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
+At the lowest depth the sensitivity and specificity is no better than random.
+However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43.
+The performance improves monotonically with read depth.
+Figures~\ref{fig:ROC}B-C for larger MAFs show a similar relationship between coverage depth and accuracy.
 
@@
 \includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance of model with Bayesian Hypothesis Test.}
+\caption{ROC curve varying read depth showing detection performance.}
 \label{fig:ROC}
@@
 
-Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 41,573$.
-The only false positive is at position 8, though posterior distributions for the case and control clearly show a difference at that position.
-The posterior mean estimates are shrunken towards the global error rate parameter $\mu_0 = 0.0023$ for the case and control due to the empirical Bayes procedure.
+Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 41,573$ and MAF = XX\%.
+All of the called positions show a clear difference between the case and control error rates.
+The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
 
@@
 
-We compare the empirical performance of RVD2 to other variant calling algorithms.
-We use the synthetic DNA data sets described in the previous section to explore performance across a dynamic range of read depths as well as minor-allele frequencies.
-We report false discovery rate, sensitivity and specificity for each algorithm to detect the 14 true positive variant positions out of 400 tested positions.
-In a research application, the false discovery rate is more relevant because the aim is generally to identify interesting variants and not spend resources following up false positives.
+We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
+In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
 The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in calling all of the true variants and none of the true negatives.
 
-\subsubsection*{Parameters setting for variant calling}\
+[DISCUSS \citet{Stead:2013fu}]
 
-The variant calling algorithms we compare RVD2 to include samtools/bcftools, GATK VarScan2-mpileup, Varscan2-somatic, Strelka, and muTect.
+\subsubsection*{Sensitivity/Specificity Comparison}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
+\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
+\label{tbl:comparison_ss}
+\end{center}
+\end{figure}
+
+Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
+They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
+VarScan2-somatic is able to call more mixed samples.
+However, as the read depth increases the specificity degrades.
+Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below.
+muTect has good performance across a wide range of MAF levels.
+But even at the highest depth only has around 0.5 sensitivity for low MAF levels.
+
+The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
+The sensitivity increases considerably with read depth at a slight expense to specificity.
+With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs.
+However, in practice one may not know the optimal $\tau^*$ a-priori.
+With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives.
+When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
+The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see supplementary table SX).
+
+\subsubsection*{False Discovery Rate Comparison}
+Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, strelka and muTect. Blank cells indicate no positive calls were made.
+
+Samtools performs well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
+GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
+VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=160mm]{pdf_figs/comparison_table_fdr.pdf}
+\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
+\label{tbl:comparison_fdr}
+\end{center}
+\end{figure}
+
+VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level.
+However, the FDR is high due to many false positives.
+Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and 1.00 for the highest read depth.
+Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF.
+However, it does not call any variants at or below 1\% MAF.
+MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
+But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\%-100\% MAF.
+
+RVD2 has a lower FDR than other algorithms when the read depth is greater than $10\times$ the MAF with $N=1$ and $\tau$ set to the default value of zero or to the optimal value.
+The FDR is higher when $N=6$ because the variance of the control error rate distribution $P(\mu_j^{\text{control}} | r^{\text{control}})$ is smaller.
+The smaller variance yields improvements in sensitivity at the expense of more false positives.
+Since the FDR only considers positive calls, the performance degrades.
+
+\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca
+
+RVD2 identified 5 variants in the XXkb PAXIP1 gene.
+We identified the same variants when the fastq data was aligned to HG19 masked for repetitive regions and when we used the unmasked iGenome reference sequence.
+Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
+The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
+
+%TODO: Add plan for discussion of these results
+
+\appendix
+
+%%%%%%%%%%%%%%%%%%%
+% Appendix A: Parameter Initialization
+%%%%%%%%%%%%%%%%%%%
+\section{Parameter Initialization}\label{sec:appendix_mom}
+Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
+Therefore the MoM estimator is
+\begin{equation}
+	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
+\end{equation}
+
+We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
+The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
+The first and second population moments are
+\begin{eqnarray}
+	E[\theta_{ji}] =& \mu_j,\\
+	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
+\end{eqnarray}
+The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
+Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
+\begin{eqnarray}
+	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
+	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
+\end{eqnarray}
+
+Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
+\begin{eqnarray}
+	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
+	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
+\end{eqnarray}
+
+%%%%%%%%%%%%%%%%%%%
+% Appendix B: Comparison Statistics
+%%%%%%%%%%%%%%%%%%%
+\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
+\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
+\label{fig:comparison_mcc}
+\end{center}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%
+% Appendix C: Data Set Statistics
+%%%%%%%%%%%%%%%%%
+\section{RVD2 Estimated Parameters}\label{sec:synthetic_estimate}
+The RVD2 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:depthM}.
+
+The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
+For the 100\% mutant data set, the read depth drops at the mutant locations.
+This is due to the parameters imposed at the alignment stage.
+The reads are 36bp long and we required no more than 2 mismatches.
+Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
+\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
+\label{fig:depthM}
+\end{center}
+\end{figure}
+
+The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
+$M_j$ measures the variance between replicates at location $j$.
+There is little variability across positions indicating that the replication variance does not change greatly across position.
+Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
+
+The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
+We see that the variation between replicates is smaller than the variation between location.
+$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
+Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
+
+%%%%%%%%%%%%%%%%%
+% Appendix D: Parameter Settings for Variant Calling Algorithms
+%%%%%%%%%%%%%%%%%
+\section*{Parameter setting for other variant calling algorithms}
 
@@
 In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap.
-Our synthetic data was mapped using BWA Version 0.7.4, -C50 option was applied to minimize the effect from reads with overestimated mapping quality.
 mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling.
@@
 
-\subsubsection*{False Discovery Rate Comparison}
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 and other variant calling algorithms.
-We computed the Matthews Correlation Coefficient (MCC) and sensitivity and specificity as a combined performance measure and present those results in Appendix B~\cite{}.
-Blank cells indicate no positive calls were made.
-The standard genotype variant calling algorithms: samtools, GATK, and VarScan2-mpileup are shown in the first columns.
-Samtools performs very well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
-GATK performs well to detect both 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
-VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_fdr.pdf}
-\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
-\label{tbl:comparison_fdr}
-\end{center}
-\end{figure}
-
-VarScan2-somatic generally has a poor performance across all MAF and read depths.
-The FDR is close to 1 for the lower MAF experiments.
-The FDR drastically increases with read depth due many more false positive calls.
-This effect is not exclusive to VarScan2-somatic and is also observed in samtools and GATK.
-Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths and the 100\% and 10\% MAF.
-However, it does not call any variants at or below 1\% MAF.
-MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
-The FDR is stable across a wide range of read depths.
-But the stable FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\%-100\% MAF.
-Indeed, muTect performs worse than samtools, GATK and varScan2 for high MAF experiments.
-
-We tested our model RVD2 using two variations of the hypothesis test.
-We set $\tau$, the hypothesis test threshold to the naive value of zero and we optimized $\tau$ for the read depth and MAF ($\tau^*$).
-These optimal values were computed from this data set, but in general may be computed once and used as a lookup table for future experiments.
-With $\tau=0$, RVD2 is not able to call variants with a MAF of 0.1\% if the read depth is less than 4,129.
-This generally agrees with previous observations and intuition that more than 1,000 reads are required to call a 1:1000 MAF.
-The FDR degrades at higher read depths with this fixed $\tau$ as more false positives are called.
-At high read depths and MAFs FDR is perfect.
-With $\tau$ set to the optimal value $\tau^*$, we find a significant improvement at all read depths and the FDR decreases with read depth.
-Comparing $\tau=0$ with $\tau^*$, we see that the $\tau=0$ parameter setting is more conservative and negatively affected by higher read depths.
-
-\subsubsection*{Sensitivity/Specificity Comparison}
-
-Figure~\ref{tbl:comparison_ss} shows the sensitivity and specificity for other variant calling algorithms. Each cell contains the sensitivity and specificity of the method separated by a slash.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_ss.pdf}
-\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
-\label{tbl:comparison_ss}
-\end{center}
-\end{figure}
-
-Samtools, GATK and VarScan2-pileup all have similar performance.
-They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
-VarScan2-somatic is able to call mixed samples as expected.
-However, as the read depth increases the specificity degrades.
-Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below.
-muTect has good performance across a wide range of MAF levels.
-But even at the highest depth only has around 0.5 sensitivity for low MAF levels.
-
-The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels.
-The sensitivity increases considerably with read depth at a slight expense to specificity.
-With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAF levels.
-However, the algorithm performance is still limited at low MAF levels and low read depth.
-\subsection{HCC1187 primary ductual carcinoma sample}
-We test our algorithm on normal-tumor paired data from sample HCC1187, a 41 years old white female with primary ductal carcinoma in TNM stage IIA, grade 3.
-The tumor DNA was extracted from HCC1187 breast ductal carcinoma cell line while the normal DNA from a matching lymphoblastoid cell line.
-
-%\subsection{Pooled Multiple-Sclerosis Samples}\label{sec:ms}
-%
-%We tested our algorithm on data from pooled samples from patients with multiple-sclerosis (MS).
-The MS patients are ``high-burden"; meaning that they are known to carry one or more of several genetic risk alleles.
-Our aim in studying this  patient population is to identify rare risk alleles that may not yet be known to be associated with MS.
-%
-%DNA was extracted from whole blood samples from 100 patients with multiple sclerosis and 100 unaffected, unrelated individuals.
-Each sample was subjected to targeted PCR amplification of IL2RA and interleukin 7 receptor (IL7R); genes with known risk alleles~\cite{HauserNEJM2007}.
-We targeted X exons covering XXbp in IL2RA and X exon covering XXbp in IL7R.
-The experiment was repeated on two independent cohorts to filter for spurious associations due to population sampling.
-%
-%We took 4000 samples from the Markov chain, removed the first 10\% for burn-in and thinned the chain by a factor of 2.
-We used 10 Metropolis samples for each Gibbs step.
-We called variants with $\tau = 0.005$ and $\alpha = 0.05$ because we expect that the minimum detectable genotype is a heterozygous mutation in 1 out of 100 samples.
-%
-%We identified by consensus of the two independent cohorts two variants in IL7R and one variant in IL2RA.
-All position information is referenced to human genome reference hg19.
-One variant in IL7R is rs1494558 at chr5:35861068 T$>$C and has a MAF of 76\% and 71\% in the two cohorts.
-This mutation causes a missense variant Ile66Thr and has been shown to be associated with abnormal T-cell response and IgA neuropathy~\citep{Hahn:2009bi}.
-The second IL7R variant is at chr5:35875593 A$>$T and has a MAF of 60\% and 55\% in the two cohorts.
-This variant has not yet been shown to be associated with any known disease.
-The IL2RA variant is at chr10:6061965 A$>$T and has a MAF  of 3\% and 5\% in the two cohorts.
-This variant has not yet been shown to be associated with any known disease.
-%
-%\begin{figure}[h]
-%\begin{center}
-%%\includegraphics[width=80mm]{pdf_figs/MuBarPlot.pdf}
-%\caption{Estimated minor allele fractions for consensus variants in MS patient pools.}
-%\label{fig:MS_MAF}
-%\end{center}
-%\end{figure}
-%
-%The validity of these loci are subject to confirmation on a larger independent sample.
-However, we note that most GWAS studies focus on common SNPs due to the cost-effective microarray technology used for those studies.
-We have instead used sequencing technology which yields and unbiased coverage of the targeted regions and has the potential to uncover rare variants that are not targeted by microarray technology.
-Indeed, the common SNP that was found is associated with an autoimmune disorder and the other two variants are not at dbSNP loci.
-
-500 ng of DNA were prepped using an early access version of Illumina’s TruSeq DNA PCR-Free kit.
-The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
-Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% of loci covered with 10 or more reads.
-\appendix
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix A: Parameter Initialization
-%%%%%%%%%%%%%%%%%%%
-\section{Parameter Initialization}\label{sec:appendix_mom}
-Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
-Therefore the MoM estimator is
-\begin{equation}
-	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
-\end{equation}
-
-We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
-The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
-The first and second population moments are
-\begin{eqnarray}
-	E[\theta_{ji}] =& \mu_j,\\
-	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
-\end{eqnarray}
-The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
-Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
-\begin{eqnarray}
-	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
-	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
-\end{eqnarray}
-
-Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
-\begin{eqnarray}
-	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
-	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
-\end{eqnarray}
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix B: Comparison Statistics
-%%%%%%%%%%%%%%%%%%%
-\section{Comparison to other algorithms}\label{sec:app_comparison}
-
-\begin{figure}[h]
-\begin{center}
-%\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
-\caption{Comparison of RVD2 with other variant calling algorithms (MCC).}
-\label{fig:comparison_mcc}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%
-% Appendix C: Data Set Statistics
-%%%%%%%%%%%%%%%%%
-\section{Estimated Parameters for Synthetic DNA Data}\label{sec:synthetic_estimate}
-The RVD2 algorithm provides estimates of the model parameters given the data. We show several of these key parameters in Figure~\ref{fig:depthM}.
-
-The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The reads are 36bp long and we required no more than 2 mismatches.
-Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
-\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
-\label{fig:depthM}
-\end{center}
-\end{figure}
-
-The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
-
-The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-We see that the variation between replicates is smaller than the variation between location.
-$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
-Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
-
 \bibliographystyle{apalike}

### 2013-11-25 doc/bioinformatics2013/RVD2.tex
update the manuscript

@@
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency.
-We identify XX mutations in the PAXXX gene in a matched clinical XXX tumor sample with 80\% tumor purity and XX median read depth.
+We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample with 80\% tumor purity and 52 median read depth.%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
 %We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
@@
 We have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-We set the proposal distribution variance $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise.
+We set the proposal distribution standard deviation $\sigma_j = 0.1$, a constant to meet the Detailed Balance requirement.
+
+%We set the proposal distribution variance $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise.
+
 We resample from the proposal if the sample is outside of the support of the posterior distribution.
@@
 \end{figure}
+% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
 
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/MuBarPlot.pdf}
+\includegraphics[width=120mm]{pdf_figs/Synthetic_MuBarPlot.pdf}
 \caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
@@
 \end{figure}
+% from folder '2013-11-24_synthetic_mu_barplot_optimalT_Qsd_0_1'
 
@@
 \end{figure}
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 
@@
 
-\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca
+\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
 
-RVD2 identified 5 variants in the XXkb PAXIP1 gene.
+RVD2 identified 5 variants in the 44716b PAXIP1 gene.
 We identified the same variants when the fastq data was aligned to HG19 masked for repetitive regions and when we used the unmasked iGenome reference sequence.
@@
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
+\label{fig:MAF}
+\end{center}
+\end{figure}
+
+% from folder 'rvd2\results\2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'
 
@@
 \end{figure}
+
+% from folder '2013-11-24_depthM_plot_avg10_Qsd_0_1', dilution rate is 0.1.
 

### 2013-11-26 doc/bioinformatics2013/RVD2.tex
added results for HCC sample and discussion section.

@@
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency.
-We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample with 80\% tumor purity and 52 median read depth.%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
+We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+
+%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
 %We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
@@
 
-We have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-We set the proposal distribution standard deviation $\sigma_j = 0.1$, a constant to meet the Detailed Balance requirement.
-
-%We set the proposal distribution variance $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise.
+We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise.
+Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
 
 We resample from the proposal if the sample is outside of the support of the posterior distribution.
-In the experiments, we discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
+We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
 Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors.
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Hypothesis Test}\label{sec:hypothesis_test}
+\subsection{Posterior Density Test}\label{sec:hypothesis_test}
 Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data.
@@
 
+% TODO: discuss alignment protocol
+
 \section{Results}
 
-We test RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
+We tested RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
 The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
@@
 
-RVD2 identified 5 variants in the 44716b PAXIP1 gene.
-We identified the same variants when the fastq data was aligned to HG19 masked for repetitive regions and when we used the unmasked iGenome reference sequence.
+RVD2 identified 5 variants in the 44,716bp PAXIP1 gene.
 Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
+None of the identified mutations are at known dnSNPX.X locations.
 
+Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
+Positions chr7:743899X>X, chr7:753635X>X and chr7:780960X>X are germline homozygous mutations.
+Positions chr7:754371 and chr7:758813 are called heterozygous in the normal sample.
+In the tumor cells we identified all five mutations chr7:743899X>X, chr7:753635X>X, chr7:754371X>X, chr7:758813X>X, and chr7:780960X>X.
+Positions chr7:754371X>X and chr7:758813X>X are identified as loss-of-heterozygosity events.
+
+% TODO: replace X>X with the reference and mutated base here
 \begin{figure}[h]
@@
 \caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
-\label{fig:MAF}
+\label{fig:brca_MAF}
 \end{center}
 \end{figure}
+%('rvd2\results\2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1')
 
-% from folder 'rvd2\results\2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'
+The original research describing this sample used Strelka to identify mutations in the same sample. They identified XXX and XXX as variant, but missed the other three variants.
 
-%TODO: Add plan for discussion of these results
+\section{Discussion}
+
+We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data.
+Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths.
+
+Our inference algorithm uses Gibbs sampling and Metropolis-Hastings sampling to estimate the empirical Bayes statistical model.
+This sampling procedure provides a guarantee to identify the global optimal parameter settings in the long run.
+However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches.
+We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.
+
+We have focused on the statistical model and hypothesis test in this study and our results do not include any pre-filtration of erroneous reads or post-filtration of mutation calls beyond a simple quality score threshold.
+Incorporation of such data-cleaning steps will likely improve the accuracy of the algorithm.
+
+Our approach does not address identification of indels, structural variants or copy number variants.
+Those mutations typically require specific data analysis models and tests that are different than those for single-nucleotide variants.
+Furthermore, analysis of RNA-seq data or other data generated on the NGS platform may require different models that are more appropriately tuned to the particular noise feature of that data.
 

### 2013-11-26 doc/bioinformatics2013/RVD2.tex
Update manuscript

@@
 % TODO: check matched cells line origin
+% double checked by Yuting, didn't find the matched cells line origin
 % TODO: mention tumor purity here
@@
 
-Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 41,573$ and MAF = XX\%.
+Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
 All of the called positions show a clear difference between the case and control error rates.
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/Synthetic_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in 0.1\% dilution.}
+\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
+\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
 \label{fig:MAF}
@@
 \end{figure}
-% from folder '2013-11-24_synthetic_mu_barplot_optimalT_Qsd_0_1'
+% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
 
@@
 % from folder 'rvd2\results\2013-09-19_operating_characteristics'
+% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
 
@@
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:743899X>X, chr7:753635X>X and chr7:780960X>X are germline homozygous mutations.
+Positions chr7:743899XC$\rightarrow$T, chr7:753635T$\rightarrow$C and chr7:780960C$\rightarrow$T are germline homozygous mutations.
 Positions chr7:754371 and chr7:758813 are called heterozygous in the normal sample.
-In the tumor cells we identified all five mutations chr7:743899X>X, chr7:753635X>X, chr7:754371X>X, chr7:758813X>X, and chr7:780960X>X.
-Positions chr7:754371X>X and chr7:758813X>X are identified as loss-of-heterozygosity events.
+In the tumor cells we identified all five mutations chr7:743899C$\rightarrow$T, chr7:753635T$\rightarrow$C, chr7:754371T$\rightarrow$C, chr7:758813G$\rightarrow$A, and chr7:780960C$\rightarrow$T.
+Positions chr7:754371T$\rightarrow$C and chr7:758813G$\rightarrow$A are identified as loss-of-heterozygosity events.
 
 % TODO: replace X>X with the reference and mutated base here
+% Done
+
 \begin{figure}[h]
@@
 \end{figure}
-%('rvd2\results\2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1')
+%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
 
-The original research describing this sample used Strelka to identify mutations in the same sample. They identified XXX and XXX as variant, but missed the other three variants.
+The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but missed the other four variants.
 
@@
 
-% from folder '2013-11-24_depthM_plot_avg10_Qsd_0_1', dilution rate is 0.1.
+% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.
 

### 2013-11-26 doc/bioinformatics2013/RVD2.tex
updated gitignore minor manuscript changes.

@@
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:743899XC$\rightarrow$T, chr7:753635T$\rightarrow$C and chr7:780960C$\rightarrow$T are germline homozygous mutations.
+Positions chr7:743899XC$>$T, chr7:753635T$>$C and chr7:780960C$>$T are germline homozygous mutations.
 Positions chr7:754371 and chr7:758813 are called heterozygous in the normal sample.
-In the tumor cells we identified all five mutations chr7:743899C$\rightarrow$T, chr7:753635T$\rightarrow$C, chr7:754371T$\rightarrow$C, chr7:758813G$\rightarrow$A, and chr7:780960C$\rightarrow$T.
-Positions chr7:754371T$\rightarrow$C and chr7:758813G$\rightarrow$A are identified as loss-of-heterozygosity events.
+In the tumor cells we identified all five mutations chr7:743899C$>$T, chr7:753635T$>$C, chr7:754371T$>$C, chr7:758813G$>$A, and chr7:780960C$>$T.
+Positions chr7:754371T$>$C and chr7:758813G$>$A are identified as loss-of-heterozygosity events.
 
@@
 %%%%%%%%%%%%%%%%%
-\section*{Parameter setting for other variant calling algorithms}
+\section{Parameter settings for other variant calling algorithms}
 

### 2013-11-26 doc/bioinformatics2013/RVD2.tex
added rs positions to HCC calls.

@@
 
-RVD2 identified 5 variants in the 44,716bp PAXIP1 gene.
+RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:XXX to chr7:XXX.
 Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
-None of the identified mutations are at known dnSNPX.X locations.
 
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:743899XC$>$T, chr7:753635T$>$C and chr7:780960C$>$T are germline homozygous mutations.
-Positions chr7:754371 and chr7:758813 are called heterozygous in the normal sample.
-In the tumor cells we identified all five mutations chr7:743899C$>$T, chr7:753635T$>$C, chr7:754371T$>$C, chr7:758813G$>$A, and chr7:780960C$>$T.
-Positions chr7:754371T$>$C and chr7:758813G$>$A are identified as loss-of-heterozygosity events.
+Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are germline homozygous mutations.
+Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
+In the tumor cells we identified all five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
+Positions chr7:154754371T$>$C and chr7:154758813G$>$A then loss-of-heterozygosity events.
 
-% TODO: replace X>X with the reference and mutated base here
-% Done
+%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
+%TODO: place these rs labels beneath the position labels on the x-axis of the figure. Remove the above sentence and add. 
+%Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding rs identities are shown in the figure.
 
@@
 
-The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but missed the other four variants.
+The original research describing this sample used Strelka to identify mutations in the same sample.
+They identified chr7:154760439 as variant, but did not call the other four variants.
+In particular strelka missed the two LOH events.
 

### 2013-11-27 doc/bioinformatics2013/RVD2.tex
Updated the manuscript

@@
 A matched normal cells were derived from lymphoblastoid cells from whole blood.
+Sequecing libraries were prepared according to the protocal described in the original techical report~\cite{}.
+The raw FASTQ read files were aligned to hg19 igenome ~\cite{} using the Isaac aligner~\cite{} to generate BAM files.
+We used samtools mpileup to generate pileup files using hg19 igenome ~\cite{} as reference sequence.
+
+% TODO:cite the paper Raczy C, Petrovski R, Saunders C, Chorny I, Kruglyak S, Margulies E, Chuang HY, Kallberg M, Kumar SA, Liao A, Little KM, Stromberg M, Tanner S.
+Isaac: Ultra-fast whole genome secondary analysis on Illumina sequencing platforms.
+2013
+% TODO:cite the technical report.
 % TODO: check matched cells line origin
@@
 
-Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
+%Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
 The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
@@
 % TODO: discuss alignment protocol
+% TODO: rvd2 parameters
+% TODO: chi2 test
+% TODO: Pubmed for peer-reviewed paper on HCC1187 sample
 
@@
 We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
+
+Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2.
+First, we compared the posterior density $\mu_j^{\Delta}$ to find out if $\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation and confidence level $1-\alpha$ was set at 0.95.
+% The difference is constant at 1.74*10^{-4}
+\begin{equation}\label{eqn:bayes_test}
+	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
+\end{equation}
+\begin{equation}\label{eqn:bayes_test}
+	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
+\end{equation}
+
+%TODO: write up the somatic test section
+%TODO: specify the results from somatic test or from the $\mu_j^{\Delta}$ posterior density test.
+
+Second, we identified the gemerline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently. We can compute
+$[0.0,0.2],[0.2,0.8],[0.8.1]$
+confidence level $1-\alpha$ was set at 0.8
+
+\begin{equation}\label{eqn:bayes_test}
+	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha,
+\end{equation}
 
@@
 
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:XXX to chr7:XXX.
+RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
 Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
@@
 In the tumor cells we identified all five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
-Positions chr7:154754371T$>$C and chr7:154758813G$>$A then loss-of-heterozygosity events.
-
-%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-%TODO: place these rs labels beneath the position labels on the x-axis of the figure. Remove the above sentence and add. 
-%Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding rs identities are shown in the figure.
+Positions chr7:154754371T$>$C and chr7:154758813G$>$A then loss-of-heterozygosity events.Some of these mutations are also found to be common population SNPs according to dbSNPv138.
+The corresponding rs identities are shown in the figure.
 
@@
 In particular strelka missed the two LOH events.
+
+% we ran strelka and it found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
 

### 2013-11-28 doc/bioinformatics2013/RVD2.tex
Update manuscript.

@@
 %TODO: Label all results/figures with comment indicating what experiment folder those results depend on and come from.
-
 \begin{abstract}
@@
 Sequecing libraries were prepared according to the protocal described in the original techical report~\cite{}.
-The raw FASTQ read files were aligned to hg19 igenome ~\cite{} using the Isaac aligner~\cite{} to generate BAM files.
-We used samtools mpileup to generate pileup files using hg19 igenome ~\cite{} as reference sequence.
+The raw FASTQ read files were aligned to hg19 igenome ~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
+We used samtools mpileup to generate pileup files using hg19 igenome ~\cite{} as reference sequence.~\citep{Navin:2010gu}.
+HCC1187 dataset is a well-recognized dataset from Illumina on which different research has been done~\citep{newman2013relative,howarth2011large,howarth2007array}.
 
-% TODO:cite the paper Raczy C, Petrovski R, Saunders C, Chorny I, Kruglyak S, Margulies E, Chuang HY, Kallberg M, Kumar SA, Liao A, Little KM, Stromberg M, Tanner S.
-Isaac: Ultra-fast whole genome secondary analysis on Illumina sequencing platforms.
-2013
 % TODO:cite the technical report.
@@
 
-% TODO: discuss alignment protocol
-% TODO: rvd2 parameters
-% TODO: chi2 test
-% TODO: Pubmed for peer-reviewed paper on HCC1187 sample
+% TODO: discuss alignment protocol. Done.
+% TODO: rvd2 parameters. Done.
+% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done
 
@@
 Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2.
-First, we compared the posterior density $\mu_j^{\Delta}$ to find out if $\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation and confidence level $1-\alpha$ was set at 0.95.
+First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95.
+We did $\chi^2$ test afterwards to further remove false callings.
 % The difference is constant at 1.74*10^{-4}
-\begin{equation}\label{eqn:bayes_test}
+\begin{equation}\label{eqn:bayes_test_positive}
 	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
 \end{equation}
-\begin{equation}\label{eqn:bayes_test}
+\begin{equation}\label{eqn:bayes_test_negative}
 	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
@@
 
-%TODO: write up the somatic test section
-%TODO: specify the results from somatic test or from the $\mu_j^{\Delta}$ posterior density test.
+%TODO: write up the somatic test section. Done.
 
-Second, we identified the gemerline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently. We can compute
-$[0.0,0.2],[0.2,0.8],[0.8.1]$
-confidence level $1-\alpha$ was set at 0.8
+Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently.
+Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
+\begin{equation}\label{eqn:bayes_test_interval}
+	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
+\end{equation}
 
-\begin{equation}\label{eqn:bayes_test}
-	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha,
-\end{equation}
+Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.
 
@@
 In particular strelka missed the two LOH events.
-
-% we ran strelka and it found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
+% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample.
+Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
 
@@
 In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap.
-mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling.
+Samtools mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling.
 Read depth for each sample was kept with -D enabled, and -u was turned on to make sure the output bcf files uncompressed.
@@
 Due to some format  incompatibility, we applied Picard to format read group and GATK for realignment as preprocessing.
-When calling variants using UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches.
+InUnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches.
 All other parameters were set at default.

### 2013-12-04 doc/bioinformatics2013/RVD2.tex
updated TODO in manuscript

--- 2013-11-26 doc/bioinformatics2013/RVD2.tex
@@
 \begin{document}
-%TODO: Label all results/figures with comment indicating what experiment folder those results depend on and come from.
 
@@
 Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
-Clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
+Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
 However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
 We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
-We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency.
+We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
 We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
@@
 
+Several studies have compared the relative performance of these algorithms.
+%TODO: [DISCUSS \citet{Stead:2013fu}] and others
+
 Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
@@
 RVD uses a two-stage approach for detecting for rare variants.
-First, it estimates the parameters of two hierarchical Bayesian models; one using data from the sample of interest (case) and one using data from a known reference sample (control).
-Then, it tests for a significant difference between model parameters in the case and control samples and returns called variant positions.
+First, it estimates the parameters of a hierarchical Bayesian model under two sequencing data sets: one from the sample of interest (case) and one from a known reference sample (control).
+Then, it tests for a significant difference between key model parameters in the case and control samples and returns called variant positions.
 
 For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-The generative process is as follows:
+The model generative process is as follows:
 
@@
 
-This process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
+The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
 The global error rate, $\mu_0$, estimates the expected error rate across all locations.
@@
 Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
-In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\cite{}.
+In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{Jordan2004}.
 A rounded box or ``plate" represents replication of the nodes within the plate.
-This graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
+The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
 
@@
 \begin{equation}\label{eqn:jointpdf}
-p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right)
+p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
 \end{equation}
 
-The individual conditional distributions are:
+where
 \begin{align}
@@
 
-Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data at location $j$ in replicate $i$, 
+Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data, 
 \begin{equation}
@@
 
-Finally, the complete log-likelihood for the data set is
+Finally, the log-likelihood of the data set is
 
@@
 The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution.
-However, there is not a closed form for the product of a Beta distribution with another Beta distribution and exact inference is intractable.
-exact inference is intractable.
+However, there is not a closed form for the product of a Beta distribution with another Beta distribution, so exact inference is intractable.
 
@@
 $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. 
-While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encounter such pathological behavior in this application.
+While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
+\subsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
 The posterior distribution over $\mu_j$ given its Markov blanket is 
@@
 
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$. The acceptance probability is then
+A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from ythe posterior distribution.
+The acceptance probability is then
 \begin{equation}
@@
 Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
+%TODO: make sure \sigma_j is correct here.
 
@@
 We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
-Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors.
+Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.
 
@@
 \subsection{Posterior Density Test}\label{sec:hypothesis_test}
-Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data.
+Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
 For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
@@
 The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ maximizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
+The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
 \begin{equation}
-	\tau^* = \arg\max_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
+	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
 \end{equation}
 
-While we are able to compute the optimal $\tau$ threshold for this test data set, in general we would not have access to $\tau^*$.
+While we are able to compute the optimal $\tau$ threshold for a test data set, in general we would not have access to $\tau^*$.
 With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest.
@@
 
-The non-reference bases may be due to a true mutation or due to a random sequencing error and we would like to differentiate these two scenarios.
-We assume non-reference read counts caused by a non-biological mechanism results in a uniform distributed among three non-reference categories.
-In contrast, the distribution of counts among three non-reference bases caused by biological mutation would be biased to one base.
+An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios.
+We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
+In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
 
@@
 where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.
+\citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
 
-At each position, we have one statistic for each replicate. We use Fisher's combined probability test to combine the $N$ p-values into a single p-value for each position,
+We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,
 
 \begin{equation}\label{eqn:fisher_combined}
-	X^2 = -2 \sum_{i=1}^N \ln(p_i).
+	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
 \end{equation}
 
-Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.
-
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test}.
+Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
+Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{}.
+%TODO: add two citations here
 
@@
 We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms.
-The synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
-The HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
+Synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
+HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
 
@@
 Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-More details of the experimental protocol are available from the original publication~\cite{}.
+Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
 We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
@@
 The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
-A matched normal cells were derived from lymphoblastoid cells from whole blood.
+A matched normal cells were derived from lymphoblastoid cells from whole blood~\citep{}.
+% TODO: cite the original publication here.
 % TODO: check matched cells line origin
@@
 We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
-For these ROC curves, we only used the Bayesian test without the $\chi^2$ test.
+For these ROC curves, we used the Bayesian test without the $\chi^2$ test.
 Adding the $\chi^2$ improves specificity at the expense of sensitivity.
@@
 The performance improves monotonically with read depth.
-Figures~\ref{fig:ROC}B-C for larger MAFs show a similar relationship between coverage depth and accuracy.
+Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
 
@@
 
-Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
+%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
 All of the called positions show a clear difference between the case and control error rates.
 The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
-\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-\label{fig:MAF}
-\end{center}
-\end{figure}
-% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
+%
+%\begin{figure}[h]
+%\begin{center}
+%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
+%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+%\label{fig:MAF}
+%\end{center}
+%\end{figure}
+%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
 
@@
 In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
-The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in calling all of the true variants and none of the true negatives.
-
-[DISCUSS \citet{Stead:2013fu}]
+The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
+GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
 
@@
 When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see supplementary table SX).
+The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see Appendix~\ref{sec:app_comparison}).
 
@@
 However, the FDR is high due to many false positives.
-Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and 1.00 for the highest read depth.
+Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth.
 Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF.
@@
 MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
-But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\%-100\% MAF.
+But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.
 
@@
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
+%TODO: add start and stop location here
 
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are germline homozygous mutations.
+Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
 Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor cells we identified all five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
-Positions chr7:154754371T$>$C and chr7:154758813G$>$A then loss-of-heterozygosity events.
-
+In the tumor sample we identified five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
+Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
+Some of these mutations are also found to be common population SNPs according to dbSNPv138.
+The corresponding identities are shown in the figure.
+%TODO: verify these rs positions
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-%TODO: place these rs labels beneath the position labels on the x-axis of the figure. Remove the above sentence and add. 
-%Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding rs identities are shown in the figure.
 
@@
 
-The original research describing this sample used Strelka to identify mutations in the same sample.
+The original research describing this sample used Strelka to identify mutations.
 They identified chr7:154760439 as variant, but did not call the other four variants.
-In particular strelka missed the two LOH events.
+In particular Strelka missed the two LOH events.
 
@@
 We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data.
-Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths.
+Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths and minor allele frequencies.
 
-Our inference algorithm uses Gibbs sampling and Metropolis-Hastings sampling to estimate the empirical Bayes statistical model.
-This sampling procedure provides a guarantee to identify the global optimal parameter settings in the long run.
+Our inference algorithm uses Gibbs sampling to estimate the hierarchical empirical Bayes model.
+This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically.
 However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches.
@@
 \section{Algorithm Comparison Statistics}\label{sec:app_comparison}
+%TODO: discuss this result
 
@@
 \section{Parameter settings for other variant calling algorithms}
+%TODO: review grammar and completeness here
 

### 2013-12-05 doc/bioinformatics2013/RVD2.tex

@@
 
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from ythe posterior distribution.
+A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from the posterior distribution.
 The acceptance probability is then
@@
 The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
-A matched normal cells were derived from lymphoblastoid cells from whole blood~\citep{}.
-% TODO: cite the original publication here.
+A matched normal cells were derived from lymphoblastoid cells from whole blood.
+Sequencing libraries were prepared according to the protocol described in the original technical report~\cite{}.
+The raw FASTQ read files were aligned to hg19 igenome~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
+We used samtools mpileup to generate pileup files using hg19 igenome~\cite{} as reference sequence~\citep{Navin:2010gu}.
+The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
+% TODO:cite the technical report.
 % TODO: check matched cells line origin
@@
 
-% TODO: discuss alignment protocol
+% TODO: discuss alignment protocol. Done.
+% TODO: rvd2 parameters. Done.
+% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done
 
@@
 We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
+
+%TODO: This section doesn't make sense to me. Lets talk about what needs to be done here.
+Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2.
+First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95.
+We did $\chi^2$ test afterwards to further remove false callings.
+% The difference is constant at 1.74*10^{-4}
+\begin{equation}\label{eqn:bayes_test_positive}
+	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
+\end{equation}
+\begin{equation}\label{eqn:bayes_test_negative}
+	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
+\end{equation}
+
+Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently.
+Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
+\begin{equation}\label{eqn:bayes_test_interval}
+	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
+\end{equation}
+
+Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.
+% End of section that is confusing to me.
 
@@
 
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:XXX to chr7:XXX.
+RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
 Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
-%TODO: add start and stop location here
 
@@
 
-The original research describing this sample used Strelka to identify mutations.
+The original research describing this sample used Strelka to identify mutations in the same sample.
 They identified chr7:154760439 as variant, but did not call the other four variants.
-In particular Strelka missed the two LOH events.
+In particular strelka missed the two LOH events.
+% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample.
+Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
 

### 2013-12-06 doc/bioinformatics2013/RVD2.tex
Updated the manuscript

@@
 
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise.
+We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
 Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-%TODO: make sure \sigma_j is correct here.
 
@@
 Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{}.
-%TODO: add two citations here
+Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
 
@@
 
+The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
 The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
-A matched normal cells were derived from lymphoblastoid cells from whole blood.
-Sequencing libraries were prepared according to the protocol described in the original technical report~\cite{}.
-The raw FASTQ read files were aligned to hg19 igenome~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
-We used samtools mpileup to generate pileup files using hg19 igenome~\cite{} as reference sequence~\citep{Navin:2010gu}.
-The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
-% TODO:cite the technical report.
-% TODO: check matched cells line origin
-% double checked by Yuting, didn't find the matched cells line origin
-% TODO: mention tumor purity here
+The estimated tumor purity was reported to be 0.8.
+A matched normal cells were derived from lymphoblastoid cells from peripheral blood.
+Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
+The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
+We used samtools mpileup to generate pileup files using hg19 ~\cite{} as reference sequence~\citep{Navin:2010gu}.
+% TODO:cite the technical report. %Not sure the author is correct
+% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
+% TODO: how to cite hg19?
 
-Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
+%Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
 The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
 Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% coverage with 10 or more reads~\cite{}.
-% TODO: write this in own words.
-% TODO: add citation here.
 
-% TODO: discuss alignment protocol. Done.
 % TODO: rvd2 parameters. Done.
-% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done
 
@@
 % End of section that is confusing to me.
+
+%TODO: adjust the posterior density interval $[0.0,0.2],[0.2,0.8],[0.8.1]$ and do the experiment
+%TODO: find the correct terminology for alpha.
 
@@
 The corresponding identities are shown in the figure.
-%TODO: verify these rs positions
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
@@
 \section{Parameter settings for other variant calling algorithms}
-%TODO: review grammar and completeness here
-
-\paragraph{\textbf{Samtools}}We used samtools/bcftools (v0.1.19) for variant calling.
-mpileup was the function to call variants and bcftools was used to save the result in standard Variant Call Format (VCF) files.
-In mpileup, we set the -d option sufficiently high at $10^6$ to avoid coverage cap.
-Samtools mpileup was able to accept multiple replicates at one time so six bam files from each replicate group were fed into mpileup for variant calling.
-Read depth for each sample was kept with -D enabled, and -u was turned on to make sure the output bcf files uncompressed.
+\paragraph{\textbf{Samtools}}We used samtools (v0.1.19) function mpileup to call variants and bcftools to save the result in standard VCF files.
+In mpileup, we set the -d option sufficiently high at $10^6$ to avoid truncating read deapth.
+Option -u was enabled to make sure the output bcf files were uncompressed.
 
 \paragraph{\textbf{GATK}}
-UnifiedGenotyper is the function we used in GATK (v2.1-8) to detect mutations.We followed the general GATK recomended workflow to detect variants.
-Due to some format  incompatibility, we applied Picard to format read group and GATK for realignment as preprocessing.
-InUnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; we change the default setting for -dcov (default max coverage cutoff of $250\times$ ) to $10^6$ to avoid downsamplig so that the coverage was comparable to other approaches.
-All other parameters were set at default.
-Six replicates was passed to GATK for variant calling.
+We used GATK (v2.1-8) UnifiedGenotyper function to detect mutations on our synthetic data following the recommended workflow.
+Due to some format incompatibility, we applied Picard to format read group and GATK for realignment.
+In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsamplig coverage within GATK.
 
 \paragraph{\textbf{VarScan2-mpileup}}
-VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.
-We set -d at $10^6$ and enabled -50 for same concern previously stated.
-In afterwards mpileup2snp function, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.
-Six replicates were all introduced for variant calling.
+VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.We assigned parameter -C value 50 as the synthetic data was aligned using BWA and set -d at $10^6$.
+In mpileup2snp, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.
 
 \paragraph{\textbf{VarScan2-somatic}}
-We tested VarScan2-somatic on our six replicates synthetic data set as well.
-Considering that RVD2 doesn't call somatic status, we combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown).
---normal-purity was set at 1.00, while --tumor-purity was set as the dilution rate.
---min-var-freq was set at $10^{-5}$.
+We tested VarScan2 somatic on our synthetic dataset.
+The parameter --normal-purity set was at 1.00, --tumor-purity at the dilution rate.
+The parameter --min-var-freq was set at $10^{-5}$.
+We combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown) to compare with performance of RVD2.
 
-\paragraph{\textbf{Strelka}}
-Configuration and Analysis for Strelka is standard so we just installed the program and ran it on our data set.
-One thing to note is that since Strelka can not accept relicate data, we used a single replicate with a read depth that most closely matched the overall median depth of the replicates.
+\paragraph{\textbf{Strelka and muTect}}
+Since configuration and Analysis for Strelka and muTect is standardized and no parameter needs to be specified, we installed these two programs and ran them on our data set separately.
 
-\paragraph{\textbf{muTect}}
-Similar to Strelka, we ran muTect with standard settings and only passed one replicate of synthetic data.
+Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
+GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
+Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
 
+% TODO: modify the following section
 GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
@@
 We find its performance on much lower MAF variants and across a wider range of coverage depths.
-Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
-
 \bibliographystyle{apalike}

### 2013-12-06 doc/bioinformatics2013/RVD2.tex
updated results section with narrative of testing procedure.

--- 2013-12-05 doc/bioinformatics2013/RVD2.tex
@@
 The raw FASTQ read files were aligned to hg19 igenome~\cite{} using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
+The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads.
 We used samtools mpileup to generate pileup files using hg19 igenome~\cite{} as reference sequence~\citep{Navin:2010gu}.
@@
 
-Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
-The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
-Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% coverage with 10 or more reads~\cite{}.
-% TODO: write this in own words.
-% TODO: add citation here.
-
-% TODO: discuss alignment protocol. Done.
 % TODO: rvd2 parameters. Done.
-% TODO: Pubmed for peer-reviewed paper on HCC1187 sample. Done
 
@@
 
-%TODO: This section doesn't make sense to me. Lets talk about what needs to be done here.
-Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2.
-First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95.
-We did $\chi^2$ test afterwards to further remove false callings.
-% The difference is constant at 1.74*10^{-4}
-\begin{equation}\label{eqn:bayes_test_positive}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
-\end{equation}
-\begin{equation}\label{eqn:bayes_test_negative}
-	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
-\end{equation}
+We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
+To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
 
-Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently.
-Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
-\begin{equation}\label{eqn:bayes_test_interval}
-	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
-\end{equation}
-
-Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.
-% End of section that is confusing to me.
+To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
+loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
+homozygous somatic mutation).
+The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$.
+The size of the test is $\alpha=0.05$.
+%TODO: correct notation from \Pr to summation because we have samples, not distributions.
 

### 2013-12-06 doc/bioinformatics2013/RVD2.tex

--- 2013-12-06 doc/bioinformatics2013/RVD2.tex
@@
 The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
-We used samtools mpileup to generate pileup files using hg19 ~\cite{} as reference sequence~\citep{Navin:2010gu}.
+The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads.
+We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
 % TODO:cite the technical report. %Not sure the author is correct
 % HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
-% TODO: how to cite hg19?
-
-%Sequencing libraries were prepared from 500 ng of DNA using an early access version of Illuminas TruSeq DNA PCR-Free kit.
-The libraries were sequenced on four and eight lanes of a HiSeq 2000 flow cell for the normal and tumor, respectively, using 100 bp paired-end reads.
-Data, analyzed using the pre-release version of the Illumina Cancer Sequencing Workflow showed an average coverage of 40x for the normal and 90x for the tumor DNA with about 96\% coverage with 10 or more reads~\cite{}.
-
-% TODO: rvd2 parameters. Done.
 
@@
 
-%TODO: This section doesn't make sense to me. Lets talk about what needs to be done here.
-Considering HCC1187 sample is diploid, we did two tests based on posterior density using RVD2.
-First, we compared the posterior density $\tilde\mu_j^{\Delta}$ to find out if $\tilde\mu_j$ in case sample was significantly higher than that in control sample and vice verse.Threshold $\tau$ was set at $1\times10^{-3}$ to avoid error brought by computer limitation; confidence level $1-\alpha$ was set at 0.95.
-We did $\chi^2$ test afterwards to further remove false callings.
-% The difference is constant at 1.74*10^{-4}
-\begin{equation}\label{eqn:bayes_test_positive}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
-\end{equation}
-\begin{equation}\label{eqn:bayes_test_negative}
-	\Pr( \tilde{\mu}_j^{\text{control}} - \tilde{\mu}_j^{\text{case}} \geq \tau ) > 1-\alpha,
-\end{equation}
+We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
+To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
 
-Second, we identified the germline status and somatic status of HCC1187 sample by looking into posterior density of control sample and case sample independently.
-Basically, We computed the empirical posterior probability of $\tilde\mu_j$ across interval $[\tau_\text{l},\tau_\text{u}]$ and compare it with confidence level $1-\alpha$,
-\begin{equation}\label{eqn:bayes_test_interval}
-	\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha.
-\end{equation}
-
-Confidence level $1-\alpha$ was set at 0.8 due to our lower confidence on germline and somatic status.Posterior density interval for reference, heterzygous mutation and homozygous mutation were set as $[0.0,0.2],[0.2,0.8],[0.8.1]$, respectively.
-% End of section that is confusing to me.
-
-%TODO: adjust the posterior density interval $[0.0,0.2],[0.2,0.8],[0.8.1]$ and do the experiment
-%TODO: find the correct terminology for alpha.
+To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
+loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
+homozygous somatic mutation).
+The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$.
+The size of the test is $\alpha=0.05$.
+%TODO: correct notation from \Pr to summation because we have samples, not distributions.
 

### 2013-12-06 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/plot_prior_posterior.py
Plot the distribution of priors and posteriors for rvd29 (Jeffreys prior) and rvd30 (lognormal prior)

@@ -1,0 +1,89 @@
+from scipy.special import gammaln, polygamma
+from scipy.stats import gaussian_kde
+import numpy as np
+import scipy.stats as ss
+import matplotlib.pyplot as plt
+import sys
+import os
+import h5py
+import logging
+import itertools
+import math
+import rvd30
+import rvd29
+
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    locationList = (100,300,104,244)
+    
+    # Same experiment condition: read depth rate = 1/100
+    h5Filename_rvd29 = './hdf5_rvd29_gibbs=20000_dc=100/Case10_0.hdf5'
+    h5Filename_rvd30 = './hdf5_rvd30_gibbs=20000_dc=100/Case10_0.hdf5'
+    M = range(1,10000)
+    fig=plt.figure(figsize=(20, 12))
+
+    """ Probability of Jeffreys' prior from rvd29"""
+    logging.debug("Plot rvd29 prior and posterior")
+    for loc in locationList:
+        logging.debug("Processing location: %d" % loc)
+        ax = fig.add_subplot(2,4,locationList.index(loc)+1)
+        (Phi, M_s, Theta, Mu, Loc, R, N) = rvd29.load_model(h5Filename_rvd29)
+        mu=np.mean(Mu, axis=0) 
+        mu=np.mean(mu)
+        p = np.sqrt(polygamma(1, np.dot(mu,M))*(mu**2) + polygamma(1, np.dot(1-mu,M))*(1-mu)**2 -polygamma(1,M))
+        sum = np.sum(p)
+        print (sum)
+        M1 = M_s[loc]       
+        # Use Gaussian Kernel Density Estimate for posterior Mj
+        kernel = gaussian_kde(M1) 
+        norm = kernel.integrate_box(0,10000)
+        print (norm)
+        ax.semilogy(M, p/sum)
+        ax.semilogy(M, kernel(M).T/norm)
+        ax.legend( ['prior','posterior'],loc=10, ncol=10)
+        subtitle='position='+str(loc)
+        ax.set_title(subtitle)
+        if locationList.index(loc)==0:
+           ax.set_ylabel('Probability(Jeffrey)',fontsize=10)
+
+    """ Probability of lognormal prior from rvd30"""
[…]

### 2013-12-06 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/plot_prior_posterior_rvd29.py

@@ -1,0 +1,52 @@
+#!/usr/bin/env python
+# Plot the probability of Jeffreys' prior and the posterior Mj over a range of M.
+
+from scipy.special import gammaln, polygamma
+from scipy.stats import gaussian_kde
+import numpy as np
+import scipy.stats as ss
+import matplotlib.pyplot as plt
+import sys
+import os
+import h5py
+import logging
+import rvd29
+import itertools
+import math
+# read depth rate = 1/100
+h5Filename = './hdf5_rvd29_gibbs=20000_dc=100/Case0_1.hdf5'
+loc = 104
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+
+    M = range(1,10000)
+
+    # Probability of Jeffreys' prior from rvd29
+    (Phi, M_s, Theta, Mu, Loc, R, N) = rvd29.load_model(h5Filename)
+
+    # mu from gibbs
+    mu=np.mean(Mu, axis=0) 
+    mu=np.mean(mu)
+    print mu
+    
+    p = np.sqrt(polygamma(1, np.dot(mu,M))*(mu**2) + polygamma(1, np.dot(1-mu,M))*(1-mu)**2 -polygamma(1,M))
+
+    M1 = M_s[loc]
+    print M1
+
+    # Use Gaussian Kernel Density Estimate for posterior Mj
+    kernel = gaussian_kde(M1)  
+    fig=plt.figure(figsize=(20,12))
+
+    plt.semilogy(M,p)
+    plt.semilogy(M,kernel(M).T)
+
+    #figname = "pdf_%s" %loc
+    plt.savefig('pdf.pdf')
+
+if __name__ == '__main__':
+    main()
+    

### 2013-12-06 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/plot_prior_posterior_rvd30.py

@@ -1,0 +1,55 @@
+#!/usr/bin/env python
+# Plot the probability of lognormal prior and the posterior Mj over a range of M.
+
+from scipy.special import gammaln, polygamma
+from scipy.stats import gaussian_kde
+import numpy as np
+import scipy.stats as ss
+import matplotlib.pyplot as plt
+import sys
+import os
+import h5py
+import logging
+import rvd30
+import itertools
+import math
+# read depth rate = 1/100
+h5Filename = '../2013-11-23_test_rvd30_synthetic_data/100/Case0_1.hdf5'
+loc = 0
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+
+    M = range(1,10000)
+
+    # Probability of lognormal prior from rvd30
+    (Phi, M_s, Theta, Mu, Loc, R, N) = rvd30.load_model(h5Filename)
+
+    # mu from gibbs
+    mu=np.mean(Mu, axis=0) 
+    mu=np.mean(mu)
+
+    print mu
+    print Phi['delta']
+    print Phi['mu_M']
+
+    p = ss.lognorm.pdf(M, Phi['delta'], scale=np.exp(Phi['mu_M']))
+    print p
+
+    M1 = M_s[loc]
+    
+    # Use Gaussian Kernel Density Estimate for posterior Mj
+    kernel = gaussian_kde(M1)  
+    fig=plt.figure(figsize=(20,12))
+
+    plt.semilogy(M,p)
+    plt.semilogy(M,kernel(M).T)
+
+    #figname = "pdf_%s" %loc
+    plt.savefig('pdf.pdf')
+
+if __name__ == '__main__':
+    main()
+    

### 2013-12-06 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/runall.py

@@ -1,0 +1,67 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the rvd30 directory at front of the path
+
+rvddir = os.path.join('./rvd30')
+#rvddir = os.path.join('../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd30
+
+##pool =None
+pool = mp.Pool(processes=10)
+tocfilename = "synthetic_toc_p1.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 20000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/100/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd30.load_depth(controlFileList)
+    phi, theta_s, mu_s, M_s = rvd30.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd30.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, M=M_s, r=r, n=n, loc=loc, refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2013-12-06 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/rvd29.py

@@ -1,0 +1,787 @@
+#!/usr/bin/env python
+
+
+from __future__ import print_function
+from __future__ import division
+
+import numpy as np
+
+import scipy.stats as ss
+from scipy.special import gammaln, polygamma
+from scipy import special
+
+import logging
+import multiprocessing as mp
+from itertools import repeat
+import h5py
+import tempfile
+
+import os
+import subprocess
+from datetime import date
+
+import re
+import pdb
+import time
+
+
+"""rvd29.py: put Jeffrey's Prior on Mj."""
+
+
+
+def main():
+    import argparse
+
+    # Populate our options, -h/--help is already there.
+    description = """
+                    RVD is a hierarchical bayesian model for identifying
+                    rare variants from short-read sequence data. """
+                    
+    # create the top-level parser
+    argp = argparse.ArgumentParser(prog='rvd', description=description)
+    argp.add_argument('--version', action='version', version='%(prog)s 2.7')
+    argp.add_argument('-v', '--verbose', dest='verbose', action='count',
+                help="increase verbosity (specify multiple times for more)")
+    
+    # argp.add_argument('cmd', action='store', nargs='*',
+    #             choices=['gen', 'gibbs'])
+    subparsers = argp.add_subparsers(help='sub-command help')
+    
+    # create subparser for gibbs fitting
+    argpGibbs = subparsers.add_parser('gibbs', 
+                        help='fit the RVD model using Gibbs sampling')
+    argpGibbs.add_argument('dcfile', nargs='+',
+                        help='depth chart file name')
+    argpGibbs.add_argument('-o', dest='outputFile', 
+                default='output.hdf5',
+                help='output HDF5 file name')
+    argpGibbs.add_argument('-p', '--pool', type=int, default=None,
+                help='number of workers in multithread pool')
+    argpGibbs.set_defaults(func=gibbs)
+                
+                
+    # create subparser to compare two model files
+    argpTest = subparsers.add_parser('test', 
+                        help='test if case error rate is greater than control by T')
+    argpTest.add_argument('controlHDF5Name',
+                help='control model file (HDF5)')
[…]

### 2013-12-06 results/2013-11-03_rvd28_Gammar_Prior/plot.py
Add notebooks.

@@ -1,0 +1,14 @@
+#!/usr/bin/env python
+
+
+import scipy.stats as ss
+import numpy as np
+import matplotlib.pyplot as plt
+
+alpha, loc, beta=2.5e4, 1, 5
+data=ss.gamma.rvs(alpha,loc=loc,scale=beta,size=5000)
+myHist = plt.hist(data, 100, normed=True)
+rv = ss.gamma(alpha,loc,beta)
+x = np.linspace(0,100) 
+h = plt.plot(x, rv.pdf(x), lw=2)
+plt.show()

### 2013-12-08 doc/bioinformatics2013/RVD2.tex
typo corrected.

@@
 Due to some format incompatibility, we applied Picard to format read group and GATK for realignment.
-In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsamplig coverage within GATK.
+In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsampling coverage within GATK.
 

### 2013-12-11 doc/ISMB2014/RVD3.tex
ISMB2014 RVD3 folder (outline and first draft).

--- 2013-12-06 doc/bioinformatics2013/RVD2.tex
@@
 We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+
+(RVD3) Sensitivity analysis Bayesian model
+Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
 
@@
 
+(RVD3)
+Improper Prior: Playing no prior on M is an implicit improper prior.
+Jeffreys Prior:Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
+\begin{equation}\label{eqn:JefferyPrior}
+\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
+\end{equation}
+
+Lognormal Prior:
+In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
+Incorporate lognormal prior on $M_{j}$.
+
 %%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%
+% Sampling Mj
+%%%%%%%%%%%%%%%%%%
+\subsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
+(RVD3) For Jeffreys prior,since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by
+
+\begin{equation}\label{equ:JefferyInference}
+I\left({M}_{j}\right)={E}_{{M}_{j}}\left[ -\frac{\delta ^{2}\log p\left(\theta _{j}|\mu_{j},M_{j}\right)}{\delta M^{2}_{j}}\right]
+\end{equation}
+
+Now to calculate the equations to acquire the Fisher information, we assume there is only one replicate,
+
+\begin{equation}\label{eqn:Betapdf}
+p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
+\end{equation}
+
+\begin{equation}\label{equ:JefferyInference1}
+\begin{split}
+\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
+& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
+\end{split}
+\end{equation}
+
+\begin{equation}
+\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
+\end{equation}
+
+\begin{equation}
+\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
+\end{equation}
+
+Now we have the Jeffreys' prior for $M_{j}$:
+
+\begin{equation}
+\pi\left({M}_{j}\right) = [-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
+\end{equation}
+
+For log-normal prior, the posterior distribution over $M_{j}$ given its Markov blanket is
+
+\begin{equation}
+	p( M_{j} |\mu, \sigma, \theta_{ji},\mu_j) \propto p(\theta_{ji} | \mu_j, M_j) p(M_{j} | \mu, \sigma)
+\end{equation}
+
+We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$.
+Since we cannot give an analytical form for this, so we sample from the posterior distribution using the Metropolis-Hastings algorithm.
+
+%%%%%%%%%%%%%%%%%%
 % Posterior Density Test
@@
 
-\subsection{Performance with read depth}\label{sec:read_depth}
+\subsection{Sensitivity analysis of priors}
+In order to examine the robustness of RVD3 to changes of different priors, we...
 
+We found that RVD3 is robust to changing the priors.
+The RVD3 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:M_loc_rvd29} and Figure ~\ref{fig:M_loc_rvd30}.
+
+The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
+For the 100\% mutant data set, the read depth drops at the mutant locations.
+This is due to the parameters imposed at the alignment stage.
+The reads are 36bp long and we required no more than 2 mismatches.
+Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=120mm]{figs/M_loc_rvd29.pdf}
+\caption{Key parameters for RVD3 model with Jeffreys prior for synthetic DNA data sets.}
+\label{fig:M_loc_rvd29}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=120mm]{figs/M_loc_rvd29.pdf}
+\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
+\label{fig:M_loc_rvd30}
+\end{center}
+\end{figure}
+% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.
+
+The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
+$M_j$ measures the variance between replicates at location $j$.
+There is little variability across positions indicating that the replication variance does not change greatly across position.
+Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
+
+The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
+We see that the variation between replicates is smaller than the variation between location.
+$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
+Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
+
+\subsection{Results of priors on synthetic data}
+Performance with read depth:
 We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
@@
 The performance improves monotonically with read depth.
-Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
+Figures~\ref{fig:ROC_29}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
 
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
+\includegraphics[width=120mm]{figs/ROC_without_chi2_rvd29.pdf}
 \caption{ROC curve varying read depth showing detection performance.}
-\label{fig:ROC}
+\label{fig:ROC_29}
 \end{center}
 \end{figure}
-% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
 
-%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
-All of the called positions show a clear difference between the case and control error rates.
-The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
-%
-%\begin{figure}[h]
-%\begin{center}
-%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
-%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-%\label{fig:MAF}
-%\end{center}
-%\end{figure}
-%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=120mm]{figs/ROC_without_chi2_rvd30.pdf}
+\caption{ROC curve varying read depth showing detection performance.}
+\label{fig:ROC_30}
+\end{center}
+\end{figure}
 
-\subsection{Empirical performance compared with other algorithms}\label{sec:comparison}
-
+Empirical performance compared with other algorithms:
 We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
@@
 \begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
-\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
-\label{tbl:comparison_ss}
+\includegraphics[width=160mm]{tables/comparison_table_ss.pdf}
+\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
+\label{tbl:table_29}
 \end{center}
 \end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
 
@@
 
-\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
+\subsection{Results of Jeffreys prior on yeast data}
 
@@
 
-%%%%%%%%%%%%%%%%%
-% Appendix C: Data Set Statistics
-%%%%%%%%%%%%%%%%%
-\section{RVD2 Estimated Parameters}\label{sec:synthetic_estimate}
-The RVD2 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:depthM}.
-
-The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The reads are 36bp long and we required no more than 2 mismatches.
-Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
-\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
-\label{fig:depthM}
-\end{center}
-\end{figure}
-
-% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.
-
-The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
-
-The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-We see that the variation between replicates is smaller than the variation between location.
-$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
-Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
-
-%%%%%%%%%%%%%%%%%
-% Appendix D: Parameter Settings for Variant Calling Algorithms
-%%%%%%%%%%%%%%%%%
-\section{Parameter settings for other variant calling algorithms}
-\paragraph{\textbf{Samtools}}We used samtools (v0.1.19) function mpileup to call variants and bcftools to save the result in standard VCF files.
-In mpileup, we set the -d option sufficiently high at $10^6$ to avoid truncating read deapth.
-Option -u was enabled to make sure the output bcf files were uncompressed.
-
-\paragraph{\textbf{GATK}}
-We used GATK (v2.1-8) UnifiedGenotyper function to detect mutations on our synthetic data following the recommended workflow.
-Due to some format incompatibility, we applied Picard to format read group and GATK for realignment.
-In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsamplig coverage within GATK.
-
-\paragraph{\textbf{VarScan2-mpileup}}
-VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.We assigned parameter -C value 50 as the synthetic data was aligned using BWA and set -d at $10^6$.
-In mpileup2snp, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.
-
-\paragraph{\textbf{VarScan2-somatic}}
-We tested VarScan2 somatic on our synthetic dataset.
-The parameter --normal-purity set was at 1.00, --tumor-purity at the dilution rate.
-The parameter --min-var-freq was set at $10^{-5}$.
-We combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown) to compare with performance of RVD2.
-
-\paragraph{\textbf{Strelka and muTect}}
-Since configuration and Analysis for Strelka and muTect is standardized and no parameter needs to be specified, we installed these two programs and ran them on our data set separately.
-
-Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
-GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
-Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
-
-% TODO: modify the following section
-GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
-Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
-\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
-We find its performance on much lower MAF variants and across a wider range of coverage depths.
 \bibliographystyle{apalike}

### 2013-12-12 doc/bioinformatics2013/RVD2.tex
Manuscript updated.

--- 2013-12-06 doc/bioinformatics2013/RVD2.tex
@@
 Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
-Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{}.
-GATK involves more sophisticate pre and post-processing steps.
+Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{li2011statistical,depristo2011framework}.
+GATK involves more sophisticate pre- and post-processing steps.
 The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
@@
 Several studies have compared the relative performance of these algorithms.
-%TODO: [DISCUSS \citet{Stead:2013fu}] and others
+%TODO: [DISCUSS \citet{Stead:2013fu}, \citet{spencer2013performance}] and others
 
@@
 Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
-In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{Jordan2004}.
+In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{jordan2004graphical}.
 A rounded box or ``plate" represents replication of the nodes within the plate.
@@
 where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-\citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
+\citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
 
-We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,
+We control for multiple hypothesis testing in two ways.
+We use Fisher's combined probability test \citep{fisher1970statistical} to combine the p-values for $N$ replicates into a single p-value at position $j$,
 
@@
 The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
-The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
+The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
 The estimated tumor purity was reported to be 0.8.
@@
 When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see Appendix~\ref{sec:app_comparison}).
 
@@
 \section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-%TODO: discuss this result
+Figure~\ref{fig:comparison_mcc} compares RVD2 with samtools, GATK, varscan, strelka and muTect using Matthew Correlation Coefficient (MCC)~citep(matthews1985homeostasis).
 
-\begin{figure}[h]
+Samtools and VarScan2-mpileup achieved MCC value generally higher than 0.90 on 100\% MAF sample across all read depthes, with 1.0 represents for a perfect prediction.
+However, both of them achieved MCC value at 0.00 when MAF is 10.0\% or lower, which indicates no better than random prediction.
+GATK, Varscan2-somatic, Strelka outperformed Samtools and VarScan2-mpile on the 10.0\% MAF sample, while performed similarly in other cases.
+GATK5, achieved relatively stable MCC between 0.65~0.88.
+Strelka achieved even better MCC on 10\% MAF sample, more specifically around 1.00 when read deapth is high but 0.53 when read depth is 22.
+There is a very obvious but unconventional decreasing trend in VarScan2-somatic MCC value across different read depth and MAF level, a phenomenon also observed by ~\citet{Stead:2013fu}.
+It is because VarScan2-somatic tends to call more false positives as read depth gets higher.
+As to Mutect, overall it seems to performs the best among all the method RVD2 is comparing to.
+It achieve postive MCC values (-0.02--0.43, with a single case at 0.60), though too low to be practically meaningful.
+Moreover, muTect achieved relatively low MCC (0.77--0.90) comparing to some other algorithms when the MAF level is 10\% and 100\%, as a counteractive of being oversensitive.
+
+RVD2 achieved MCC value at 1.00 when the MAF is 100.0\% at all read depth and 10\% when read depth is not lower than 260.
+The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
+
+\begin{figure}[ht]
 \begin{center}
@@
 \end{figure}
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 % from folder 'rvd2\results\2013-09-19_operating_characteristics'
@@
 
-Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
-GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
+Samtools mpileup and GATK can accept multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
+VarScan2-mpileup takes multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
 Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.

### 2013-12-13 doc/ISMB2014/RVD2.tex
Draft update.

--- 2013-12-12 doc/bioinformatics2013/RVD2.tex
@@
 \section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-Figure~\ref{fig:comparison_mcc} compares RVD2 with samtools, GATK, varscan, strelka and muTect using Matthew Correlation Coefficient (MCC)~citep(matthews1985homeostasis).
+Figure~\ref{fig:comparison_mcc} compares RVD2 with samtools, GATK, varscan, strelka and muTect using Matthew Correlation Coefficient (MCC)~\citep{matthews1985homeostasis}.
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
+\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
+\label{fig:comparison_mcc}
+\end{center}
+\end{figure}
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 
@@
 The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
-
-\begin{figure}[ht]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
-\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
-\label{fig:comparison_mcc}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 

### 2013-12-18 doc/ISMB2014/RVD3.tex
Update RVD3 Draft

@@
 
-\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
+\title[RVD3]{RVD3: Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data}
 \author{}
@@
 \begin{abstract}
-Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
-Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
-However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
-We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
-We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
-
-(RVD3) Sensitivity analysis Bayesian model
+Sensitivity analysis Bayesian model
 Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
 
-%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
-%We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
 \end{abstract}
@@
 
+%%%%%%%%%%%%%%%%%%
+% Introduction
+%%%%%%%%%%%%%%%%%%
 \section{Introduction}
 
-Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}.
-Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}.
-As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
-NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.
+%%%%%%%%%%%%%%%%%%
+% Data Sets
+%%%%%%%%%%%%%%%%%%
+\section{Data Sets}
 
-Increasingly, NGS is being used to interrogate mutations in heterogeneous clinical samples.
-For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
-Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
-DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
+\subsection{Synthetic DNA Sequence Data}
 
-However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
-Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
-Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{}.
-GATK involves more sophisticate pre and post-processing steps.
-The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
+Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
+Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
 
-Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
-VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
-Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}.
-With the joint distribution available, one is able to identify locations with dissimilar allele frequencies.
-muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.
+To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
+The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
 
-Several studies have compared the relative performance of these algorithms.
-%TODO: [DISCUSS \citet{Stead:2013fu}] and others
-
-Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
-These steps may improve algorithm performance, but their presence makes it difficult to isolate what step is most responsible for the performance.
-Instead, we focus exclusively on the model structure and statistical inference steps.
-Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study.
-We have isolated the statistical inference and hypothesis testing steps so that they may be used independently or as part of a larger variant calling pipeline.
-
-We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}.
-In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
-We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
-Finally, in Section~\ref{sec:brca} we present results on detecting mutations in the PAXIP1 gene from matched tumor-normal data from the HCC1187 cell line.
+\subsection{Yeast Data}
 
 %%%%%%%%%%%%%%%%%%
-% Model Structure
+% RVD3 Model
 %%%%%%%%%%%%%%%%%%
-\section{Model Structure}\label{sec:model_structure}
+\section{RVD3 Model}
 
-RVD uses a two-stage approach for detecting for rare variants.
-First, it estimates the parameters of a hierarchical Bayesian model under two sequencing data sets: one from the sample of interest (case) and one from a known reference sample (control).
-Then, it tests for a significant difference between key model parameters in the case and control samples and returns called variant positions.
-
-For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-The model generative process is as follows:
-
-\begin{enumerate}[noitemsep]
-	\item For each location $j$:
-	\begin{enumerate}
-		\item Draw an error rate $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
-		\item For each replicate $i$:
-		\begin{enumerate}
-			\item Draw $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$
-			\item Draw $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
-		\end{enumerate}
-	\end{enumerate}
-\end{enumerate}
-
-The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
-The global error rate, $\mu_0$, estimates the expected error rate across all locations.
-The global precision, $M_0$, estimates the variation in the error rate  across locations.
-The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.
-
-RVD2 has three levels of sampling.
-First, a global error rate and global precision are chosen once for the entire data set.
-Then, at each location, a local precision is chosen and a local error rate is sampled from a Beta distribution.
-Finally, the error rate for replicate $i$ at location $j$ is drawn from a Beta distribution and the number of non-reference reads is drawn from a binomial.
-
-RVD2 hierarchically partitions sources of variation in the data.
-$r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
-$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility.
+\subsection{Model Structure}\label{sec:model_structure}
+RVD3 is based on a hierarchical Bayesian model.
+Through hypothesis test on case and control samples by RVD3, we can call the variants sucsessfully.
+Now the definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
+Three hyperparameters are:  $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision; $\mu_0$, the global error rate expected across all positions; $M_0$, the global position which estimates the variation in the error rate across the positioins.
+$M_j$, the local precision measures the alteration in the error rate across replicates at position $j$.
+RVD3 hierarchically includes four levels of samplings:  $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
+$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation caused by experimental repeatability.
 The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-Importantly, increasing the read depth $n_{ji}$ only reduces the sampling error, but does nothing to reduce experimental variation or variation due to sequence context.
+Two kinds of priors for $M_{j}$ are Log-normal prior and Jeffreys prior, which enhanced the former RVD model [XX].
+$ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ is the model for Log-normal prior, and $pi\left({M}_{j}\right)$ for Jeffreys prior is based on Fisher information.
+The graphical chart for RVD3 with different priors are shown in Figure~\ref{fig:graphical_model}.
 
@@
 \begin{center}
-\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
-\caption{RVD2 Graphical Model}
+\includegraphics[width=40mm]{figs/RVD3_model.pdf}
+\caption{RVD3 Graphical Model}
 \label{fig:graphical_model}
@@
 
-Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
-In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{Jordan2004}.
-A rounded box or ``plate" represents replication of the nodes within the plate.
-The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
-
-The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
-
-\begin{equation}\label{eqn:jointpdf}
-p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
-\end{equation}
-
-where
-\begin{align}
-p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
-p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
-p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
-\end{align}
-
-Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
-\begin{equation}
-p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
-
-Finally, the log-likelihood of the data set is
-
-\begin{equation}
-\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
-
-(RVD3)
-Improper Prior: Playing no prior on M is an implicit improper prior.
-Jeffreys Prior:Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
-\begin{equation}\label{eqn:JefferyPrior}
-\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
-\end{equation}
-
-Lognormal Prior:
-In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
-Incorporate lognormal prior on $M_{j}$.
-
 %%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%
-\section{Inference and Hypothesis Testing}
+\subsection{Inference and Hypothesis Testing}
 
@@
 %%%%%%%%%%%%%
-\subsection{Initialization}
+\subsubsection{Initialization}
 The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
+\subsubsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
+\subsubsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
 The posterior distribution over $\mu_j$ given its Markov blanket is
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
-(RVD3) For Jeffreys prior,since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by
+\subsubsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
+For Jeffreys prior,since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Posterior Density Test}\label{sec:hypothesis_test}
+\subsubsection{Posterior Density Test}\label{sec:hypothesis_test}
 Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{$\chi^2$ test for non-uniform base distribution}
+\subsubsection{$\chi^2$ test for non-uniform base distribution}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%
+% Priors for precision parameter
+%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Priors for precision parameter}
+\subsubsection{Improper Prior}
+Playing no prior on M is an implicit improper prior.
+
+\subsubsection{Jeffreys Prior}
+Jeffreys Prior:Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
+\begin{equation}\label{eqn:JefferyPrior}
+\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
+\end{equation}
+
+\subsubsection{Log-normal Prior}
+In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
+Incorporate lognormal prior on $M_{j}$.
+
 %%%%%%%%%%%%%%%%%%
-% Data Sets
+% Results
 %%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms.
-Synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
-HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
-
-To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
-The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
-
-\subsection{HCC1187 Sequence Data}
-
-The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
-The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
-The estimated tumor purity was reported to be 0.8.
-A matched normal cells were derived from lymphoblastoid cells from peripheral blood.
-Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
-The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
-The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads.
-We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
-% TODO:cite the technical report. %Not sure the author is correct
-% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
-
 \section{Results}
 
-We tested RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
-The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
-We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
-
-We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
-
-To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
-loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
-homozygous somatic mutation).
-The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$.
-The size of the test is $\alpha=0.05$.
-%TODO: correct notation from \Pr to summation because we have samples, not distributions.
-
+%%%%%%%%%%%%%%%%%%
+% Sensitivity analysis of priors
+%%%%%%%%%%%%%%%%%%
 \subsection{Sensitivity analysis of priors}
-In order to examine the robustness of RVD3 to changes of different priors, we...
+In order to examine the robustness of RVD3 to changes of different priors, we perform information prior (Log-normal) and noninformation prior (Jeffreys).
 
 We found that RVD3 is robust to changing the priors.
-The RVD3 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:M_loc_rvd29} and Figure ~\ref{fig:M_loc_rvd30}.
+The RVD3 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:M_jeffrey} and Figure ~\ref{fig:M_lognormal}.
 
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/M_loc_rvd29.pdf}
+\includegraphics[width=120mm]{figs/M_jeffrey.pdf}
 \caption{Key parameters for RVD3 model with Jeffreys prior for synthetic DNA data sets.}
-\label{fig:M_loc_rvd29}
+\label{fig:M_jeffrey}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/M_loc_rvd29.pdf}
+\includegraphics[width=120mm]{figs/M_lognormal.pdf}
 \caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
-\label{fig:M_loc_rvd30}
+\label{fig:M_lognormal}
 \end{center}
 \end{figure}
-% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.
 
-The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
+The right column of Figure~\ref{fig:M_jeffrey} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
 $M_j$ measures the variance between replicates at location $j$.
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Results of priors on synthetic data
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Results of priors on synthetic data}
-Performance with read depth:
+\subsubsection{Performance with read depth}
 We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
@@
 Adding the $\chi^2$ improves specificity at the expense of sensitivity.
-Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
+Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
 Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
@@
 The performance improves monotonically with read depth.
-Figures~\ref{fig:ROC_29}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
+Figures~\ref{fig:ROC_jeffrey}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
 
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/ROC_without_chi2_rvd29.pdf}
+\includegraphics[width=120mm]{figs/ROC_without_chi2_jeffrey.pdf}
 \caption{ROC curve varying read depth showing detection performance.}
-\label{fig:ROC_29}
+\label{fig:ROC_jeffrey}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/ROC_without_chi2_rvd30.pdf}
+\includegraphics[width=120mm]{figs/ROC_without_chi2_lognormal.pdf}
 \caption{ROC curve varying read depth showing detection performance.}
-\label{fig:ROC_30}
+\label{fig:ROC_lognormal}
 \end{center}
@@
 
-Empirical performance compared with other algorithms:
-We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
-In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
-The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
-GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
-
-\subsubsection*{Sensitivity/Specificity Comparison}
+\subsubsection{Sensitivity/Specificity/FDR}
 
@@
 \begin{center}
-\includegraphics[width=160mm]{tables/comparison_table_ss.pdf}
+\includegraphics[width=160mm]{tables/Sen_Speci.pdf}
 \caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
-\label{tbl:table_29}
+\label{tbl:SS}
 \end{center}
@@
 
-Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
-They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
-VarScan2-somatic is able to call more mixed samples.
-However, as the read depth increases the specificity degrades.
-Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below.
-muTect has good performance across a wide range of MAF levels.
-But even at the highest depth only has around 0.5 sensitivity for low MAF levels.
-
-The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
-The sensitivity increases considerably with read depth at a slight expense to specificity.
-With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs.
-However, in practice one may not know the optimal $\tau^*$ a-priori.
-With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives.
-When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see Appendix~\ref{sec:app_comparison}).
-
-\subsubsection*{False Discovery Rate Comparison}
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, strelka and muTect. Blank cells indicate no positive calls were made.
-
-Samtools performs well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
-GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
-VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
-
 \begin{figure}[h]
 \begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_fdr.pdf}
-\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
-\label{tbl:comparison_fdr}
+\includegraphics[width=160mm]{tables/FDR_ACC_MCC.pdf}
+\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
+\label{tbl:FDR_ACC_MCC}
 \end{center}
 \end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 
-VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level.
-However, the FDR is high due to many false positives.
-Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth.
-Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF.
-However, it does not call any variants at or below 1\% MAF.
-MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
-But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.
-
-RVD2 has a lower FDR than other algorithms when the read depth is greater than $10\times$ the MAF with $N=1$ and $\tau$ set to the default value of zero or to the optimal value.
-The FDR is higher when $N=6$ because the variance of the control error rate distribution $P(\mu_j^{\text{control}} | r^{\text{control}})$ is smaller.
-The smaller variance yields improvements in sensitivity at the expense of more false positives.
-Since the FDR only considers positive calls, the performance degrades.
-
+Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 with different priors. Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
+Figure~\ref{tbl:FDR_ACC_MCC} shows the false discovery rate, accuracy, anc corelation coefficient of the RVD3 with different priors.
+Jeffreys prior shows a smaller false discovery rate and higher accuracy.
+Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropariate choice for the precision prior.
+Here for we chose Jeffreys prior because it needs more attention to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Results of Jeffreys prior on yeast data
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Results of Jeffreys prior on yeast data}
-
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
-The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
-
-Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
-Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor sample we identified five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
-Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
-Some of these mutations are also found to be common population SNPs according to dbSNPv138.
-The corresponding identities are shown in the figure.
-%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
-\label{fig:brca_MAF}
-\end{center}
-\end{figure}
-%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
 
@@
 
+%%%%%%%%%%%%
+% Discussion
+%%%%%%%%%%%%
 \section{Discussion}
-
-We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data.
-Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths and minor allele frequencies.
-
-Our inference algorithm uses Gibbs sampling to estimate the hierarchical empirical Bayes model.
-This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically.
-However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches.
-We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.
-
-We have focused on the statistical model and hypothesis test in this study and our results do not include any pre-filtration of erroneous reads or post-filtration of mutation calls beyond a simple quality score threshold.
-Incorporation of such data-cleaning steps will likely improve the accuracy of the algorithm.
-
-Our approach does not address identification of indels, structural variants or copy number variants.
-Those mutations typically require specific data analysis models and tests that are different than those for single-nucleotide variants.
-Furthermore, analysis of RNA-seq data or other data generated on the NGS platform may require different models that are more appropriately tuned to the particular noise feature of that data.
 
@@
 %%%%%%%%%%%%%%%%%%%
-\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-%TODO: discuss this result
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
-\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
-\label{fig:comparison_mcc}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 

### 2013-12-18 doc/ISMB2014/RVD3.tex
Update RVD3 Draft

@@
 \subsubsection{Improper Prior}
-Playing no prior on M is an implicit improper prior.
+Playing no prior on M is an implicit improper prior. So we coniderated about two typical priors: noninformation prior - Jeffreys prior, and information prior - log-normal prior.
 
 \subsubsection{Jeffreys Prior}
-Jeffreys Prior:Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
+Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
 \begin{equation}\label{eqn:JefferyPrior}
@@
 \subsection{Sensitivity analysis of priors}
-In order to examine the robustness of RVD3 to changes of different priors, we perform information prior (Log-normal) and noninformation prior (Jeffreys).
+In order to examine the robustness of RVD3 to changes of different priors, we performed information prior (Log-normal) and noninformation prior (Jeffreys).
 
@@
 
-The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+The left column of Figure~\ref{fig:M_jeffrey} and Figure~\ref{fig:M_lognormal} show the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
 Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
@@
 This is due to the parameters imposed at the alignment stage.
-The reads are 36bp long and we required no more than 2 mismatches.
-Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
 
@@
 
-The right column of Figure~\ref{fig:M_jeffrey} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
+The right column of Figure~\ref{fig:M_jeffrey} and Figure~\ref{fig:M_lognormal} show the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
 $M_j$ measures the variance between replicates at location $j$.
 There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
-
-The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-We see that the variation between replicates is smaller than the variation between location.
-$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
-Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
+Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] the M value drops at the mutant positions when the data is 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}].
+Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
+$M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
 
@@
 \subsubsection{Performance with read depth}
-We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
-For these ROC curves, we used the Bayesian test without the $\chi^2$ test.
-Adding the $\chi^2$ improves specificity at the expense of sensitivity.
-Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
-Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
-At the lowest depth the sensitivity and specificity is no better than random.
-However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43.
-The performance improves monotonically with read depth.
-Figures~\ref{fig:ROC_jeffrey}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
+To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
+Here the Bayesian test is used without the $\chi^2$ test.
+Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
+The performance improves when the read depth goes up.
+Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model without priors for $M_J$, which definitely demonstrates the advantage of priors.
 
@@
 
-\subsubsection{Sensitivity/Specificity/FDR}
-
 \begin{figure}[h]
@@
 \begin{center}
-\includegraphics[width=160mm]{tables/FDR_ACC_MCC.pdf}
+\includegraphics[width=160mm]{tables/FDR.pdf}
 \caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
-\label{tbl:FDR_ACC_MCC}
+\label{tbl:FDR}
 \end{center}
@@
 
-Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 with different priors. Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{tbl:FDR_ACC_MCC} shows the false discovery rate, accuracy, anc corelation coefficient of the RVD3 with different priors.
-Jeffreys prior shows a smaller false discovery rate and higher accuracy.
+\subsubsection{Sensitivity/Specificity/FDR}
+
+Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model without priors.
+Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
+Figure~\ref{tbl:FDR} shows the false discovery rate of the RVD3 with different priors.
+Jeffreys prior shows a smaller false discovery rate and higher accuracy than others.
+It is obvious no matter which pirors, the variant detection performance acquires lower FDR than the model without priors.
 Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropariate choice for the precision prior.
-Here for we chose Jeffreys prior because it needs more attention to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+Here we chose Jeffreys prior because it needs more attention to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The original research describing this sample used Strelka to identify mutations in the same sample.
-They identified chr7:154760439 as variant, but did not call the other four variants.
-In particular strelka missed the two LOH events.
-% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample.
-Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
+We demonstrated our RVD3 model with Jeffreys piror on yeast data to identify the variants.
 

### 2013-12-19 doc/ISMB2014/RVD3.tex
Update draft

@@
 
-\title[RVD3]{RVD3: Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data}
+\title[RVD3]{Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data}
 \author{}
@@
 \begin{abstract}
-Sensitivity analysis Bayesian model
-Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
+
+We analyzed the Sensitivity of the Bayesian model to different priors. Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
 
@@
 Now the definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
-Three hyperparameters are:  $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision; $\mu_0$, the global error rate expected across all positions; $M_0$, the global position which estimates the variation in the error rate across the positioins.
-$M_j$, the local precision measures the alteration in the error rate across replicates at position $j$.
+Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positioins; priors for $M_j$, two kinds of priors are log-normal prior and Jeffreys prior, which enhanced the former RVD model [XXX].
+Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$.
+The graphical chart for RVD3 are shown in Figure~\ref{fig:graphical_model}.
 RVD3 hierarchically includes four levels of samplings: $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
@@
 The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-Two kinds of priors for $M_{j}$ are Log-normal prior and Jeffreys prior, which enhanced the former RVD model [XX].
-$ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ is the model for Log-normal prior, and $pi\left({M}_{j}\right)$ for Jeffreys prior is based on Fisher information.
-The graphical chart for RVD3 with different priors are shown in Figure~\ref{fig:graphical_model}.
+And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and $\pi\left({M}_{j}\right)$ (Jeffreys prior).
 
@@
 \begin{center}
-\includegraphics[width=40mm]{figs/RVD3_model.pdf}
+\includegraphics[width=120mm]{figs/RVD3_model.pdf}
 \caption{RVD3 Graphical Model}
@@
 \subsection{Inference and Hypothesis Testing}
-
-The primary object of inference in this model is the joint posterior distribution function over the latent variables,
-\begin{equation}
-	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
-\end{equation}
-where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.
-
-The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution.
-However, there is not a closed form for the product of a Beta distribution with another Beta distribution, so exact inference is intractable.
-
-Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
-First, the hyperparameters are initialized using method-of-moments (MoM).
-Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule.
-Finally, we sample from the marginal posterior distribution for $\theta_{ji}$ given its Markov blanket.
-Samples from $\theta_{ji}$ can be drawn from the posterior distribution directly  because the prior and likelihood form a conjugate pair.
-This sampling procedure is repeated until the chain converges to a stationary distribution then we draw samples from the posterior distribution over latent variables.
+Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are illustrated.
 
@@
 \Repeat
-\For {each location j} \Comment{Sample $\mu_j$}
-  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
+\For {each location j}
+  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H \Comment{Sample $\mu_j$}
   \State Set $\mu_j$ to the sample median for the T samples
+  \State Draw T samples from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$ using M--H \Comment{Sample $M_j$}
 
-  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
-	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
+  \For {each replicate i} 
+	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$  \Comment{Sample $\theta_{ji}$}
   \EndFor
@@
 Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from the posterior distribution.
-The acceptance probability is then
-\begin{equation}
-	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
-\end{equation}
 
@@
 We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
-\begin{equation}
-	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
-\end{equation}
-
-While we are able to compute the optimal $\tau$ threshold for a test data set, in general we would not have access to $\tau^*$.
-With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest.
-Absent this information we set $\tau = 0$.
+We set $\tau = 0$ in our experiment.
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsubsection{$\chi^2$ test for non-uniform base distribution}
-
-An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios.
-We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
-In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
-
-We use a $\chi^2$ goodness-of-fit test on a multinomial distribution over the non-reference bases to distinguish these two possible scenarios.
-The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
-Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
-The test statistic is
-
-\begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
-\end{equation}
-
-where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-\citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
-
-We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,
-
-\begin{equation}\label{eqn:fisher_combined}
-	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
-\end{equation}
-
-Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
+% Limited in the 9 pages, so maybe considering neglect the Chi^2 Test since we show the results without Chi^2 Test.
 
@@
 In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
-Incorporate lognormal prior on $M_{j}$.
+Incorporate log-normal prior on $M_{j}$.
 
@@
 In order to examine the robustness of RVD3 to changes of different priors, we performed information prior (Log-normal) and noninformation prior (Jeffreys).
+We found that RVD3 is robust to changing the priors.
+The RVD3 algorithm provides estimates of model parameters and latent variables given the data.
 
-We found that RVD3 is robust to changing the priors.
-The RVD3 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:M_jeffrey} and Figure ~\ref{fig:M_lognormal}.
-
-The left column of Figure~\ref{fig:M_jeffrey} and Figure~\ref{fig:M_lognormal} show the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+We show several of these parameters in Figure~\ref{fig:M_jeffrey} and Figure ~\ref{fig:M_lognormal}.
+The left column of these two figures show the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
 Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
@@
 There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] the M value drops at the mutant positions when the data is 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}].
+Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}].
 Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
 $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=170mm]{figs/prior_post_dilu_1.pdf}
+\caption{Distribution of priors and posteriors when dilution is 1\%}
+\label{fig:dilu_1}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=170mm]{figs/prior_post_dilu_10.pdf}
+\caption{Distribution of priors and posteriors when dilution is 10\%}
+\label{fig:dilu_10}
+\end{center}
+\end{figure}
+
+Statistically, We analyzed the posteriors and priors for Jeffreys prior and log-normal prior.
+Figure~\ref{fig:dilu_1} and Figure~\ref{fig:dilu_10} show the distribution over different M values focused on 4 different locations (position 100 and 300 are mutant, and position 104 and 244 are non-mutant).
+The positions are chosen in the middle of the dase length.
+From the posterior distributions, non-mutant positions display normal and stable without a peak nor a strange shape.
+The Jeffreys' prior shows more information than the lognormal prior (flat) from the posterior curve.
+From the prior distribution, it shows our model wants to search for a small value for M.
 
@@
 \subsubsection{Performance with read depth}
-To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
-Here the Bayesian test is used without the $\chi^2$ test.
-Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
-The performance improves when the read depth goes up.
-Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model without priors for $M_J$, which definitely demonstrates the advantage of priors.
 
@@
 \includegraphics[width=120mm]{figs/ROC_without_chi2_jeffrey.pdf}
-\caption{ROC curve varying read depth showing detection performance.}
+\caption{ROC curve for variants detection performance by Jeffreys prior.}
 \label{fig:ROC_jeffrey}
@@
 \includegraphics[width=120mm]{figs/ROC_without_chi2_lognormal.pdf}
-\caption{ROC curve varying read depth showing detection performance.}
+\caption{ROC curve for variants detection performance by log-normal prior.}
 \label{fig:ROC_lognormal}
@@
 \end{figure}
+
+To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
+Here the Bayesian test is used without the $\chi^2$ test.
+Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
+The performance improves when the read depth goes up.
+(ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
+Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model without priors for $M_J$, which definitely demonstrates the advantage of priors.
 
@@
 \includegraphics[width=160mm]{tables/FDR.pdf}
-\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
+\caption{False Discovery Rate comparison of RVD3 with different priors.}
 \label{tbl:FDR}

### 2013-12-30 doc/ISMB2014/RVD3.tex
Paper draft ISMB2014.

@@
 
-Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
-
-To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
-The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
+Two 400bp DNA sequences differ only 14 single nucleotide positions.
+Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
+The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+We used BWA v0.7.5a to align the reads to the reference sequence with the -C50 option to remove the reads of high mapping quality.
+BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104.
+The final data set contains read pairs for $N=6$ replicates for the control at different MAF levels.
 
 \subsection{Yeast Data}
+We first mapped the GSY1135 to Chromosome 10 in S288c reference genome by BWA v0.7.5a [XXX].
+Then called SNPs by GATK UnifiedGenotyper and created a FASTA GSY1135 reference using GATK FastaAlternative [XXX].
+Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from [XXX], and removed WT population using FASTX Barcode Splitter [XXX] and trimed the Nextera tag by Cutadapt v1.3.
+Finally, the FASTQ files of case and control were mapped to GSY1135 reference and pileup files were generated using Samtools v0.1.19.
 
@@
 Now the definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
-Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positioins; priors for $M_j$, two kinds of priors are log-normal prior and Jeffreys prior, which enhanced the former RVD model [XXX].
+Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions; priors for $M_j$, two kinds of priors are log-normal prior and Jeffreys prior, which enhanced the former RVD model [XXX].
 Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$.
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/RVD3_model.pdf}
+\includegraphics[width=100mm]{figs/RVD3_model.pdf}
 \caption{RVD3 Graphical Model}
@@
 \subsection{Inference and Hypothesis Testing}
-Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are illustrated.
 
 \begin{algorithm}[ht]
-\caption{Metropolis-within-Gibbs Algorithm}
+\caption{Inference process for Metropolis-within-Gibbs}
 \label{alg:metro_gibbs}
@@
 
-\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
+\State Initialize $\theta$, $\mu$, $M_j$, $\mu_0$, $M_0$
 \Repeat
@@
 \end{algorithm}
+
+Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are also illustrated.
 
@@
 
-We resample from the proposal if the sample is outside of the support of the posterior distribution.
+We resampled from the proposal if the sample is outside of the support of the posterior distribution.
 We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
@@
 \subsubsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
-For Jeffreys prior,since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by
+For Jeffreys prior is from the Fisher information and ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$,
 
@@
 
-Now to calculate the equations to acquire the Fisher information, we assume there is only one replicate,
+We calculated the equations (Appendix B) and obtained the Jeffreys' prior for $M_j$:
 
-\begin{equation}\label{eqn:Betapdf}
-p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
-\end{equation}
-
-\begin{equation}\label{equ:JefferyInference1}
-\begin{split}
-\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
-& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
-\end{split}
-\end{equation}
-
+% \pi\left({M}_{j}\right) =
 \begin{equation}
-\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
-\end{equation}
-
-\begin{equation}
-\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
-\end{equation}
-
-Now we have the Jeffreys' prior for $M_{j}$:
-
-\begin{equation}
-\pi\left({M}_{j}\right) = [-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
+[-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
 \end{equation}
@@
 We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$.
-Since we cannot give an analytical form for this, so we sample from the posterior distribution using the Metropolis-Hastings algorithm.
+Since we cannot provide an intepretive form for it, so Metropolis-Hastings algorithm was used to sample from the posterior distribution.
 
@@
 \subsubsection{Posterior Density Test}\label{sec:hypothesis_test}
-Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
-For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
+Posterior distributions of $\mu_j$ for the control and case are achieved-  $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$, by Metropolis-within-Gibbs.
+So we called a variant when $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with $1-\alpha$ confidence,
 \begin{equation}\label{eqn:bayes_test}
@@
 \end{equation}
-where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
-We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-We set $\tau = 0$ in our experiment.
+where $\tau$ is a detection threshold and $1-\alpha$ is the confidence level. We set $\tau = 0$ in our experiment [XXX].
 
@@
 \subsection{Priors for precision parameter}
-\subsubsection{Improper Prior}
-Playing no prior on M is an implicit improper prior. So we coniderated about two typical priors: noninformation prior - Jeffreys prior, and information prior - log-normal prior.
+%\subsubsection{Improper Prior}
+Playing no prior on $ M_{j}$ is an implicit improper prior.
+An improper prior integrates to infinity, and may cause an improper posterior which results in an invalid inferences [XXX].
+Because an improper prior is a bad choice since a step has to be taken to confirm the posterior is proper.
+We considered about two typical priors: non-information prior - Jeffreys prior, and information prior - log-normal prior.
 
-\subsubsection{Jeffreys Prior}
-Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
-\begin{equation}\label{eqn:JefferyPrior}
-\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
-\end{equation}
-
-\subsubsection{Log-normal Prior}
-In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
-Incorporate log-normal prior on $M_{j}$.
+%\subsubsection{Jeffreys Prior}
+Jeffreys prior is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys (1946).
+It is defined in terms of the Fisher information and works well with a single parameter.
+In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
+Log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
+So log-normal prior was incorporated in the model.
+The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
 
@@
 \subsection{Sensitivity analysis of priors}
-In order to examine the robustness of RVD3 to changes of different priors, we performed information prior (Log-normal) and noninformation prior (Jeffreys).
+In order to examine the robustness of RVD3 to changes of different priors, we performed information prior (log-normal) and non-information prior (Jeffreys).
 We found that RVD3 is robust to changing the priors.
@@
 
-We show several of these parameters in Figure~\ref{fig:M_jeffrey} and Figure ~\ref{fig:M_lognormal}.
-The left column of these two figures show the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+We show several of these parameters of the model with Jeffreys prior in Figure~\ref{fig:M_jeffrey}.
+The left column of is shows the read depth for each of the six BAM files (three replicates each with two read pairs) for each data set.
 Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
@@
 This is due to the parameters imposed at the alignment stage.
+The right column of Figure~\ref{fig:M_jeffrey} reveals the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
+$M_j$ measures the variance between replicates at location $j$.
+There is little variability across positions indicating that the replication variance does not change greatly across position.
+Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix C.
+Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
+$M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
 
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/M_lognormal.pdf}
-\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
-\label{fig:M_lognormal}
-\end{center}
-\end{figure}
-
-The right column of Figure~\ref{fig:M_jeffrey} and Figure~\ref{fig:M_lognormal} show the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}].
-Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-$M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=170mm]{figs/prior_post_dilu_1.pdf}
-\caption{Distribution of priors and posteriors when dilution is 1\%}
-\label{fig:dilu_1}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=170mm]{figs/prior_post_dilu_10.pdf}
+\includegraphics[width=170mm]{figs/post_prior_dilu_10.pdf}
 \caption{Distribution of priors and posteriors when dilution is 10\%}
@@
 Statistically, We analyzed the posteriors and priors for Jeffreys prior and log-normal prior.
-Figure~\ref{fig:dilu_1} and Figure~\ref{fig:dilu_10} show the distribution over different M values focused on 4 different locations (position 100 and 300 are mutant, and position 104 and 244 are non-mutant).
-The positions are chosen in the middle of the dase length.
-From the posterior distributions, non-mutant positions display normal and stable without a peak nor a strange shape.
+Figure~\ref{fig:dilu_10} shows the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
+The positions are chosen in the middle of the base length- position 100 and 300 are mutant, and position 104 and 244 are non-mutant.
+From the posterior probability distributions are estimated by Gaussian Kernel Density [XXX].
+From the figure non-mutant positions display normal and stable without a peak nor a strange shape.
 The Jeffreys' prior shows more information than the lognormal prior (flat) from the posterior curve.
@@
 
-To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
-Here the Bayesian test is used without the $\chi^2$ test.
-Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
-The performance improves when the read depth goes up.
-(ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
-Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model without priors for $M_J$, which definitely demonstrates the advantage of priors.
-
-\begin{figure}[h]
+\begin{figure}[htbp]
 \begin{center}
@@
 
-\begin{figure}[h]
+\begin{figure}[htbp]
 \begin{center}
@@
 
+To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
+Here the Bayesian test is used without the $\chi^2$ test.
+Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
+The performance improves when the read depth goes up.
+(ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
+Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model with improper priors for $M_j$, which definitely demonstrates the advantage of priors.
+
 \subsubsection{Sensitivity/Specificity/FDR}
 
-Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model without priors.
+Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
 Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
@@
 Jeffreys prior shows a smaller false discovery rate and higher accuracy than others.
-It is obvious no matter which pirors, the variant detection performance acquires lower FDR than the model without priors.
-Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropariate choice for the precision prior.
-Here we chose Jeffreys prior because it needs more attention to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+It is obvious no matter Jeffreys or log-normal, the variant detection performance acquires lower FDR than the model with improper priors.
+Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
+Here we chose Jeffreys prior model because it's a non-information prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We demonstrated our RVD3 model with Jeffreys piror on yeast data to identify the variants.
+We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants.
 
@@
 %%%%%%%%%%%%%%%%%%%
-% Appendix B: Comparison Statistics
+% Appendix B: Inference of Jeffreys prior
 %%%%%%%%%%%%%%%%%%%
+\section{Inference of Jeffreys prior}\label{sec:appendix_mom}
+We assume there is only one replicate,
+
+\begin{equation}\label{eqn:Betapdf}
+p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
+\end{equation}
+
+\begin{equation}\label{equ:JefferyInference1}
+\begin{split}
+\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
+& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
+\end{split}
+\end{equation}
+
+\begin{equation}
+\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
+\end{equation}
+
+\begin{equation}
+\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
+\end{equation}
+
+Now we have the Jeffreys' prior for $M_{j}$:
+
+\begin{equation}
+[-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
+\end{equation}
+
+%%%%%%%%%%%%%%%%%%%
+% Appendix C: Key parameters for RVD3 model with log-normal prior
+%%%%%%%%%%%%%%%%%%%
+\section{Key parameters for RVD3 model with log-normal prior}\label{sec:appendix_mom}
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=120mm]{figs/M_lognormal.pdf}
+\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
+\label{fig:M_lognormal}
+\end{center}
+\end{figure}
 

### 2013-12-30 results/2013-12-20_Yeast_Data_Processing/GenomeAnalysisTK-2.5-2-gf57256b/resources/Pileup.java
Yeast data processing.

@@ -1,0 +1,174 @@
+/*
+* Copyright (c) 2012 The Broad Institute
+* 
+* Permission is hereby granted, free of charge, to any person
+* obtaining a copy of this software and associated documentation
+* files (the "Software"), to deal in the Software without
+* restriction, including without limitation the rights to use,
+* copy, modify, merge, publish, distribute, sublicense, and/or sell
+* copies of the Software, and to permit persons to whom the
+* Software is furnished to do so, subject to the following
+* conditions:
+* 
+* The above copyright notice and this permission notice shall be
+* included in all copies or substantial portions of the Software.
+* 
+* THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+* EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES
+* OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+* NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT
+* HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY,
+* WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
+* FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR
+* THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+*/
+
+package org.broadinstitute.sting.gatk.walkers.qc;
+
+import org.broad.tribble.Feature;
+import org.broadinstitute.sting.commandline.Argument;
+import org.broadinstitute.sting.commandline.Input;
+import org.broadinstitute.sting.commandline.Output;
+import org.broadinstitute.sting.commandline.RodBinding;
+import org.broadinstitute.sting.gatk.CommandLineGATK;
+import org.broadinstitute.sting.gatk.contexts.AlignmentContext;
+import org.broadinstitute.sting.gatk.contexts.ReferenceContext;
+import org.broadinstitute.sting.gatk.refdata.RefMetaDataTracker;
+import org.broadinstitute.sting.gatk.walkers.LocusWalker;
+import org.broadinstitute.sting.gatk.walkers.NanoSchedulable;
+import org.broadinstitute.sting.gatk.walkers.TreeReducible;
+import org.broadinstitute.sting.utils.Utils;
[…]

### 2013-12-30 results/2013-12-20_Yeast_Data_Processing/GenomeAnalysisTK-2.5-2-gf57256b/resources/PrintReads.java

@@ -1,0 +1,284 @@
+/*
+* Copyright (c) 2012 The Broad Institute
+* 
+* Permission is hereby granted, free of charge, to any person
+* obtaining a copy of this software and associated documentation
+* files (the "Software"), to deal in the Software without
+* restriction, including without limitation the rights to use,
+* copy, modify, merge, publish, distribute, sublicense, and/or sell
+* copies of the Software, and to permit persons to whom the
+* Software is furnished to do so, subject to the following
+* conditions:
+* 
+* The above copyright notice and this permission notice shall be
+* included in all copies or substantial portions of the Software.
+* 
+* THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+* EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES
+* OF MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+* NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT
+* HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY,
+* WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
+* FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR
+* THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+*/
+
+package org.broadinstitute.sting.gatk.walkers.readutils;
+
+import net.sf.samtools.SAMFileWriter;
+import net.sf.samtools.SAMReadGroupRecord;
+import org.broadinstitute.sting.commandline.Argument;
+import org.broadinstitute.sting.commandline.Hidden;
+import org.broadinstitute.sting.commandline.Output;
+import org.broadinstitute.sting.gatk.CommandLineGATK;
+import org.broadinstitute.sting.gatk.GenomeAnalysisEngine;
+import org.broadinstitute.sting.gatk.contexts.ReferenceContext;
+import org.broadinstitute.sting.gatk.io.StingSAMFileWriter;
+import org.broadinstitute.sting.gatk.iterators.ReadTransformer;
+import org.broadinstitute.sting.gatk.iterators.ReadTransformersMode;
+import org.broadinstitute.sting.gatk.refdata.RefMetaDataTracker;
+import org.broadinstitute.sting.gatk.walkers.*;
[…]

### 2013-12-31 doc/ISMB2014/RVD3.tex
Updated draft and notebook.

@@
 \section{Introduction}
+
+Massively parallel sequencing data were generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
+One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants(SNV) candidates.
+Currently SNV are deduced on NGS platform by variants calling algorithms.
+Typically, SAMtools[XX], Genome Analysis Toolkit (GATK) [XX], VarScan2[XX], and SPLINTER[XX] are all highly used in detecting low-frequency variants.
+However identifying true SNVs remains challenging because of high false positive rates or high false discovery rate.
+Another statistical method-RVD, was proposed based on beta-binomial model for ultrasensitive rare single nucleotide variant detection.
+So based on the RVD hierarchical model we built an improved robustness and accuracy model for low-depth variants detection.
 

### 2014-01-07 src/python/rvd27/rvd27.py
Set Qsd constant in Gibbs sampling

@@ -138,8 +138,8 @@ def test(controlHDF5Name, caseHDF5Name, T=0, N=1000, outputFile=None):
     
     # Sample from the posterior Z = muCase - muControl
     # Adjusting for baseline error rate for case and control
-    #(Z, caseMuS, controlMuS) = sample_post_diff(caseMu-casePhi['mu0'], controlMu-controlPhi['mu0'], N)
-    (Z, caseMuS, controlMuS) = sample_post_diff(caseMu, controlMu, N)
+    (Z, caseMuS, controlMuS) = sample_post_diff(caseMu-casePhi['mu0'], controlMu-controlPhi['mu0'], N)
+##    (Z, caseMuS, controlMuS) = sample_post_diff(caseMu, controlMu, N)
 
     # Posterior Prob that muCase is greater than muControl by T
     postP = bayes_test(Z, [(T, np.inf)]) 
@@ -405,7 +405,7 @@ def sampleLocMuMH(args):
         mu = mu_p
     return mu
     
-def sampleMuMH(theta, mu0, M0, M, mu=ss.beta.rvs(1, 1), burnin=0, mh_nsample=1, thin=0, pool=None):
+def sampleMuMH(theta, mu0, M0, M, mu=ss.beta.rvs(1, 1), Qsd=0.1, burnin=0, mh_nsample=1, thin=0, pool=None):
     """ Return a sample of mu with parameters mu0 and M0.
     """
     if np.ndim(theta) == 1: (N, J) = (1, np.shape(theta)[0])
@@ -414,23 +414,6 @@ def sampleMuMH(theta, mu0, M0, M, mu=ss.beta.rvs(1, 1), burnin=0, mh_nsample=1, 
     alpha0 = mu0*M0 + np.finfo(np.float).eps
     beta0 = (1-mu0)*M0 + np.finfo(np.float).eps
 
-    # set Qsd as the higher value between mu[j]/10 and 1.0E-4 for each position
-    def bound(mu):
-        bound = 0.001
-        if bound < mu < 1-bound:
-            return mu*(1-mu)
-        else:
-            return bound
-        
-    Qsd = map(bound, mu)       
-
-##    bound = 1.0E-4*np.ones_like(mu)
-##    Qsd = [max(bound[i],0.1*mu[i]) for i in range(np.shape(mu)[0])]
-
-    # rvd272
-    # Qsd = 0.1*np.ones_like(mu)
-    # rvd273
-    # Qsd = 0.01*np.ones_like(mu)
[…]

### 2014-01-07 doc/bioinformatics2014/document.tex
Bioinformatics2014 folder for rvd27

--- 2013-12-08 doc/bioinformatics2013/RVD2.tex
@@
-\documentclass[11pt,reqno]{amsart}
-\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-%\usepackage{algorithm, algorithmic}
+\documentclass{bioinfo}
+\copyrightyear{2005}
+\pubyear{2005}
+\usepackage{amsmath}
+\usepackage{afterpage}
+\usepackage{algorithm}
+\usepackage{bbm}
 
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
-\usepackage{graphicx}
-
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-\usepackage{enumitem}
-
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{natbib}
-
-%\usepackage{epstopdf}
-%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
+%\usepackage{placeins}
+\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{xr}
+\externaldocument{supplementary_info}
+\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}
 
@@
 \DeclareMathOperator{\diag}{diag}
-
-\algnewcommand{\Inputs}[1]{%
-  \State \textbf{Inputs:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
-\algnewcommand{\Initialize}[1]{%
-  \State \textbf{Initialize:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
+\begin{document}
+\firstpage{1}
 
 \title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
-\author{}
-%\date{}                                           % Activate to display a given date or no date
+\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{to whom correspondence should be addressed}}
+\address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}
 
-\begin{document}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
+\maketitle
 
 \begin{abstract}
+
+\section{Motivation:}
 Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
@@
 However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
+
+\section{Results:}
 We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+We identify six mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
 
-%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
-%We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
+\section{Availability:}
+Text
+
+\section{Contact:} \href{pjflaherty@wpi.edu}{pjflaherty}
 \end{abstract}
-
-\maketitle
 
@@
 However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
-Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
-Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{}.
-GATK involves more sophisticate pre and post-processing steps.
-The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
+Samtools/bcftools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}.
+GATK involves more sophisticate pre- and post-processing steps wherein the genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
 
@@
 muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.
+RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{}.
+However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.
 
 Several studies have compared the relative performance of these algorithms.
-%TODO: [DISCUSS \citet{Stead:2013fu}] and others
+\citet{spencer2013performance} demonstrated  that VarScan-somatic performed the best comparing with SAMtools, GATK and SPLINTER in detecting minor allele frequencies (MAFs) of 1\% to 8\%, with $ > $500× coverage required for optimal performance.
+However, \citet{spencer2013performance} also highlighted the fact that VarScan2 yielded more false positives at high read depth.
+\citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
+We test their performance on much lower MAF variants and across a wider range of coverage depths.
 
-Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
-These steps may improve algorithm performance, but their presence makes it difficult to isolate what step is most responsible for the performance.
-Instead, we focus exclusively on the model structure and statistical inference steps.
-Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study.
-We have isolated the statistical inference and hypothesis testing steps so that they may be used independently or as part of a larger variant calling pipeline.
-
-We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}.
-In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
-We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
-Finally, in Section~\ref{sec:brca} we present results on detecting mutations in the PAXIP1 gene from matched tumor-normal data from the HCC1187 cell line.
+The remainder of this article is organized as follows.
+In the next section we describe the statistical model structure of our new algorithm, RVD2.
+Then, we derive a sampling algorithm for computing the posterior distribution over latent variables in the model and use those samples in a Bayesian posterior distribution hypothesis test to call variants.
+We compare the performance of RVD2 to several other variant calling algorithms for a range of read depths and minor allele fractions.
+Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.
 
@@
 
-The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
+The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters can be factorized as
 
-\begin{equation}\label{eqn:jointpdf}
-p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
-\end{equation}
+\begin{multline}\label{eqn:jointpdf}
+p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) \\ = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
+\end{multline}
 
 where
+
 \begin{align}
-p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
-p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
-p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
+  \begin{split}
+  p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \cdot \\
+  &\quad \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
+  \end{split}\\
+ %
+  \begin{split}
+  p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \cdot \\
+  &\quad \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
+  \end{split}\\
+ %
+  \begin{split}
+  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \\
+  &\quad \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber\\
+  \end{split}\\
 \end{align}
@@
 Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
-\begin{equation}
-p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
+\begin{multline}
+p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
+\end{multline}
 
@@
 
-\begin{equation}
-\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
+\begin{multline}
+\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
+\end{multline}
 
@@
   \State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
+
   \Repeat
-\For {each location j} \Comment{Sample $\mu_j$}
+
+    \For {each location j}
+
     \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H 
+   % \Comment{Sample $\mu_j$}
+
     \State Set $\mu_j$ to the sample median for the T samples
 
-  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
+      \For {each replicate i}
+
 	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
+	%\Comment{Sample $\theta_{ji}$}
+
       \EndFor
@@
     \EndFor
+
   \Until {sample size sufficient}
+
   \end{algorithmic}
@@
 
-We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are 
+We present the initial parameter estimates here and provide the derivations in Supplementary Information. The MoM estimate for replicate-level parameters are
 $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
@@
 $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
-While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
+%While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
 
@@
 
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
+We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \tilde{\mu}_j\cdot (1-\tilde{\mu}_j)$ if $\tilde{\mu}_j \in (10^{-3},1-10^{-3})$ and $\sigma_j = 10^{-4}$ otherwise, where $\tilde{\mu}_j$ is the MoM estimator of $\mu_j$.
 Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
@@
 %%%%%%%%%%%%%%%%%%
-% Posterior Density Test
+% Posterior Distribution Test
 %%%%%%%%%%%%%%%%%%
-\subsection{Posterior Density Test}\label{sec:hypothesis_test}
+\subsection{Posterior Distribution Test}\label{sec:hypothesis_test}
 Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
-For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
+For notational simplicity, we define the random variables associated with these two distributions $\mu_j^{\text{case}}$ and $\mu_j^{\text{control}}$ and the associated samples as $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
 
-A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
+A variant is called if $\mu_j^{\text{case}} > \mu_j^{\text{control}}$ with high confidence,
 \begin{equation}\label{eqn:bayes_test}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
+	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{case}} - \tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha,
 \end{equation}
@@
 The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
+The optimal $\tau$ maximizes the Matthews Correlation Coefficient (MCC) to perfect classification in the ROC curve plot,
 \begin{equation}
-	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
+	\tau^* = \arg\max_\tau \left\{ \text{MCC}(\tau)\right\}.
 \end{equation}
@@
 where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-\citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
+\citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available; we choose that value.
 
-We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,
+We control for multiple hypothesis testing in two ways.
+We use Fisher's combined probability test \citep{fisher1970statistical} to combine the p-values for $N$ replicates into a single p-value at position $j$,
 
@@
 Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
+We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
 
@@
 Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+Sample of the case and control DNA were mixed at defined fractions to yield defined MAFs of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
 More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
@@
 The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
-The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
+The HCC1187 cell line was derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
 The estimated tumor purity was reported to be 0.8.
-A matched normal cells were derived from lymphoblastoid cells from peripheral blood.
+Matched normal cells were derived from lymphoblastoid cells from peripheral blood.
 Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
@@
 We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
+To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j^{\text{control}} \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.2]$, heterozygous mutant $[0.1, 0.9]$, and homozygous mutant $[0.8, 1.0]$ and the size of the test is $\alpha=0.05$.
 
@@
 homozygous somatic mutation).
-The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$.
+The two hypothesis tests are then $\Pr( \mu_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\Delta} \leq \tau ) > 1-\alpha$.
 The size of the test is $\alpha=0.05$.
-%TODO: correct notation from \Pr to summation because we have samples, not distributions.
 
@@
 
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
+\caption{ROC curve varying read depth showing detection performance.}
+\label{fig:ROC}
+\end{center}
+\end{figure}
+
+% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
 We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
-For these ROC curves, we used the Bayesian test without the $\chi^2$ test.
+For these ROC curves, we used the posterior distribution test without the $\chi^2$ test.
 Adding the $\chi^2$ improves specificity at the expense of sensitivity.
@@
 Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance.}
-\label{fig:ROC}
-\end{center}
-\end{figure}
-% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
-
-%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
-All of the called positions show a clear difference between the case and control error rates.
-The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
-%
-%\begin{figure}[h]
-%\begin{center}
-%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
-%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-%\label{fig:MAF}
-%\end{center}
-%\end{figure}
-%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
 
@@
 
-\begin{figure}[h]
+\begin{figure*}[!tpb]
 \begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
+\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
 \caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
@@
 \end{center}
-\end{figure}
+\end{figure*}
+
 % from folder 'rvd2\results\2013-09-19_operating_characteristics'
@@
 They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
+GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
+Therefore, those algorithms are expected to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
 VarScan2-somatic is able to call more mixed samples.
@@
 
-Samtools performs well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
+Samtools performs well on 100\% MAF sample and performance improves for read depths 3,089 and 30,590.
 GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
@@
 
-\begin{figure}[h]
+\begin{figure*}[!tpb]
 \begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_fdr.pdf}
+\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
 \caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
@@
 \end{center}
-\end{figure}
+\end{figure*}
 % from folder 'rvd2\results\2013-09-19_operating_characteristics'
+%\afterpage{\clearpage}
 
@@
 
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
+RVD2 identified six variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+Of the six called variants, three were homozygous germline variants and two were heterozygous germline variants.
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
+The remaining one is a homozygous somatic variant.
 
+%\iffalse
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
+Positions chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF.
 Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
 Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor sample we identified five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
+In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T.
+Through manual checking, position chr7:154743899C$>$T and chr7:154753635T$>$C are likely false positives due to insufficient sampling in the Metropolis\_within-Gibbs algorithm.
 Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
@@
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
+%\fi
 
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\includegraphics[width=0.5\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
 \caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
@@
 The original research describing this sample used Strelka to identify mutations in the same sample.
-They identified chr7:154760439 as variant, but did not call the other four variants.
+They identified chr7:154760439 as variant, but did not call the other five variants.
 In particular strelka missed the two LOH events.
@@
 \section{Discussion}
-
 We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data.
@@
 
-Our inference algorithm uses Gibbs sampling to estimate the hierarchical empirical Bayes model.
+Our inference algorithm uses Gibbs sampling to estimate the RVD2 hierarchical empirical Bayes model.
 This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically.
 However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches.
-We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.
+We opted for this balance of speed and accuracy because computational time is often not limiting and the cost of a false positive or false negative greatly outweighs the cost of more computation.
 
@@
 
-\appendix
+The availability of clinical sequence data is increasing as the technical capability to sequence clinical samples at low cost improves.
+Consequently, we require statistically accurate algorithms that are able to call germline and somatic point mutations in heterogeneous samples with low purity.
+Such accurate algorithms are a step towards greater access to genomics for clinical diagnostics.
 
-%%%%%%%%%%%%%%%%%%%
-% Appendix A: Parameter Initialization
-%%%%%%%%%%%%%%%%%%%
-\section{Parameter Initialization}\label{sec:appendix_mom}
-Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
-Therefore the MoM estimator is
-\begin{equation}
-	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
-\end{equation}
+\section*{Acknowledgement}
+We would like to thank Fan Zhang for careful review of this manuscript and helpful comments.
 
-We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
-The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
-The first and second population moments are
-\begin{eqnarray}
-	E[\theta_{ji}] =& \mu_j,\\
-	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
-\end{eqnarray}
-The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
-Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
-\begin{eqnarray}
-	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
-	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
-\end{eqnarray}
+\paragraph{Funding\textcolon}
 
-Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
-\begin{eqnarray}
-	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
-	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
-\end{eqnarray}
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix B: Comparison Statistics
-%%%%%%%%%%%%%%%%%%%
-\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-%TODO: discuss this result
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
-\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
-\label{fig:comparison_mcc}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-
-%%%%%%%%%%%%%%%%%
-% Appendix C: Data Set Statistics
-%%%%%%%%%%%%%%%%%
-\section{RVD2 Estimated Parameters}\label{sec:synthetic_estimate}
-The RVD2 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:depthM}.
-
-The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The reads are 36bp long and we required no more than 2 mismatches.
-Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
-\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
-\label{fig:depthM}
-\end{center}
-\end{figure}
-
-% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.
-
-The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
-
-The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-We see that the variation between replicates is smaller than the variation between location.
-$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
-Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
-
-%%%%%%%%%%%%%%%%%
-% Appendix D: Parameter Settings for Variant Calling Algorithms
-%%%%%%%%%%%%%%%%%
-\section{Parameter settings for other variant calling algorithms}
-\paragraph{\textbf{Samtools}}We used samtools (v0.1.19) function mpileup to call variants and bcftools to save the result in standard VCF files.
-In mpileup, we set the -d option sufficiently high at $10^6$ to avoid truncating read deapth.
-Option -u was enabled to make sure the output bcf files were uncompressed.
-
-\paragraph{\textbf{GATK}}
-We used GATK (v2.1-8) UnifiedGenotyper function to detect mutations on our synthetic data following the recommended workflow.
-Due to some format incompatibility, we applied Picard to format read group and GATK for realignment.
-In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsampling coverage within GATK.
-
-\paragraph{\textbf{VarScan2-mpileup}}
-VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.We assigned parameter -C value 50 as the synthetic data was aligned using BWA and set -d at $10^6$.
-In mpileup2snp, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.
-
-\paragraph{\textbf{VarScan2-somatic}}
-We tested VarScan2 somatic on our synthetic dataset.
-The parameter --normal-purity set was at 1.00, --tumor-purity at the dilution rate.
-The parameter --min-var-freq was set at $10^{-5}$.
-We combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown) to compare with performance of RVD2.
-
-\paragraph{\textbf{Strelka and muTect}}
-Since configuration and Analysis for Strelka and muTect is standardized and no parameter needs to be specified, we installed these two programs and ran them on our data set separately.
-
-Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
-GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
-Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
-
-% TODO: modify the following section
-GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
-Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
-\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
-We find its performance on much lower MAF variants and across a wider range of coverage depths.
-\bibliographystyle{apalike}
+\bibliographystyle{natbib}
 \bibliography{bioinfo}

### 2014-01-07 doc/bioinformatics2014/document_syntax.tex
Bioinformatics2014 folder for rvd27

--- 2013-12-08 doc/bioinformatics2013/RVD2.tex
@@
-\documentclass[11pt,reqno]{amsart}
-\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-%\usepackage{algorithm, algorithmic}
+\documentclass{bioinfo}
+\copyrightyear{2005}
+\pubyear{2005}
 
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
-\usepackage{graphicx}
-
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-\usepackage{enumitem}
-
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{natbib}
-
-%\usepackage{epstopdf}
-%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
-\newcommand{\RR}{I\!\!R} %real numbers
-\DeclareMathOperator{\diag}{diag}
-
-\algnewcommand{\Inputs}[1]{%
-  \State \textbf{Inputs:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
-\algnewcommand{\Initialize}[1]{%
-  \State \textbf{Initialize:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
+\begin{document}
+\firstpage{1}
 
 \title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
-\author{}
-%\date{}                                           % Activate to display a given date or no date
+\author[Sample \textit{et~al}]{Corresponding Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^2$\footnote{to whom correspondence should be addressed}}
+\address{$^{1}$Department of XXXXXXX, Address XXXX etc.\\
+$^{2}$Department of XXXXXXXX, Address XXXX etc.}
 
-\begin{document}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
+\maketitle
 
 \begin{abstract}
+
+\section{Motivation:}
 Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
@@
 However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
+
+\section{Results:}
 We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
@@
 %We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
+\section{Availability:}
+Text
+
+\section{Contact:} \href{name@bio.com}{name@bio.com}
 \end{abstract}
-
-\maketitle
 
@@
 
-Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}.
-Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}.
-As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
-NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.
+Text Figure~\ref{fig:01} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.  \citep{Bag01} wants to know about …… text follows.
 
-Increasingly, NGS is being used to interrogate mutations in heterogeneous clinical samples.
-For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
-Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
-DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
+\begin{equation}
+\sum x+ y =Z\label{eq:01}
+\end{equation}
 
-However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
-Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
-Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{}.
-GATK involves more sophisticate pre and post-processing steps.
-The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
+\section{Approach}
 
-Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
-VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
-Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}.
-With the joint distribution available, one is able to identify locations with dissimilar allele frequencies.
-muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.
+Equation~(\ref{eq:01}) Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citealp{Boffelli03} might want to know about  text text text text ……
 
-Several studies have compared the relative performance of these algorithms.
-%TODO: [DISCUSS \citet{Stead:2013fu}] and others
+\begin{methods}
+\section{Methods}
 
-Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
-These steps may improve algorithm performance, but their presence makes it difficult to isolate what step is most responsible for the performance.
-Instead, we focus exclusively on the model structure and statistical inference steps.
-Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study.
-We have isolated the statistical inference and hypothesis testing steps so that they may be used independently or as part of a larger variant calling pipeline.
+Text
 
-We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}.
-In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
-We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
-Finally, in Section~\ref{sec:brca} we present results on detecting mutations in the PAXIP1 gene from matched tumor-normal data from the HCC1187 cell line.
+\end{methods}
 
-%%%%%%%%%%%%%%%%%%
-% Model Structure
-%%%%%%%%%%%%%%%%%%
-\section{Model Structure}\label{sec:model_structure}
-
-RVD uses a two-stage approach for detecting for rare variants.
-First, it estimates the parameters of a hierarchical Bayesian model under two sequencing data sets: one from the sample of interest (case) and one from a known reference sample (control).
-Then, it tests for a significant difference between key model parameters in the case and control samples and returns called variant positions.
-
-For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-The model generative process is as follows:
-
-\begin{enumerate}[noitemsep]
-	\item For each location $j$: 
-	\begin{enumerate}
-		\item Draw an error rate $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
-		\item For each replicate $i$:
-		\begin{enumerate}
-			\item Draw $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$
-			\item Draw $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
-		\end{enumerate}
-	\end{enumerate}
-\end{enumerate}
-
-The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
-The global error rate, $\mu_0$, estimates the expected error rate across all locations.
-The global precision, $M_0$, estimates the variation in the error rate  across locations.
-The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.
-
-RVD2 has three levels of sampling.
-First, a global error rate and global precision are chosen once for the entire data set.
-Then, at each location, a local precision is chosen and a local error rate is sampled from a Beta distribution.
-Finally, the error rate for replicate $i$ at location $j$ is drawn from a Beta distribution and the number of non-reference reads is drawn from a binomial.
-
-RVD2 hierarchically partitions sources of variation in the data.
-$r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
-$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility.
-The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-Importantly, increasing the read depth $n_{ji}$ only reduces the sampling error, but does nothing to reduce experimental variation or variation due to sequence context.
+\begin{figure}[!tpb]%figure1
+%\centerline{\includegraphics{pdf_figs/comparison_table_fdr.pdf}}
+\caption{Caption, caption.}\label{fig:01}
+\end{figure}
 
@@
 
-Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
-In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{Jordan2004}.
-A rounded box or ``plate" represents replication of the nodes within the plate.
-The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
-
-The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
-
-\begin{equation}\label{eqn:jointpdf}
-p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
-\end{equation}
-
-where
-\begin{align}
-p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
-p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
-p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
-\end{align}
-
-Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data, 
-\begin{equation}
-p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
-
-Finally, the log-likelihood of the data set is
-
-\begin{equation}
-\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%
-% Inference & Hypothesis Testing
-%%%%%%%%%%%%%%%%%%
-\section{Inference and Hypothesis Testing}
-
-The primary object of inference in this model is the joint posterior distribution function over the latent variables,
-\begin{equation}
-	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
-\end{equation}
-where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.
-
-The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution.
-However, there is not a closed form for the product of a Beta distribution with another Beta distribution, so exact inference is intractable.
-
-Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
-First, the hyperparameters are initialized using method-of-moments (MoM).
-Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule.
-Finally, we sample from the marginal posterior distribution for $\theta_{ji}$ given its Markov blanket.
-Samples from $\theta_{ji}$ can be drawn from the posterior distribution directly  because the prior and likelihood form a conjugate pair.
-This sampling procedure is repeated until the chain converges to a stationary distribution then we draw samples from the posterior distribution over latent variables.
-
-\begin{algorithm}[ht]
-\caption{Metropolis-within-Gibbs Algorithm}
-\label{alg:metro_gibbs}
-\begin{algorithmic}[1]
-
-\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
-\Repeat
-\For {each location j} \Comment{Sample $\mu_j$}
-  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
-  \State Set $\mu_j$ to the sample median for the T samples
-
-  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
-	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
-  \EndFor
-
-\EndFor
-\Until {sample size sufficient}
-\end{algorithmic}
-\end{algorithm}
-
-%%%%%%%%%%%%%
-% Initialization
-%%%%%%%%%%%%%
-\subsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
-MoM works by setting the population moment equal to the sample moment.
-A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
-We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.
-
-We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are 
-$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. 
-The estimates for the position-level parameters are 
-$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ 
-and 
-$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. 
-The estimates for the genome-level parameters are 
-$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ 
-and 
-$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. 
-While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
-
-%%%%%%%%%%%%%%%%%%
-% Sampling theta
-%%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
-
-Samples from the posterior distribution 
-$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ 
-are drawn analytically because of the Bayesian conjugacy between the prior 
-$p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ 
-and the likelihood 
-$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$. 
-The posterior distribution is 
-\begin{equation}
-	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%
-% Sampling mu
-%%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
-The posterior distribution over $\mu_j$ given its Markov blanket is 
-\begin{equation}
-	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
-\end{equation}
-
-Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution.
-Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from the posterior distribution.
-The acceptance probability is then
-\begin{equation}
-	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
-\end{equation}
-
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
-Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-
-We resample from the proposal if the sample is outside of the support of the posterior distribution.
-We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
-Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.
-
-%%%%%%%%%%%%%%%%%%
-% Posterior Density Test
-%%%%%%%%%%%%%%%%%%
-\subsection{Posterior Density Test}\label{sec:hypothesis_test}
-Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
-For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
-\begin{equation}\label{eqn:bayes_test}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
-\end{equation}
-where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
-We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
-\begin{equation}
-	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
-\end{equation}
-
-While we are able to compute the optimal $\tau$ threshold for a test data set, in general we would not have access to $\tau^*$.
-With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest.
-Absent this information we set $\tau = 0$.
-
-%%%%%%%%%%%%%%%%%%
-% Chi^2 Test
-%%%%%%%%%%%%%%%%%%
-\subsection{$\chi^2$ test for non-uniform base distribution}
-
-An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios.
-We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
-In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
-
-We use a $\chi^2$ goodness-of-fit test on a multinomial distribution over the non-reference bases to distinguish these two possible scenarios.
-The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
-Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
-The test statistic is
-
-\begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
-\end{equation}
-
-where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-\citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
-
-We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,
-
-\begin{equation}\label{eqn:fisher_combined}
-	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
-\end{equation}
-
-Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
-
-%%%%%%%%%%%%%%%%%%
-% Data Sets
-%%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms.
-Synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
-HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
-
-To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
-The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
-
-\subsection{HCC1187 Sequence Data}
-
-The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
-The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
-The estimated tumor purity was reported to be 0.8.
-A matched normal cells were derived from lymphoblastoid cells from peripheral blood.
-Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
-The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
-The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads.
-We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
-% TODO:cite the technical report. %Not sure the author is correct
-% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
-
-\section{Results}
-
-We tested RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
-The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
-We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
-
-We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
-
-To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
-loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
-homozygous somatic mutation).
-The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$.
-The size of the test is $\alpha=0.05$.
-%TODO: correct notation from \Pr to summation because we have samples, not distributions.
-
-\subsection{Performance with read depth}\label{sec:read_depth}
-
-We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
-For these ROC curves, we used the Bayesian test without the $\chi^2$ test.
-Adding the $\chi^2$ improves specificity at the expense of sensitivity.
-Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
-Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
-At the lowest depth the sensitivity and specificity is no better than random.
-However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43.
-The performance improves monotonically with read depth.
-Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance.}
-\label{fig:ROC}
-\end{center}
-\end{figure}
-% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
-
-%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
-All of the called positions show a clear difference between the case and control error rates.
-The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
-%
-%\begin{figure}[h]
-%\begin{center}
-%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
-%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-%\label{fig:MAF}
-%\end{center}
-%\end{figure}
-%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
-
-\subsection{Empirical performance compared with other algorithms}\label{sec:comparison}
-
-We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
-In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
-The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
-GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
-
-\subsubsection*{Sensitivity/Specificity Comparison}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
-\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
-\label{tbl:comparison_ss}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
-
-Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
-They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
-VarScan2-somatic is able to call more mixed samples.
-However, as the read depth increases the specificity degrades.
-Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below.
-muTect has good performance across a wide range of MAF levels.
-But even at the highest depth only has around 0.5 sensitivity for low MAF levels.
-
-The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
-The sensitivity increases considerably with read depth at a slight expense to specificity.
-With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs.
-However, in practice one may not know the optimal $\tau^*$ a-priori.
-With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives.
-When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see Appendix~\ref{sec:app_comparison}).
-
-\subsubsection*{False Discovery Rate Comparison}
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, strelka and muTect. Blank cells indicate no positive calls were made.
-
-Samtools performs well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
-GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
-VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_fdr.pdf}
-\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
-\label{tbl:comparison_fdr}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-
-VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level.
-However, the FDR is high due to many false positives.
-Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth.
-Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF.
-However, it does not call any variants at or below 1\% MAF.
-MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
-But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.
-
-RVD2 has a lower FDR than other algorithms when the read depth is greater than $10\times$ the MAF with $N=1$ and $\tau$ set to the default value of zero or to the optimal value.
-The FDR is higher when $N=6$ because the variance of the control error rate distribution $P(\mu_j^{\text{control}} | r^{\text{control}})$ is smaller.
-The smaller variance yields improvements in sensitivity at the expense of more false positives.
-Since the FDR only considers positive calls, the performance degrades.
-
-\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
-
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
-The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
-
-Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
-Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor sample we identified five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
-Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
-Some of these mutations are also found to be common population SNPs according to dbSNPv138.
-The corresponding identities are shown in the figure.
-%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
-\label{fig:brca_MAF}
-\end{center}
-\end{figure}
-%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
-
-The original research describing this sample used Strelka to identify mutations in the same sample.
-They identified chr7:154760439 as variant, but did not call the other four variants.
-In particular strelka missed the two LOH events.
-% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample.
-Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
-
 \section{Discussion}
 
-We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data.
-Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths and minor allele frequencies.
+Text
 
-Our inference algorithm uses Gibbs sampling to estimate the hierarchical empirical Bayes model.
-This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically.
-However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches.
-We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.
+\section{Conclusion}
 
-We have focused on the statistical model and hypothesis test in this study and our results do not include any pre-filtration of erroneous reads or post-filtration of mutation calls beyond a simple quality score threshold.
-Incorporation of such data-cleaning steps will likely improve the accuracy of the algorithm.
+\begin{enumerate}
+\item this is item, use enumerate
+\item this is item, use enumerate
+\item this is item, use enumerate
+\end{enumerate}
 
-Our approach does not address identification of indels, structural variants or copy number variants.
-Those mutations typically require specific data analysis models and tests that are different than those for single-nucleotide variants.
-Furthermore, analysis of RNA-seq data or other data generated on the NGS platform may require different models that are more appropriately tuned to the particular noise feature of that data.
+\section*{Acknowledgement}
+Text Text Text Text Text Text  Text Text.  \citealp{Boffelli03} might want to know about  text text text text
 
-\appendix
+\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
 
-%%%%%%%%%%%%%%%%%%%
-% Appendix A: Parameter Initialization
-%%%%%%%%%%%%%%%%%%%
-\section{Parameter Initialization}\label{sec:appendix_mom}
-Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
-Therefore the MoM estimator is
-\begin{equation}
-	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
-\end{equation}
+%\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+%
+%\bibliographystyle{plain}
+%
+%\bibliography{Document}
 
-We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
-The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
-The first and second population moments are
-\begin{eqnarray}
-	E[\theta_{ji}] =& \mu_j,\\
-	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
-\end{eqnarray}
-The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
-Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
-\begin{eqnarray}
-	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
-	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
-\end{eqnarray}
+\begin{thebibliography}{}
+\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03} Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
 
-Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
-\begin{eqnarray}
-	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
-	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
-\end{eqnarray}
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix B: Comparison Statistics
-%%%%%%%%%%%%%%%%%%%
-\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-%TODO: discuss this result
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
-\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
-\label{fig:comparison_mcc}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-
-%%%%%%%%%%%%%%%%%
-% Appendix C: Data Set Statistics
-%%%%%%%%%%%%%%%%%
-\section{RVD2 Estimated Parameters}\label{sec:synthetic_estimate}
-The RVD2 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:depthM}.
-
-The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The reads are 36bp long and we required no more than 2 mismatches.
-Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
-\caption{Key parameters for RVD2 model for synthetic DNA data sets.}
-\label{fig:depthM}
-\end{center}
-\end{figure}
-
-% from folder '22013-11-15_depthM_plot_avg10', dilution rate is 0.1.
-
-The right column of Figure~\ref{fig:depthM} shows the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
-
-The error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-We see that the variation between replicates is smaller than the variation between location.
-$M_j$ and $M_0$ are precision parameters, they are inversely proportional to the variance.
-Where $M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
-
-%%%%%%%%%%%%%%%%%
-% Appendix D: Parameter Settings for Variant Calling Algorithms
-%%%%%%%%%%%%%%%%%
-\section{Parameter settings for other variant calling algorithms}
-\paragraph{\textbf{Samtools}}We used samtools (v0.1.19) function mpileup to call variants and bcftools to save the result in standard VCF files.
-In mpileup, we set the -d option sufficiently high at $10^6$ to avoid truncating read deapth.
-Option -u was enabled to make sure the output bcf files were uncompressed.
-
-\paragraph{\textbf{GATK}}
-We used GATK (v2.1-8) UnifiedGenotyper function to detect mutations on our synthetic data following the recommended workflow.
-Due to some format incompatibility, we applied Picard to format read group and GATK for realignment.
-In UnifiedGenotyper, -ploid (Number of samples in each pool $\times$ Sample Ploidy) was set at 1 because our synthetic data is haploid; -dcov was set at $10^6$ to avoid downsampling coverage within GATK.
-
-\paragraph{\textbf{VarScan2-mpileup}}
-VarScan2 (v2.3.4) mpileup2snp is a SNP calling program which takes multi-samples from samtools mpileup pipeline.We assigned parameter -C value 50 as the synthetic data was aligned using BWA and set -d at $10^6$.
-In mpileup2snp, --min-var-freq, the only non-default parameter, was set low enough at $10^{-5}$ because the variant frequency can be as low as $10^{-3}$.
-
-\paragraph{\textbf{VarScan2-somatic}}
-We tested VarScan2 somatic on our synthetic dataset.
-The parameter --normal-purity set was at 1.00, --tumor-purity at the dilution rate.
-The parameter --min-var-freq was set at $10^{-5}$.
-We combined all the positions VarScan2-somatic called regardless the somatic status (Germline/LOH/Somatic/Unknown) to compare with performance of RVD2.
-
-\paragraph{\textbf{Strelka and muTect}}
-Since configuration and Analysis for Strelka and muTect is standardized and no parameter needs to be specified, we installed these two programs and ran them on our data set separately.
-
-Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
-GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
-Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
-
-% TODO: modify the following section
-GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
-Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
-\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
-We find its performance on much lower MAF variants and across a wider range of coverage depths.
-\bibliographystyle{apalike}
-\bibliography{bioinfo}
+\end{thebibliography}
 \end{document}

### 2014-01-07 doc/bioinformatics2014/supplementary_info.tex
Bioinformatics2014 folder for rvd27

--- 2013-12-08 doc/bioinformatics2013/RVD2.tex
@@
 \documentclass[11pt,reqno]{amsart}
-\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
+\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
 \geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
 \usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-%\usepackage{algorithm, algorithmic}
-
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
+\usepackage{algorithm, algorithmic}
 \usepackage{graphicx}
-
-\usepackage{verbatim}
+\usepackage{natbib}
 \usepackage{amssymb}
 \usepackage{amsmath}
-
-\usepackage{enumitem}
-
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{natbib}
-
 %\usepackage{epstopdf}
@@
 
-\algnewcommand{\Inputs}[1]{%
-  \State \textbf{Inputs:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
-\algnewcommand{\Initialize}[1]{%
-  \State \textbf{Initialize:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
-
-\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
-\author{}
+\title[RVD2: Supplementary Information]{Supplementary Information for
+  RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
 %\date{}                                           % Activate to display a given date or no date
@@
 \begin{document}
-
-\begin{abstract}
-Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
-Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
-However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
-We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
-We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
-
-%Unmasked positions has median read depth 51 for control sample, and 69 for case sample (from folder '2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1'; masked positions has median read depth 52 for control sample, and 70 for case sample( from foler '2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1')
-%We use our algorithm to call variants in a pooled sample of 100 patients with multiple sclerosis and identify novel variants in IL7RA and IL2R.
-\end{abstract}
-
 \maketitle
-
-\section{Introduction}
-
-Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}.
-Protocol and platform engineering improvements have enabled the generation of $1\times10^9$ bases of sequence data in 27 hours for approximately \$1000~\citep{Quail:2012hf}.
-As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
-NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.
-
-Increasingly, NGS is being used to interrogate mutations in heterogeneous clinical samples.
-For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
-Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
-DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
-
-However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
-Most analysis pipelines make use of preprocessing or postprocessing or both to eliminate error prone reads and false positives.
-Samtools/bcftools and GATK a naive Bayes decision rule to call a variant~\citep{}.
-GATK involves more sophisticate pre and post-processing steps.
-The genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
-
-Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
-VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
-Strelka uses a hierarchical Bayesian approach to model the joint distribution of the allele frequency in the tumor and normal samples at each locus~\citep{Saunders:2012fh}.
-With the joint distribution available, one is able to identify locations with dissimilar allele frequencies.
-muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.
-
-Several studies have compared the relative performance of these algorithms.
-%TODO: [DISCUSS \citet{Stead:2013fu}] and others
-
-Typically such a variant calling algorithms are comprised of multiple pre- and post-filtration steps around model estimation and hypothesis testing steps~\citep{Pabinger:2013dl}.
-These steps may improve algorithm performance, but their presence makes it difficult to isolate what step is most responsible for the performance.
-Instead, we focus exclusively on the model structure and statistical inference steps.
-Additional filtration steps may improve performance considerably, but those aspects are subjects of intense research and separate study.
-We have isolated the statistical inference and hypothesis testing steps so that they may be used independently or as part of a larger variant calling pipeline.
-
-We present the hierarchical Bayesian model in Section~\ref{sec:model_structure} and a Bayesian hypothesis test to detect mutations in Section~\ref{sec:hypothesis_test}.
-In Section~\ref{sec:read_depth} we present sensitivity and specificity results of our method on known, pure DNA samples mixed at defined fractions.
-We compare our algorithm to the most accurate methods available to date in Section~\ref{sec:comparison}.
-Finally, in Section~\ref{sec:brca} we present results on detecting mutations in the PAXIP1 gene from matched tumor-normal data from the HCC1187 cell line.
-
-%%%%%%%%%%%%%%%%%%
-% Model Structure
-%%%%%%%%%%%%%%%%%%
-\section{Model Structure}\label{sec:model_structure}
-
-RVD uses a two-stage approach for detecting for rare variants.
-First, it estimates the parameters of a hierarchical Bayesian model under two sequencing data sets: one from the sample of interest (case) and one from a known reference sample (control).
-Then, it tests for a significant difference between key model parameters in the case and control samples and returns called variant positions.
-
-For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-The model generative process is as follows:
-
-\begin{enumerate}[noitemsep]
-	\item For each location $j$: 
-	\begin{enumerate}
-		\item Draw an error rate $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
-		\item For each replicate $i$:
-		\begin{enumerate}
-			\item Draw $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$
-			\item Draw $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
-		\end{enumerate}
-	\end{enumerate}
-\end{enumerate}
-
-The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
-The global error rate, $\mu_0$, estimates the expected error rate across all locations.
-The global precision, $M_0$, estimates the variation in the error rate  across locations.
-The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.
-
-RVD2 has three levels of sampling.
-First, a global error rate and global precision are chosen once for the entire data set.
-Then, at each location, a local precision is chosen and a local error rate is sampled from a Beta distribution.
-Finally, the error rate for replicate $i$ at location $j$ is drawn from a Beta distribution and the number of non-reference reads is drawn from a binomial.
-
-RVD2 hierarchically partitions sources of variation in the data.
-$r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
-$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility.
-The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-Importantly, increasing the read depth $n_{ji}$ only reduces the sampling error, but does nothing to reduce experimental variation or variation due to sequence context.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
-\caption{RVD2 Graphical Model}
-\label{fig:graphical_model}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
-In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{Jordan2004}.
-A rounded box or ``plate" represents replication of the nodes within the plate.
-The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
-
-The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters is
-
-\begin{equation}\label{eqn:jointpdf}
-p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
-\end{equation}
-
-where
-\begin{align}
-p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
-p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
-p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
-\end{align}
-
-Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data, 
-\begin{equation}
-p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
-
-Finally, the log-likelihood of the data set is
-
-\begin{equation}
-\log p \left( r | n ; \mu_0, M_0, M \right) = \sum_{j=1}^J \sum_{i=1}^N \log \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%
-% Inference & Hypothesis Testing
-%%%%%%%%%%%%%%%%%%
-\section{Inference and Hypothesis Testing}
-
-The primary object of inference in this model is the joint posterior distribution function over the latent variables,
-\begin{equation}
-	p(\mu, \theta | r, n; \phi)  = \frac{ p(\mu, \theta, r | n; \phi) } {p ( r | n; \phi)},
-\end{equation}
-where the parameters are $\phi \triangleq \{\mu_0, M_0, M\}$.
-
-The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution.
-However, there is not a closed form for the product of a Beta distribution with another Beta distribution, so exact inference is intractable.
-
-Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
-First, the hyperparameters are initialized using method-of-moments (MoM).
-Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule.
-Finally, we sample from the marginal posterior distribution for $\theta_{ji}$ given its Markov blanket.
-Samples from $\theta_{ji}$ can be drawn from the posterior distribution directly  because the prior and likelihood form a conjugate pair.
-This sampling procedure is repeated until the chain converges to a stationary distribution then we draw samples from the posterior distribution over latent variables.
-
-\begin{algorithm}[ht]
-\caption{Metropolis-within-Gibbs Algorithm}
-\label{alg:metro_gibbs}
-\begin{algorithmic}[1]
-
-\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
-\Repeat
-\For {each location j} \Comment{Sample $\mu_j$}
-  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H
-  \State Set $\mu_j$ to the sample median for the T samples
-
-  \For {each replicate i} \Comment{Sample $\theta_{ji}$}
-	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$
-  \EndFor
-
-\EndFor
-\Until {sample size sufficient}
-\end{algorithmic}
-\end{algorithm}
-
-%%%%%%%%%%%%%
-% Initialization
-%%%%%%%%%%%%%
-\subsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
-MoM works by setting the population moment equal to the sample moment.
-A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
-We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.
-
-We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are 
-$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. 
-The estimates for the position-level parameters are 
-$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ 
-and 
-$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. 
-The estimates for the genome-level parameters are 
-$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ 
-and 
-$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. 
-While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
-
-%%%%%%%%%%%%%%%%%%
-% Sampling theta
-%%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
-
-Samples from the posterior distribution 
-$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$ 
-are drawn analytically because of the Bayesian conjugacy between the prior 
-$p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$ 
-and the likelihood 
-$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$. 
-The posterior distribution is 
-\begin{equation}
-	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%
-% Sampling mu
-%%%%%%%%%%%%%%%%%%
-\subsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
-The posterior distribution over $\mu_j$ given its Markov blanket is 
-\begin{equation}
-	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
-\end{equation}
-
-Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution.
-Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-
-A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from the posterior distribution.
-The acceptance probability is then
-\begin{equation}
-	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
-\end{equation}
-
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
-Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-
-We resample from the proposal if the sample is outside of the support of the posterior distribution.
-We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
-Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.
-
-%%%%%%%%%%%%%%%%%%
-% Posterior Density Test
-%%%%%%%%%%%%%%%%%%
-\subsection{Posterior Density Test}\label{sec:hypothesis_test}
-Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
-For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
-\begin{equation}\label{eqn:bayes_test}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
-\end{equation}
-where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
-We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
-\begin{equation}
-	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
-\end{equation}
-
-While we are able to compute the optimal $\tau$ threshold for a test data set, in general we would not have access to $\tau^*$.
-With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest.
-Absent this information we set $\tau = 0$.
-
-%%%%%%%%%%%%%%%%%%
-% Chi^2 Test
-%%%%%%%%%%%%%%%%%%
-\subsection{$\chi^2$ test for non-uniform base distribution}
-
-An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios.
-We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
-In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
-
-We use a $\chi^2$ goodness-of-fit test on a multinomial distribution over the non-reference bases to distinguish these two possible scenarios.
-The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
-Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
-The test statistic is
-
-\begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
-\end{equation}
-
-where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-\citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
-
-We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,
-
-\begin{equation}\label{eqn:fisher_combined}
-	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
-\end{equation}
-
-Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
-
-%%%%%%%%%%%%%%%%%%
-% Data Sets
-%%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms.
-Synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
-HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
-
-To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
-The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
-
-\subsection{HCC1187 Sequence Data}
-
-The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
-The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma.
-The estimated tumor purity was reported to be 0.8.
-A matched normal cells were derived from lymphoblastoid cells from peripheral blood.
-Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
-The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
-The aligned data had an average read depth of 40x for the normal sample and 90x for the tumor sample with about 96\% coverage with 10 or more reads.
-We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
-% TODO:cite the technical report. %Not sure the author is correct
-% HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
-
-\section{Results}
-
-We tested RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
-The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
-We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
-
-We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
-
-To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
-loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
-homozygous somatic mutation).
-The two hypothesis tests are then $\Pr( \tilde{\mu}_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \tilde{\mu}_j^{\Delta} \leq \tau ) > 1-\alpha$.
-The size of the test is $\alpha=0.05$.
-%TODO: correct notation from \Pr to summation because we have samples, not distributions.
-
-\subsection{Performance with read depth}\label{sec:read_depth}
-
-We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
-For these ROC curves, we used the Bayesian test without the $\chi^2$ test.
-Adding the $\chi^2$ improves specificity at the expense of sensitivity.
-Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
-Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths.
-At the lowest depth the sensitivity and specificity is no better than random.
-However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43.
-The performance improves monotonically with read depth.
-Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance.}
-\label{fig:ROC}
-\end{center}
-\end{figure}
-% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
-
-%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
-All of the called positions show a clear difference between the case and control error rates.
-The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
-%
-%\begin{figure}[h]
-%\begin{center}
-%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
-%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-%\label{fig:MAF}
-%\end{center}
-%\end{figure}
-%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
-
-\subsection{Empirical performance compared with other algorithms}\label{sec:comparison}
-
-We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
-In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
-The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
-GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
-
-\subsubsection*{Sensitivity/Specificity Comparison}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_ss.pdf}
-\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
-\label{tbl:comparison_ss}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
-
-Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
-They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
-VarScan2-somatic is able to call more mixed samples.
-However, as the read depth increases the specificity degrades.
-Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below.
-muTect has good performance across a wide range of MAF levels.
-But even at the highest depth only has around 0.5 sensitivity for low MAF levels.
-
-The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
-The sensitivity increases considerably with read depth at a slight expense to specificity.
-With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs.
-However, in practice one may not know the optimal $\tau^*$ a-priori.
-With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives.
-When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see Appendix~\ref{sec:app_comparison}).
-
-\subsubsection*{False Discovery Rate Comparison}
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, strelka and muTect. Blank cells indicate no positive calls were made.
-
-Samtools performs well on 100\% MAF sample and performance improves for read depths 3089 and 30590.
-GATK performs well on both the 10\% and 100\% variants, but makes a false positive call at the 100\% MAF level for all read depth levels.
-VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=160mm]{pdf_figs/comparison_table_fdr.pdf}
-\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
-\label{tbl:comparison_fdr}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-
-VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level.
-However, the FDR is high due to many false positives.
-Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth.
-Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF.
-However, it does not call any variants at or below 1\% MAF.
-MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
-But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.
-
-RVD2 has a lower FDR than other algorithms when the read depth is greater than $10\times$ the MAF with $N=1$ and $\tau$ set to the default value of zero or to the optimal value.
-The FDR is higher when $N=6$ because the variance of the control error rate distribution $P(\mu_j^{\text{control}} | r^{\text{control}})$ is smaller.
-The smaller variance yields improvements in sensitivity at the expense of more false positives.
-Since the FDR only considers positive calls, the performance degrades.
-
-\subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
-
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
-The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
-
-Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
-Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor sample we identified five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T.
-Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
-Some of these mutations are also found to be common population SNPs according to dbSNPv138.
-The corresponding identities are shown in the figure.
-%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
-\label{fig:brca_MAF}
-\end{center}
-\end{figure}
-%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
-
-The original research describing this sample used Strelka to identify mutations in the same sample.
-They identified chr7:154760439 as variant, but did not call the other four variants.
-In particular strelka missed the two LOH events.
-% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample.
-Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
-
-\section{Discussion}
-
-We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data.
-Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths and minor allele frequencies.
-
-Our inference algorithm uses Gibbs sampling to estimate the hierarchical empirical Bayes model.
-This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically.
-However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches.
-We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.
-
-We have focused on the statistical model and hypothesis test in this study and our results do not include any pre-filtration of erroneous reads or post-filtration of mutation calls beyond a simple quality score threshold.
-Incorporation of such data-cleaning steps will likely improve the accuracy of the algorithm.
-
-Our approach does not address identification of indels, structural variants or copy number variants.
-Those mutations typically require specific data analysis models and tests that are different than those for single-nucleotide variants.
-Furthermore, analysis of RNA-seq data or other data generated on the NGS platform may require different models that are more appropriately tuned to the particular noise feature of that data.
-
-\appendix
 
@@
 
-%%%%%%%%%%%%%%%%%%%
-% Appendix B: Comparison Statistics
-%%%%%%%%%%%%%%%%%%%
-\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-%TODO: discuss this result
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=140mm]{pdf_figs/comparison_table_mcc.pdf}
-\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
-\label{fig:comparison_mcc}
-\end{center}
-\end{figure}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-
 %%%%%%%%%%%%%%%%%
-% Appendix C: Data Set Statistics
+% Appendix B: Data Set Statistics
 %%%%%%%%%%%%%%%%%
@@
 The RVD2 algorithm provides estimates of model parameters and latent variables given the data. We show several of these parameters in Figure~\ref{fig:depthM}.
-
-The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The reads are 36bp long and we required no more than 2 mismatches.
-Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
 
@@
 \end{figure}
+
+The left column of Figure~\ref{fig:depthM} shows the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
+Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
+For the 100\% mutant data set, the read depth drops at the mutant locations.
+This is due to the parameters imposed at the alignment stage.
+The reads are 36bp long and we required no more than 2 mismatches.
+Therefore, reads that overlapped two mutations (spaced 20bp apart by design) and included one additional mutation would not align.
 
@@
 
+%%%%%%%%%%%%%%%%%%%
+% Appendix C: Comparison Statistics
+%%%%%%%%%%%%%%%%%%%
+\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
+Figure~\ref{fig:comparison_mcc} compares RVD2 with samtools, GATK, varscan, strelka and muTect using Matthews Correlation Coefficient (MCC)~\citep{matthews1985homeostasis}.
+\begin{figure}[ht]
+\begin{center}
+\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_mcc.pdf}
+\caption{Matthews correlation coefficient (MCC) comparison with other variant calling algorithms.}
+\label{fig:comparison_mcc}
+\end{center}
+\end{figure}
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
+
+Samtools and VarScan2-mpileup achieved MCC value generally higher than 0.90 on 100\% MAF sample across all read depthes, with 1.0 represents for a perfect prediction.
+However, both of them detected no variant when MAF is 10.0\% or lower, with only one exception for samtools when MAF is 10.0\% and read depth 22.
+GATK, Varscan2-somatic, Strelka and GATK outperformed Samtools and VarScan2-mpile on the 10.0\% MAF sample, while approximately tied in other cases.
+Strelka achieved best MCC on 10\% MAF sample comparing to Varscan2-somatic and GATK, more specifically around 1.00 when read depth is 260 or higher.
+There is a very obvious but unconventional decreasing trend in VarScan2-somatic MCC value across different read depth and MAF level, a phenomenon also observed by~\citealp{Stead:2013fu}.
+It is because VarScan2-somatic tends to call more false positives as read depth gets higher.
+Mutect seems to performs the best among all the algorithms expect RVD2 when MAF is 1.0\% or lower.
+It achieves MCC values varying from -0.02 to 0.43, though too low to be practically meaningful.
+However, muTect achieved relatively lower MCC values when the MAF level is 10\% and 100\%, as a counteractive of being oversensitive.
+
+RVD2 achieved MCC value 1.00 when the MAF is 100.0\% at all read depth and 10\% when read depth is not lower than 260.
+This indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
 %%%%%%%%%%%%%%%%%
-% Appendix D: Parameter Settings for Variant Calling Algorithms
+% Appendix D: Estimated minor allele fraction for called variants in synthetic gene.
+%%%%%%%%%%%%%%%%%
+\section{Estimated MAF for called variants in synthetic gene.}
+Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
+All of the called positions show a clear difference between the case and control error rates.
+The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
+\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\label{fig:MAF}
+\end{center}
+\end{figure}
+% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
+
+%%%%%%%%%%%%%%%%%
+% Appendix E: Parameter Settings for Variant Calling Algorithms
 %%%%%%%%%%%%%%%%%
@@
 
-Samtools mpileup takes multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
-GATK, VarScan2-mpileup accept multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
+Samtools mpileup and GATK can accept multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
+VarScan2-mpileup takes multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
 Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
 
-% TODO: modify the following section
-GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
-Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
-\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
-We find its performance on much lower MAF variants and across a wider range of coverage depths.
-\bibliographystyle{apalike}
+\bibliographystyle{natbib}
 \bibliography{bioinfo}

### 2014-01-07 doc/bioinformatics2013/RVD2.tex
update manuscript in bioinformatics2013

@@
 \usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-%\usepackage{algorithm, algorithmic}
+%\usepackage{algorithm}
 
@@
 %TODO: [DISCUSS \citet{Stead:2013fu}, \citet{spencer2013performance}] and others
+%TODO: modify the following section
+GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
+Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
+\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
+We find its performance on much lower MAF variants and across a wider range of coverage depths.
 
@@
 We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.5]$, heterozygous mutant $[0.25, 0.75]$, and homozygous mutant $[0.5, 1.0]$ and the size of the test is $\alpha=0.05$.
+To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.2]$, heterozygous mutant $[0.1, 0.9]$, and homozygous mutant $[0.8, 1.0]$ and the size of the test is $\alpha=0.05$.
 
@@
 
-%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
-All of the called positions show a clear difference between the case and control error rates.
-The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
-%
-%\begin{figure}[h]
-%\begin{center}
-%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
-%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-%\label{fig:MAF}
-%\end{center}
-%\end{figure}
-%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
-
 \subsection{Empirical performance compared with other algorithms}\label{sec:comparison}
@@
 
+%TODO: double check with the following sentence
 The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
@@
 
-RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants.
+RVD2 identified six variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+Of the six called variants, three were homozygous germline variants and two were heterozygous germline variants.
 The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
+The remaining one is a homozygous somatic variant.
 
@@
 \section{Algorithm Comparison Statistics}\label{sec:app_comparison}
-Figure~\ref{fig:comparison_mcc} compares RVD2 with samtools, GATK, varscan, strelka and muTect using Matthew Correlation Coefficient (MCC)~citep(matthews1985homeostasis).
-
-Samtools and VarScan2-mpileup achieved MCC value generally higher than 0.90 on 100\% MAF sample across all read depthes, with 1.0 represents for a perfect prediction.
-However, both of them achieved MCC value at 0.00 when MAF is 10.0\% or lower, which indicates no better than random prediction.
-GATK, Varscan2-somatic, Strelka outperformed Samtools and VarScan2-mpile on the 10.0\% MAF sample, while performed similarly in other cases.
-GATK5, achieved relatively stable MCC between 0.65~0.88.
-Strelka achieved even better MCC on 10\% MAF sample, more specifically around 1.00 when read deapth is high but 0.53 when read depth is 22.
-There is a very obvious but unconventional decreasing trend in VarScan2-somatic MCC value across different read depth and MAF level, a phenomenon also observed by ~\citet{Stead:2013fu}.
-It is because VarScan2-somatic tends to call more false positives as read depth gets higher.
-As to Mutect, overall it seems to performs the best among all the method RVD2 is comparing to.
-It achieve postive MCC values (-0.02--0.43, with a single case at 0.60), though too low to be practically meaningful.
-Moreover, muTect achieved relatively low MCC (0.77--0.90) comparing to some other algorithms when the MAF level is 10\% and 100\%, as a counteractive of being oversensitive.
-
-RVD2 achieved MCC value at 1.00 when the MAF is 100.0\% at all read depth and 10\% when read depth is not lower than 260.
-The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
-
+Figure~\ref{fig:comparison_mcc} compares RVD2 with samtools, GATK, varscan, strelka and muTect using Matthews Correlation Coefficient (MCC)~\citep{matthews1985homeostasis}.
 \begin{figure}[ht]
@@
 % from folder 'rvd2\results\2013-09-19_operating_characteristics'
+
+Samtools and VarScan2-mpileup achieved MCC value generally higher than 0.90 on 100\% MAF sample across all read depthes, with 1.0 represents for a perfect prediction.
+However, both of them detected no variant when MAF is 10.0\% or lower, with only one exception for samtools when MAF is 10.0\% and read depth 22.
+GATK, Varscan2-somatic, Strelka and GATK outperformed Samtools and VarScan2-mpile on the 10.0\% MAF sample, while approximately tied in other cases.
+Strelka achieved best MCC on 10\% MAF sample comparing to Varscan2-somatic and GATK, more specifically around 1.00 when read depth is 260 or higher.
+There is a very obvious but unconventional decreasing trend in VarScan2-somatic MCC value across different read depth and MAF level, a phenomenon also observed by ~\citet{Stead:2013fu}.
+It is because VarScan2-somatic tends to call more false positives as read depth gets higher.
+Mutect seems to performs the best among all the algorithms expect RVD2 when MAF is 1.0\% or lower.
+It achieves MCC values varying from -0.02 to 0.43, though too low to be practically meaningful.
+However, muTect achieved relatively lower MCC values when the MAF level is 10\% and 100\%, as a counteractive of being oversensitive.
+
+RVD2 achieved MCC value at 1.00 when the MAF is 100.0\% at all read depth and 10\% when read depth is not lower than 260.
+This indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF.
 
@@
 %%%%%%%%%%%%%%%%%
-% Appendix D: Parameter Settings for Variant Calling Algorithms
+% Appendix D: Estimated minor allele fraction for called variants in synthetic gene.
+%%%%%%%%%%%%%%%%%
+\section{Estimated MAF for called variants in synthetic gene.}
+Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%.
+All of the called positions show a clear difference between the case and control error rates.
+The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
+%
+\begin{figure}[h]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
+\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\label{fig:MAF}
+\end{center}
+\end{figure}
+% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.
+
+%%%%%%%%%%%%%%%%%
+% Appendix E: Parameter Settings for Variant Calling Algorithms
 %%%%%%%%%%%%%%%%%
@@
 
-% TODO: modify the following section
-GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
-Therefore, we expect those algorithms to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
-\citet{Stead:2013fu} showed that varscan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
-We find its performance on much lower MAF variants and across a wider range of coverage depths.
 \bibliographystyle{apalike}

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/Compute_ROC_Synthetic/plotROCbyDilution_skew.py
Experiment folder for all the results in the paper(ROC curves, barplots, measure statistics tables)

@@ -1,0 +1,208 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0)
+
+    folderList = ('six_replicates_synthetic_avg10',\
+                  'six_replicates_synthetic_avg100',\
+                  'six_replicates_synthetic_avg1000',\
+                  'six_replicates_synthetic_avg10000')
+    
+    N=1000 # Z sampling size  
+    fig=plt.figure(figsize=(10, 9))
+    chi2=False
+    lstyle=('-','--','-.',':')
+    lstyle=('-.','--','-',':')
+    lstyle=('--','-.','-',':')
+    lcolor=('c','r','g','b')
+    for d in dilutionList:
+        logging.debug("Processing dilution: %0.1f%%" % d)
+        ax = fig.add_subplot(2,2,dilutionList.index(d)+1)
+        label=[]
+        nscrewc = -2
+        for f in folderList:
+            nscrewc = nscrewc+1
+            path='vcf/%s' % str(10**(folderList.index(f)+1))
+            if not os.path.exists(path):
+                os.makedirs(path)
+            controlFile = "../%s/Control.hdf5" %f
+            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
+            caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
+             # ROC
+            [fpr,tpr,cov, T] = ROCpoints(controlFile,caseFile, path, d, P=0.95,chi2=chi2)
+##            ax.plot(fpr,tpr, color=lcolor[folderList.index(f)], label='%d' % cov)
+##            ax.plot(fpr+0.006*nscrewc,tpr-0.006*nscrewc,linestyle=lstyle[folderList.index(f)], color=lcolor[folderList.index(f)], linewidth=2.0, label='%d' % cov)
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_genome/runall.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import random
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=5)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["../../2013-11-07_HCC1187_PAXIP1_genome_fa/depth_chart/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():
+    random.seed(199096)
+    # Estimate the model for the cases
+    logging.debug("Processing control data.")
+    controlList = ['HCC1187BL_S1.dc',]
+    fit_model(controlList, 'Control.hdf5')
+
+
+##    # Estimate the model for the cases
+##    logging.debug("Processing case data.")
+##    caseList = ['HCC1187C_S1.dc',]
+##    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_genome/runall2.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import random
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=5)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["../../2013-11-07_HCC1187_PAXIP1_genome_fa/depth_chart/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():
+    random.seed(199096)
+##    # Estimate the model for the cases
+##    logging.debug("Processing control data.")
+##    controlList = ['HCC1187BL_S1.dc',]
+##    fit_model(controlList, 'Control.hdf5')
+
+
+    # Estimate the model for the cases
+    logging.debug("Processing case data.")
+    caseList = ['HCC1187C_S1.dc',]
+    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_genome_600/MuBarPlot.py

@@ -1,0 +1,144 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd271 as rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    #model samples
+    
+    controlFile = "Control.hdf5"
+    caseFile = "Case.hdf5"
+    
+    hdf5name = 'HCC1187_call_unmasked.hdf5'
+    
+    try:
+        with h5py.File(hdf5name, 'r') as f:
+                call01 = f['call01'][...]
+                call00 = f['call00'][...]
+                call1 = f['call1'][...]
+                call2 = f['call2'][...]
+                call3 = f['call3'][...]
+                call4 = f['call4'][...]
+                call5 = f['call5'][...]
+                call6 = f['call6'][...]
+                call7 = f['call7'][...]
+                call8 = f['call8'][...]
+                controlMu = f['controlMu'][...]
+                caseMu = f['caseMu'][...]
+                caseN_median = f['caseN_median'][...]
+                controlN_median = f['controlN_median'][...]
+                loc = f['loc'][...]
+                f.close()
+    except IOError as e:
+        
+        diffalpha = 0.95
+        diffroi = (0.00001,np.inf)
+        N = 1000
+        outputFile = 'control_case_diff_plus.vcf'
+        [loc, call00, controlMu, caseMu, controlN, caseN, postP, chi2P ]= rvd27.diff_test(0.95, controlFile, caseFile, \
+                                                                     diffroi, N, outputFile = 'control_case_diff_plus.vcf')
+        ## median read depth
+        caseN_median = np.median(caseN)
+        controlN_median = np.median(controlN)
+        
+        diffroi = (-np.inf, 0)
+        [_,call01,_,_,_,_,_,_] = rvd27.diff_test(0.95, controlFile, caseFile, \
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_genome_600/runall.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import random
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=5)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["../../../data/HCC1187_PAXIP1_genome_depth_chart_600/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():
+##    random.seed(199096)
+##    # Estimate the model for the cases
+##    logging.debug("Processing control data.")
+##    controlList = ['HCC1187BL_S1.dc',]
+##    fit_model(controlList, 'Control.hdf5')
+
+
+    # Estimate the model for the cases
+    logging.debug("Processing case data.")
+    caseList = ['HCC1187C_S1.dc',]
+    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_hg19masked/MuBarPlot.py

@@ -1,0 +1,143 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd271 as rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    #model samples
+    
+    controlFile = "Control.hdf5"
+    caseFile = "Case.hdf5"
+    
+    hdf5name = 'HCC1187_call_unmasked.hdf5'
+    
+    try:
+        with h5py.File(hdf5name, 'r') as f:
+                call01 = f['call01'][...]
+                call00 = f['call00'][...]
+                call1 = f['call1'][...]
+                call2 = f['call2'][...]
+                call3 = f['call3'][...]
+                call4 = f['call4'][...]
+                call5 = f['call5'][...]
+                call6 = f['call6'][...]
+                call7 = f['call7'][...]
+                call8 = f['call8'][...]
+                controlMu = f['controlMu'][...]
+                caseMu = f['caseMu'][...]
+                caseN_median = f['caseN_median'][...]
+                controlN_median = f['controlN_median'][...]
+                loc = f['loc'][...]
+                f.close()
+    except IOError as e:
+        
+        diffalpha = 0.95
+        diffroi = (0.00001,np.inf)
+        N = 1000
+        outputFile = 'control_case_diff_plus.vcf'
+        [loc, call00, controlMu, caseMu, controlN, caseN, postP, chi2P ]= rvd27.diff_test(0.95, controlFile, caseFile, \
+                                                                     diffroi, N, outputFile = 'control_case_diff_plus.vcf')
+        ## median read depth
+        caseN_median = np.median(caseN)
+        controlN_median = np.median(controlN)
+        
+        diffroi = (-np.inf, 0)
+        [_,call01,_,_,_,_,_,_] = rvd27.diff_test(0.95, controlFile, caseFile, \
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_hg19masked/runall.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import random
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=5)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["../../2013-11-07_HCC1187_PAXIP1_hg19masked/depth_chart/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():
+    random.seed(199096)
+    # Estimate the model for the cases
+    logging.debug("Processing control data.")
+    controlList = ['HCC1187BL_S1.dc',]
+    fit_model(controlList, 'Control.hdf5')
+
+##
+##    # Estimate the model for the cases
+##    logging.debug("Processing case data.")
+##    caseList = ['HCC1187C_S1.dc',]
+##    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_hg19masked/runall2.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import random
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=5)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["../../2013-11-07_HCC1187_PAXIP1_hg19masked/depth_chart/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():
+    random.seed(199096)
+##    # Estimate the model for the cases
+##    logging.debug("Processing control data.")
+##    controlList = ['HCC1187BL_S1.dc',]
+##    fit_model(controlList, 'Control.hdf5')
+
+
+    # Estimate the model for the cases
+    logging.debug("Processing case data.")
+    caseList = ['HCC1187C_S1.dc',]
+    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_hg19masked_500/runall.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import random
+
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+
+pool = mp.Pool(processes=5)
+#pool = None
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+def fit_model(fileNameList, h5FileName):
+    fileList = ["../../../data/HCC1187_PAXIP1_hg19masked_depth_chart_500/%s" % filename for filename in fileNameList]
+    (r, n, loc, refb) = rvd27.load_depth(fileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+       refb=refb)
+
+def main():
+    random.seed(199096)
+    # Estimate the model for the cases
+    logging.debug("Processing control data.")
+    controlList = ['HCC1187BL_S1.dc',]
+    fit_model(controlList, 'Control.hdf5')
+
+
+    # Estimate the model for the cases
+    logging.debug("Processing case data.")
+    caseList = ['HCC1187C_S1.dc',]
+    fit_model(caseList, 'Case.hdf5')
+
+if __name__ == "__main__":
+	main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/one_replicate_synthetic_T0/optimalT_0.py

@@ -1,0 +1,41 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0,100.0)
+
+    folderList = ('one_replicate_synthetic_avg10',\
+                  'one_replicate_synthetic_avg100',\
+                  'one_replicate_synthetic_avg1000',\
+                  'one_replicate_synthetic_avg10000')
+
+    for d in dilutionList:
+        logging.debug("Processing dilution: %0.1f%%" % d)
+    	for f in folderList:
+            path='vcf/%s' % str(10**(folderList.index(f)+1))
+            if not os.path.exists(path):
+                os.makedirs(path)
+            controlFile = "../%s/Control.hdf5" %f
+            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
+            caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
+            
+            outputFile='%(path)s/vcf%(dilution)s' %{'path':path,'dilution':str(d).replace('.','_')}
+           
+            rvd27.test(controlFile,caseFile,T=0,N=1000,outputFile=outputFile)
+if __name__ == '__main__':
+    main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/one_replicate_synthetic_avg10/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p3.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/10/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/one_replicate_synthetic_avg100/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p3.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/100/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/one_replicate_synthetic_avg1000/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p3.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/one_replicate_synthetic_avg10000/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p3.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/10000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/one_replicate_synthetic_optT/optimalT.py

@@ -1,0 +1,236 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0,100.0)
+
+    folderList = ('one_replicate_synthetic_avg10',\
+                  'one_replicate_synthetic_avg100',\
+                  'one_replicate_synthetic_avg1000',\
+                  'one_replicate_synthetic_avg10000')
+
+    outputFile='optimalT.hdf5'
+    try:
+        with h5py.File(outputFile, 'r') as f:
+                cov = f['cov'][...]
+                T_ED = f['T_ED'][...]
+                T_L1 = f['T_L1'][...]
+                T_MCC = f['T_MCC'][...]           
+                f.close()
+                logging.debug('optimalT.hdf5 exits already')
+    except IOError as e:
+        logging.debug('Generate optimal dataset...')
+        N=1000 # Z sampling size  
+        chi2=True
+        times=8
+        T_ED=np.zeros((times,len(dilutionList),len(folderList)))
+        T_L1=np.zeros((times,len(dilutionList),len(folderList)))
+        T_MCC=np.zeros((times,len(dilutionList),len(folderList)))
+        cov=np.zeros((times,len(dilutionList),len(folderList)))
+        for i in xrange(times):
+            logging.debug('Runing the %(i)d times out of %(total)d...' % {'i':i+1, 'total':times})
+            for d in dilutionList:
+                logging.debug("Processing dilution: %0.1f%%" % d)
+                for f in folderList:
+                    path='%s' % str(10**(folderList.index(f)+1))
+##                    pdb.set_trace()
+                    controlFile = "../%s/Control.hdf5" %f
+                    caseFile = "Case%s.hdf5" % str(d).replace(".","_")
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/six_replicates_synthetic_T0/optimalT_0.py

@@ -1,0 +1,40 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0,100.0)
+
+    folderList = ('six_replicates_synthetic_avg10',\
+                  'six_replicates_synthetic_avg100',\
+                  'six_replicates_synthetic_avg1000',\
+                  'six_replicates_synthetic_avg10000')
+    for d in dilutionList:
+        logging.debug("Processing dilution: %0.1f%%" % d)
+    	for f in folderList:
+            path='vcf/%s' % str(10**(folderList.index(f)+1))
+            if not os.path.exists(path):
+                os.makedirs(path)
+            controlFile = "../%s/Control.hdf5" %f
+            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
+            caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
+            
+            outputFile='%(path)s/vcf%(dilution)s' %{'path':path,'dilution':str(d).replace('.','_')}
+           
+            rvd27.test(controlFile,caseFile,T=0,N=1000,outputFile=outputFile)
+if __name__ == '__main__':
+    main()
+

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/six_replicates_synthetic_avg10/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p1.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/10/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/six_replicates_synthetic_avg100/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p1.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/100/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/six_replicates_synthetic_avg1000/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p1.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/six_replicates_synthetic_avg10000/runall.py

@@ -1,0 +1,66 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+##pool =None
+pool = mp.Pool(processes=5)
+tocfilename = "../../../data/synthetic_toc_p1.txt"
+toc = pd.read_table(tocfilename)
+
+gibbs_nsample = 4000
+mh_nsample = 10
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    controlFileList = ["../../2013-08-06_Downsample_Read_Depth/depth_chart/10000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
+    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
+    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+
+# Estimate the model for the cases
+for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
+    logging.debug("Processing dilution: %0.1f" % dilution)
+    
+    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
+
+    try:
+        with h5py.File(h5FileName, 'r') as f:
+            pass
+    except IOError as e:
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/six_replicates_synthetic_optT/optimalT.py

@@ -1,0 +1,236 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0,100.0)
+
+    folderList = ('six_replicates_synthetic_avg10',\
+                  'six_replicates_synthetic_avg100',\
+                  'six_replicates_synthetic_avg1000',\
+                  'six_replicates_synthetic_avg10000')
+
+    outputFile='optimalT.hdf5'
+    try:
+        with h5py.File(outputFile, 'r') as f:
+                cov = f['cov'][...]
+                T_ED = f['T_ED'][...]
+                T_L1 = f['T_L1'][...]
+                T_MCC = f['T_MCC'][...]           
+                f.close()
+                logging.debug('optimalT.hdf5 exits already')
+    except IOError as e:
+        logging.debug('Generate optimal dataset...')
+        N=1000 # Z sampling size  
+        chi2=True
+        times=8
+        T_ED=np.zeros((times,len(dilutionList),len(folderList)))
+        T_L1=np.zeros((times,len(dilutionList),len(folderList)))
+        T_MCC=np.zeros((times,len(dilutionList),len(folderList)))
+        cov=np.zeros((times,len(dilutionList),len(folderList)))
+        for i in xrange(times):
+            logging.debug('Runing the %(i)d times out of %(total)d...' % {'i':i+1, 'total':times})
+            for d in dilutionList:
+                logging.debug("Processing dilution: %0.1f%%" % d)
+                for f in folderList:
+                    path='%s' % str(10**(folderList.index(f)+1))
+                    controlFile = "../%s/Control.hdf5" %f
+                    caseFile = "Case%s.hdf5" % str(d).replace(".","_")
[…]

### 2014-01-07 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/synthetic_mu_barplot_optimalT/MuBarPlot.py

@@ -1,0 +1,199 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+
+# Insert the src/python/rvd27 directory at front of the path
+rvddir = os.path.join('../../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    dilutionList = (0.1,0.3,1.0,10.0)
+
+    folderList = ('six_replicates_synthetic_avg10',\
+                  'six_replicates_synthetic_avg100',\
+                  'six_replicates_synthetic_avg1000',\
+                  'six_replicates_synthetic_avg10000')
+    
+    N=1000 # Z sampling size  
+    fig=plt.figure(figsize=(10, 9))
+    chi2=False
+    plt.suptitle('Comparing ROC plot for different read depth' )
+    for d in dilutionList:
+        logging.debug("Processing dilution: %0.1f%%" % d)
+        ax = fig.add_subplot(2,2,dilutionList.index(d)+1)
+        label=[]
+        for f in folderList:            
+            controlFile = "../%s/Control.hdf5" %f
+            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
+            caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
+             # ROC          
+            [fpr,tpr, cov, optimalT, caseMu1, controlMu1, loc1] = ROCpoints(controlFile,caseFile,P=0.95,chi2=chi2)
+            title='Dilution%(dilution)0.1f_Depth%(depth)d_Variant_Call.pdf'%{'dilution':d,'depth':cov}
+            MuBarPlot(controlMu1,caseMu1,loc1,optimalT,title)
+            
+            ax.plot(fpr,tpr, label='%d' % cov)
+
+        l = ax.legend(loc=4,prop={'size':9},title='Read depth')
+        l.get_title().set_fontsize(9)
+        ax.plot([0,1],[0,1],color='k',linestyle='dashed')
+        ax.set_title('%0.1f%% Mutant Mixture' % d)
+        ax.set_xlim((-0.01,1.01))
[…]

### 2014-01-07 results/2013-11-22_HCC1187_PAXIP1_genome_Qsd_0_1/MuBarPlot.py
Delete folders no longer needed

@@ -1,133 +1,0 @@
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import h5py
-import logging
-import pdb
-import scipy.stats as ss
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd271 as rvd27
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(message)s')
-
-def main():
-    #model samples
-    controlFile = "Control.hdf5"
-    caseFile = "Case.hdf5"
-
-    hdf5name = 'HCC1187_call.hdf5'
-    
-    try:
-        with h5py.File(hdf5name, 'r') as f:
-                call01 = f['call01'][...]
-                call00 = f['call00'][...]
-                call1 = f['call1'][...]
-                call2 = f['call2'][...]
-                call3 = f['call3'][...]
-                call4 = f['call4'][...]
-                call5 = f['call5'][...]
-                call6 = f['call6'][...]
-                call7 = f['call7'][...]
-                call8 = f['call8'][...]
-                controlMu = f['controlMu'][...]
-                caseMu = f['caseMu'][...]
-                caseN_median = f['caseN_median'][...]
-                controlN_median = f['controlN_median'][...]
-                loc = f['loc'][...]
-                f.close()
-    except IOError as e:
-        
-        diffalpha = 0.95
-        diffroi = (0,np.inf)
-        N = 1000
-        outputFile = 'control_case_diff_plus.vcf'
-        [loc, call00, controlMu, caseMu, controlN, caseN, postP, chi2P ]= rvd27.diff_test(0.95, controlFile, caseFile, \
-                                                                     diffroi, N, outputFile = 'control_case_diff_plus.vcf')
-        ## median read depth
-        caseN_median = np.median(caseN)
-        controlN_median = np.median(controlN)
-        
-        diffroi = (-np.inf, 0)
-        [_,call01,_,_,_,_,_,_] = rvd27.diff_test(0.95, controlFile, caseFile, \
[…]

### 2014-01-07 results/2013-11-22_HCC1187_PAXIP1_hg19masked_Qsd_0_1/MuBarPlot.py

@@ -1,133 +1,0 @@
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import h5py
-import logging
-import pdb
-import scipy.stats as ss
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd271 as rvd27
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(message)s')
-
-def main():
-    #model samples
-    controlFile = "Control.hdf5"
-    caseFile = "Case.hdf5"
-
-    hdf5name = 'HCC1187_call.hdf5'
-    
-    try:
-        with h5py.File(hdf5name, 'r') as f:
-                call01 = f['call01'][...]
-                call00 = f['call00'][...]
-                call1 = f['call1'][...]
-                call2 = f['call2'][...]
-                call3 = f['call3'][...]
-                call4 = f['call4'][...]
-                call5 = f['call5'][...]
-                call6 = f['call6'][...]
-                call7 = f['call7'][...]
-                call8 = f['call8'][...]
-                controlMu = f['controlMu'][...]
-                caseMu = f['caseMu'][...]
-                caseN_median = f['caseN_median'][...]
-                controlN_median = f['controlN_median'][...]
-                loc = f['loc'][...]
-                f.close()
-    except IOError as e:
-        
-        diffalpha = 0.95
-        diffroi = (0,np.inf)
-        N = 1000
-        outputFile = 'control_case_diff_plus.vcf'
-        [loc, call00, controlMu, caseMu, controlN, caseN, postP, chi2P ]= rvd27.diff_test(0.95, controlFile, caseFile, \
-                                                                     diffroi, N, outputFile = 'control_case_diff_plus.vcf')
-        ## median read depth
-        caseN_median = np.median(caseN)
-        controlN_median = np.median(controlN)
-        
-        diffroi = (-np.inf, 0)
-        [_,call01,_,_,_,_,_,_] = rvd27.diff_test(0.95, controlFile, caseFile, \
[…]

### 2014-01-07 results/2013-11-22_one_replicates_T_zero_with_chi2_Qsd_0_01/optimalT_0.py

@@ -1,41 +1,0 @@
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import h5py
-import logging
-import pdb
-import scipy.stats as ss
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(message)s')
-
-def main():
-    dilutionList = (0.1,0.3,1.0,10.0,100.0)
-
-    folderList = ('2013-11-22_six_replicates_synthetic_avg10_Qsd_0_01',\
-                  '2013-11-22_six_replicates_synthetic_avg100_Qsd_0_01',\
-                  '2013-11-22_six_replicates_synthetic_avg1000_Qsd_0_01',\
-                  '2013-11-22_six_replicates_synthetic_avg10000_Qsd_0_01')
-
-    for d in dilutionList:
-        logging.debug("Processing dilution: %0.1f%%" % d)
-    	for f in folderList:
-            path='vcf/%s' % str(10**(folderList.index(f)+1))
-            if not os.path.exists(path):
-                os.makedirs(path)
-            controlFile = "../%s/Control.hdf5" %f
-            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
-            caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
-            
-            outputFile='%(path)s/vcf%(dilution)s' %{'path':path,'dilution':str(d).replace('.','_')}
-           
-            rvd27.test(controlFile,caseFile,T=0,N=1000,outputFile=outputFile)
-if __name__ == '__main__':
-    main()
-

### 2014-01-07 results/2013-11-22_one_replicates_T_zero_with_chi2_Qsd_0_1/optimalT_0.py

@@ -1,41 +1,0 @@
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import h5py
-import logging
-import pdb
-import scipy.stats as ss
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(message)s')
-
-def main():
-    dilutionList = (0.1,0.3,1.0,10.0,100.0)
-
-    folderList = ('2013-11-22_six_replicates_synthetic_avg10_Qsd_0_1',\
-                  '2013-11-22_six_replicates_synthetic_avg100_Qsd_0_1',\
-                  '2013-11-22_six_replicates_synthetic_avg1000_Qsd_0_1',\
-                  '2013-11-22_six_replicates_synthetic_avg10000_Qsd_0_1')
-
-    for d in dilutionList:
-        logging.debug("Processing dilution: %0.1f%%" % d)
-    	for f in folderList:
-            path='vcf/%s' % str(10**(folderList.index(f)+1))
-            if not os.path.exists(path):
-                os.makedirs(path)
-            controlFile = "../%s/Control.hdf5" %f
-            caseFile = "Case%s.hdf5" % str(d).replace(".","_")
-            caseFile = "../%(folder)s/%(file)s" %{'folder':f,'file':caseFile}
-            
-            outputFile='%(path)s/vcf%(dilution)s' %{'path':path,'dilution':str(d).replace('.','_')}
-           
-            rvd27.test(controlFile,caseFile,T=0,N=1000,outputFile=outputFile)
-if __name__ == '__main__':
-    main()
-

### 2014-01-07 results/2013-11-22_one_replicates_optT_with_chi2_Qsd_0_1/optimalT.py

@@ -1,235 +1,0 @@
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import h5py
-import logging
-import pdb
-import scipy.stats as ss
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(message)s')
-
-def main():
-    dilutionList = (0.1,0.3,1.0,10.0,100.0)
-
-    folderList = ('2013-11-22_one_replicates_synthetic_avg10_Qsd_0_1',\
-                  '2013-11-22_one_replicates_synthetic_avg100_Qsd_0_1',\
-                  '2013-11-22_one_replicates_synthetic_avg1000_Qsd_0_1',\
-                  '2013-11-22_one_replicates_synthetic_avg10000_Qsd_0_1')
-    outputFile='optimalT.hdf5'
-    try:
-        with h5py.File(outputFile, 'r') as f:
-                cov = f['cov'][...]
-                T_ED = f['T_ED'][...]
-                T_L1 = f['T_L1'][...]
-                T_MCC = f['T_MCC'][...]           
-                f.close()
-                logging.debug('optimalT.hdf5 exits already')
-    except IOError as e:
-        logging.debug('Generate optimal dataset...')
-        N=1000 # Z sampling size  
-        chi2=True
-        times=8
-        T_ED=np.zeros((times,len(dilutionList),len(folderList)))
-        T_L1=np.zeros((times,len(dilutionList),len(folderList)))
-        T_MCC=np.zeros((times,len(dilutionList),len(folderList)))
-        cov=np.zeros((times,len(dilutionList),len(folderList)))
-        for i in xrange(times):
-            logging.debug('Runing the %(i)d times out of %(total)d...' % {'i':i+1, 'total':times})
-            for d in dilutionList:
-                logging.debug("Processing dilution: %0.1f%%" % d)
-                for f in folderList:
-                    path='%s' % str(10**(folderList.index(f)+1))
-##                    pdb.set_trace()
-                    controlFile = "../%s/Control.hdf5" %f
[…]

### 2014-01-07 results/2013-11-22_one_replicates_synthetic_avg10000_Qsd_0_01/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd273 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_one_replicates_synthetic_avg1000_Qsd_0_01/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd273 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_one_replicates_synthetic_avg1000_Qsd_0_1/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd272 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_one_replicates_synthetic_avg100_Qsd_0_1/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd272 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/100/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_one_replicates_synthetic_avg10_Qsd_0_01/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd273 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_one_replicates_synthetic_avg10_Qsd_0_1/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd272 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_six_replicates_synthetic_avg10000_Qsd_0_1/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd272 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p1.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_six_replicates_synthetic_avg1000_Qsd_0_1/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd272 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p1.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-11-22_six_replicates_synthetic_avg100_Qsd_0_1/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd272 as rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p1.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/100/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-12-05__HCC1187_PAXIP1_genome_Qsd_two_end_bounded/runall.py

@@ -1,59 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import random
-
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../../rvd2/src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-
-pool = mp.Pool(processes=5)
-#pool = None
-
-gibbs_nsample = 4000
-mh_nsample = 5
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-def fit_model(fileNameList, h5FileName):
-    fileList = ["depth_chart/%s" % filename for filename in fileNameList]
-    (r, n, loc, refb) = rvd27.load_depth(fileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-       refb=refb)
-
-def main():
-    random.seed(199096)
-    # Estimate the model for the cases
-    logging.debug("Processing control data.")
-    controlList = ['HCC1187BL_S1.dc',]
-    fit_model(controlList, 'Control.hdf5')
-
-
-    # Estimate the model for the cases
-    logging.debug("Processing case data.")
-    caseList = ['HCC1187C_S1.dc',]
-    fit_model(caseList, 'Case.hdf5')
-
-if __name__ == "__main__":
-	main()
-

### 2014-01-07 results/2013-12-05_one_replicates_synthetic_avg10000_Qsd_0_1 _bounded/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-12-05_one_replicates_synthetic_avg1000_Qsd_0_1 _bounded/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-12-05_one_replicates_synthetic_avg100_Qsd_0_1 _bounded/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/100/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-12-05_one_replicates_synthetic_avg10_Qsd_0_1 _bounded/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p3.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-12-05_six_replicates_synthetic_avg1000_Qsd_0_1_bounded/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p1.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/1000/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 results/2013-12-05_six_replicates_synthetic_avg10_Qsd_0_1_bounded/runall.py

@@ -1,66 +1,0 @@
-# -*- coding: utf-8 -*-
-# <nbformat>3.0</nbformat>
-
-# <codecell>
-
-import sys
-import os
-import numpy as np
-import matplotlib.pyplot as plt
-import pandas as pd
-import h5py
-import multiprocessing as mp
-import logging
-import pdb
-# <codecell>
-
-logging.basicConfig(level=logging.DEBUG,
-                    format='%(levelname)s:%(module)s:%(message)s')
-
-# Insert the src/python/rvd27 directory at front of the path
-rvddir = os.path.join('../../src/python/rvd27')
-sys.path.insert(0, rvddir)
-import rvd27
-
-##pool =None
-pool = mp.Pool(processes=10)
-tocfilename = "../../data/synthetic_toc_p1.txt"
-toc = pd.read_table(tocfilename)
-
-gibbs_nsample = 4000
-mh_nsample = 10
-
-logging.debug("Gibbs step size=%d" % gibbs_nsample)
-logging.debug("mh sample size=%d" % mh_nsample)
-
-# Estimate the model for the cases
-logging.debug("Processing control data.")
-h5FileName = "Control.hdf5"
-try:
-    with h5py.File(h5FileName, 'r') as f:
-        pass
-except IOError as e:
-    controlFileList = ["../2013-08-06_Downsample_Read_Depth/depth_chart/10/%s" % filename for filename in toc.Filename[toc.isRef=='Y']]
-    (r, n, loc, refb) = rvd27.load_depth(controlFileList)
-    phi, theta_s, mu_s = rvd27.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
-    logging.debug("Saving model in %s" % h5FileName)
-    rvd27.save_model(h5FileName, phi, mu=mu_s, theta=theta_s, r=r, n=n, loc=loc,
-           refb=refb)
-
-# Estimate the model for the cases
-for dilution in np.unique(toc[toc.isRef=='N'].Dilution):
-    logging.debug("Processing dilution: %0.1f" % dilution)
-    
-    h5FileName = "Case%s.hdf5" % str(dilution).replace(".", "_", 1)
-
-    try:
-        with h5py.File(h5FileName, 'r') as f:
-            pass
-    except IOError as e:
[…]

### 2014-01-07 doc/bioinformatics2014/document.tex
Add the command line citation in to the manuscript

@@
 For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
-Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
+Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja,cushing2013rvd, Ghedin:2010ie}.
 DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.

### 2014-01-07 doc/ISMB2014/RVD3_1.tex

--- 2013-12-30 doc/ISMB2014/RVD3.tex
@@
-\documentclass[11pt,reqno]{amsart}
-\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
-\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
-%\usepackage{algorithm, algorithmic}
+\documentclass{bioinfo}
+\copyrightyear{2005}
+\pubyear{2005}
+\usepackage{amsmath}
+\usepackage{afterpage}
+\usepackage{algorithm}
 
-\usepackage{algorithm}
-\usepackage{algpseudocode}
-
-\usepackage{graphicx}
-
-\usepackage{verbatim}
-\usepackage{amssymb}
-\usepackage{amsmath}
-
-\usepackage{enumitem}
-
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{natbib}
-
-%\usepackage{epstopdf}
-%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
+%\usepackage{placeins}
+\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{xr}
+\externaldocument{supplementary_info}
+\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}
 
@@
 \DeclareMathOperator{\diag}{diag}
-
-\algnewcommand{\Inputs}[1]{%
-  \State \textbf{Inputs:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
-\algnewcommand{\Initialize}[1]{%
-  \State \textbf{Initialize:}
-  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
-}
+\begin{document}
+\firstpage{1}
 
 \title[RVD3]{Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data}
-\author{}
-%\date{}                                           % Activate to display a given date or no date
+\author[Sample \textit{et~al}]{Corresponding Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^2$\footnote{to whom correspondence should be addressed}}
+\address{$^{1}$Department of XXXXXXX, Address XXXX etc.\\
+$^{2}$Department of XXXXXXXX, Address XXXX etc.}
 
-\begin{document}
+\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
+
+\editor{Associate Editor: XXXXXXX}
+
+\maketitle
 
@@
 
+\section{Motivation:}
+Including prior information in the Bayesian approach is difficult but meaningful.
+
+\section{Results:}
 We analyzed the Sensitivity of the Bayesian model to different priors. Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
 
+\section{Availability:}
+
+\section{Contact:} \href{name@bio.com}{name@bio.com}
 \end{abstract}
 
-\maketitle
+\section{Introduction}
+Massively parallel sequencing data were generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
+One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants (SNV) candidates.
 
-%%%%%%%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-%%%%%%%%%%%%%%%%%%
-% Data Sets
-%%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences differ only 14 single nucleotide positions.
-Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
-The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We used BWA v0.7.5a to align the reads to the reference sequence with the -C50 option to remove the reads of high mapping quality.
-BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104.
-The final data set contains read pairs for $N=6$ replicates for the control at different MAF levels.
-
-\subsection{Yeast Data}
-We first mapped the GSY1135 to Chromosome 10 in S288c reference genome by BWA v0.7.5a [XXX].
-Then called SNPs by GATK UnifiedGenotyper and created a FASTA GSY1135 reference using GATK FastaAlternative [XXX].
-Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from [XXX], and removed WT population using FASTX Barcode Splitter [XXX] and trimed the Nextera tag by Cutadapt v1.3.
-Finally, the FASTQ files of case and control were mapped to GSY1135 reference and pileup files were generated using Samtools v0.1.19.
+Currently SNV are deduced on NGS platform by variants calling algorithms.
+Typically, SAMtools, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, VarScan2 \citep{Koboldt:2012cga}, Strelka \citep{Saunders:2012fh}, and muTect \citep{Cibulskis:2013ta} are all highly used in detecting low-frequency variants.
+However identifying true SNVs remains challenging because of high false positive rates or high false discovery rate.
+Another statistical method-RVD \citep{Flaherty:2011ja}, was proposed based on beta-binomial model for ultrasensitive rare single nucleotide variant detection.
+So based on the RVD hierarchical model we built an improved robustness and accuracy model for low-depth variants detection.
 
@@
 \subsection{Model Structure}\label{sec:model_structure}
-RVD3 is based on a hierarchical Bayesian model.
-Through hypothesis test on case and control samples by RVD3, we can call the variants sucsessfully.
-Now the definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
-Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions; priors for $M_j$, two kinds of priors are log-normal prior and Jeffreys prior, which enhanced the former RVD model [XXX].
+RVD is based on a two-stage hierarchical Bayesian model for variant detection \citep{Flaherty:2011ja}.
+Through hypothesis test on case and control samples by RVD, we can call the variants successfully.
+Now RVD3 has three-stage model including priors built on the former RVD model.
+The definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
+Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions; two kinds of priors for $M_j$ are log-normal prior and Jeffreys prior \citep{jeffreys1946invariant}, which enhanced the former RVD model \citep{Flaherty:2011ja}.
 Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$.
-The graphical chart for RVD3 are shown in Figure~\ref{fig:graphical_model}.
-RVD3 hierarchically includes four levels of samplings: $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
+The graphical chart for RVD3 is shown in Figure~\ref{fig:graphical_model}.
+
+RVD3 hierarchically includes three levels of samplings: $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
 $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation caused by experimental repeatability.
 The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and $\pi\left({M}_{j}\right)$ (Jeffreys prior).
+And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and Jeffreys prior for $M_j$.
+%$\pi\left({M}_{j}\right)$
 
@@
 \begin{center}
-\includegraphics[width=100mm]{figs/RVD3_model.pdf}
+\includegraphics[width=85mm]{figs/RVD3_model.pdf}
 \caption{RVD3 Graphical Model}
@@
 
+Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are also illustrated.
+
 \begin{algorithm}[ht]
@@
 \For {each location j}
-  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H \Comment{Sample $\mu_j$}
-  \State Set $\mu_j$ to the sample median for the T samples
-  \State Draw T samples from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$ using M--H \Comment{Sample $M_j$}
+  \State Samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ \Comment{Sample $\mu_j$}
+  \State Set $\mu_j$ to the sample median for the samples
+  \State Samples from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$ \Comment{Sample $M_j$}
 
@@
 
-Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are also illustrated.
-
 %%%%%%%%%%%%%
@@
 
-We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are
+The initial parameter estimates and derivations are provided in Appendix~\ref{sec:appendix_mom}. Below is the MoM estimate for replicate-level parameters
 $\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
@@
 $\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
-While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
 
@@
 
-Samples from the posterior distribution
-$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$
-are drawn analytically because of the Bayesian conjugacy between the prior
+Because the Bayesian conjugacy between the prior
 $p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$
 and the likelihood
-$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
+$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$,
+we draw the samples from the posterior distribution
+$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$
 The posterior distribution is
@@
 \subsubsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
-The posterior distribution over $\mu_j$ given its Markov blanket is
+Based on Markov blanket, the posterior distribution over $\mu_j$ is
 \begin{equation}
@@
 
-Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution.
-Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
+Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
+By experience when $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$, the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration is set to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ ; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
+We have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
 
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
-Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-
-We resampled from the proposal if the sample is outside of the support of the posterior distribution.
+(Copy) We resampled from the proposal if the sample is outside of the support of the posterior distribution.
 We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
@@
 \subsubsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
-For Jeffreys prior is from the Fisher information and ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$,
+Since Jeffreys prior is from the Fisher information and in RVD3 model ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$,
 
@@
 
-We calculated the equations (Appendix B) and obtained the Jeffreys' prior for $M_j$:
+We calculated the equations Appendix~\ref{sec:appendix_Jeffreys} and obtained the Jeffreys' prior for $M_j$:
 
@@
 We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$.
-Since we cannot provide an intepretive form for it, so Metropolis-Hastings algorithm was used to sample from the posterior distribution.
+Since we cannot provide an interpretive form for it, Metropolis-Hastings algorithm was still used to sample from the posterior distribution.
 
@@
 \subsubsection{Posterior Density Test}\label{sec:hypothesis_test}
-Posterior distributions of $\mu_j$ for the control and case are achieved-  $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$, by Metropolis-within-Gibbs.
+Posterior distributions of $\mu_j$ for the control and case are achieved -  $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$, by Metropolis-within-Gibbs.
 So we called a variant when $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with $1-\alpha$ confidence,
@@
 %%%%%%%%%%%%%%%%%%
-% Limited in the 9 pages, so maybe considering neglect the Chi^2 Test since we show the results without Chi^2 Test.
+\subsubsection{$\chi^2$ test for non-uniform base distribution}
+($\chi^2$ test is totally Copied!)
+An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios.
+We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
+In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
+
+We use a $\chi^2$ goodness-of-fit test on a multinomial distribution over the non-reference bases to distinguish these two possible scenarios.
+The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
+Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
+The test statistic is
+
+\begin{equation}
+ 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
+\end{equation}
+
+where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
+\citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
+
+We control for multiple hypothesis testing in two ways.
+We use Fisher's combined probability test \citep{fisher1970statistical} to combine the p-values for $N$ replicates into a single p-value at position $j$,
+
+\begin{equation}\label{eqn:fisher_combined}
+	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
+\end{equation}
+
+Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
+Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
 
@@
 %\subsubsection{Improper Prior}
-Playing no prior on $ M_{j}$ is an implicit improper prior.
-An improper prior integrates to infinity, and may cause an improper posterior which results in an invalid inferences [XXX].
-Because an improper prior is a bad choice since a step has to be taken to confirm the posterior is proper.
-We considered about two typical priors: non-information prior - Jeffreys prior, and information prior - log-normal prior.
+The prior distribution characterizes the knowledge of the parameters in the model.
+An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}.
+Furthermore when the Markov chain Monte Carlo method is taken to derive the posterior, it is possibly hard to sniff out the improper posterior.
+Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}.
+Playing no prior on $ M_{j}$ is exactly an implicit improper prior.
+Therefore we considered about non-information prior and information prior.
 
 %\subsubsection{Jeffreys Prior}
-Jeffreys prior is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys (1946).
+Various noninformative prior distributions have been suggested for parameters in hierarchical models.
+Jeffreys prior, as a typical and influential one, is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys \citep{jeffreys1946invariant}.
 It is defined in terms of the Fisher information and works well with a single parameter.
 In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
-Log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
-So log-normal prior was incorporated in the model.
+
+A good informative prior is imperative to promote accurate posterior estimates.
+In our study, log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
 The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
+
+%%%%%%%%%%%%%%%%%%
+% Data Sets
+%%%%%%%%%%%%%%%%%%
+\section{Data Sets}
+
+\subsection{Synthetic DNA Sequence Data}
+
+Two 400bp DNA sequences differ only 14 single nucleotide positions.
+Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
+The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+We used BWA v0.7.5a to align the reads to the reference sequence with the -C50 option to remove the reads of high mapping quality.
+BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104 (http://picard.sourceforge.net).
+The final data set contains read pairs for $N=6$ replicates for the control at different MAF levels.
+
+\subsection{Yeast Data}
+We first mapped the wild-tpye strain GSY1135 \citep{kvitek2011reciprocal} to Chromosome 10 in S288c reference genome (SGD; http://www.yeastgenome.org/) by BWA v0.7.5a \citep{li2009fast}.
+Then called SNPs by GATK v2.5 UnifiedGenotyper \citep{McKenna:2010bva, depristo2011framework} and created a FASTA GSY1135 reference using GATK FastaAlternative.
+Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from \citep{kvitek2013whole}, and removed WT population using FASTX Barcode Splitter and trimed the Nextera tag by Cutadapt v1.3.
+Finally, the FASTQ files of case and control were mapped to the corresponding reference genome created before.
+And pileup files were generated using Samtools v0.1.19 and depth chart file were derived by randomly down sampling.
 
@@
 There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix C.
+Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix~\ref{sec:appendix_parameters}.
 Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/M_jeffrey.pdf}
-\caption{Key parameters for RVD3 model with Jeffreys prior for synthetic DNA data sets.}
+\includegraphics[width=90mm]{figs/M_jeffrey.pdf}
+\caption{Key parameters for RVD3 model with Jeffreys prior on synthetic DNA data sets.}
 \label{fig:M_jeffrey}
@@
 
-\begin{figure}[htbp]
+\begin{figure*}[!tbp]
 \begin{center}
-\includegraphics[width=170mm]{figs/post_prior_dilu_10.pdf}
+\includegraphics[width=190mm]{figs/post_prior_dilu_10.pdf}
 \caption{Distribution of priors and posteriors when dilution is 10\%}
@@
 \end{center}
-\end{figure}
+\end{figure*}
 
@@
 Figure~\ref{fig:dilu_10} shows the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
-The positions are chosen in the middle of the base length- position 100 and 300 are mutant, and position 104 and 244 are non-mutant.
-From the posterior probability distributions are estimated by Gaussian Kernel Density [XXX].
+The positions are chosen in the middle of the base length- position 104 and 244 are mutant, and position 100 and 300 are non-mutant.
+From the posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
 From the figure non-mutant positions display normal and stable without a peak nor a strange shape.
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/ROC_without_chi2_jeffrey.pdf}
+\includegraphics[width=90mm]{figs/ROC_without_chi2_jeffrey.pdf}
 \caption{ROC curve for variants detection performance by Jeffreys prior.}
@@
 \begin{center}
-\includegraphics[width=120mm]{figs/ROC_without_chi2_lognormal.pdf}
+\includegraphics[width=90mm]{figs/ROC_without_chi2_lognormal.pdf}
 \caption{ROC curve for variants detection performance by log-normal prior.}
 \label{fig:ROC_lognormal}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=160mm]{tables/Sen_Speci.pdf}
-\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
-\label{tbl:SS}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=160mm]{tables/FDR.pdf}
-\caption{False Discovery Rate comparison of RVD3 with different priors.}
-\label{tbl:FDR}
 \end{center}
@@
 
-Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
+Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
 Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{tbl:FDR} shows the false discovery rate of the RVD3 with different priors.
+Figure~\ref{fid:FDR} shows the false discovery rate of the RVD3 with different priors.
 Jeffreys prior shows a smaller false discovery rate and higher accuracy than others.
@@
 Here we chose Jeffreys prior model because it's a non-information prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=80mm]{tables/Sen_Speci.pdf}
+\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
+\label{fig:SS}
+\end{center}
+\end{figure}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=80mm]{tables/FDR.pdf}
+\caption{False Discovery Rate comparison of RVD3 with different priors.}
+\label{fig:FDR}
+\end{center}
+\end{figure}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%
-% Discussion
-%%%%%%%%%%%%
 \section{Discussion}
+
+Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that the above method  Text
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+%     as it may ignore the figures.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Conclusion}
+
+(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
+Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
+\citealp{Boffelli03} might want to know about  text text text text
+
+\begin{enumerate}
+\item this is item, use enumerate
+\item this is item, use enumerate
+\item this is item, use enumerate
+\end{enumerate}
+
+\section*{Acknowledgement}
+Text Text
+
+\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
+
+\bibliographystyle{natbib}
+%\bibliographystyle{achemnat}
+%\bibliographystyle{plainnat}
+%\bibliographystyle{abbrv}
+%\bibliographystyle{bioinformatics}
+%\bibliographystyle{plain}
+
+\bibliography{bioinfo}
 
@@
 %%%%%%%%%%%%%%%%%%%
-\section{Inference of Jeffreys prior}\label{sec:appendix_mom}
+\section{Inference of Jeffreys prior}\label{sec:appendix_Jeffreys}
 We assume there is only one replicate,
@@
 %%%%%%%%%%%%%%%%%%%
-\section{Key parameters for RVD3 model with log-normal prior}\label{sec:appendix_mom}
+
+\section{Key parameters for RVD3 model with log-normal prior}\label{sec:appendix_parameters}
+Key parameters of the RVD3 with log-normal prior is ploted in Figure~\ref{fig:M_lognormal}.
+
 \begin{figure}[htbp]
 \begin{center}
-\includegraphics[width=120mm]{figs/M_lognormal.pdf}
+\includegraphics[width=90mm]{figs/M_lognormal.pdf}
 \caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
@@
 
-\bibliographystyle{apalike}
-\bibliography{bioinfo}
 \end{document}

### 2014-01-09 doc/bioinformatics2014/document.tex
update manuscript in bioinformatics2014

--- 2014-01-07 doc/bioinformatics2014/document.tex
@@
 For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
-Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
+Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, cushing2013rvd, Ghedin:2010ie}.
 DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
@@
 The remaining one is a homozygous somatic variant.
+Because the median read depth for the case sample is 60, we set the threshold of the test for $\mu^\Delta$ to $\tau = 16\%$ to minimize the false discovery rate (Figure~\ref{tbl:comparison_fdr}).
 
-%\iffalse
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF.
+Positions chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF.
 Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
@@
 In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T.
-Through manual checking, position chr7:154743899C$>$T and chr7:154753635T$>$C are likely false positives due to insufficient sampling in the Metropolis\_within-Gibbs algorithm.
 Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
@@
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-%\fi
 
@@
 \includegraphics[width=0.5\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in PAXIP1 gene.}
+\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}} - \mu^{\text{control}} \geq \tau$.}
 \label{fig:brca_MAF}
@@
 %('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
+%TODO: add star to test that have mudelta > .16 Done.
+%TODO: set tau=0.16 for mudelta test.  Done
 

### 2014-01-09 results/2013-12-20_experiment_set_gibbs_Qsd_mu_1_mu_over_10_minus_mu0/HCC1187_PAXIP1_genome_600/MuBarPlot.py
update Mubarplot program to add a start on positions passed posterior density difference test.

@@ -10,7 +10,7 @@ import scipy.stats as ss
 # Insert the src/python/rvd27 directory at front of the path
 rvddir = os.path.join('../../../src/python/rvd27')
 sys.path.insert(0, rvddir)
-import rvd271 as rvd27
+import rvd27
 
 logging.basicConfig(level=logging.DEBUG,
                     format='%(levelname)s:%(message)s')
@@ -44,7 +44,7 @@ def main():
     except IOError as e:
         
         diffalpha = 0.95
-        diffroi = (0.00001,np.inf)
+        diffroi = (0.16,np.inf)
         N = 1000
         outputFile = 'control_case_diff_plus.vcf'
         [loc, call00, controlMu, caseMu, controlN, caseN, postP, chi2P ]= rvd27.diff_test(0.95, controlFile, caseFile, \
@@ -53,7 +53,7 @@ def main():
         caseN_median = np.median(caseN)
         controlN_median = np.median(controlN)
         
-        diffroi = (-np.inf, 0)
+        diffroi = (-np.inf, -0.16)
         [_,call01,_,_,_,_,_,_] = rvd27.diff_test(0.95, controlFile, caseFile, \
                                                 diffroi, N, outputFile = 'control_case_diff_minus.vcf')
         Interval=[[0.0,0.2],[0.1,0.9],[0.8,1.0]]
@@ -121,10 +121,13 @@ def MuBarPlot(CallControlMu,CallCaseMu,Loc,title):
         ax.set_xlabel("HG19 Genomic Location [chr7:154,000,000+X]")
         ax.set_xticks(ind+width)
         
+        for i in xrange(3):
+            #ax.text(i+width+2-width/9.0, 1.0021,'*',color='b',fontsize=15)
+            ax.text(i+width+2-width/9.0, 1.0021,'*',fontsize=15)
+            
         Loclabel = [x.split(':')[1][3:] for x in Loc]
         rslabel = ['rs1239326','rs1239324','rs71534174','rs35505514',' ','rs4398858']
 
-        pdb.set_trace()
         label = ['%s\n%s' %(Loclabel[i], rslabel[i]) for i in xrange(len(Loclabel))]
 
         ax.set_xticklabels(label)
@@ -134,7 +137,6 @@ def MuBarPlot(CallControlMu,CallCaseMu,Loc,title):
                    loc=4,ncol=1, mode="expand", borderaxespad=0., prop={'size':12})
[…]

### 2014-01-09 doc/bioinformatics2014/document.tex
update manucript

@@
 \section{Availability:}
-Text
+\href{http://genomics.wpi.edu/rvd2/rvd2.html}{http://genomics.wpi.edu/rvd2/rvd2.html}
 
-\section{Contact:} \href{pjflaherty@wpi.edu}{pjflaherty}
+\section{Contact:} \href{pjflaherty@wpi.edu}{pjflaherty@wpi.edu}
 \end{abstract}
@@
 For example, NGS-based non-invasive fetal DNA testing uses maternal blood sample to sequence the minority fraction of cell-free fetal DNA~\citep{Fan:2008di}.
-Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, cushing2013rvd, Ghedin:2010ie}.
+Infectious diseases such as HIV and influenza may contain many genetically heterogeneous sub-populations~\citep{Flaherty:2011ja, Ghedin:2010ie}.
 DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
@@
 muTect uses a Bayesian posterior probability in its decision rule to evaluate the likelihood of a mutation~\citep{Cibulskis:2013ta}.
-RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{}.
+RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}.
 However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.
@@
 \citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
-We test their performance on much lower MAF variants and across a wider range of coverage depths.
+%We test their performance on much lower MAF variants and across a wider range of coverage depths.
 
@@
 
-\paragraph{Funding\textcolon}
+\paragraph{Funding\textcolon}  P.F. and Y.E. were supported by seed funding from Worcester Polytechnic Institute.
 

### 2014-01-13 doc/GLBIO2014/GLBIO2014_Abstract_Poster.tex
GLBIO2014 Abstract Draft.

--- 2013-12-19 doc/ISMB2014/RVD3.tex
@@
 \author{}
-%\date{}                                           % Activate to display a given date or no date
-
+%\date{}                                           
 \begin{document}
@@
 
-We analyzed the Sensitivity of the Bayesian model to different priors. Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
+Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
+To address the detection of SNVs, a variety of methods are being under-represented.
+However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.
+
+We propose a novel empirical Bayesian model to call variants accurately.
+And the sensitivity of this hierarchical model is analyzed by the improper prior, information prior and non-information prior on the synthetic sequence data with varying read depth and a range of allele frequencies.
+Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
+Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
+For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
+Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
 
@@
 \maketitle
-
-%%%%%%%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-%%%%%%%%%%%%%%%%%%
-% Data Sets
-%%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
-More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
-
-To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
-The final data set contains read pairs for three replicates of each case and pairs of reads three replicates for the control sample giving $N=6$ replicates for the control and each MAF level.
-
-\subsection{Yeast Data}
-
-%%%%%%%%%%%%%%%%%%
-% RVD3 Model
-%%%%%%%%%%%%%%%%%%
-\section{RVD3 Model}
-
-\subsection{Model Structure}\label{sec:model_structure}
-RVD3 is based on a hierarchical Bayesian model.
-Through hypothesis test on case and control samples by RVD3, we can call the variants sucsessfully.
-Now the definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
-Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positioins; priors for $M_j$, two kinds of priors are log-normal prior and Jeffreys prior, which enhanced the former RVD model [XXX].
-Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$.
-The graphical chart for RVD3 are shown in Figure~\ref{fig:graphical_model}.
-RVD3 hierarchically includes four levels of samplings: $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
-$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation caused by experimental repeatability.
-The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and $\pi\left({M}_{j}\right)$ (Jeffreys prior).
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=120mm]{figs/RVD3_model.pdf}
-\caption{RVD3 Graphical Model}
-\label{fig:graphical_model}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%
-% Inference & Hypothesis Testing
-%%%%%%%%%%%%%%%%%%
-\subsection{Inference and Hypothesis Testing}
-Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are illustrated.
-
-\begin{algorithm}[ht]
-\caption{Metropolis-within-Gibbs Algorithm}
-\label{alg:metro_gibbs}
-\begin{algorithmic}[1]
-
-\State Initialize $\theta$, $\mu$, $M$, $\mu_0$, $M_0$
-\Repeat
-\For {each location j}
-  \State Draw T samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ using M--H \Comment{Sample $\mu_j$}
-  \State Set $\mu_j$ to the sample median for the T samples
-  \State Draw T samples from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$ using M--H \Comment{Sample $M_j$}
-
-  \For {each replicate i} 
-	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$  \Comment{Sample $\theta_{ji}$}
-  \EndFor
-
-\EndFor
-\Until {sample size sufficient}
-\end{algorithmic}
-\end{algorithm}
-
-%%%%%%%%%%%%%
-% Initialization
-%%%%%%%%%%%%%
-\subsubsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
-MoM works by setting the population moment equal to the sample moment.
-A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
-We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.
-
-We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are
-$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
-The estimates for the position-level parameters are
-$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$
-and
-$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
-The estimates for the genome-level parameters are
-$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$
-and
-$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
-While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
-
-%%%%%%%%%%%%%%%%%%
-% Sampling theta
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
-
-Samples from the posterior distribution
-$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$
-are drawn analytically because of the Bayesian conjugacy between the prior
-$p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$
-and the likelihood
-$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
-The posterior distribution is
-\begin{equation}
-	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%
-% Sampling mu
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
-The posterior distribution over $\mu_j$ given its Markov blanket is
-\begin{equation}
-	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
-\end{equation}
-
-Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we cannot write an analytical form for the posterior distribution.
-Instead, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
-Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-
-We resample from the proposal if the sample is outside of the support of the posterior distribution.
-We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
-Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.
-
-%%%%%%%%%%%%%%%%%%
-% Sampling Mj
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
-For Jeffreys prior,since ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$, the Fisher information is given by
-
-\begin{equation}\label{equ:JefferyInference}
-I\left({M}_{j}\right)={E}_{{M}_{j}}\left[ -\frac{\delta ^{2}\log p\left(\theta _{j}|\mu_{j},M_{j}\right)}{\delta M^{2}_{j}}\right]
-\end{equation}
-
-Now to calculate the equations to acquire the Fisher information, we assume there is only one replicate,
-
-\begin{equation}\label{eqn:Betapdf}
-p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
-\end{equation}
-
-\begin{equation}\label{equ:JefferyInference1}
-\begin{split}
-\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
-& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
-\end{split}
-\end{equation}
-
-\begin{equation}
-\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
-\end{equation}
-
-\begin{equation}
-\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
-\end{equation}
-
-Now we have the Jeffreys' prior for $M_{j}$:
-
-\begin{equation}
-\pi\left({M}_{j}\right) = [-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
-\end{equation}
-
-For log-normal prior, the posterior distribution over $M_{j}$ given its Markov blanket is
-
-\begin{equation}
-	p( M_{j} |\mu, \sigma, \theta_{ji},\mu_j) \propto p(\theta_{ji} | \mu_j, M_j) p(M_{j} | \mu, \sigma)
-\end{equation}
-
-We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$.
-Since we cannot give an analytical form for this, so we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-
-%%%%%%%%%%%%%%%%%%
-% Posterior Density Test
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Posterior Density Test}\label{sec:hypothesis_test}
-Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
-For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-
-A variant is called if $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with high confidence,
-\begin{equation}\label{eqn:bayes_test}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
-\end{equation}
-where $\tau$ is a detection threshold and $1-\alpha$ is a confidence level.
-We draw a sample from the posterior distribution $\tilde{\mu}_j^{\Delta} \triangleq \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ by simple random sampling with replacement from $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.
-We set $\tau = 0$ in our experiment.
-
-%%%%%%%%%%%%%%%%%%
-% Chi^2 Test
-%%%%%%%%%%%%%%%%%%
-% Limited in the 9 pages, so maybe considering neglect the Chi^2 Test since we show the results without Chi^2 Test.
-
-%%%%%%%%%%%%%%%%%%%%%%%%
-% Priors for precision parameter
-%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Priors for precision parameter}
-\subsubsection{Improper Prior}
-Playing no prior on M is an implicit improper prior. So we coniderated about two typical priors: noninformation prior - Jeffreys prior, and information prior - log-normal prior.
-
-\subsubsection{Jeffreys Prior}
-Jeffreys prior is proposed for invariance by Harold Jeffreys (1946), and defined in terms of the Fisher information. For our ploblem,
-\begin{equation}\label{eqn:JefferyPrior}
-\pi\left({M}_{j}\right)= I\left({M}_{j}\right)^{\frac{1}{2}}
-\end{equation}
-
-\subsubsection{Log-normal Prior}
-In log-normal distribution, the parameters denoted $\mu$ and $\sigma$, the mean and standard deviation respectively. Log-normal prior was used on $ M_{j}$.
-Incorporate log-normal prior on $M_{j}$.
-
-%%%%%%%%%%%%%%%%%%
-% Results
-%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-%%%%%%%%%%%%%%%%%%
-% Sensitivity analysis of priors
-%%%%%%%%%%%%%%%%%%
-\subsection{Sensitivity analysis of priors}
-In order to examine the robustness of RVD3 to changes of different priors, we performed information prior (Log-normal) and noninformation prior (Jeffreys).
-We found that RVD3 is robust to changing the priors.
-The RVD3 algorithm provides estimates of model parameters and latent variables given the data.
-
-We show several of these parameters in Figure~\ref{fig:M_jeffrey} and Figure ~\ref{fig:M_lognormal}.
-The left column of these two figures show the read depth for each of the six bam files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{figs/M_jeffrey.pdf}
-\caption{Key parameters for RVD3 model with Jeffreys prior for synthetic DNA data sets.}
-\label{fig:M_jeffrey}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{figs/M_lognormal.pdf}
-\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
-\label{fig:M_lognormal}
-\end{center}
-\end{figure}
-
-The right column of Figure~\ref{fig:M_jeffrey} and Figure~\ref{fig:M_lognormal} show the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}].
-Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-$M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=170mm]{figs/prior_post_dilu_1.pdf}
-\caption{Distribution of priors and posteriors when dilution is 1\%}
-\label{fig:dilu_1}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=170mm]{figs/prior_post_dilu_10.pdf}
-\caption{Distribution of priors and posteriors when dilution is 10\%}
-\label{fig:dilu_10}
-\end{center}
-\end{figure}
-
-Statistically, We analyzed the posteriors and priors for Jeffreys prior and log-normal prior.
-Figure~\ref{fig:dilu_1} and Figure~\ref{fig:dilu_10} show the distribution over different M values focused on 4 different locations (position 100 and 300 are mutant, and position 104 and 244 are non-mutant).
-The positions are chosen in the middle of the dase length.
-From the posterior distributions, non-mutant positions display normal and stable without a peak nor a strange shape.
-The Jeffreys' prior shows more information than the lognormal prior (flat) from the posterior curve.
-From the prior distribution, it shows our model wants to search for a small value for M.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results of priors on synthetic data
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Results of priors on synthetic data}
-\subsubsection{Performance with read depth}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{figs/ROC_without_chi2_jeffrey.pdf}
-\caption{ROC curve for variants detection performance by Jeffreys prior.}
-\label{fig:ROC_jeffrey}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=120mm]{figs/ROC_without_chi2_lognormal.pdf}
-\caption{ROC curve for variants detection performance by log-normal prior.}
-\label{fig:ROC_lognormal}
-\end{center}
-\end{figure}
-
-To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
-Here the Bayesian test is used without the $\chi^2$ test.
-Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
-The performance improves when the read depth goes up.
-(ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
-Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model without priors for $M_J$, which definitely demonstrates the advantage of priors.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=160mm]{tables/Sen_Speci.pdf}
-\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
-\label{tbl:SS}
-\end{center}
-\end{figure}
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=160mm]{tables/FDR.pdf}
-\caption{False Discovery Rate comparison of RVD3 with different priors.}
-\label{tbl:FDR}
-\end{center}
-\end{figure}
-
-\subsubsection{Sensitivity/Specificity/FDR}
-
-Figure~\ref{tbl:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model without priors.
-Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{tbl:FDR} shows the false discovery rate of the RVD3 with different priors.
-Jeffreys prior shows a smaller false discovery rate and higher accuracy than others.
-It is obvious no matter which pirors, the variant detection performance acquires lower FDR than the model without priors.
-Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropariate choice for the precision prior.
-Here we chose Jeffreys prior because it needs more attention to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results of Jeffreys prior on yeast data
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Results of Jeffreys prior on yeast data}
-
-We demonstrated our RVD3 model with Jeffreys piror on yeast data to identify the variants.
-
-%%%%%%%%%%%%
-% Discussion
-%%%%%%%%%%%%
-\section{Discussion}
-
-\appendix
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix A: Parameter Initialization
-%%%%%%%%%%%%%%%%%%%
-\section{Parameter Initialization}\label{sec:appendix_mom}
-Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
-Therefore the MoM estimator is
-\begin{equation}
-	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
-\end{equation}
-
-We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
-The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
-The first and second population moments are
-\begin{eqnarray}
-	E[\theta_{ji}] =& \mu_j,\\
-	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
-\end{eqnarray}
-The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
-Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
-\begin{eqnarray}
-	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
-	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
-\end{eqnarray}
-
-Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
-\begin{eqnarray}
-	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
-	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
-\end{eqnarray}
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix B: Comparison Statistics
-%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\bibliography{bioinfo}
 \end{document}

### 2014-01-13 doc/variational_notes.tex
Some refactoring of docs folder.

--- 2013-02-22 doc/notes.tex
@@
 \end{multline}
-
+C
 The computation of $p(r;u_0,\sigma_0^2,M_0)$ is intractable due to the coupling in the latent variables induced by the integral so we must use an approximation or sampling method.

### 2014-01-13 doc/2014ismb_he/document.tex
refactor docs folder.

--- 2014-01-09 doc/bioinformatics2014/document.tex
@@
 We present the initial parameter estimates here and provide the derivations in Supplementary Information. The MoM estimate for replicate-level parameters are
-$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
+$\hat{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
 The estimates for the position-level parameters are
-$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$
+$\hat{\mu}_j = \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}$
 and
-$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
+$\hat{M_j} = \frac{ \hat{\mu}_j (1 - \hat{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}^2 } -1$.
 The estimates for the genome-level parameters are
-$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$
+$\hat{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \hat{\mu}_j$
 and
-$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
+$\hat{M}_0 = \frac{ \hat{\mu}_0 (1 - \hat{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \hat{\mu}_j^2 } -1$.
 %While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.
@@
 
-We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \tilde{\mu}_j\cdot (1-\tilde{\mu}_j)$ if $\tilde{\mu}_j \in (10^{-3},1-10^{-3})$ and $\sigma_j = 10^{-4}$ otherwise, where $\tilde{\mu}_j$ is the MoM estimator of $\mu_j$.
+We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \hat{\mu}_j\cdot (1-\hat{\mu}_j)$ if $\hat{\mu}_j \in (10^{-3},1-10^{-3})$ and $\sigma_j = 10^{-4}$ otherwise, where $\hat{\mu}_j$ is the MoM estimator of $\mu_j$.
 Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.

### 2014-01-23 doc/2014bioinformatics_zhang/RVD3_1.tex

--- 2014-01-07 doc/ISMB2014/RVD3_1.tex
@@
 \section{Motivation:}
-Including prior information in the Bayesian approach is difficult but meaningful.
+Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
+To address the detection of SNVs, a variety of methods are being under-represented.
+However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.
 
 \section{Results:}
-We analyzed the Sensitivity of the Bayesian model to different priors. Jeffreys prior improves robustness and accuracy of Single Nucleotide Variant detection algorithm.
-
+We propose a novel empirical Bayesian model to call variants accurately.
+And the sensitivity of this hierarchical model is analyzed by the improper prior, information prior and non-information prior on the synthetic sequence data with varying read depth and a range of allele frequencies.
+Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
+Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
+For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
+Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
 \section{Availability:}
@@
 \section{Introduction}
-Massively parallel sequencing data were generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
-One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants (SNV) candidates.
+Massively parallel sequencing data has been generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
+One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants (SNVs) candidates.
+Somatic SNVs are detected by comparing the tumor and corresponding normal samples.
 
-Currently SNV are deduced on NGS platform by variants calling algorithms.
-Typically, SAMtools, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, VarScan2 \citep{Koboldt:2012cga}, Strelka \citep{Saunders:2012fh}, and muTect \citep{Cibulskis:2013ta} are all highly used in detecting low-frequency variants.
-However identifying true SNVs remains challenging because of high false positive rates or high false discovery rate.
-Another statistical method-RVD \citep{Flaherty:2011ja}, was proposed based on beta-binomial model for ultrasensitive rare single nucleotide variant detection.
-So based on the RVD hierarchical model we built an improved robustness and accuracy model for low-depth variants detection.
+To address the detection of SNVs at low allele frequencies, a number of algorithms from NGS platform are being under-represented.
+Strelka \citep{Saunders:2012fh}, VarScan2 \citep{Koboldt:2012cga}, JointSNVMix \citep{roth2012jointsnvmix} are highly used to differentiate somatic SNVs from germline cells.
+Also, SAMtools, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, and muTect \citep{Cibulskis:2013ta} broadly concentrate on detecting low-frequency variants.
+However identifying the true SNVs remains challenging because of the high false positive rate or high false negative rate which are mainly caused by the clonal heterogeneity.
 
+Recently empirical Bayesian approaches have been made use to identify SNVs.
+A empirical Bayesian framework for somatic mutation detection from cancer genome sequencing data - EBCall, enables accurate mutations calling with low allele frequencies (less than 10\%) in a minor tumour subpopulation \citep{shiraishi2013empirical}.
+Another method based on empirical Bayesian hierarchical model - RVD, was proposed for ultrasensitive rare SNV detection using beta-binomial model \citep{Flaherty:2011ja}.
+RVD method is demonstrated to robustly detect mutations at 0.1\% fractional representation, which means accurately call one mutant per every 1000 wild-type alleles.
+With the shortcoming of high read depth estimation in RVD, we originally built an improved robustness and accuracy model - RVD3 for low-depth SNVs detection.
+Variants calling ability is tested and analyzed both on the synthetic sequence data and the true yeast sequencing data with different read depths.
+
+In this article, RVD3 model - a novel Bayesian structure to accurately identify SNVs with small false discovery rate is first described in detail.
+Secondly, Metropolis-within-Gibbs sampling is evolved for inference. And then to detect variants, a Bayesian posterior distribution is taken for hypothesis test.
+Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffreys prior and log-normal prior, and apply the different priors on the synthetic sequence data.
+Finally, we choose Jeffreys prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
+The analysis for the results reveals the enhanced robustness and improved accuracy of our algorithm.
 %%%%%%%%%%%%%%%%%%
@@
 The prior distribution characterizes the knowledge of the parameters in the model.
+Including prior information in the Bayesian approach is difficult but meaningful.
 An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}.
@@
 
-We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants.
+We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants \citep{Flaherty:2011ja}.
 
 \section{Discussion}
-
-Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text. Figure \ref{fig:02} shows that the above method  Text
+analysis of the tradeoff between experimental precision and sequencing depth to achieve ultrasensitive variant detection.
 
@@
 
-(Table~\ref{Tab:01}) Text Text Text Text Text Text  Text Text Text Text Text Text Text Text Text  Text Text Text Text Text Text.
-Figure \ref{fig:02} shows that the above method  Text Text Text Text  Text Text Text Text Text Text  Text Text.
-\citealp{Boffelli03} might want to know about  text text text text
-
 \begin{enumerate}
@@
 
-\paragraph{Funding\textcolon} Text Text Text Text Text Text  Text Text.
+\paragraph{Funding\textcolon} Text Text  Text Text.
 

### 2014-01-23 doc/2014bioinformatics_zhang/RVD3.tex
Updated draft.

--- 2014-01-23 doc/2014bioinformatics_zhang/RVD3_1.tex
@@
-\documentclass{bioinfo}
-\copyrightyear{2005}
-\pubyear{2005}
+\documentclass[11pt,reqno]{amsart}
+\usepackage[top=1in, left=1in, right=1in, bottom=1in]{geometry}                % See geometry.pdf to learn the layout options. There are lots.
+\geometry{letterpaper}                   % ... or a4paper or a5paper or ...
+\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
+%\usepackage{algorithm, algorithmic}
+
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+
+\usepackage{graphicx}
+
+\usepackage{verbatim}
+\usepackage{amssymb}
 \usepackage{amsmath}
-\usepackage{afterpage}
-\usepackage{algorithm}
 
-%\usepackage{placeins}
-\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
-\usepackage{xr}
-\externaldocument{supplementary_info}
-\algnewcommand{\LineComment}[1]{\State \(\triangleright\) #1}
+\usepackage{enumitem}
+
+\usepackage{setspace}
+\doublespacing
+
+\usepackage{natbib}
+
+%\usepackage{epstopdf}
+%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
 
@@
 \DeclareMathOperator{\diag}{diag}
-\begin{document}
-\firstpage{1}
+
+\algnewcommand{\Inputs}[1]{%
+  \State \textbf{Inputs:}
+  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
+}
+\algnewcommand{\Initialize}[1]{%
+  \State \textbf{Initialize:}
+  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
+}
 
 \title[RVD3]{Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data}
-\author[Sample \textit{et~al}]{Corresponding Author\,$^{1,*}$, Co-Author\,$^{2}$ and Co-Author\,$^2$\footnote{to whom correspondence should be addressed}}
-\address{$^{1}$Department of XXXXXXX, Address XXXX etc.\\
-$^{2}$Department of XXXXXXXX, Address XXXX etc.}
+\author{}
+%\date{}                                           % Activate to display a given date or no date
 
-\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-\editor{Associate Editor: XXXXXXX}
-
-\maketitle
+\begin{document}
 
 \begin{abstract}
-
-\section{Motivation:}
 Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
@@
 
-\section{Results:}
-We propose a novel empirical Bayesian model to call variants accurately.
-And the sensitivity of this hierarchical model is analyzed by the improper prior, information prior and non-information prior on the synthetic sequence data with varying read depth and a range of allele frequencies.
+We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
+We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior funstion.
 Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
@@
 For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
-Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
-\section{Availability:}
+This model is able o detect the emergence of a beneficial SNV earlier than was previously shown.
 
-\section{Contact:} \href{name@bio.com}{name@bio.com}
 \end{abstract}
 
+\maketitle
+
+%%%%%%%%%%%%%%%%%%
+% Introduction
+%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+
 Massively parallel sequencing data has been generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
@@
 Strelka \citep{Saunders:2012fh}, VarScan2 \citep{Koboldt:2012cga}, JointSNVMix \citep{roth2012jointsnvmix} are highly used to differentiate somatic SNVs from germline cells.
-Also, SAMtools, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, and muTect \citep{Cibulskis:2013ta} broadly concentrate on detecting low-frequency variants.
+Also, SAMtools \citep{li2009sequence}, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, and MuTect \citep{Cibulskis:2013ta} broadly concentrate on detecting low-frequency variants.
+Through the comparision of these somatic mutation callers \citep{wang2013detecting}, VarScan2 excelled at the detection of high coverage and allele frequecy, while MuTect outperformed the other methods in detecting the low allelic fraction SNVs.
 However identifying the true SNVs remains challenging because of the high false positive rate or high false negative rate which are mainly caused by the clonal heterogeneity.
 
-Recently empirical Bayesian approaches have been made use to identify SNVs.
+Recently empirical Bayesian approaches have been made use to identify SNVs, which can automatically adjust for multiple testing and selection bias \citep{liao2014prior}.
 A empirical Bayesian framework for somatic mutation detection from cancer genome sequencing data - EBCall, enables accurate mutations calling with low allele frequencies (less than 10\%) in a minor tumour subpopulation \citep{shiraishi2013empirical}.
@@
 Finally, we choose Jeffreys prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
-The analysis for the results reveals the enhanced robustness and improved accuracy of our algorithm.
+Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
+
+%%%%%%%%%%%%%%%%%%
+% Data Sets
+%%%%%%%%%%%%%%%%%%
+\section{Data Sets}
+
+\subsection{Synthetic DNA Sequence Data}
+
+Two 400bp DNA sequences differ only 14 single nucleotide positions. Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
+The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+We utilized BWA v0.7.5a to align the short sequencing reads to the reference sequence. The -C50 option of BWA was taken to remove the reads of high mapping quality.
+BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104 (http://picard.sourceforge.net).
+The final data set contains read pairs for $N=6$ replicates for the control at different MAF levels.
+
+\subsection{Yeast Data}
+We first mapped the wild-tpye strain GSY1135 \citep{kvitek2011reciprocal} to Chromosome 10 in S288c reference genome (SGD; http://www.yeastgenome.org/) by BWA v0.7.5a \citep{li2009fast}.
+Then called SNPs by GATK v2.5 UnifiedGenotyper \citep{McKenna:2010bva, depristo2011framework} and created a FASTA GSY1135 reference using GATK FastaAlternative.
+Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from \citep{kvitek2013whole}, and removed WT population using FASTX Barcode Splitter and cut down the pair ends accordingly.
+The FASTQ files of case and control were mapped to the corresponding reference genome created before.
+Then wen used SAMtools v0.1.19 \citep{li2009sequence} to convert the alignment files to the binary alignment map (BAM) format.
+Next, pileup files were generated by SAMtools and depth chart file were derived for further SNVs detection.
+
 %%%%%%%%%%%%%%%%%%
@@
 The definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
-Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions; two kinds of priors for $M_j$ are log-normal prior and Jeffreys prior \citep{jeffreys1946invariant}, which enhanced the former RVD model \citep{Flaherty:2011ja}.
-Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$.
-The graphical chart for RVD3 is shown in Figure~\ref{fig:graphical_model}.
+Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions;
+to choose a priori distribution for $M_j$, log-normal prior and Jeffreys prior \citep{jeffreys1946invariant} are employed, which enhances the former RVD model \citep{Flaherty:2011ja}.
+Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$. The graphical chart for RVD3 is shown in Figure~\ref{fig:graphical_model}.
 
@@
 \subsubsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
-MoM works by setting the population moment equal to the sample moment.
+The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure. MoM works by setting the population moment equal to the sample moment.
 A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
@@
 Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-By experience when $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$, the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration is set to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ ; otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
+By experience when $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$, the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration is set to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ ;
+otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
 We have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
 
-(Copy) We resampled from the proposal if the sample is outside of the support of the posterior distribution.
-We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
-Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.
+We resampled from the proposal if the sample is outside of the support of the posterior distribution.
+We throw away a burn-in period - 20\% of the samples, and thin the chain by a factor 2 to reduce autocorrelation among samples, resulting in a sample with size 1600 from the posterior distribution.
 
@@
 
-% \pi\left({M}_{j}\right) =
 \begin{equation}
@@
 We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$.
-Since we cannot provide an interpretive form for it, Metropolis-Hastings algorithm was still used to sample from the posterior distribution.
+Instead of computing the posterior distribution directly, Metropolis-Hastings algorithm was taken to sample from the posterior distribution.
 
@@
 ($\chi^2$ test is totally Copied!)
-An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios.
-We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
+An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error;
+we would like to differentiate these two scenarios. We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
 In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
@@
 The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
-Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$, that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
+Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$,
+that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
 The test statistic is
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Priors for precision parameter
-%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Priors for precision parameter}
 %\subsubsection{Improper Prior}
-The prior distribution characterizes the knowledge of the parameters in the model.
-Including prior information in the Bayesian approach is difficult but meaningful.
+The prior distribution characterizes the knowledge of the parameters in the statistical model. Including prior information in the Bayesian approach is difficult but meaningful.
 An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}.
@@
 Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}.
-Playing no prior on $ M_{j}$ is exactly an implicit improper prior.
-Therefore we considered about non-information prior and information prior.
+Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-information prior and information prior for sensitivity analysis.
 
 %\subsubsection{Jeffreys Prior}
-Various noninformative prior distributions have been suggested for parameters in hierarchical models.
+Non-information prior seems to be more unbiased and objective.
+Various non-informative prior distributions have been suggested for parameters in hierarchical models.
 Jeffreys prior, as a typical and influential one, is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys \citep{jeffreys1946invariant}.
-It is defined in terms of the Fisher information and works well with a single parameter.
-In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
+It is defined in terms of the Fisher information and works well with a single parameter. In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
 
-A good informative prior is imperative to promote accurate posterior estimates.
+Informative prior distribution is the most specific type of prior. A good informative prior is imperative to promote accurate posterior estimates.
 In our study, log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
 The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
-
-%%%%%%%%%%%%%%%%%%
-% Data Sets
-%%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences differ only 14 single nucleotide positions.
-Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
-The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We used BWA v0.7.5a to align the reads to the reference sequence with the -C50 option to remove the reads of high mapping quality.
-BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104 (http://picard.sourceforge.net).
-The final data set contains read pairs for $N=6$ replicates for the control at different MAF levels.
-
-\subsection{Yeast Data}
-We first mapped the wild-tpye strain GSY1135 \citep{kvitek2011reciprocal} to Chromosome 10 in S288c reference genome (SGD; http://www.yeastgenome.org/) by BWA v0.7.5a \citep{li2009fast}.
-Then called SNPs by GATK v2.5 UnifiedGenotyper \citep{McKenna:2010bva, depristo2011framework} and created a FASTA GSY1135 reference using GATK FastaAlternative.
-Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from \citep{kvitek2013whole}, and removed WT population using FASTX Barcode Splitter and trimed the Nextera tag by Cutadapt v1.3.
-Finally, the FASTQ files of case and control were mapped to the corresponding reference genome created before.
-And pileup files were generated using Samtools v0.1.19 and depth chart file were derived by randomly down sampling.
 
@@
 \subsection{Sensitivity analysis of priors}
-In order to examine the robustness of RVD3 to changes of different priors, we performed information prior (log-normal) and non-information prior (Jeffreys).
-We found that RVD3 is robust to changing the priors.
-The RVD3 algorithm provides estimates of model parameters and latent variables given the data.
+To examing the robustness of our Bayesian model, different plausible prior distributions can be utilized \citep{gelman2006prior}.
+For this purpose, we performed information prior (log-normal) and non-information prior (Jeffreys).
+We found that RVD3 is robust to changing of the priors. The RVD3 algorithm provides estimates of model parameters and latent variables given the data.
 
@@
 The left column of is shows the read depth for each of the six BAM files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries.
-For the 100\% mutant data set, the read depth drops at the mutant locations.
+Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries. For the 100\% mutant data set, the read depth drops at the mutant locations.
 This is due to the parameters imposed at the alignment stage.
-The right column of Figure~\ref{fig:M_jeffrey} reveals the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set.
-$M_j$ measures the variance between replicates at location $j$.
+The right column of Figure~\ref{fig:M_jeffrey} reveals the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set. $M_j$ measures the variance between replicates at location $j$.
 There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix~\ref{sec:appendix_parameters}.
+Interestingly, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix~\ref{sec:appendix_parameters}.
 Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-$M_j$ is greater than $M_0$ the precision between replicates is higher than the precision across positions.
+$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision across positions.
 
@@
 
-Statistically, We analyzed the posteriors and priors for Jeffreys prior and log-normal prior.
-Figure~\ref{fig:dilu_10} shows the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
+The Bayesian posterior predictive distributions and the priors distributions are shown in Figure~\ref{fig:dilu_10}.
+These plots indicate the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
 The positions are chosen in the middle of the base length- position 104 and 244 are mutant, and position 100 and 300 are non-mutant.
-From the posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
-From the figure non-mutant positions display normal and stable without a peak nor a strange shape.
-The Jeffreys' prior shows more information than the lognormal prior (flat) from the posterior curve.
-From the prior distribution, it shows our model wants to search for a small value for M.
+The posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
+Generally the non-mutant positions display normal and stable without a peak nor a strange shape.
+The two distributions show different ways of the prior and posterior.
+The log-normal prior attributes a higher prior probability to M values between 0 and 2000.
+The Jeffreys prior assigns more information than the log-normal prior (flat) from the posterior curve.
+They both want to search for a small value for M from the prior curve.
 
@@
 \subsection{Results of priors on synthetic data}
+The RVD3 model is analyzed by the selecting two reasonable prior distributions, and the corresponding results are compared from the aspects of performance with different read depths, sensitivity and specificity, and FDR.
+
 \subsubsection{Performance with read depth}
@@
 To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
-Here the Bayesian test is used without the $\chi^2$ test.
-Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
-The performance improves when the read depth goes up.
-(ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
+Here the Bayesian test is used without the $\chi^2$ test. Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
+The performance improves when the read depth goes up. (ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
 Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model with improper priors for $M_j$, which definitely demonstrates the advantage of priors.
@@
 Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{fid:FDR} shows the false discovery rate of the RVD3 with different priors.
-Jeffreys prior shows a smaller false discovery rate and higher accuracy than others.
-It is obvious no matter Jeffreys or log-normal, the variant detection performance acquires lower FDR than the model with improper priors.
+Figure~\ref{fid:FDR} shows the false discovery rate of the RVD3 with different priors. Jeffreys prior shows a smaller false discovery rate than others.
+It is obvious no matter Jeffreys or log-normal, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, campared with improper priors.
 Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
@@
 \includegraphics[width=80mm]{tables/FDR.pdf}
-\caption{False Discovery Rate comparison of RVD3 with different priors.}
+\caption{False Discovery Rate comparison with different priors on RVD3.}
 \label{fig:FDR}
@@
 
-We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants \citep{Flaherty:2011ja}.
+We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants \citep{kvitek2013whole}.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The empirical Bayesian commonly causes the false discovery rate (FDR) \citep{efron2010large}.
 
@@
 
-\begin{enumerate}
-\item this is item, use enumerate
-\item this is item, use enumerate
-\item this is item, use enumerate
-\end{enumerate}
-
-\section*{Acknowledgement}
-Text Text
-
-\paragraph{Funding\textcolon} Text Text  Text Text.
-
-\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%\bibliographystyle{plain}
-
-\bibliography{bioinfo}
-
 \appendix
-
 %%%%%%%%%%%%%%%%%%%
@@
 We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
-The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
-The first and second population moments are
+The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$. The first and second population moments are
 \begin{eqnarray}
@@
 \section{Inference of Jeffreys prior}\label{sec:appendix_Jeffreys}
-We assume there is only one replicate,
+We assume there is only one replicate (i=1),
 
@@
 \begin{split}
-\log p\left(\theta_{j}|\mu_{j},M_{j}\right)& =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)- \log \Gamma\left(1-\mu_{j},M_{j}\right)\\
-& + (\mu_{j}M_{j}-1)\log\theta_{j} + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
+&\log p\left(\theta_{j}|\mu_{j},M_{j}\right) =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)\\
+&- \log \Gamma\left(1-\mu_{j},M_{j}\right) + (\mu_{j}M_{j}-1)\log\theta_{j}\\
+& + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
 \end{split}
@@
 \begin{equation}
-\frac{\delta\log p(\theta_{j})}{\delta M_{j}} = \Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j} - \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})
+\begin{split}
+&\frac{\delta\log p(\theta_{j})}{\delta M_{j}} =\Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j}\\
+&- \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})\
+\end{split}
 \end{equation}
@@
 
-Now we have the Jeffreys' prior for $M_{j}$:
+Now we have the Jeffreys' prior $\pi\left({M}_{j}\right)$ for $M_{j}$:
 
@@
 
+\bibliographystyle{apalike}
+\bibliography{bioinfo}
 \end{document}

### 2014-01-23 doc/2014bioinformatics_zhang/RVD3_1.tex
Updated draft.

@@
 \section{Results:}
-We propose a novel empirical Bayesian model to call variants accurately.
-And the sensitivity of this hierarchical model is analyzed by the improper prior, information prior and non-information prior on the synthetic sequence data with varying read depth and a range of allele frequencies.
+We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
+We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior funstion.
 Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
@@
 For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
-Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
+This model is able o detect the emergence of a beneficial SNV earlier than was previously shown.
+
 \section{Availability:}
@@
 Finally, we choose Jeffreys prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
-The analysis for the results reveals the enhanced robustness and improved accuracy of our algorithm.
+Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
+
 %%%%%%%%%%%%%%%%%%
@@
 Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}.
-Playing no prior on $ M_{j}$ is exactly an implicit improper prior.
-Therefore we considered about non-information prior and information prior.
+Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-information prior and information prior for sensitivity analysis.
 
 %\subsubsection{Jeffreys Prior}
-Various noninformative prior distributions have been suggested for parameters in hierarchical models.
+Non-information prior seems to be more unbiased and objective. 
+Various non-informative prior distributions have been suggested for parameters in hierarchical models.
 Jeffreys prior, as a typical and influential one, is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys \citep{jeffreys1946invariant}.
-It is defined in terms of the Fisher information and works well with a single parameter.
-In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
+It is defined in terms of the Fisher information and works well with a single parameter. In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
 
-A good informative prior is imperative to promote accurate posterior estimates.
+Informative prior distribution is the most specific type of prior. A good informative prior is imperative to promote accurate posterior estimates.
 In our study, log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
@@
 
-Two 400bp DNA sequences differ only 14 single nucleotide positions.
-Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
+Two 400bp DNA sequences differ only 14 single nucleotide positions. Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
 The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We used BWA v0.7.5a to align the reads to the reference sequence with the -C50 option to remove the reads of high mapping quality.
+We utilized BWA v0.7.5a to align the short sequencing reads to the reference sequence. The -C50 option of BWA was taken to remove the reads of high mapping quality.
 BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104 (http://picard.sourceforge.net).
@@
 Then called SNPs by GATK v2.5 UnifiedGenotyper \citep{McKenna:2010bva, depristo2011framework} and created a FASTA GSY1135 reference using GATK FastaAlternative.
-Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from \citep{kvitek2013whole}, and removed WT population using FASTX Barcode Splitter and trimed the Nextera tag by Cutadapt v1.3.
-Finally, the FASTQ files of case and control were mapped to the corresponding reference genome created before.
-And pileup files were generated using Samtools v0.1.19 and depth chart file were derived by randomly down sampling.
+Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from \citep{kvitek2013whole}, and removed WT population using FASTX Barcode Splitter and cut down the pair ends accordingly.
+The FASTQ files of case and control were mapped to the corresponding reference genome created before.
+Then wen used SAMtools v0.1.19 \citep{li2009sequence} to convert the alignment files to the binary alignment map (BAM) format.
+Next, pileup files were generated by SAMtools and depth chart file were derived for further SNVs detection.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The empirical Bayesian commonly causes the false discovery rate (FDR) \citep{efron2010large}.
 
 \section{Conclusion}
-
-\begin{enumerate}
-\item this is item, use enumerate
-\item this is item, use enumerate
-\item this is item, use enumerate
-\end{enumerate}
 

### 2014-01-23 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/plot_prior_posterior_test.py
Yeast data experiment with log-normal prior.

@@ -1,0 +1,92 @@
+from scipy.special import gammaln, polygamma
+from scipy.stats import gaussian_kde
+import numpy as np
+import scipy.stats as ss
+import matplotlib.pyplot as plt
+import sys
+import os
+import h5py
+import logging
+import itertools
+import math
+import rvd30
+import rvd29
+
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    locationList = (100,300,104,244)
+    
+    # Same experiment condition: read depth rate = 1/100
+    h5Filename_rvd29 = './hdf5_rvd29_gibbs=20000_dc=100/Case10_0.hdf5'
+ #   h5Filename_rvd30 = './hdf5_rvd30_gibbs=20000_dc=100/Case10_0.hdf5'
+    M = range(1,10000)
+    fig=plt.figure(figsize=(20, 12))
+
+    """ Probability of Jeffreys' prior from rvd29"""
+    logging.debug("Plot rvd29 prior and posterior")
+    for loc in locationList:
+        logging.debug("Processing location: %d" % loc)
+        ax = fig.add_subplot(2,4,locationList.index(loc)+1)
+        (Phi, M_s, Theta, Mu, Loc, R, N) = rvd29.load_model(h5Filename_rvd29)
+        mu=np.mean(Mu, axis=0) 
+        mu=np.mean(mu)
+        p = np.sqrt(polygamma(1, np.dot(mu,M))*(mu**2) + polygamma(1, np.dot(1-mu,M))*(1-mu)**2 -polygamma(1,M))
+        sum = np.sum(p)
+        M1 = M_s[loc]       
+        print (M1)
+        # Use Gaussian Kernel Density Estimate for posterior Mj
+        kernel = gaussian_kde(M1) 
+        norm = kernel.integrate_box(0,10000)
+        ax.semilogy(M, p/sum)
+        ax.semilogy(M, kernel(M).T/norm)
+        ax.semilogy(M1)
+        ax.legend( ['prior','posterior'],loc=9, ncol=9)
+        subtitle='position='+str(loc)
+        ax.set_title(subtitle)
+        if locationList.index(loc)==0:
+           ax.set_ylabel('Probability of Jeffreys prior',fontsize=15)
+
+        
+    
+    title='test'
+    figformat='.pdf'
[…]

### 2014-01-23 results/2014-01-20-Yeast_data_chr10_390000_470000_lognormal_experiment/character.py

@@ -11,7 +11,7 @@ import xlwt
 # Insert the source directory at front of the path
 rvddir = os.path.join('./')
 sys.path.insert(0, rvddir)
-import rvd29
+import rvd30
 
 def main():
     book=xlwt.Workbook(encoding="utf-8")
@@ -31,7 +31,7 @@ def main():
     sheet4.write(0, 0, "MAF")
     sheet4.write(0, 1, "Median Depth")
     
-    method = {'RVD3(T=0)':'./vcf'}
+    method = {'RVD3 with lognormal prior(T=0)':'./vcf'}
     
     DilutionList = (0,)
     
@@ -62,7 +62,7 @@ def main():
                 print hdf5Dir
                 caseFile = 'Case%s.hdf5' %str(d).replace('.','_')
                 caseFile = "%(dir)s/%(file)s" %{'dir':hdf5Dir,'file':caseFile}
-                (_, _, _, _, _, _, caseN) = rvd29.load_model(caseFile)
+                (_, _, _, _, _, _, caseN) = rvd30.load_model(caseFile)
                 cov = int(np.median(caseN))
 
                 # print the median coverage

### 2014-01-23 results/2014-01-20-Yeast_data_chr10_390000_470000_lognormal_experiment/test_rvd30.py

@@ -16,11 +16,13 @@ import rvd30
 logging.basicConfig(level=logging.DEBUG,
                     format='%(levelname)s:%(message)s')
 
+    
 def main():
     controlFile = "./1/Control.hdf5"
-    caseFile = "./1/Case0.hdf5" 
+    caseFile = "./1/Case0.hdf5"
     outputFile='./vcf'
     rvd30.test(controlFile,caseFile,T=0,N=1000,outputFile=outputFile)
+    
 if __name__ == '__main__':
     main()
 

### 2014-02-05 results/2014-02-03_variational_inference/rvd_var.py
Update variational approach inplementation

@@ -48,12 +48,12 @@ def EqMu(delta, gamma):
 def EqlogMu(delta, gamma):
 	return psi(delta) - psi(delta + gamma)
 
-def Eqlog1_mu(delta, gamma):
+def Eqlog1_Mu(delta, gamma):
 	return psi(gamma) - psi(delta + gamma)
 
 def EqlogGamma(delta, gamma, M):
 	# Eqlog\Gamma(muM)
-	N = 6
+	N = 7
 	mu = range(0, 1 , 2^N+1)
 	y = [kernel(delta, gamma, x, M) for x in mu]
 
@@ -63,18 +63,49 @@ def EqlogGamma(delta, gamma, M):
 def kernel(delta, gamma, mu, M):
 		return ss.beta.pdf(mu, delta, gamma)*betaln(mu*M, (1-mu)*M)
 
+
 ## compute entropy
 def BetaEntropy(a, b):
+	# To compute EqlogQmu and EqlogQtheta
 	#return betaln(delta,gamma) - (delta-1)psi(delta) - (gamma - 1) psi(gamma) +(delta + gamma -2) psi(delta + gamma)
 	return betaln(a,b) - (a-1)psi(a) - (b - 1) psi(b) +(a + b -2) psi(a + b)
 
+
 ## compute ELBO
-def ELBO():
-	EqlogPr = 0
-	EqlogPtheta = 
-	EqlogPmu = 
-	EqlogQtheta = 
-	EqlogQmu = 
+def ELBO(r, n, theta, mu, M, mu0, M0, alpha, beta, delta, gamma):
+	(J,N) = n.shape
+
+	Mu = [EqMu(delta[j], gamma[j]) for j in xrange(J)]
+	logMu = [EqlogMu(delta[j], gamma[j]) for j in xrange(J)]
+	log1_Mu = [Eqlog1_Mu(delta[j], gamma[j]) for j in xrange(J)]
+	logTheta = [EqlogTheta(alpha[j,i], beta[j,i]) for j in xrange(J) for i in xrange(N)]
+	log1_Theta = [Eqlog1_Theta(alpha[j,i], beta[j,i]) for j in xrange(J) for i in xrange(N)]
+
+	EqlogPr = 0.0
+	for j in xrange(J):
+		for i in xrange(N):
[…]

### 2014-02-09 results/2014-02-03_variational_inference/rvd_var.py
Update codes for variational approach. finished opt functions

@@ -23,12 +23,14 @@ def main():
     logging.basicConfig(level=log_level,
         format='%(levelname)s:%(module)s:%(message)s')
     
-    (y, x, theta) = gen1()
+  ## load the data
     
     pool = None
     #pool = mp.Pool(processes=2)
-    
+  ## model optimization  
     (phi, q) = varlap_opt(y, K=2, seed=10241978, pool=pool)
+
+  ## save the parameters.
 
     save_model('model.hdf5', y, phi, q)
     
@@ -36,50 +38,50 @@ def main():
     
 
 ## compute sufficient statistics
-def EqlogTheta(alpha, beta):
-	return psi(alpha) - psi(alpha + beta)
+def EqlogTheta(delta):
+	return psi(delta[0]) - psi(np.sum(delta))
 
-def Eqlog1_Theta(alpha, beta):
-	return psi(beta) - psi(alpha + beta)
+def Eqlog1_Theta(delta):
+	return psi(delta[1]) - psi(np.sum(delta))
 
-def EqMu(delta, gamma):
-	return delta / (delta + gamma) # eps?
+def EqMu(gam):
+	return gam[0] / (np.sum(gam)) # eps?
 
-def EqlogMu(delta, gamma):
-	return psi(delta) - psi(delta + gamma)
+def EqlogMu(gam):
+	return psi(gam[0]) - psi(np.sum(gam))
 
-def Eqlog1_Mu(delta, gamma):
-	return psi(gamma) - psi(delta + gamma)
+def Eqlog1_Mu(gam):
+	return psi(gam[1]) - psi(np.sum(gam))
 
-def EqlogGamma(delta, gamma, M):
-	# Eqlog\Gamma(muM)
+def EqlogGamma(gam, M):
+	# Eqlog\gam(muM)
 	N = 7
 	mu = range(0, 1 , 2^N+1)
-	y = [kernel(delta, gamma, x, M) for x in mu]
+	y = [kernel(gam[0], gam[1], x, M) for x in mu]
 
 	return integrate.romb(y, dx = 0.1/(2^N+1))
 	
 
-def kernel(delta, gamma, mu, M):
-		return ss.beta.pdf(mu, delta, gamma)*betaln(mu*M, (1-mu)*M)
+def kernel(gam, mu, M):
+		return ss.beta.pdf(mu, gam[0], gam[1])*betaln(mu*M, (1-mu)*M)
 
 
 ## compute entropy
-def BetaEntropy(a, b):
+def BetaEntropy(x):
[…]

### 2014-02-10 doc/2014bioinformatics_zhang/RVD3.tex
RVD3 test on HCC1187 data.

@@
 We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
-We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior funstion.
+We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior function.
 Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
@@
 For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
-This model is able o detect the emergence of a beneficial SNV earlier than was previously shown.
+This model is able to detect the emergence of a beneficial SNV earlier than was previously shown.
 

### 2014-02-10 results/2014-02-07_RVD3_test_on_HCC1187/Jeffreys/runall_control.py

@@ -1,0 +1,48 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+# Insert the src/python/rvd29 directory at front of the path
+
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+import rvd29
+
+##pool =None
+pool = mp.Pool(processes=30)
+
+gibbs_nsample = 4000
+mh_nsample = 5
+
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the control
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    filename = "HCC1187BL_S1.dc"
+    controlFileList = ["../../../../../yhe2/Research/rvd2/results/2013-11-07_HCC1187_PAXIP1_hg19masked/depth_chart/%s" % filename]
+    (r, n, loc, refb) = rvd29.load_depth(controlFileList)
+    phi, theta_s, mu_s, M_s = rvd29.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd29.save_model(h5FileName, phi, mu=mu_s, M=M_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)

### 2014-02-10 results/2014-02-07_RVD3_test_on_HCC1187/Jeffreys/test_rvd29.py

@@ -1,0 +1,32 @@
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import scipy.stats as ss
+import multiprocessing as mp
+
+# Insert the source directory at front of the path
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+import rvd29
+
+pool = mp.Pool(processes=40)
+# T=10/Depth
+T=10/50
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+    controlFile = "./Control.hdf5"
+    caseFile = "./Case.hdf5"
+    outputFile='./vcf'
+    rvd29.test(controlFile,caseFile,T=T,N=1000,outputFile=outputFile, pool=pool)
+
+if __name__ == '__main__':
+    main()
+

### 2014-02-10 results/2014-02-07_RVD3_test_on_HCC1187/lognormal/runall_control.py

@@ -1,0 +1,45 @@
+# -*- coding: utf-8 -*-
+# <nbformat>3.0</nbformat>
+
+# <codecell>
+
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import pandas as pd
+import h5py
+import multiprocessing as mp
+import logging
+import pdb
+# <codecell>
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(module)s:%(message)s')
+
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+import rvd30
+
+pool = mp.Pool(processes=30)
+gibbs_nsample = 4000
+mh_nsample = 5
+logging.debug("Gibbs step size=%d" % gibbs_nsample)
+logging.debug("mh sample size=%d" % mh_nsample)
+
+# Estimate the model for the cases
+logging.debug("Processing control data.")
+h5FileName = "Control.hdf5"
+
+try:
+    with h5py.File(h5FileName, 'r') as f:
+        pass
+except IOError as e:
+    filename = "HCC1187BL_S1.dc"
+    controlFileList = ["../../../../../../yhe2/Research/rvd2/results/2013-11-07_HCC1187_PAXIP1_hg19masked/depth_chart/%s" % filename]
+    (r, n, loc, refb) = rvd30.load_depth(controlFileList)
+    phi, theta_s, mu_s, M_s = rvd30.mh_sample(r, n, gibbs_nsample=gibbs_nsample,mh_nsample=mh_nsample, burnin=0.2, pool=pool)
+    logging.debug("Saving model in %s" % h5FileName)
+    rvd30.save_model(h5FileName, phi, mu=mu_s, M=M_s, theta=theta_s, r=r, n=n, loc=loc,
+           refb=refb)
+

### 2014-02-12 doc/2014bioinformatics_zhang/RVD3.tex
Updated RVD3 manuscript.

--- 2014-01-23 doc/2014bioinformatics_zhang/RVD3.tex
@@
 \begin{abstract}
-Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
-To address the detection of SNVs, a variety of methods are being under-represented.
-However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.
+%Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
+%To address the detection of SNVs, a variety of methods are being under-represented.
+%However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.
 
-We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
-We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior funstion.
-Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
-Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
-For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
-This model is able o detect the emergence of a beneficial SNV earlier than was previously shown.
+%We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
+We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior function.
+%Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
+%Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
+%For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
+%This model is able to detect the emergence of a beneficial SNV earlier than was previously shown.
+%Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data
+
+% The same with the Abstract poster of RECOMB2014.
+Massively parallel sequencing data generated by next-generation sequencing (NGS) technology is routinely used to detect single nucleotide variants (SNVs) in research samples. 
+An emerging challenge for this technology is the identification of SNVs in heterogeneous cell populations with low read-depth data.
+We have developed a Bayesian statistical model is able to share information between correlated positions and call low-frequency variants in heterogeneous samples. 
+We present a Bayesian sensitivity analysis of the model to variations in the prior function. 
+Our model with different priors both performs a high accuracy, and a Jeffreys prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within 39 low read depth compared with an improper prior.
+In an analysis of a directed evolution experiment, we are able to detect the emergence of a beneficial SNV earlier than was previously shown.
 
@@
 \subsubsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure. MoM works by setting the population moment equal to the sample moment.
+[This paragrah COPIED]The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
+MoM works by setting the population moment equal to the sample moment.
 A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
@@
 
-Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
+[This paragrah COPIED]Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
 By experience when $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$, the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration is set to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ ;
@@
 \subsubsection{$\chi^2$ test for non-uniform base distribution}
-($\chi^2$ test is totally Copied!)
+[This part is COPIED]
+
 An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error;
@@
 Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{fid:FDR} shows the false discovery rate of the RVD3 with different priors. Jeffreys prior shows a smaller false discovery rate than others.
+Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffreys prior shows a smaller false discovery rate than others.
 It is obvious no matter Jeffreys or log-normal, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, campared with improper priors.
@@
 \caption{False Discovery Rate comparison with different priors on RVD3.}
-\label{fig:FDR}
+\label{fig:fdr}
 \end{center}
@@
 \section{Discussion}
-analysis of the tradeoff between experimental precision and sequencing depth to achieve ultrasensitive variant detection.
-
-The empirical Bayesian commonly causes the false discovery rate (FDR) \citep{efron2010large}.
 
@@
 \section{Parameter Initialization}\label{sec:appendix_mom}
+[This part is COPIED]
+
 Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.

### 2014-02-21 doc/2014bioinformatics_zhang/RVD3.tex
Upload all the manuscript and figures in pdfs of RVD3.

@@
 We present a Bayesian sensitivity analysis of the model to variations in the prior function.
-Our model with different priors both performs a high accuracy, and a Jeffreys prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within 39 low read depth compared with an improper prior.
+Our model with different priors both performs a high accuracy, and a Jeffreys prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with an improper prior.
 In an analysis of a directed evolution experiment, we are able to detect the emergence of a beneficial SNV earlier than was previously shown.
@@
 
-Two 400bp DNA sequences differ only 14 single nucleotide positions. Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
+Two 400bp DNA sequences(control/case) were synthesized with only 14 different single nucleotide positions.
+Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
 The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We utilized BWA v0.7.5a to align the short sequencing reads to the reference sequence. The -C50 option of BWA was taken to remove the reads of high mapping quality.
+We used BWA v0.7.5a to align the short sequencing reads to the reference sequence. The -C50 option of BWA was taken to remove the reads of low mapping quality.
 BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104 (http://picard.sourceforge.net).

### 2014-02-24 doc/bioinformatics2014/document.tex
Create folder bioinformatics2014 for the manuscript

@@
 \documentclass{bioinfo}
-\copyrightyear{2005}
-\pubyear{2005}
+\copyrightyear{2014}
+\pubyear{2014}
 \usepackage{amsmath}
@@
 
-RVD uses a two-stage approach for detecting for rare variants.
+RVD2 uses a two-stage approach for detecting rare variants.
 First, it estimates the parameters of a hierarchical Bayesian model under two sequencing data sets: one from the sample of interest (case) and one from a known reference sample (control).
@@
 
-The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision.
+The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision;  $ \mu_j $, a local error rate.
+$M_j$, a local precision.
 The global error rate, $\mu_0$, estimates the expected error rate across all locations.
 The global precision, $M_0$, estimates the variation in the error rate  across locations.
+The local error rate, $ \mu_j $, estimates the exepected error rate across replicates at location $ j $.
 The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.
@@
 \subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
-
+%TODO: double check if the posterior is correct
 Samples from the posterior distribution
@@
 
-\paragraph{Funding\textcolon}  P.F. and Y.E. were supported by seed funding from Worcester Polytechnic Institute.
+\paragraph{Funding\textcolon}  P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute.
 

### 2014-02-24 results/2014-02-23_HCC1187_PAXIP1_using_mutect/SNP_muTect_HCC1187.sh
Variant calling in clinical data HCC1187 PAXIP1 gene using MuTect

@@ -1,0 +1,21 @@
+#!/bin/sh
+
+FASTAFILE=../../data/hg19/Homo_sapiens/UCSC/hg19/Sequence/WholeGenomeFasta/genome.fa
+
+MUTECT=muTect.jar
+
+BAMCONTROL=../../data/HCC1187/HCC1187BL_S1.bam
+BAMCASE=../../data/HCC1187/HCC1187C_S1.bam
+
+
+WORKDIR=work/
+mkdir -p $WORKDIR
+
+java -Xmx2g -jar $MUTECT \
+	--analysis_type MuTect \
+	--reference_sequence $FASTAFILE \
+	--intervals chr7:154738059-154782774 \
+	--input_file:normal $BAMCONTROL \
+	--input_file:tumor $BAMCASE \
+	--out $WORKDIR/HCC.call_stats.txt \
+	--coverage_file $WORKDIR/HCC.coverage.wig.txt

### 2014-03-03 doc/2014bioinformatics_zhang/Infer_Hypo_Test.tex
Updated draft of RVD3.

--- 2014-02-10 doc/2014bioinformatics_zhang/RVD3.tex
@@
 
-\begin{abstract}
-Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
-To address the detection of SNVs, a variety of methods are being under-represented.
-However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.
-
-We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
-We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior function.
-Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
-Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
-For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
-This model is able to detect the emergence of a beneficial SNV earlier than was previously shown.
-
-\end{abstract}
-
-\maketitle
-
-%%%%%%%%%%%%%%%%%%
-% Introduction
-%%%%%%%%%%%%%%%%%%
-\section{Introduction}
-
-Massively parallel sequencing data has been generated by Next-generation sequencing (NGS) technology to benefit clinical diagnostics and sequencing based phylogenetic analyses.
-One primary application of NGS is variation detection among related populations and separate novel single-nucleotide variants (SNVs) candidates.
-Somatic SNVs are detected by comparing the tumor and corresponding normal samples.
-
-To address the detection of SNVs at low allele frequencies, a number of algorithms from NGS platform are being under-represented.
-Strelka \citep{Saunders:2012fh}, VarScan2 \citep{Koboldt:2012cga}, JointSNVMix \citep{roth2012jointsnvmix} are highly used to differentiate somatic SNVs from germline cells.
-Also, SAMtools \citep{li2009sequence}, Genome Analysis Toolkit (GATK) \citep{McKenna:2010bva}, and MuTect \citep{Cibulskis:2013ta} broadly concentrate on detecting low-frequency variants.
-Through the comparision of these somatic mutation callers \citep{wang2013detecting}, VarScan2 excelled at the detection of high coverage and allele frequecy, while MuTect outperformed the other methods in detecting the low allelic fraction SNVs.
-However identifying the true SNVs remains challenging because of the high false positive rate or high false negative rate which are mainly caused by the clonal heterogeneity.
-
-Recently empirical Bayesian approaches have been made use to identify SNVs, which can automatically adjust for multiple testing and selection bias \citep{liao2014prior}.
-A empirical Bayesian framework for somatic mutation detection from cancer genome sequencing data - EBCall, enables accurate mutations calling with low allele frequencies (less than 10\%) in a minor tumour subpopulation \citep{shiraishi2013empirical}.
-Another method based on empirical Bayesian hierarchical model - RVD, was proposed for ultrasensitive rare SNV detection using beta-binomial model \citep{Flaherty:2011ja}.
-RVD method is demonstrated to robustly detect mutations at 0.1\% fractional representation, which means accurately call one mutant per every 1000 wild-type alleles.
-With the shortcoming of high read depth estimation in RVD, we originally built an improved robustness and accuracy model - RVD3 for low-depth SNVs detection.
-Variants calling ability is tested and analyzed both on the synthetic sequence data and the true yeast sequencing data with different read depths.
-
-In this article, RVD3 model - a novel Bayesian structure to accurately identify SNVs with small false discovery rate is first described in detail.
-Secondly, Metropolis-within-Gibbs sampling is evolved for inference. And then to detect variants, a Bayesian posterior distribution is taken for hypothesis test.
-Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffreys prior and log-normal prior, and apply the different priors on the synthetic sequence data.
-Finally, we choose Jeffreys prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
-Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
-
-%%%%%%%%%%%%%%%%%%
-% Data Sets
-%%%%%%%%%%%%%%%%%%
-\section{Data Sets}
-
-\subsection{Synthetic DNA Sequence Data}
-
-Two 400bp DNA sequences differ only 14 single nucleotide positions. Sample of the case and control DNA were mixed to yield 0.1\%, 0.3\%, 1\%, 10\%, and 100\% defined minor allele frequencies (MAFs).
-The details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
-We utilized BWA v0.7.5a to align the short sequencing reads to the reference sequence. The -C50 option of BWA was taken to remove the reads of high mapping quality.
-BAM files were sampled by $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.104 (http://picard.sourceforge.net).
-The final data set contains read pairs for $N=6$ replicates for the control at different MAF levels.
-
-\subsection{Yeast Data}
-We first mapped the wild-tpye strain GSY1135 \citep{kvitek2011reciprocal} to Chromosome 10 in S288c reference genome (SGD; http://www.yeastgenome.org/) by BWA v0.7.5a \citep{li2009fast}.
-Then called SNPs by GATK v2.5 UnifiedGenotyper \citep{McKenna:2010bva, depristo2011framework} and created a FASTA GSY1135 reference using GATK FastaAlternative.
-Secondly, we downloaded generation 7 as control and generation 133 as case in experiment 1 from \citep{kvitek2013whole}, and removed WT population using FASTX Barcode Splitter and cut down the pair ends accordingly.
-The FASTQ files of case and control were mapped to the corresponding reference genome created before.
-Then wen used SAMtools v0.1.19 \citep{li2009sequence} to convert the alignment files to the binary alignment map (BAM) format.
-Next, pileup files were generated by SAMtools and depth chart file were derived for further SNVs detection.
-
-%%%%%%%%%%%%%%%%%%
-% RVD3 Model
-%%%%%%%%%%%%%%%%%%
-\section{RVD3 Model}
-
-\subsection{Model Structure}\label{sec:model_structure}
-RVD is based on a two-stage hierarchical Bayesian model for variant detection \citep{Flaherty:2011ja}.
-Through hypothesis test on case and control samples by RVD, we can call the variants successfully.
-Now RVD3 has three-stage model including priors built on the former RVD model.
-The definitions for sample data are given: $r_{ji}$ is the number of reads with a non-reference base at position $j$ in replicate $i$, and $n_{ji}$ is the total number of reads at position $j$ in replicate $i$.
-Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions;
-to choose a priori distribution for $M_j$, log-normal prior and Jeffreys prior \citep{jeffreys1946invariant} are employed, which enhances the former RVD model \citep{Flaherty:2011ja}.
-Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$. The graphical chart for RVD3 is shown in Figure~\ref{fig:graphical_model}.
-
-RVD3 hierarchically includes three levels of samplings: $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
-$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation caused by experimental repeatability.
-The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and Jeffreys prior for $M_j$.
-%$\pi\left({M}_{j}\right)$
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=85mm]{figs/RVD3_model.pdf}
-\caption{RVD3 Graphical Model}
-\label{fig:graphical_model}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%
-% Inference & Hypothesis Testing
-%%%%%%%%%%%%%%%%%%
-\subsection{Inference and Hypothesis Testing}
+\section{Inference and Hypothesis Testing}
 
@@
 
-\begin{algorithm}[ht]
+\begin{algorithm}
 \caption{Inference process for Metropolis-within-Gibbs}
@@
 %%%%%%%%%%%%%
-\subsubsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure. MoM works by setting the population moment equal to the sample moment.
+\subsection{Initialization}
+[This paragrah COPIED]The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
+MoM works by setting the population moment equal to the sample moment.
 A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
@@
 %%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
+\subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
 
@@
 %%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
+\subsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
 Based on Markov blanket, the posterior distribution over $\mu_j$ is
@@
 %%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
+\subsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
 Since Jeffreys prior is from the Fisher information and in RVD3 model ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$,
@@
 %%%%%%%%%%%%%%%%%%
-\subsubsection{Posterior Density Test}\label{sec:hypothesis_test}
+\subsection{Posterior Density Test}\label{sec:hypothesis_test}
 Posterior distributions of $\mu_j$ for the control and case are achieved -  $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$, by Metropolis-within-Gibbs.
@@
 %%%%%%%%%%%%%%%%%%
-\subsubsection{$\chi^2$ test for non-uniform base distribution}
-($\chi^2$ test is totally Copied!)
+\subsection{$\chi^2$ test for non-uniform base distribution}
+[This part is COPIED]
+
 An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error;
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Priors for precision parameter
-%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Priors for precision parameter}
-%\subsubsection{Improper Prior}
-The prior distribution characterizes the knowledge of the parameters in the statistical model. Including prior information in the Bayesian approach is difficult but meaningful.
-An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}.
-Furthermore when the Markov chain Monte Carlo method is taken to derive the posterior, it is possibly hard to sniff out the improper posterior.
-Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}.
-Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-information prior and information prior for sensitivity analysis.
-
-%\subsubsection{Jeffreys Prior}
-Non-information prior seems to be more unbiased and objective.
-Various non-informative prior distributions have been suggested for parameters in hierarchical models.
-Jeffreys prior, as a typical and influential one, is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys \citep{jeffreys1946invariant}.
-It is defined in terms of the Fisher information and works well with a single parameter. In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
-
-Informative prior distribution is the most specific type of prior. A good informative prior is imperative to promote accurate posterior estimates.
-In our study, log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
-The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
-
-%%%%%%%%%%%%%%%%%%
-% Results
-%%%%%%%%%%%%%%%%%%
-\section{Results}
-
-%%%%%%%%%%%%%%%%%%
-% Sensitivity analysis of priors
-%%%%%%%%%%%%%%%%%%
-\subsection{Sensitivity analysis of priors}
-To examing the robustness of our Bayesian model, different plausible prior distributions can be utilized \citep{gelman2006prior}.
-For this purpose, we performed information prior (log-normal) and non-information prior (Jeffreys).
-We found that RVD3 is robust to changing of the priors. The RVD3 algorithm provides estimates of model parameters and latent variables given the data.
-
-We show several of these parameters of the model with Jeffreys prior in Figure~\ref{fig:M_jeffrey}.
-The left column of is shows the read depth for each of the six BAM files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries. For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The right column of Figure~\ref{fig:M_jeffrey} reveals the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set. $M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Interestingly, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix~\ref{sec:appendix_parameters}.
-Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision across positions.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=90mm]{figs/M_jeffrey.pdf}
-\caption{Key parameters for RVD3 model with Jeffreys prior on synthetic DNA data sets.}
-\label{fig:M_jeffrey}
-\end{center}
-\end{figure}
-
-\begin{figure*}[!tbp]
-\begin{center}
-\includegraphics[width=190mm]{figs/post_prior_dilu_10.pdf}
-\caption{Distribution of priors and posteriors when dilution is 10\%}
-\label{fig:dilu_10}
-\end{center}
-\end{figure*}
-
-The Bayesian posterior predictive distributions and the priors distributions are shown in Figure~\ref{fig:dilu_10}.
-These plots indicate the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
-The positions are chosen in the middle of the base length- position 104 and 244 are mutant, and position 100 and 300 are non-mutant.
-The posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
-Generally the non-mutant positions display normal and stable without a peak nor a strange shape.
-The two distributions show different ways of the prior and posterior.
-The log-normal prior attributes a higher prior probability to M values between 0 and 2000.
-The Jeffreys prior assigns more information than the log-normal prior (flat) from the posterior curve.
-They both want to search for a small value for M from the prior curve.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results of priors on synthetic data
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Results of priors on synthetic data}
-The RVD3 model is analyzed by the selecting two reasonable prior distributions, and the corresponding results are compared from the aspects of performance with different read depths, sensitivity and specificity, and FDR.
-
-\subsubsection{Performance with read depth}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=90mm]{figs/ROC_without_chi2_jeffrey.pdf}
-\caption{ROC curve for variants detection performance by Jeffreys prior.}
-\label{fig:ROC_jeffrey}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=90mm]{figs/ROC_without_chi2_lognormal.pdf}
-\caption{ROC curve for variants detection performance by log-normal prior.}
-\label{fig:ROC_lognormal}
-\end{center}
-\end{figure}
-
-To evaluate the performance of RVD3 model with priors, we generated receiver-operating characteristic curves (ROCs) for median read depth and minor allele frequencies (MAFs).
-Here the Bayesian test is used without the $\chi^2$ test. Figure~\ref{fig:ROC_jeffrey} and Figure~\ref{fig:ROC_lognormal} shows ROC curves with a fixed $\alpha=0.05$.
-The performance improves when the read depth goes up. (ROC shows that the model with priors performs better than improper prior situation especially on the small read depth).
-Noticed at the lowest depth (22) with 10.0\% mutant mixture, the sensitivity and specificity value are 1 and much better than the model with improper priors for $M_j$, which definitely demonstrates the advantage of priors.
-
-\subsubsection{Sensitivity/Specificity/FDR}
-
-Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
-Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{fid:FDR} shows the false discovery rate of the RVD3 with different priors. Jeffreys prior shows a smaller false discovery rate than others.
-It is obvious no matter Jeffreys or log-normal, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, campared with improper priors.
-Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
-Here we chose Jeffreys prior model because it's a non-information prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=80mm]{tables/Sen_Speci.pdf}
-\caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
-\label{fig:SS}
-\end{center}
-\end{figure}
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=80mm]{tables/FDR.pdf}
-\caption{False Discovery Rate comparison with different priors on RVD3.}
-\label{fig:FDR}
-\end{center}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results of Jeffreys prior on yeast data
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Results of Jeffreys prior on yeast data}
-
-We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants \citep{kvitek2013whole}.
-
-\section{Discussion}
-analysis of the tradeoff between experimental precision and sequencing depth to achieve ultrasensitive variant detection.
-
-The empirical Bayesian commonly causes the false discovery rate (FDR) \citep{efron2010large}.
-
-\section{Conclusion}
-
-\appendix
-%%%%%%%%%%%%%%%%%%%
-% Appendix A: Parameter Initialization
-%%%%%%%%%%%%%%%%%%%
-\section{Parameter Initialization}\label{sec:appendix_mom}
-Since $r_{ji} \thicksim \text{Binomial}(n_{ji}, \theta_{ji})$, the first population moment is  $E[r_{ji}] = \theta_{ji} n_{ji}$ and the first sample moment is simply $m_1 = r_{ji}$.
-Therefore the MoM estimator is
-\begin{equation}
-	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
-\end{equation}
-
-We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
-The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$. The first and second population moments are
-\begin{eqnarray}
-	E[\theta_{ji}] =& \mu_j,\\
-	\text{Var}[\theta_{ji}] =& \frac{\mu_j(1-\mu_j)} { M_j + 1 }.
-\end{eqnarray}
-The first and second sample moments are $m_1 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}$ and $m_2 = \frac{1}{N}\sum_{i=1}^N \theta_{ji}^2$.
-Setting the population moments equal to the sample moments and solving for $\mu_j$ and $M_j$ gives
-\begin{eqnarray}
-	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
-	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
-\end{eqnarray}
-
-Following the same procedure for the parameters of $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$ gives the following MoM estimates
-\begin{eqnarray}
-	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
-	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
-\end{eqnarray}
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix B: Inference of Jeffreys prior
-%%%%%%%%%%%%%%%%%%%
-\section{Inference of Jeffreys prior}\label{sec:appendix_Jeffreys}
-We assume there is only one replicate (i=1),
-
-\begin{equation}\label{eqn:Betapdf}
-p\left({\theta }_{j} \right)= \frac{\Gamma \left({M}_{j} \right)}{\Gamma \left({\mu }_{j} {M}_{j}\right)\Gamma \left(( 1-{\mu }_{j}){M}_{j}\right)} {{\theta}_{j}}^{{\mu}_{j}{M}_{j}-1}{\left(1-\theta\right)_{j}}^{\left(1-{\mu}_{j}\right){M}_{j}-1}
-\end{equation}
-
-\begin{equation}\label{equ:JefferyInference1}
-\begin{split}
-&\log p\left(\theta_{j}|\mu_{j},M_{j}\right) =\log \Gamma \left(M_{j}\right)-\log \Gamma\left(\mu_{j},M_{j}\right)\\
-&- \log \Gamma\left(1-\mu_{j},M_{j}\right) + (\mu_{j}M_{j}-1)\log\theta_{j}\\
-& + ((1-\,u_{j})M_{j}-1)\log(1-\theta_{j})\
-\end{split}
-\end{equation}
-
-\begin{equation}
-\begin{split}
-&\frac{\delta\log p(\theta_{j})}{\delta M_{j}} =\Psi(M_{j}) - \Psi(\mu_{j} M_{j})\mu_{j}\\
-&- \Psi((1-\mu_{j})M_{j})(1-\mu_{j}) +\mu_{j}\log\theta_{j} + (1-\mu_{j})\log(1-\theta_{j})\
-\end{split}
-\end{equation}
-
-\begin{equation}
-\frac{\delta^{2}\log p(\theta_{j})}{\delta M_{j}^{2}}  = \Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j})(1-\mu_{j})^{2}
-\end{equation}
-
-Now we have the Jeffreys' prior $\pi\left({M}_{j}\right)$ for $M_{j}$:
-
-\begin{equation}
-[-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%%
-% Appendix C: Key parameters for RVD3 model with log-normal prior
-%%%%%%%%%%%%%%%%%%%
-
-\section{Key parameters for RVD3 model with log-normal prior}\label{sec:appendix_parameters}
-Key parameters of the RVD3 with log-normal prior is ploted in Figure~\ref{fig:M_lognormal}.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=90mm]{figs/M_lognormal.pdf}
-\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
-\label{fig:M_lognormal}
-\end{center}
-\end{figure}
-
-\bibliographystyle{apalike}
-\bibliography{bioinfo}
 \end{document}

### 2014-03-03 doc/2014bioinformatics_zhang/RVD3.tex
Updated draft of RVD3.

@@
 We present a Bayesian sensitivity analysis of the model to variations in the prior function.
-Our model with different priors both performs a high accuracy, and a Jeffreys prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with an improper prior.
+Our model with different priors both performs a high accuracy, and a Jeffrey's prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with an improper prior.
 In an analysis of a directed evolution experiment, we are able to detect the emergence of a beneficial SNV earlier than was previously shown.
@@
 Secondly, Metropolis-within-Gibbs sampling is evolved for inference. And then to detect variants, a Bayesian posterior distribution is taken for hypothesis test.
-Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffreys prior and log-normal prior, and apply the different priors on the synthetic sequence data.
-Finally, we choose Jeffreys prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
+Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffrey's prior and log-normal prior, and apply the different priors on the synthetic sequence data.
+Finally, we choose Jeffrey's prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
 Thus our Bayesian model achieves a enhanced robustness and accuracy when calling variants for the low read depth and minor allele frequencies.
@@
 Three parameters of the model are: $\mu_0$, a global error rate; $M_0$, the global position which estimates the variation in the error rate across the positions;
-to choose a priori distribution for $M_j$, log-normal prior and Jeffreys prior \citep{jeffreys1946invariant} are employed, which enhances the former RVD model \citep{Flaherty:2011ja}.
+to choose a priori distribution for $M_j$, log-normal prior and Jeffrey's prior \citep{jeffreys1946invariant} are employed, which enhances the former RVD model \citep{Flaherty:2011ja}.
 Here $M_j$ is the local precision measures the alteration in the error rate across replicates at position $j$. The graphical chart for RVD3 is shown in Figure~\ref{fig:graphical_model}.
@@
 The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
-And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and Jeffreys prior for $M_j$.
+And the local precision is modeled by $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$ (log-normal prior), and Jeffrey's prior for $M_j$.
 %$\pi\left({M}_{j}\right)$
@@
 
-Metropolis-within-Gibbs sampling is evolved for inference. Algorithm 1 shows the inference process and the detail are also illustrated.
-
-\begin{algorithm}[ht]
-\caption{Inference process for Metropolis-within-Gibbs}
-\label{alg:metro_gibbs}
-\begin{algorithmic}[1]
-
-\State Initialize $\theta$, $\mu$, $M_j$, $\mu_0$, $M_0$
-\Repeat
-\For {each location j}
-  \State Samples from $p \left( \mu_j |\theta_{ij},\mu_0,M_0\right)$ \Comment{Sample $\mu_j$}
-  \State Set $\mu_j$ to the sample median for the samples
-  \State Samples from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$ \Comment{Sample $M_j$}
-
-  \For {each replicate i}
-	\State Sample from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$ \Comment{Sample $\theta_{ji}$}
-  \EndFor
-
-\EndFor
-\Until {sample size sufficient}
-\end{algorithmic}
-\end{algorithm}
-
-%%%%%%%%%%%%%
-% Initialization
-%%%%%%%%%%%%%
-\subsubsection{Initialization}
-[This paragrah COPIED]The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
-MoM works by setting the population moment equal to the sample moment.
-A system of equations is formed such that the number of moment equations is equal to the number of unknown parameters and the equations are solved simultaneously to give the parameter estimates.
-We simply start with the data matrices $r$ and $n$ and work up the hierarchy of the graphical model solving for the parameters of each conditional distribution in turn.
-
-The initial parameter estimates and derivations are provided in Appendix~\ref{sec:appendix_mom}. Below is the MoM estimate for replicate-level parameters
-$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
-The estimates for the position-level parameters are
-$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$
-and
-$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$.
-The estimates for the genome-level parameters are
-$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$
-and
-$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$.
-
-%%%%%%%%%%%%%%%%%%
-% Sampling theta
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
-
-Because the Bayesian conjugacy between the prior
-$p(\theta_{ji} | \mu_j, M_j) \thicksim \text{Beta}(\mu_j, M_j)$
-and the likelihood
-$p(r_{ji} | n_{ji}, \theta_{ji}) \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$,
-we draw the samples from the posterior distribution
-$p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)$
-The posterior distribution is
-\begin{equation}
-	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
-\end{equation}
-
-%%%%%%%%%%%%%%%%%%
-% Sampling mu
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
-Based on Markov blanket, the posterior distribution over $\mu_j$ is
-\begin{equation}
-	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
-\end{equation}
-
-[This paragrah COPIED]Since the prior, $p(\mu_j | \mu_0, M_0)$, is not conjugate to the likelihood, $p(\theta_{ji} | \mu_j, M_j)$, we sample from the posterior distribution using the Metropolis-Hastings algorithm.
-By experience when $\mu_j^{(p)} \in (10^{-3},1-10^{-3})$, the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration is set to $\sigma_j = 0.1 \cdot \mu_j^{(p)}$ ;
-otherwise, we set $\sigma_j = 10^{-4}$ if $\mu_j^{(p)}< 10^{-3}$ and $\sigma_j = 10^{-1}-10^{-4}$ if $\mu_j^{(p)}>1-10^{-3}$.
-We have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
-
-We resampled from the proposal if the sample is outside of the support of the posterior distribution.
-We throw away a burn-in period - 20\% of the samples, and thin the chain by a factor 2 to reduce autocorrelation among samples, resulting in a sample with size 1600 from the posterior distribution.
-
-%%%%%%%%%%%%%%%%%%
-% Sampling Mj
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Sampling from $p \left( M_{j} |\mu,\sigma, \theta_{ji},\mu_j\right)$}
-Since Jeffreys prior is from the Fisher information and in RVD3 model ${\theta }_{ji}\sim Beta\left( {\mu }_{j},{M}_{j}\right)$,
-
-\begin{equation}\label{equ:JefferyInference}
-I\left({M}_{j}\right)={E}_{{M}_{j}}\left[ -\frac{\delta ^{2}\log p\left(\theta _{j}|\mu_{j},M_{j}\right)}{\delta M^{2}_{j}}\right]
-\end{equation}
-
-We calculated the equations Appendix~\ref{sec:appendix_Jeffreys} and obtained the Jeffreys' prior for $M_j$:
-
-\begin{equation}
-[-\left(\Psi_{1}(M_{j}) - \Psi_{1}(\mu_{j} M_{j})\mu_{j}^{2} - \Psi_{1}((1-\mu_{j})M_{j}){(1-\mu_{j})^{2}}\right)]^{\frac{1}{2}}
-\end{equation}
-
-For log-normal prior, the posterior distribution over $M_{j}$ given its Markov blanket is
-
-\begin{equation}
-	p( M_{j} |\mu, \sigma, \theta_{ji},\mu_j) \propto p(\theta_{ji} | \mu_j, M_j) p(M_{j} | \mu, \sigma)
-\end{equation}
-
-We have $ \theta_{ji}\thicksim\text{Beta}(\mu_{j},M_{j})$, and $ M_{j} \thicksim \text{log-normal}(\mu, \sigma)$.
-Instead of computing the posterior distribution directly, Metropolis-Hastings algorithm was taken to sample from the posterior distribution.
-
-%%%%%%%%%%%%%%%%%%
-% Posterior Density Test
-%%%%%%%%%%%%%%%%%%
-\subsubsection{Posterior Density Test}\label{sec:hypothesis_test}
-Posterior distributions of $\mu_j$ for the control and case are achieved -  $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$, by Metropolis-within-Gibbs.
-So we called a variant when $\tilde{\mu}_j^{\text{case}} > \tilde{\mu}_j^{\text{control}}$ with $1-\alpha$ confidence,
-\begin{equation}\label{eqn:bayes_test}
-	\Pr( \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}} \geq \tau ) > 1-\alpha,
-\end{equation}
-where $\tau$ is a detection threshold and $1-\alpha$ is the confidence level. We set $\tau = 0$ in our experiment [XXX].
-
-%%%%%%%%%%%%%%%%%%
-% Chi^2 Test
-%%%%%%%%%%%%%%%%%%
-\subsubsection{$\chi^2$ test for non-uniform base distribution}
-[This part is COPIED]
-
-An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error;
-we would like to differentiate these two scenarios. We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases.
-In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.
-
-We use a $\chi^2$ goodness-of-fit test on a multinomial distribution over the non-reference bases to distinguish these two possible scenarios.
-The null hypothesis is $H_0: p = (p_1, p_2, p_3)$ where $p_1=p_2=p_3=1/3$.
-Cressie and Read (1984) identified a power-divergence family of statistics, indexed by $\lambda$,
-that includes as special cases Pearson's $\chi^2 (\lambda = 1)$ statistic, the log likelihood ratio statistic $(\lambda = 0)$, the Freeman-Tukey statistic $(\lambda = -1/2)$, and the Neyman modified statistic $X^2 (\lambda = -2)$.
-The test statistic is
-
-\begin{equation}
- 2nI^\lambda = \frac{2}{\lambda(\lambda+1)}\sum_{k=1}^3 r_{ji}^{(k)} \left[\left(\frac{r_{ji}^{(k)}}{E_{ji}^{(k)}}\right)^\lambda-1\right];\lambda \in R,
-\end{equation}
-
-where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
- \citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
-
-We control for multiple hypothesis testing in two ways.
-We use Fisher's combined probability test \citep{fisher1970statistical} to combine the p-values for $N$ replicates into a single p-value at position $j$,
-
-\begin{equation}\label{eqn:fisher_combined}
-	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
-\end{equation}
-
-Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{benjamini1995controlling, efron2010large}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Priors for precision parameter}
-%\subsubsection{Improper Prior}
 The prior distribution characterizes the knowledge of the parameters in the statistical model. Including prior information in the Bayesian approach is difficult but meaningful.
@@
 Even though no problems happened in estimation by improper prior, other troubles could be caused in the Bayesian inference and analysis by it \citep{stein1965approximation}.
-Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-information prior and information prior for sensitivity analysis.
+Playing no prior on $ M_{j}$ is exactly an implicit improper prior. Therefore we considered about non-informative prior and informative prior for sensitivity analysis.
 
-%\subsubsection{Jeffreys Prior}
 Non-information prior seems to be more unbiased and objective.
 Various non-informative prior distributions have been suggested for parameters in hierarchical models.
-Jeffreys prior, as a typical and influential one, is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys \citep{jeffreys1946invariant}.
-It is defined in terms of the Fisher information and works well with a single parameter. In our research Jeffreys prior for $M_j$ is the square root of Fisher information of $M_j$.
+Jeffrey's prior, as a typical and influential one, is proposed to establish a least informative prior that is automatically invariant to transformations by Harold Jeffreys \citep{jeffreys1946invariant}.
+It is defined in terms of the Fisher information and works well with a single parameter. In our research Jeffrey's prior for $M_j$ is the square root of Fisher information of $M_j$.
 
 Informative prior distribution is the most specific type of prior. A good informative prior is imperative to promote accurate posterior estimates.
-In our study, log-normal prior is a typical prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
+In our study, log-normal prior is a proper prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
 The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
@@
 %%%%%%%%%%%%%%%%%%
-\subsection{Sensitivity analysis of priors}
-To examing the robustness of our Bayesian model, different plausible prior distributions can be utilized \citep{gelman2006prior}.
-For this purpose, we performed information prior (log-normal) and non-information prior (Jeffreys).
-We found that RVD3 is robust to changing of the priors. The RVD3 algorithm provides estimates of model parameters and latent variables given the data.
-
-We show several of these parameters of the model with Jeffreys prior in Figure~\ref{fig:M_jeffrey}.
-The left column of is shows the read depth for each of the six BAM files (three replicates each with two read pairs) for each data set.
-Because the DNA was not sheared and ligated prior to sequencing, the read depth drops to zero at the boundaries. For the 100\% mutant data set, the read depth drops at the mutant locations.
-This is due to the parameters imposed at the alignment stage.
-The right column of Figure~\ref{fig:M_jeffrey} reveals the parameter estimates $\hat{M}_j$ and $\hat{M}_0$ for each data set. $M_j$ measures the variance between replicates at location $j$.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-Interestingly, $M_j$ from Jeffreys prior [Figure~\ref{fig:M_jeffrey}] across the M value drops at the mutant positions when the data is 10\% and 100\% mutant, which is not shown in log-normal prior [Figure~\ref{fig:M_lognormal}] in Appendix~\ref{sec:appendix_parameters}.
-Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots in the right column.
-$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision across positions.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=90mm]{figs/M_jeffrey.pdf}
-\caption{Key parameters for RVD3 model with Jeffreys prior on synthetic DNA data sets.}
-\label{fig:M_jeffrey}
-\end{center}
-\end{figure}
-
-\begin{figure*}[!tbp]
-\begin{center}
-\includegraphics[width=190mm]{figs/post_prior_dilu_10.pdf}
-\caption{Distribution of priors and posteriors when dilution is 10\%}
-\label{fig:dilu_10}
-\end{center}
-\end{figure*}
+\subsection{Sensitivity analysis}
+A way to choose the prior function is to use the Schwarz criterion or Bayesian information criterion (BIC) \citep{weakliem1999critique}.
+Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}.
+To consider the Bayesian sensitivity analysis, another way is to consider the range of posterior quantities of interest, such as the posterior mean or posterior probability \citep{bayarri1998robust}.
+Based on this theory, we use analysis the sensitivity of our model by analyzing the posterior and prior probability.
+Different plausible prior distributions can be utilized for this purpose\citep{gelman2006prior}.
+We performed two main classes of priors: informative prior (log-normal) and non-informative prior (Jeffrey's).
 
@@
 The log-normal prior attributes a higher prior probability to M values between 0 and 2000.
-The Jeffreys prior assigns more information than the log-normal prior (flat) from the posterior curve.
+The Jeffrey's prior assigns more information than the log-normal prior (flat) from the posterior curve.
 They both want to search for a small value for M from the prior curve.
+
+There is little variability across positions indicating that the replication variance does not change greatly across position.
+The proper prior (log-normal) and non-informative prior (Jeffrey¡¯s) are performed to compare with the improper prior.
+We show several key parameters of the model in Figure~\ref{fig:priors_compare}.
+A key model parameter M is less variable with a log-normal and improper prior compared to the Jeffrey¡¯s prior.
+Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots.
+$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision across positions.
+
+\begin{figure*}[!tbp]
+\begin{center}
+\includegraphics[width=185mm]{figs/post_prior_dilu_10.pdf}
+\caption{Distribution of priors and posteriors when dilution is 10\%}
+\label{fig:dilu_10}
+\end{center}
+\end{figure*}
+
+\begin{figure}[htbp]
+\begin{center}
+\includegraphics[width=160mm]{figs/priors_compare.pdf}
+\caption{Key parameters for RVD3 model with different priors.}
+\label{fig:priors_compare}
+\end{center}
+\end{figure}
 
@@
 \includegraphics[width=90mm]{figs/ROC_without_chi2_jeffrey.pdf}
-\caption{ROC curve for variants detection performance by Jeffreys prior.}
+\caption{ROC curve for variants detection performance by Jeffrey's prior.}
 \label{fig:ROC_jeffrey}
@@
 Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
-Log-normal prior shows a higher sensitivity and specificity value than the Jeffreys prior.
-Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffreys prior shows a smaller false discovery rate than others.
-It is obvious no matter Jeffreys or log-normal, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, campared with improper priors.
-Based on the various advantages for Jeffreys and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
-Here we chose Jeffreys prior model because it's a non-information prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+Log-normal prior shows a higher sensitivity and specificity value than the Jeffrey's prior.
+Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffrey's prior shows a smaller false discovery rate than others.
+It is obvious no matter Jeffrey's or log-normal, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, compared with improper priors.
+Based on the various advantages for Jeffrey's and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
+Here we chose Jeffrey's prior model because it's a non-information prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Results of Jeffreys prior on yeast data
+% Results of Jeffrey's prior on yeast data
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Results of Jeffreys prior on yeast data}
+\subsection{Results of Jeffrey's prior on yeast data}
 
-We demonstrated our RVD3 model with Jeffreys prior on yeast data to identify the variants \citep{kvitek2013whole}.
+We demonstrated our RVD3 model with Jeffrey's prior on yeast data to identify the variants \citep{kvitek2013whole}.
 
@@
 %%%%%%%%%%%%%%%%%%%
-\section{Inference of Jeffreys prior}\label{sec:appendix_Jeffreys}
+\section{Inference of Jeffrey's prior}\label{sec:appendix_Jeffreys}
 We assume there is only one replicate (i=1),
@@
 
-%%%%%%%%%%%%%%%%%%%
-% Appendix C: Key parameters for RVD3 model with log-normal prior
-%%%%%%%%%%%%%%%%%%%
-
-\section{Key parameters for RVD3 model with log-normal prior}\label{sec:appendix_parameters}
-Key parameters of the RVD3 with log-normal prior is ploted in Figure~\ref{fig:M_lognormal}.
-
-\begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=90mm]{figs/M_lognormal.pdf}
-\caption{Key parameters for RVD3 model with log-normal prior for synthetic DNA data sets.}
-\label{fig:M_lognormal}
-\end{center}
-\end{figure}
-
 \bibliographystyle{apalike}

### 2014-03-04 doc/2014bioinformatics_zhang/RVD3.tex
Updated draft of RVD3 (add the Bayesian sensitivity analysis review)

@@
 Secondly, Metropolis-within-Gibbs sampling is evolved for inference. And then to detect variants, a Bayesian posterior distribution is taken for hypothesis test.
-Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffrey's prior and log-normal prior, and apply the different priors on the synthetic sequence data.
+Furthermore, we analyze the sensitivity of the Bayesian model to different priors - Jeffrey's prior, log-normal prior, and improper prior.
 Finally, we choose Jeffrey's prior for the RVD3 model and demonstrate its performance on the yeast sequence data.
@@
 \subsection{Priors for precision parameter}
-The prior distribution characterizes the knowledge of the parameters in the statistical model. Including prior information in the Bayesian approach is difficult but meaningful.
+The prior distribution characterizes the knowledge of the parameters in the statistical model.
+Including prior information in the Bayesian approach is difficult but meaningful.
+A way to choose the prior function is to use the Schwarz criterion or Bayesian information criterion (BIC) \citep{weakliem1999critique}.
+Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}.
+To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust}.
+Based on this theory, we analyze the posterior and prior probability of the precision hyper-parameter.
+For this purpose, a bunch of different priors should be chosen for analysis \citep{gelman2006prior}.
+In a polygenic modeling with Bayesian sparse linear mixed model research, the results don't change dramatically by changing the prior measurement which reveals that the prior specification for the hyper-parameters are fine \citep{zhou2013polygenic}.
+So we performed three different plausible prior distributions: improper prior, informative prior (log-normal), and non-informative prior (Jeffrey's).
+
 An improper prior is the prior distribution integrates to infinity, and may cause an improper posterior which results in an invalid inferences \citep{lesaffre2012bayesian}.
@@
 
-Non-information prior seems to be more unbiased and objective.
+Informative prior distribution is the most specific type of prior. A good informative prior is imperative to promote accurate posterior estimates.
+In our study, log-normal prior is a proper prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
+The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
+
+Non-informative prior seems to be more unbiased and objective.
 Various non-informative prior distributions have been suggested for parameters in hierarchical models.
@@
 It is defined in terms of the Fisher information and works well with a single parameter. In our research Jeffrey's prior for $M_j$ is the square root of Fisher information of $M_j$.
-
-Informative prior distribution is the most specific type of prior. A good informative prior is imperative to promote accurate posterior estimates.
-In our study, log-normal prior is a proper prior for the beta density's parameters, $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
-The parameters of it denoted $\mu$ (mean) and $\sigma$ (standard deviation) respectively.
 
@@
 \subsection{Sensitivity analysis}
-A way to choose the prior function is to use the Schwarz criterion or Bayesian information criterion (BIC) \citep{weakliem1999critique}.
-Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}.
-To consider the Bayesian sensitivity analysis, another way is to consider the range of posterior quantities of interest, such as the posterior mean or posterior probability \citep{bayarri1998robust}.
-Based on this theory, we use analysis the sensitivity of our model by analyzing the posterior and prior probability.
-Different plausible prior distributions can be utilized for this purpose\citep{gelman2006prior}.
-We performed two main classes of priors: informative prior (log-normal) and non-informative prior (Jeffrey's).
 
@@
 The posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
-Generally the non-mutant positions display normal and stable without a peak nor a strange shape.
+Generally the distribution plots display normal and stable without a peak nor a strange shape.
 The two distributions show different ways of the prior and posterior.
@@
 
+The proper prior (log-normal) and non-informative prior (Jeffrey¡¯s) are performed to compare with the improper prior.
 There is little variability across positions indicating that the replication variance does not change greatly across position.
-The proper prior (log-normal) and non-informative prior (Jeffrey¡¯s) are performed to compare with the improper prior.
 We show several key parameters of the model in Figure~\ref{fig:priors_compare}.
@@
 Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffrey's prior shows a smaller false discovery rate than others.
-It is obvious no matter Jeffrey's or log-normal, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, compared with improper priors.
+It is obvious no matter Jeffrey's or log-normal prior, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, compared with improper priors, which seems desirable of our model for extending for a broad application.
 Based on the various advantages for Jeffrey's and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
-Here we chose Jeffrey's prior model because it's a non-information prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+Here we chose Jeffrey's prior model because it's a non-informative prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
 

### 2014-03-04 doc/2014bioinformatics_zhang/RVD3.tex
Refactor draft folder.

@@
 Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}.
-To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust}.
+To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust, stocker2006noise, harrison2013bayesian}.
 Based on this theory, we analyze the posterior and prior probability of the precision hyper-parameter.

### 2014-03-04 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/plot_prior_posterior.py

@@ -43,8 +43,8 @@ def main():
         kernel = gaussian_kde(M1) 
         norm = kernel.integrate_box(0,10000)
         print (norm)
-        ax.semilogy(M, p/sum)
-        ax.semilogy(M, kernel(M).T/norm)
+        ax.semilogy(M, p/sum, 'r--')
+        ax.semilogy(M, kernel(M).T/norm) 
         ax.legend( ['prior','posterior'],loc=9, ncol=9)
         subtitle='position='+str(loc)
         ax.set_title(subtitle)
@@ -66,8 +66,8 @@ def main():
         # Use Gaussian Kernel Density Estimate for posterior Mj
         kernel = gaussian_kde(M1)
         norm = kernel.integrate_box(0,10000)		
-        ax.semilogy(M, p/sum)
-        ax.semilogy(M, kernel(M).T/norm)
+        ax.semilogy(M, p/sum,'r--')
+        ax.semilogy(M, kernel(M).T/norm) 
         ax.set_title('location %d' % loc, fontsize=15)
         subtitle='position='+str(loc)
         ax.set_title(subtitle)

### 2014-03-07 doc/2014bioinformatics_zhang/RVD3.tex
2014-03-05-RVD3_sensitivity_analysis, and updated draft.

--- 2014-03-04 doc/2014bioinformatics_zhang/RVD3.tex
@@
 \begin{abstract}
-%Massively parallel sequencing data has been generated by next-generation sequencing (NGS) technology for single nucleotide variants (SNVs) identification among related populations.
-%To address the detection of SNVs, a variety of methods are being under-represented.
-%However, by the reason of the error rate of the clonal heterogeneity and the limitation of the algorithms, identifying the true variants at minor allele frequencies for the very low sequencing read depth remains challenging.
-
-%We propose a novel empirical Bayesian model to call variants in heterogeneous samples accurately.
-We present a Bayeisan sensitivity analysis of this hierarchical model to variations of the prior function.
-%Our model with information prior (log-normal prior) and non-information prior (Jeffreys prior) both performs a high accuracy (96\%) when applied to a known 0.1\% minor allele frequency within very low read depth (39).
-%Moreover, the model with Jeffreys prior presents much lower false discovery rate (FDR) to a known 0.1\% minor allele frequency event, compared with using improper prior.
-%For further validation, our statistical model was applied on the yeast sequence data which also shows a high specificity and sensitivity over a wide range of read depth.
-%This model is able to detect the emergence of a beneficial SNV earlier than was previously shown.
-%Variant detection model with improved robustness and accuracy for low-depth targeted next-generation sequencing data
 
@@
 We present a Bayesian sensitivity analysis of the model to variations in the prior function.
-Our model with different priors both performs a high accuracy, and a Jeffrey's prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with an improper prior.
+Our model with different priors both performs a high accuracy, and a Jeffrey's prior gives a lower false discovery rate (FDR) to detect a 0.1\% minor allele frequency event within minor read depth compared with the log-normal prior and an improper prior.
+And log-normal prior performs a higher specificity than the other two. 
 In an analysis of a directed evolution experiment, we are able to detect the emergence of a beneficial SNV earlier than was previously shown.
@@
 Besides, the Bayes factor does tend to be more sensitive to access the prior distributions on the model parameters \citep{kass1995bayes}.
-To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust}.
+To consider the Bayesian sensitivity analysis, a more practical way is to consider the range of posterior quantities of interest- the posterior mean or posterior probability \citep{bayarri1998robust, harrison2013bayesian}.
 Based on this theory, we analyze the posterior and prior probability of the precision hyper-parameter.
@@
 
-The Bayesian posterior predictive distributions and the priors distributions are shown in Figure~\ref{fig:dilu_10}.
+The Bayesian posterior distributions and the priors distributions are shown in Figure~\ref{fig:dilu_10}.
 These plots indicate the probability distribution over different M values when dilution is 10\% at $100\times$ read depth rate.
-The positions are chosen in the middle of the base length- position 104 and 244 are mutant, and position 100 and 300 are non-mutant.
-The posterior probability distributions are estimated by Gaussian Kernel Density \citep{silverman1986density}.
-Generally the distribution plots display normal and stable without a peak nor a strange shape.
-The two distributions show different ways of the prior and posterior.
-The log-normal prior attributes a higher prior probability to M values between 0 and 2000.
-The Jeffrey's prior assigns more information than the log-normal prior (flat) from the posterior curve.
-They both want to search for a small value for M from the prior curve.
+The positions are chosen in the middle of the base length- position 185 and 265 are mutant, and position 200 and 300 are non-mutant.
+The posterior probability distributions for $M_j$ are estimated by Gaussian Kernel Density \citep{silverman1986density}.
+Generally the distribution curves display normal and stable.
+Jeffrey's posterior distribution changes a lot which shows Jeffrey's piror is more sensitive than the log-normal prior between the mutant and non-mutant positions.
 
-The proper prior (log-normal) and non-informative prior (Jeffrey¡¯s) are performed to compare with the improper prior.
-There is little variability across positions indicating that the replication variance does not change greatly across position.
-We show several key parameters of the model in Figure~\ref{fig:priors_compare}.
-A key model parameter M is less variable with a log-normal and improper prior compared to the Jeffrey¡¯s prior.
+Figure~\ref{fig:priors_compare} shows that across all priors the non-mutant positions indicate that the replicate variance is small. 
+At the known mutant positions, the replicate parameter $M_j$ is less variable compared to the reference position with the log-normal and improper prior compared with Jeffrey's prior, which corresponds to the information in Figure~\ref{fig:dilu_10}
+However, at the begining and the end of the sequence which are the low read-depth positions, the improper and log-normal prior are more variable compared to Jeffrey's prior.
 Additionally, the error rate across positions is captured by the $M_0$ parameter shown as a horizontal dotted line in the plots. 
-$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision across positions.
+$M_j$ is greater than $M_0$ shows that the precision between replicates is higher than the precision between positions. 
+Here $\widetilde{E}[M|D]$ represents the expectation of empirical posterior approximation.
 
@@
 \begin{center}
-\includegraphics[width=185mm]{figs/post_prior_dilu_10.pdf}
-\caption{Distribution of priors and posteriors when dilution is 10\%}
+\includegraphics[width=170mm]{figs/post_prior_10.pdf}
+\caption{Distribution of priors and posteriors when MAF is 10\%}
 \label{fig:dilu_10}
@@
 \begin{center}
-\includegraphics[width=160mm]{figs/priors_compare.pdf}
-\caption{Key parameters for RVD3 model with different priors.}
+\includegraphics[width=140mm]{figs/priors_compare.pdf}
+\caption{Precision parameter for RVD3 model with different priors. $M_0$ is the global precision. X indicates mutant position.}
 \label{fig:priors_compare}
@@
 
-Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors compared with the model with improper priors.
-Log-normal prior shows a higher sensitivity and specificity value than the Jeffrey's prior.
-Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffrey's prior shows a smaller false discovery rate than others.
+Figure~\ref{fig:SS} shows that the sensitivity and specificity of the RVD3 of different priors.
+Log-normal prior performs a higher sensitivity and specificity value, and the specificity of log-normal prior is a little better than the improper prior and much better than the performance of Jeffrey's prior.
+It shows that log-normal prior performs a smaller false positive rate than the other two priors.
+
+Figure~\ref{fid:fdr} shows the false discovery rate of the RVD3 with different priors. Jeffrey's prior shows a smaller false discovery rate than others over all the MAF levels at various read depths.
 It is obvious no matter Jeffrey's or log-normal prior, the variant detection performance acquires lower FDR to a known 0.1\% minor allele frequency event, compared with improper priors, which seems desirable of our model for extending for a broad application.
-Based on the various advantages for Jeffrey's and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter.
-Here we chose Jeffrey's prior model because it's a non-informative prior, and more attention should be paid to false discovery rate and accuracy for variants calling research, compared with the clinical experiment or diagnosis which cares more on true positive rate and true negative rate.
+Based on the various advantages for Jeffrey's and log-normal prior, RVD3 can afford a more appropriate choice for the precision parameter to accommodate to the target.
+When we do the variants calling research, we should choose Jeffrey's prior model because it's a non-informative prior, and can produce a smaller false discovery rate and high accuracy; when we do the clinical experiment or diagnosis, we prefer to choose log-normal prior which cares more on true positive rate and true negative rate.
 
@@
 \begin{center}
-\includegraphics[width=80mm]{tables/Sen_Speci.pdf}
+\includegraphics[width=110mm]{tables/Sen_Speci.pdf}
 \caption{Sensitivity/Specificity comparison of RVD3 with different priors.}
@@
 \begin{center}
-\includegraphics[width=80mm]{tables/FDR.pdf}
+\includegraphics[width=100mm]{tables/FDR.pdf}
 \caption{False Discovery Rate comparison with different priors on RVD3.}

### 2014-03-07 results/2013-11-26-plot_posterior_prior_over_M_rvd29_rvd30/plot_prior_posterior.py

@@ -7,7 +7,6 @@ import sys
 import os
 import h5py
 import logging
-import itertools
 import math
 import rvd30
 import rvd29
@@ -19,11 +18,13 @@ logging.basicConfig(level=logging.DEBUG,
                     format='%(levelname)s:%(message)s')
 
 def main():
-    locationList = (100,300,104,244)
+    # Actual positions are 200,300,205,265
+    locationList = (199,299,184,264)
     
     # Same experiment condition: read depth rate = 1/100
-    h5Filename_rvd29 = './hdf5_rvd29_gibbs=20000_dc=100/Case10_0.hdf5'
-    h5Filename_rvd30 = './hdf5_rvd30_gibbs=20000_dc=100/Case10_0.hdf5'
+    h5Filename_rvd29 = 'Case10_jeffreys.hdf5'
+    h5Filename_rvd30 = 'Case10_lognormal.hdf5'
+    
     M = range(1,10000)
     fig=plt.figure(figsize=(20, 12))
 
@@ -37,19 +38,20 @@ def main():
         mu=np.mean(mu)
         p = np.sqrt(polygamma(1, np.dot(mu,M))*(mu**2) + polygamma(1, np.dot(1-mu,M))*(1-mu)**2 -polygamma(1,M))
         sum = np.sum(p)
-        print (sum)
         M1 = M_s[loc]       
-        # Use Gaussian Kernel Density Estimate for posterior Mj
+    # Use Gaussian Kernel Density Estimate for posterior Mj
         kernel = gaussian_kde(M1) 
         norm = kernel.integrate_box(0,10000)
-        print (norm)
         ax.semilogy(M, p/sum, 'r--')
         ax.semilogy(M, kernel(M).T/norm) 
-        ax.legend( ['prior','posterior'],loc=9, ncol=9)
-        subtitle='position='+str(loc)
+        ax.legend( ['prior','posterior'],loc=1)
+        subtitle='position='+str(loc+1)
         ax.set_title(subtitle)
         if locationList.index(loc)==0:
-           ax.set_ylabel('Probability of Jeffreys prior',fontsize=15)
+           ax.set_ylabel('Jeffrey\'s prior',fontsize=15)
+        ax.set_ylim(0.000001,0.01)
+        ax.yaxis.grid()
+        ax.xaxis.grid()
 
     """ Probability of lognormal prior from rvd30"""
     logging.debug("Plot rvd30 prior and posterior")
@@ -61,9 +63,8 @@ def main():
[…]

### 2014-03-07 results/2014-03-05-RVD3_sensitivity_analysis/priors_compare.py

@@ -1,0 +1,82 @@
+import sys
+import os
+import numpy as np
+import matplotlib.pyplot as plt
+import h5py
+import logging
+import pdb
+import itertools as it
+
+rvddir = os.path.join('./')
+sys.path.insert(0, rvddir)
+import rvd29
+import rvd30
+import rvd2
+
+# Sensitivity analysis: Compare the M value in the depth chart = 10, dilution = 100%
+
+logging.basicConfig(level=logging.DEBUG,
+                    format='%(levelname)s:%(message)s')
+
+def main():
+
+    N=1000 # Z sampling size  
+    (n_gibbs, nmh) = (4000, 50)
+
+# Plot E[M|Data] for RVD3 with Jeffreys prior
+    logging.debug("Processing M for Jeffreys prior")
+    caseFile = "./Case100_Jeffreys.hdf5" 
+    (casePhi, caseM, caseTheta, caseMu, caseLoc, caseR, caseN) = rvd29.load_model(caseFile)
+    #fig = plt.figure()
+    fig=plt.figure(figsize=(20, 12))
+    ax1 = plt.subplot(131)
+    caseLoc = [int(x.split(':')[1]) for x in caseLoc]
+    #ax1.plot(caseLoc,caseN.T)
+    ax1.semilogy(caseLoc,np.mean(caseM,axis=1),color='b',label="M")
+    ax1.set_title('Jeffrey\'s prior')
+    ax1.semilogy(caseLoc, list(it.repeat(casePhi['M0'],400)),linewidth=2,color='r',ls='--',label="M0")
+    ax1.legend(loc='upper left')
+    ax1.set_xlabel('Position',fontsize=15)
+    ax1.set_ylabel('E(M|x)',fontsize=15)
+    ax1.set_ylim(0.1, 100000000)
+    ax1.plot([85,105,125,145,165,185,205,225,245,265,285,305,325,345],[1,1,1,1,1,1,1,1,1,1,1,1,1,1],'kx')
+
+
+# Plot E[M|Data] for RVD3 with log-normal prior
+    logging.debug("Processing M for log-normal prior")
+    caseFile = "./Case100_lognormal.hdf5" 
+    (casePhi, caseM, caseTheta, caseMu, caseLoc, caseR, caseN) = rvd30.load_model(caseFile)
+    ax2 = plt.subplot(132)
[…]

### 2014-03-07 results/2014-03-05-RVD3_sensitivity_analysis/rvd2.py

@@ -1,0 +1,788 @@
+#!/usr/bin/env python
+
+
+from __future__ import print_function
+from __future__ import division
+
+import numpy as np
+
+import scipy.stats as ss
+from scipy.special import gammaln, polygamma
+from scipy import special
+
+import logging
+import multiprocessing as mp
+from itertools import repeat
+import h5py
+import tempfile
+
+import os
+import subprocess
+from datetime import date
+
+import re
+import pdb
+import time
+
+
+"""rvd29.py: put Jeffrey's Prior on Mj."""
+
+
+
+def main():
+    import argparse
+
+    # Populate our options, -h/--help is already there.
+    description = """
+                    RVD is a hierarchical bayesian model for identifying
+                    rare variants from short-read sequence data. """
+                    
+    # create the top-level parser
+    argp = argparse.ArgumentParser(prog='rvd', description=description)
+    argp.add_argument('--version', action='version', version='%(prog)s 2.7')
+    argp.add_argument('-v', '--verbose', dest='verbose', action='count',
+                help="increase verbosity (specify multiple times for more)")
+    
+    # argp.add_argument('cmd', action='store', nargs='*',
+    #             choices=['gen', 'gibbs'])
+    subparsers = argp.add_subparsers(help='sub-command help')
+    
+    # create subparser for gibbs fitting
+    argpGibbs = subparsers.add_parser('gibbs', 
+                        help='fit the RVD model using Gibbs sampling')
+    argpGibbs.add_argument('dcfile', nargs='+',
+                        help='depth chart file name')
+    argpGibbs.add_argument('-o', dest='outputFile', 
+                default='output.hdf5',
+                help='output HDF5 file name')
+    argpGibbs.add_argument('-p', '--pool', type=int, default=None,
+                help='number of workers in multithread pool')
+    argpGibbs.set_defaults(func=gibbs)
+                
+                
+    # create subparser to compare two model files
+    argpTest = subparsers.add_parser('test', 
+                        help='test if case error rate is greater than control by T')
+    argpTest.add_argument('controlHDF5Name',
+                help='control model file (HDF5)')
[…]

### 2014-03-13 results/2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset/mutation_dotplot.py
Experiment on plotting the mutation detection result from multiple algorithms.

@@ -1,0 +1,110 @@
+# Function created by Yuting He on 03/13/2014
+from __future__ import print_function
+from __future__ import division
+
+import sys
+import os
+import numpy as np
+
+import logging
+import pdb
+
+import matplotlib.pyplot as plt
+import matplotlib.gridspec as gridspec
+from matplotlib.ticker import MultipleLocator
+
+import pandas as pd
+import vcf
+
+# Insert the src/python/directory at front of the path
+rvddir = os.path.join('../../src/python/rvd27')
+sys.path.insert(0, rvddir)
+import rvd27
+
+def main():
+	vcfFile = 'Varscan2_somatic_HCC1187.vcf'
+	vcf_reader = vcf.Reader(open(vcfFile, 'r'))
+	varscan_position =[record.POS for record in vcf_reader]
+
+	muTect_position = [154743899, 154749704,154753635,154754371,154758813,
+	154760439,154766700,154766732, 154777014,154777118, 154780960]
+
+	RVD2_somatic_position = [154743899,154749704, 154753635, 
+	154754371,154757241,154758813,154760439,154766700,
+	154766732,154766832,154777118,154780960,154781769]
+
+	RVD2_diff_position = [154749704, 154753635, 154754371,
+	154758813,154760439,154766732,154766832,154777118]
+
+	Strelka_position = [154760439]
+
+	position = list(set(varscan_position+muTect_position+
+		RVD2_somatic_position+RVD2_diff_position+Strelka_position))
+
+	# plot varscan
+	varscan_idx = [position.index(pos)+1 for pos in varscan_position]
+
+	muTect_idx = [position.index(pos)+1 for pos in muTect_position]
+
+	RVD2_somatic_idx = [position.index(pos)+1 for pos in RVD2_somatic_position]
+
+	RVD2_diff_idx = [position.index(pos)+1 for pos in RVD2_diff_position]
+
+	Strelka_idx = [position.index(pos)+1 for pos in Strelka_position]
+
+	fig = plt.figure(figsize=(10,4))
+	gs = gridspec.GridSpec(1, 2, width_ratios=[4, 1]) 
[…]

### 2014-03-13 doc/bioinformatics2014/document.tex
Update manuscript

@@
 
-\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
+\title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
 \author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{to whom correspondence should be addressed}}
@@
 For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
+$ J $ is the length of gene sequence and $ N $ is the number of technical replicate gene sequence in the sample.
+\label{replicate definition} Technical replicates are used to establish experimental variability in Next Generation Sequencing procedure~\citep{quackenbush2002microarray,robasky2013role}, even though multiple replicates are not indispensable for RVD2.
+
 The model generative process is as follows:
@@
 \end{multline}
+
+\label{Improvement of RVD2}
+The improvement of RVD2 upon RVD firstly lies in the $ \text{Beta}(\mu_0, M_0) $ prior on local error rate $ \mu_j $, which accurately captured the global across-position error rate distribution.
+The prior distribution also  allows to $ \mu_j $ borrow information from adjacent positions and prevent RVD2 from failure when the read depth is too low to bring out none-zero $ r_{ji} $.
+Besides the improvement on model structure, RVD2 developed a novel Bayesian Hypothesis Testing algorithm comparing to a frequentist normal z-test in RVD.
+The performance comparison between RVD and RVD2 will be elaborated in Section~\ref{sec:comparison}.
 
@@
 
+\begin{figure*}[!tpb]
+\begin{center}
+\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}}$  is significantly different from $ \mu^{\text{control}}$.}
+\label{fig:brca_MAF}
+\end{center}
+\end{figure*}
+%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
+%TODO: add star to test that have mudelta > .16 Done.
+%TODO: set tau=0.16 for mudelta test.  Done
+
 RVD2 identified six variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
@@
 Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T.
+In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, \\ chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T.
 Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
@@
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-
-\begin{figure}[h]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}} - \mu^{\text{control}} \geq \tau$.}
-\label{fig:brca_MAF}
-\end{center}
-\end{figure}
-%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
-%TODO: add star to test that have mudelta > .16 Done.
-%TODO: set tau=0.16 for mudelta test.  Done
 

### 2014-03-13 doc/bioinformatics2014/supplementary_info.tex
update manuscript in bioinformatics2014
Update manuscript

@@
 \begin{equation}
-	\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
+	\hat{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}
 \end{equation}
 
-We take the MoM estimate, $\tilde{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
+We take the MoM estimate, $\hat{\theta}_{ji}$, as data for the next conditional distribution in the hierarchical model.
 The distribution is $\theta_{ji} \thicksim \text{Beta}(\mu_jM_j, (1-\mu_j)M_j)$.
@@
 \begin{eqnarray}
-	\tilde{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \theta_{ji}, \\
-	\tilde{M_j} =& \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1.
+	\hat{\mu}_j =& \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}, \\
+	\hat{M_j} =& \frac{ \hat{\mu}_j (1 - \hat{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}^2 } -1.
 \end{eqnarray}
@@
 \begin{eqnarray}
-	\tilde{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \mu_j \\
-	\tilde{M}_0 =& \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1.
+	\hat{\mu}_0 =& \frac{1}{J} \sum_{j=1}^J \hat{\mu}_j \\
+	\hat{M}_0 =& \frac{ \hat{\mu}_0 (1 - \hat{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \hat{\mu}_j^2 } -1.
 \end{eqnarray}
@@
 
+\section{Histogram of $ \mu_j $ for mutations detected by RVD2 in HCC1187 dataset}
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
+\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\label{fig:hist_neg}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
+\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\label{fig:hist_pos}
+\end{center}
+\end{figure}
+
+\section{Comparison between RVD2, VarScan2 somatic and muTect in Clincal HCC1187 dataset}
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_RVD2.pdf}
+\caption{RVD2.}
+\label{fig:rd_RVD2}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_mutect.pdf}
+\caption{muTect.}
+\label{fig:rd_mutect}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_true_positives.pdf}
+\caption{HCC1187 varscan2 true positives.}
+\label{fig:rd_varscan2_TP}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_false_negatives.pdf}
+\caption{ HCC1187 varscan2 false negatives.}
+\label{fig:rd_varscan2_FN}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_false_positives.pdf}
+\caption{HCC1187 varscan2 false positives}
+\label{fig:rd_varscan2_FP}
+\end{center}
+\end{figure}
+
 \bibliographystyle{natbib}

### 2014-03-14 doc/bioinformatics2014/document.tex
Update the manuscript and supplementary information

@@
 \usepackage{bbm}
+\usepackage{subfig}
 
@@
 \section{Results:}
+
+%TODO Update the chinical dataset result
 We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
@@
 
+%TODO: change the result
 We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j^{\text{control}} \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.2]$, heterozygous mutant $[0.1, 0.9]$, and homozygous mutant $[0.8, 1.0]$ and the size of the test is $\alpha=0.05$.
+To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j^{\text{control}} \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.05)$, heterozygous mutant $[0.5, 0.0.75)$, and homozygous mutant $[0.75, 1.0]$.
 
@@
 \begin{center}
-\includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
+\includegraphics[width=0.45\textwidth]{pdf_figs/ROC_without_chi2.pdf}
 \caption{ROC curve varying read depth showing detection performance.}
@@
 \begin{center}
-\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
 \caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}}$  is significantly different from $ \mu^{\text{control}}$.}
@@
 %TODO: set tau=0.16 for mudelta test.  Done
+
+\begin{figure*}[!tpb]
+    \centering
+    \subfloat[Estimated minor allele fraction for called variants in PAXIP1 gene.  $ \mu^{\text{control}}$.]{{\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
+    %
+    \subfloat[Depth Chart for Positions called by RVD2 and muTect]{{\includegraphics[width=0.2\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
+    \caption{2 Figures side by side}%
+    \label{fig:example}%
+\end{figure*}
 

### 2014-03-14 doc/bioinformatics2014/supplementary_info.tex
Update the manuscript and supplementary information

@@
 \usepackage{amsmath}
+\usepackage{pdfpages}
 %\usepackage{epstopdf}
@@
 \begin{center}
+\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
+\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\label{fig:hist_pos}
+\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
 \includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
@@
 
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
-\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
-\label{fig:hist_pos}
-\end{center}
-\end{figure}
-
 \section{Comparison between RVD2, VarScan2 somatic and muTect in Clincal HCC1187 dataset}
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_RVD2.pdf}
-\caption{RVD2.}
-\label{fig:rd_RVD2}
-\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_mutect.pdf}
-\caption{muTect.}
-\label{fig:rd_mutect}
-\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_true_positives.pdf}
-\caption{HCC1187 varscan2 true positives.}
-\label{fig:rd_varscan2_TP}
-\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_false_negatives.pdf}
-\caption{ HCC1187 varscan2 false negatives.}
-\label{fig:rd_varscan2_FN}
-\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_varscan2_false_positives.pdf}
-\caption{HCC1187 varscan2 false positives}
-\label{fig:rd_varscan2_FP}
-\end{center}
-\end{figure}
+\includepdf[pages={1-},scale=0.9]{pdf_figs/position_lookup_chart_all.pdf}
 

### 2014-03-17 doc/bioinformatics2014/document.tex
Update manuscript. Addressed all the comments from ISMB.

--- 2014-03-13 doc/bioinformatics2014/document.tex
@@
 \usepackage{bbm}
-
+\usepackage{subfig}
+%\usepackage[font={small}]{caption}
+\usepackage[skip=2pt,font=scriptsize]{caption}
 %\usepackage{placeins}
@@
 
-\section{Motivation:}
-Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
+\section{Motivation:}Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
 Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
@@
 
-\section{Results:}
-We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
+%TODO Update the chinical dataset result
+\section{Results:}We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
@@
 \section{Introduction}
-
+\label{sec:intro}
 Next-generation sequencing (NGS) technology has enabled the systematic interrogation of the genome for a fraction of the cost of traditional assays~\citep{Koboldt:2013kw}.
@@
 Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
+Some approaches attempt to estimate the tumor sample purity then adjust the calling algorithm with that estimation.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP.
+However, RVD2 and some other algorithms are able to call variations in heterogeneous samples without estimating the tumor purity.
 VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
@@
 Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.
-
+%TODO: Fixed this part
 %%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}[h]
+\begin{figure}[!bpth]
 \begin{center}
 \includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
-\caption{RVD2 Graphical Model}
+\caption{RVD2 Graphical Model.}
 \label{fig:graphical_model}
@@
 Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-We use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
-
+We use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
+%TODO: Bonferroni correction
 %%%%%%%%%%%%%%%%%%
@@
 
+%TODO: change the result
 We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we used the control data in the following hypothesis test $\Pr( \tilde{\mu}_j^{\text{control}} \in [\tau_\text{l},\tau_\text{u}] ) > 1-\alpha$, where the intervals are: homozygous reference $[0, 0.2]$, heterozygous mutant $[0.1, 0.9]$, and homozygous mutant $[0.8, 1.0]$ and the size of the test is $\alpha=0.05$.
+To identify germline mutations, we computed the empirical posterior mean $ \bar{\mu}_j^{control} $ in the control data and test which interval it falls in, where the intervals are: homozygous reference $[0, 0.05)$, heterozygous mutant $[0.05, 0.95)$, and homozygous mutant $[0.95, 1.0]$.
+We set the threshold between homozygous reference and heterozygous mutant at a relatively high 0.05 considering the low average read depth (40x for the normal sample and 90x for the tumor sample), likewise for the border between heterozygous and homozygous mutant at 0.95.
+We reported the 95\% credible interval to shown the dominant region for $ \hat{\mu}_j^{control} $ in called positions afterwards.
 
@@
 
+$ \chi^2 $ test was performed with p-value set at 0.05 after posterior distribution test.
+In the synthetic data we used Bejamini-Hochberg method to control the family-wise error rate over positions called by posterior distribution test.
+However, we didn't apply Benjamini-Hochberg method to clinical HCC1187 sequence data as Benjamini-Hochberg method is well-recognized very conservative, and it restraints RVD2 from detecting low rate mutated positions when read depth is as low as 90x.
+
+%TODO: No bonferoni correction for the chi2 test
+
 \subsection{Performance with read depth}\label{sec:read_depth}
@@
 \includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance.}
+\caption{ROC curve varying read depth showing detection performance.
+Each subfigure shows ROC curves across four different read depths for one MAF level.
+Within one subfigure it can be seen that the performance improves monotonically with read depth.
+Across different subfigures, it can be seen that the performance improves with MAF level.}
 \label{fig:ROC}
@@
 We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
+Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2 Somatic, Strelka and muTect are designed to call variants in heterogeneous samples, which serve as better comparison to RVD2.
 In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
 The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
-GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
+GATK, Varscan2, Strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
 
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
+% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
+
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
+%\afterpage{\clearpage}
 \subsubsection*{Sensitivity/Specificity Comparison}
-
-\begin{figure*}[!tpb]
+\begin{figure*}[!htpb]
 \begin{center}
 \includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
-\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms.}
+\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms using synthetic sequence data.}
 \label{tbl:comparison_ss}
@@
 
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-% rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
-
+\begin{figure*}[!tbhp]
+\begin{center}
+\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
+\caption{False discovery rate comparison of RVD2 with other variant calling algorithms using synthetic sequence data. Blank cells indicate no locations were called variant.}
+\label{tbl:comparison_fdr}
+\end{center}
+\end{figure*}
 Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
@@
 
+The statistics for RVD is an average from six replicates as RVD uses six control replicates to estimate the model but returns six sets of statistics separately for six case replicates.
+
+RVD uses multiple control replicates to estimate the model but only one pair-end case sample (two replicates) to do hypothesis testing.
+Given a data set of three pair-end replicates for control sample and three pair-end replicates for case sample, three set of mutation calling results are returned.
+The statistics for RVD shown in Figure~\ref{tbl:comparison_ss} and Figure~\ref{tbl:comparison_fdr} is an average from three set of statistics.
+RVD performed the best among all algorithms when the read depth is as high as 40000x.
+RVD correctly called all the mutated positions across all MAF levels with no false positives when MAF level is 0.3\% or lower.
+RVD tends to be a little bit over sensitive when the MAF level is 1.0\% or higher, which causes specificity slightly lower than 1.00.
+However, RVD program fails when the depth is at 5000x or lower and is not able to call any mutations.
+
 The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
@@
 When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
+\begin{figure*}[!bpth]
+    \centering
+    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.65\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf} }}%
+    %
+    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.3\textwidth]{pdf_figs/RVD2calltable.pdf} }}%
+    \caption{Germline and Somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+HG19 Genomic locations are chr7:154,700,000+X.
+(a) Estimated minor allele fraction for called variants in PAXIP1 gene.
+Star (*) indicates $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$.
+The vertical lines represent 95\% credible interval around mean MAF.
+(b) Mutation type classified by RVD2.
+RVD2 identified twelve variants in the 44kbp PAXIP1 gene.
+There were eight germline variants and eight somatic variants.
+Especially, three heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
+The dbSNPv138 column shows the identities for positions found to be common populations SNPs according to dbSNPv138.}%
+    \label{fig:brca_MAF}%
+\end{figure*}
+\subsubsection*{False Discovery Rate Comparison}
 
-\subsubsection*{False Discovery Rate Comparison}
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, strelka and muTect. Blank cells indicate no positive calls were made.
+Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, Strelka and muTect. Blank cells indicate no positive calls were made.
 
@@
 VarScan2-pileup performs perfectly for all but the lowest depth for the 100\% MAF.
-
-\begin{figure*}[!tpb]
-\begin{center}
-\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
-\caption{False discovery rate comparison of RVD2 with other variant calling algorithms. Blank cells indicate no locations were called variant.}
-\label{tbl:comparison_fdr}
-\end{center}
-\end{figure*}
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-%\afterpage{\clearpage}
 
@@
 However, it does not call any variants at or below 1\% MAF.
-MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
+muTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths.
 But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.
+RVD has best FDR performance in the high read depth for 0.1\% -- 1\% MAF levels.
+The FDR increases to around 0.3 for 10\% -- 100\% MAF in the high read depth.
 
@@
 \subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
+\subsubsection{Performance of RVD2}
 
-\begin{figure*}[!tpb]
-\begin{center}
-\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for called variants in PAXIP1 gene. Star (*) indicates $\mu^{\text{case}}$  is significantly different from $ \mu^{\text{control}}$.}
-\label{fig:brca_MAF}
-\end{center}
+\begin{figure*}[!pthb]
+    \centering
+    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
+    %
+    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
+    \caption{(a)Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+VarScan2-somatic reported 60 positions, muTect 11 positions, RVD2 12 positions and Strelka only 1 position.
+This figure uses position index to show the correspondence of positions called by different algorithms for notation simplicity.
+Complete actual positions and depth distribution are provided in Supplementary Table 1 for validation.
+(b) Depth distribution for positions called by RVD2 and muTect.}%
+    \label{fig:brca_dot}%
 \end{figure*}
-%('rvd2\results\2013-11-21_HCC1187_PAXIP1_mu_barplot_T0')
-%TODO: add star to test that have mudelta > .16 Done.
-%TODO: set tau=0.16 for mudelta test.  Done
 
-RVD2 identified six variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Of the six called variants, three were homozygous germline variants and two were heterozygous germline variants.
-The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.
-The remaining one is a homozygous somatic variant.
-Because the median read depth for the case sample is 60, we set the threshold of the test for $\mu^\Delta$ to $\tau = 16\%$ to minimize the false discovery rate (Figure~\ref{tbl:comparison_fdr}).
+RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+There were eight germline variants and eight somatic mutations in the twelve variants.
+Especially, three heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
 
-Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C are called significantly different in tumor and normal sample MAF.
-Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations.
-Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample.
-In the tumor sample we identified six mutations chr7:154743899C$>$T, chr7:154753635T$>$C, \\ chr7:154754371T$>$C, chr7:154758813G$>$A, chr7:154760439A$>$C and chr7:154780960C$>$T.
-Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events.
+Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
+Positions chr7:154743899 C$>$T, chr7:154753635 T$>$C and chr7:154780960 C$>$T were called germline homozygous mutations.
+Positions chr7:154749704, chr7:154754371 and chr7:154758813, chr7:154766700, chr7:154781769 were called heterozygous in the normal sample.
+Positions chr7:154749704, chr7:154753635 T$>$C, chr7:154754371 T$>$C, chr7:154758813 G$>$A, chr7:154760439 A$>$C, chr7:154766732,  chr7:154766832, chr7:154777118 were called significantly different in tumor and normal sample MAF.
+Positions chr7:154749704, chr7:154754371 T$>$C and chr7:154758813 G$>$A were called loss-of-heterozygosity events.
 Some of these mutations are also found to be common population SNPs according to dbSNPv138.
-The corresponding identities are shown in the figure.
+The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}.
+The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1 for specificity validation.
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
 
-The original research describing this sample used Strelka to identify mutations in the same sample.
-They identified chr7:154760439 as variant, but did not call the other five variants.
-In particular strelka missed the two LOH events.
-% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample.
-Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
+\subsubsection{Performance comparison with other algorithms}
+
+For further detection validation and performance comparison, we ran muTect and VarScan2-somatic to call mutations in the PAXIP1 gene in HCC1187 sample.
+We also compared to the result shown in original research report where Strelka was used to identify mutations in the same sample ~\citep{hcc1187techreport}.
+Figure~\ref{fig:brca_dot:a} shows mutation detection result from Strelka, RVD2, muTect, and VarScan2-somatic, the state-of-art algorithms able to call mutation from heterogeneous samples.
+For notation simplicity, we use position index to present actual positions in Figure~\ref{fig:brca_dot}, while the correspondence is provided in Supplementary Table 1.
+
+The mutations called by RVD2 and muTect are the most consistent among all the techniques.
+RVD2 detected twelve germline and somatic mutations, while muTect reported eleven, ten in common.
+In the disagreements, RVD2 did not call position 50 while muTect did not call position 37 and 58.
+Referring to the depth distribution shown in Figure~\ref{fig:brca_dot:b}, it can be seen that position 37 and 58 are more likely mutated while position 50 is less likely mutated.
+
+Strelka was the least sensitive algorithms among all the algorithms.
+According to the technical report, Strelka identified position 22 (chr7:154760439) as variant, but did not call any other variants.
+In particular Strelka missed the three LOH events called by RVD2.
+On the contrary, VarScan2-somatic was the most sensitive algorithms, and called sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
+VarScan2-somatic  detected all the positions called by RVD2 except position 35, which turns out to be a very likely mutation given the depth distribution in Figure~\ref{fig:brca_dot:b}.
+This fact reveals the imperfect though high sensitivity for VarScan2-somatic.
+On the other side, VarScan2-somatic reported fifty positions which were not called by any other three algorithms.
+The read depth in Supplementary Table 1 suggestions that these positions are very likely to be false positives.
+The fact that VarScan2-somatic can be over-sensitive has appeared in synthetic dataset analysis.
+As shown in Figure~\ref{tbl:comparison_fdr}, the False Discovery Rate for VarScan2-somatic at read depth 53 MAF level 1.0\% is as high as 1.00.
+\citet{spencer2013performance} also mentioned that VarScan2 has tendency to be over-sensitive at high read depth.
+% To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
 
@@
 We opted for this balance of speed and accuracy because computational time is often not limiting and the cost of a false positive or false negative greatly outweighs the cost of more computation.
+Another factor that can affect the speed of RVD2 is the number of Metropolis-Hasting sample within one Gibbs sampling run.
+However, RVD2 is able to use N cores in parallel ( no more than the number of Metropolis-Hasting samples), which can significantly decrease the running time to $ \dfrac{1}{N} $ of the running time in sequence.
+Finally, the running time for RVD2 is nearly linearly proportional to sequence length of the sample.
+Taking our 400bp long synthetic data as an example, it takes around 1.5 seconds when we run the program in sequence with 10 Metropolis-Hasting samples in one Gibbs sample.
+Therefore, if we take 4000 Gibbs samples, it takes around $ \dfrac{1.5\times4000}{60} = 100 $mins to run the program.
+If we use 10 cores for parallel processing, it will take around $ \dfrac{100}{10}=10 $mins to run the program.
+Considering the running time for RVD2 can be much longer for long sequences, we are currently developing a variational method to estimate RVD2 model.
 

### 2014-03-17 doc/bioinformatics2014/supplementary_info.tex
Update manuscript. Addressed all the comments from ISMB.

@@
 \usepackage{pdfpages}
+\usepackage[skip=2pt,font=scriptsize]{caption}
 %\usepackage{epstopdf}
@@
 \title[RVD2: Supplementary Information]{Supplementary Information for
-  RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data}
+  RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
 %\date{}      % Activate to display a given date or no date
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/depthM.pdf}
+\includegraphics[width=0.85\textwidth]{pdf_figs/depthM.pdf}
 \caption{Key parameters for RVD2 model for synthetic DNA data sets.}
@@
 
-\section{Histogram of $ \mu_j $ for mutations detected by RVD2 in HCC1187 dataset}
+\section{histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
 \begin{figure}[H]
@@
 \includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
-\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
 \label{fig:hist_pos}
@@
 \includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
-\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
+\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
 \label{fig:hist_neg}
@@
 
-\section{Comparison between RVD2, VarScan2 somatic and muTect in Clincal HCC1187 dataset}
-\includepdf[pages={1-},scale=0.9]{pdf_figs/position_lookup_chart_all.pdf}
+\bigskip
+\bigskip
+
+\includepdf[pages={1-},scale=1]{pdf_figs/position_lookup_chart_all.pdf}
 

### 2014-03-17 results/2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset/mutation_dotplot.py
Minor updates on codes

@@ -96,10 +96,10 @@ def main():
 	ax1.text(xval[3]+2,ypos[3]-0.05,'%d' %xval[3])
 	ax1.barh(ypos[4],xval[4], color='g', align='center',height=0.1)
 	ax1.text(xval[4]+2,ypos[4]-0.05,'%d' %xval[4])
-	ax1.set_xlim([0,75])
+	ax1.set_xlim([0,65])
 	# make these tick labels invisible
 	plt.setp( ax1.get_yticklabels(), visible=False)
-	ax1.set_xlabel('Frequency')
+	ax1.set_xlabel('Total')
 
 	plt.tight_layout()
 	# plt.show()

### 2014-03-17 results/2014-03-13_gemline_and_somatic_test_correction/MuBarPlot.py

@@ -37,7 +37,7 @@ def main():
         diff_tau = 0.0
         N = 1000
 
-        roi = [0.05, 0.75]
+        roi = [0.05, 0.95]
 
         hdf5name = 'germlineccall.hdf5'
         try:
@@ -72,7 +72,7 @@ def main():
     
     call = [somatic_call,germline_call]
     call = np.sum(call,0) > 0
-    MuBarPlot(controlMu[call], caseMu[call], loc[call],'HCC1187_mu.pdf')
+    MuBarPlot(controlMu[call], caseMu[call], loc[call],'HCC1187_MuBarPlot.pdf')
     
 
 def MuBarPlot(CallControlMu,CallCaseMu,Loc,title):
@@ -104,19 +104,26 @@ def MuBarPlot(CallControlMu,CallCaseMu,Loc,title):
         for i in diffcall:
             ax.text(i+width-width/9.0, 1.0021,'*',color='r',fontsize=15)
 
+        rsSNP = [0, 2, 3, 4, 10]
+
+        # for i in rsSNP:
+        #     ax.text(i+width-width/5.0, 1.05,r'$\diamond$',color = 'b', fontsize=13)
+
+
         Loclabel = [x.split(':')[1][4:] for x in Loc]
-        rslabel = ['rs1239326', ' ' ,'rs1239324', '\nrs71534174' ,'rs35505514', ' ' , ' ' , ' ' , ' ' , ' ' ,'rs4398858', ' ']
+
+        # rslabel = ['rs1239326', ' ' ,'rs1239324', '\nrs71534174' ,'rs35505514', ' ' , ' ' , ' ' , ' ' , ' ' ,'rs4398858', ' ']
  
 
-        label = ['%s\n%s' %(Loclabel[i], rslabel[i]) for i in xrange(len(Loclabel))]
+        # label = ['%s\n%s' %(Loclabel[i], rslabel[i]) for i in xrange(len(Loclabel))]
 
-        ax.set_xticklabels(label)
+        ax.set_xticklabels(Loclabel)
         
 
         lgd = ax.legend( (rects1[0], rects2[0]), ('Control', 'Case'), bbox_to_anchor=(0.85, 1.02, 0.15, .102),
-                   loc=4,ncol=1, mode="expand", borderaxespad=0., prop={'size':9})
+                   loc=4,ncol=1, mode="expand", borderaxespad=0., prop={'size':11})
 
[…]

### 2014-03-17 src/python/rvd27/rvd27.py

@@ -65,7 +65,7 @@ def main():
     argpTest.add_argument('caseHDF5Name', default=None,
                 help='case model file (HDF5)')
     
-    argpTest.add_argument('somatic_tau', default=(0.05,0.75),
+    argpTest.add_argument('somatic_tau', default=(0.05,0.95),
                 help='Two thresholds for somatic test. \
                 Mu lower than the lower threshold will be classified as reference (non-mutation), \
                 Mu between the two thresholds will be classified as heterozygote, \
@@ -122,7 +122,7 @@ def test_main(args):
          args.N, args.outputFile)
 
 def test(alpha=0.95, controlHDF5Name=None, caseHDF5Name=None,
-         somatic_tau=(0.05,0.75),diffroi=(0,np.inf),
+         somatic_tau=(0.05,0.95),diffroi=(0,np.inf),
          N=1000, outputFile=None):
 
     if outputfile == None:
@@ -132,7 +132,7 @@ def test(alpha=0.95, controlHDF5Name=None, caseHDF5Name=None,
         diff_test(alpha, controlHDF5Name, caseHDF5Name, diffroi, N, outputFile)
         somatic_test(controlHDF5Name, caseHDF5Name,somatic_tau, outputFile)
 
-def germline_test(controlHDF5Name, caseHDF5Name, germline_roi = [0.05, 0.75], outputFile='germlinecalltable.vcf'):
+def germline_test(controlHDF5Name, caseHDF5Name, germline_roi = [0.05, 0.95], outputFile='germlinecalltable.vcf'):
     # only test if control sample is mutated
     
     loc, refb, altb, controlMu, controlMu0, controlR, controlN, \
@@ -193,7 +193,7 @@ def germline_test(controlHDF5Name, caseHDF5Name, germline_roi = [0.05, 0.75], ou
 
     return loc, call, controlMu, caseMu, controlN, caseN
 
-def somatic_test(alpha, controlHDF5Name, caseHDF5Name, diff_tau= 0, somatic_roi = [0.05, 0.75], N=1000,  outputFile='somaticcalltable.vcf'):
+def somatic_test(alpha, controlHDF5Name, caseHDF5Name, diff_tau= 0, somatic_roi = [0.05, 0.95], N=1000,  outputFile='somaticcalltable.vcf'):
     # Two sided difference test. Somatic status will be more specific classified. Threshold hold and alpha should be assign
[…]

### 2014-03-18 doc/bioinformatics2014/supplementary_info.tex
Finished manuscript, supplementary information and review reply.

@@
 \usepackage{pdfpages}
+\usepackage{subfig}
+\usepackage{enumitem}
 \usepackage[skip=2pt,font=scriptsize]{caption}
+\newcommand{\RR}{I\!\!R} %real numbers
 %\usepackage{epstopdf}
@@
 
-\newcommand{\RR}{I\!\!R} %real numbers
 \DeclareMathOperator{\diag}{diag}
@@
 \maketitle
+
+%%%%%%%%%%%%%%%%%%%
+% Appendix A: An simulation example to illustate the idea of proposal graphical model and v.
+%%%%%%%%%%%%%%%%%%%
+\section{An simulation example to illustrate the proposal graphical model and generative process.}\label{sec:appendix_example}
+
+\begin{figure}[!bpth]
+    \centering
+    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/RVD2_model.pdf} }}%
+    \quad
+    \quad
+    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/GMexample.pdf} }}%
+    \caption{(A) RVD2 Graphical Model.(B) An simulation example to illustrate the proposal graphical model and generative process.
+A generative process is to generate randomly observable data following your generative model, which is the reverse process of doing inference from the actual data.
+Here we want to generate a simulation sample with 3 replicates, and each replicate is a 5bp long sequence.
+Therefore, we have $ J = 5, N=3.
+$ The numbers in orange are known, including $\mu_0 \in \RR^{1 \times 1}$, $M_0 \in \RR^{1 \times 1}$, $M \in \RR^{5 \times 1}$, $n \in \RR^{5 \times 3}$.
+(1) We sample local error rate $\mu \in \RR^{5 \times 1}$ by drawing random numbers $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$; (2) and then use $\mu $ and $ M $ to sample $\theta \in \RR^{5 \times 3}$ by drawing random samples $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$; (3) finally we use $\theta \in \RR^{5 \times 3}$ and $n \in \RR^{5 \times 3}$ to sample $r \in \RR^{5 \times 3}$ by drawing random numbers $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.
+In actual practice, non-reference read counts $ r $ and coverage $ n $ are the data we use to do inference.}
+    \label{fig:brca_MAF}%
+\end{figure}
 
@@
 %%%%%%%%%%%%%%%%%
+\newpage
 \section{Parameter settings for other variant calling algorithms}

### 2014-03-24 doc/bioinformatics2014/document.tex
Finished manuscript, supplementary information and review reply.
Update the manuscript

@@
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We identify six mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+We identify twelve mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
 
@@
 
-Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
-Some approaches attempt to estimate the tumor sample purity then adjust the calling algorithm with that estimation.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP.
-However, RVD2 and some other algorithms are able to call variations in heterogeneous samples without estimating the tumor purity.
+Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP.
 VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
@@
 However, \citet{spencer2013performance} also highlighted the fact that VarScan2 yielded more false positives at high read depth.
-\citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
+\citet{Stead:2013fu} showed that VarScan-somatic outperformed Strelka and had performance on-par with muTect in detecting a 5\% MAF for read depths between 100 and 1000.
 %We test their performance on much lower MAF variants and across a wider range of coverage depths.
@@
 For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-$ J $ is the length of gene sequence and $ N $ is the number of technical replicate gene sequence in the sample.
-\label{replicate definition} Technical replicates are used to establish experimental variability in Next Generation Sequencing procedure~\citep{quackenbush2002microarray,robasky2013role}, even though multiple replicates are not indispensable for RVD2.
+$ J $ is the region of interest length and $ N $ is the number of technical replicates in the sample.\label{replicate definition} Technical replicates are used to establish experimental variability in next-generation sequencing procedure~\citep{quackenbush2002microarray,robasky2013role}, though multiple replicates are not necessary for RVD2.
 
-The model generative process is as follows:
+The model generative process given hyperparameters $ \mu_0, M_0$ and $ M $is as follows:
 
@@
 			\item Draw $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$
-			\item Draw $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
+			\item Draw $r_{ji}|n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$
 		\end{enumerate}
@@
 RVD2 hierarchically partitions sources of variation in the data.
-$r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
-$\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility.
+The distribution $r_{ji} | n_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$ models the variation due to sampling the pool of DNA molecules on the sequencer.
+The distribution $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$ models the variation due to experimental reproducibility.
 The variation in error rate due to sequence context is modeled by $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
@@
 \label{Improvement of RVD2}
-The improvement of RVD2 upon RVD firstly lies in the $ \text{Beta}(\mu_0, M_0) $ prior on local error rate $ \mu_j $, which accurately captured the global across-position error rate distribution.
-The prior distribution also  allows to $ \mu_j $ borrow information from adjacent positions and prevent RVD2 from failure when the read depth is too low to bring out none-zero $ r_{ji} $.
-Besides the improvement on model structure, RVD2 developed a novel Bayesian Hypothesis Testing algorithm comparing to a frequentist normal z-test in RVD.
-The performance comparison between RVD and RVD2 will be elaborated in Section~\ref{sec:comparison}.
+RVD2 improves on RVD in three ways.
+First, RVD2 has a $ \text{Beta}(\mu_0, M_0) $ prior on local error rate $ \mu_j $, which captures the global across-position error rate.
+The prior distribution allows $ \mu_j $ to borrow information from adjacent positions and allows RVD2 to handle low read depths.
+second, this method smoothly handles multiple replicates in case samples.
+Third, RVD2 has a more accurate Bayesian hypothesis testing method compared to a frequentist normal z-test in RVD.
+We show a performance comparison between RVD and RVD2 in Section~\ref{sec:comparison}.
 
@@
 \subsection{Posterior Distribution Test}\label{sec:hypothesis_test}
+\subsubsection*{Posterior Somatic Test.}
 Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
@@
 
-The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit.
-The optimal $\tau$ maximizes the Matthews Correlation Coefficient (MCC) to perfect classification in the ROC curve plot,
+The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ maximizes the Matthews Correlation Coefficient (MCC),
 \begin{equation}
@@
 Absent this information we set $\tau = 0$.
+
+\subsubsection*{Posterior Germline Test.}
 
@@
 Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-We use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
-%TODO: Bonferroni correction
+If the sample average is higher than 200, we use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
+%TODO: Bonferroni correction, set an approapriate threshold for read depth.
 %%%%%%%%%%%%%%%%%%
@@
 \subsection{Synthetic DNA Sequence Data}
-
+\subsubsection*{Experimental process.}
 Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
@@
 More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+\subsubsection*{Computational process.}
  We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
-
 To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.
@@
 \subsection{HCC1187 Sequence Data}
-
+\subsubsection*{Experimental process.}
 The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
@@
 Matched normal cells were derived from lymphoblastoid cells from peripheral blood.
+\subsubsection*{Computational process.}
 Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
@@
 We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
-For these ROC curves, we used the posterior distribution test without the $\chi^2$ test.
-Adding the $\chi^2$ improves specificity at the expense of sensitivity.
+For these ROC curves, we used the posterior density test without the $\chi^2$ test to evaluate the performance of posterior density test individually.
 Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$.
@@
 
-The statistics for RVD is an average from six replicates as RVD uses six control replicates to estimate the model but returns six sets of statistics separately for six case replicates.
-
-RVD uses multiple control replicates to estimate the model but only one pair-end case sample (two replicates) to do hypothesis testing.
-Given a data set of three pair-end replicates for control sample and three pair-end replicates for case sample, three set of mutation calling results are returned.
-The statistics for RVD shown in Figure~\ref{tbl:comparison_ss} and Figure~\ref{tbl:comparison_fdr} is an average from three set of statistics.
+The statistics for RVD is an average statistics as RVD uses six control replicates to estimate the model but returns statistics separately for three set of pair-end case replicates.
 RVD performed the best among all algorithms when the read depth is as high as 40000x.
@@
 RVD tends to be a little bit over sensitive when the MAF level is 1.0\% or higher, which causes specificity slightly lower than 1.00.
-However, RVD program fails when the depth is at 5000x or lower and is not able to call any mutations.
+However, RVD program fails when the depth is unable to measure the baseline error rate and is not able to call any mutations.
 
@@
 \subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
-\subsubsection{Performance of RVD2}
+\subsubsection{Performance of RVD2.}
 
@@
 
-\subsubsection{Performance comparison with other algorithms}
+\subsubsection{Performance comparison with other algorithms.}
 
-For further detection validation and performance comparison, we ran muTect and VarScan2-somatic to call mutations in the PAXIP1 gene in HCC1187 sample.
+We ran muTect and VarScan2-somatic to call mutations in the PAXIP1 gene in HCC1187 sample.
 We also compared to the result shown in original research report where Strelka was used to identify mutations in the same sample ~\citep{hcc1187techreport}.
@@
 In particular Strelka missed the three LOH events called by RVD2.
-On the contrary, VarScan2-somatic was the most sensitive algorithms, and called sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
+On the contrary, VarScan2-somatic called most positions among all algorithms, sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
 VarScan2-somatic  detected all the positions called by RVD2 except position 35, which turns out to be a very likely mutation given the depth distribution in Figure~\ref{fig:brca_dot:b}.
-This fact reveals the imperfect though high sensitivity for VarScan2-somatic.
 On the other side, VarScan2-somatic reported fifty positions which were not called by any other three algorithms.
@@
 Another factor that can affect the speed of RVD2 is the number of Metropolis-Hasting sample within one Gibbs sampling run.
-However, RVD2 is able to use N cores in parallel ( no more than the number of Metropolis-Hasting samples), which can significantly decrease the running time to $ \dfrac{1}{N} $ of the running time in sequence.
-Finally, the running time for RVD2 is nearly linearly proportional to sequence length of the sample.
-Taking our 400bp long synthetic data as an example, it takes around 1.5 seconds when we run the program in sequence with 10 Metropolis-Hasting samples in one Gibbs sample.
-Therefore, if we take 4000 Gibbs samples, it takes around $ \dfrac{1.5\times4000}{60} = 100 $mins to run the program.
-If we use 10 cores for parallel processing, it will take around $ \dfrac{100}{10}=10 $mins to run the program.
-Considering the running time for RVD2 can be much longer for long sequences, we are currently developing a variational method to estimate RVD2 model.
+However, RVD2 is able to use multiple cores in parallel ( no more than the number of Metropolis-Hasting samples), which can significantly improve time efficiency.
+on a computational mode with 64 2.34GHz processor and 256Gb of RAM, RVd2 takes around 10 minutes to analyze the 400bp long synthetic data for one dilution rate and one downsampling rate.
+The memory requirement is no higher than the size of the gene sequence times the number of Gibbs samples.
+We are currently developing a variational method to estimate RVD2, aiming at decreasing time complexity.
 

### 2014-03-27 doc/bioinformatics2014/document.tex
Updated manuscript and figures.

@@
 \usepackage{algorithm}
+\usepackage{float}
 \usepackage{bbm}
@@
 
-%TODO Update the chinical dataset result
 \section{Results:}We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples.
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We identify twelve mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+We apply our model and identify twelve mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
 
@@
 \subsection{Sampling from $p \left( \theta_{ij} |r_{ij},n_{ij},\mu_j,M \right)$}
-%TODO: double check if the posterior is correct
+
 Samples from the posterior distribution
@@
 The posterior distribution is
-\begin{equation}
-	p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j) \thicksim \text{Beta}\left( \frac{r_{ji} + M_j \mu_j}{n_{ji} + M_j} , n_{ji} + M_j\right).
-\end{equation}
+\begin{flalign}
+& p(\theta_{ji} | r_{ji}, n_{ji}, \mu_j, M_j)& \nonumber \\
+& \thicksim\text {Beta}\left( r_{ji} + M_j \mu_j , n_{ji}- r_{ji} + M_j (1-\mu_j)\right).&
+\end{flalign}
 
@@
 \subsection{Posterior Distribution Test}\label{sec:hypothesis_test}
-\subsubsection*{Posterior Somatic Test.}
+\subsubsection*{Posterior Difference Test.}
 Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data.
@@
 \begin{equation}\label{eqn:bayes_test}
-	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{case}} - \tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha,
+	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}}  > \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{case}} - \tilde{\mu}_{jk}^{\text{control}} > \tau } > 1-\alpha,
 \end{equation}
@@
 
+\subsubsection*{Posterior Somatic Test.}
+
+Posterior Somatic Test is a two-sided posterior difference test using paired control-case sample.
+To identify somatic mutations, we consider scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
+loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
+homozygous somatic mutation).
+The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < \tau ) > 1-\alpha$.
+Threshold $ \tau $ is set at zero.
+
 \subsubsection*{Posterior Germline Test.}
+
+Posterior Germline Test is a one-sided posterior distribution test using a single control sample. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,
+
+\begin{equation}\label{eqn:germ_test}
+	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha,
+\end{equation}
 
@@
 Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true.
-If the sample average is higher than 200, we use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
-%TODO: Bonferroni correction, set an approapriate threshold for read depth.
+If the sample average depth is higher than 500, we use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
+The average depth threshold is set because Benjamini-Hochberg method is a highly conservative method and will remove many true calls when the read depth is not high enough.
+
 %%%%%%%%%%%%%%%%%%
@@
 The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
-We drew $1,000$ samples from $\mu^{\Delta}$ to estimate the posterior probability of a variant.
+We drew $1,000$ samples from $ \tilde{\mu}^{\Delta} =  \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ to estimate the posterior probability of a variant.
 
-%TODO: change the result
-We used RVD2 to identify germline and somatic mutations in the diploid HCC1187 sample.
-To identify germline mutations, we computed the empirical posterior mean $ \bar{\mu}_j^{control} $ in the control data and test which interval it falls in, where the intervals are: homozygous reference $[0, 0.05)$, heterozygous mutant $[0.05, 0.95)$, and homozygous mutant $[0.95, 1.0]$.
-We set the threshold between homozygous reference and heterozygous mutant at a relatively high 0.05 considering the low average read depth (40x for the normal sample and 90x for the tumor sample), likewise for the border between heterozygous and homozygous mutant at 0.95.
-We reported the 95\% credible interval to shown the dominant region for $ \hat{\mu}_j^{control} $ in called positions afterwards.
-
-To identify somatic mutations, we considered scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
-loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
-homozygous somatic mutation).
-The two hypothesis tests are then $\Pr( \mu_j^{\Delta} \geq \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\Delta} \leq \tau ) > 1-\alpha$.
-The size of the test is $\alpha=0.05$.
-
-$ \chi^2 $ test was performed with p-value set at 0.05 after posterior distribution test.
-In the synthetic data we used Bejamini-Hochberg method to control the family-wise error rate over positions called by posterior distribution test.
-However, we didn't apply Benjamini-Hochberg method to clinical HCC1187 sequence data as Benjamini-Hochberg method is well-recognized very conservative, and it restraints RVD2 from detecting low rate mutated positions when read depth is as low as 90x.
-
-%TODO: No bonferoni correction for the chi2 test
+We performed posterior disparity test on synthetic data to identify mutations given it is a haploid DNA sequence.
+We set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
+We used RVD2 to identify somatic and germline mutations in the diploid HCC1187 sample.
+In the posterior somatic test, we set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
+In the posterior germline test, We set  the threshold $ \tau =0.05 $ considering the low average coverage (40x).
+The size of the test is set at $ \alpha = 0.15$, higher than the size of somatic test.
+We are less confident on the germline test because we only use control sample in the germline test compared to normal-tumor paired sample in somatic test.
+We performed $ \chi^2 $ non-uniformity test along all the posterior density test.
 
@@
 Each subfigure shows ROC curves across four different read depths for one MAF level.
-Within one subfigure it can be seen that the performance improves monotonically with read depth.
-Across different subfigures, it can be seen that the performance improves with MAF level.}
+Within one subfigure，the performance improves monotonically with read depth.
+Across different subfigures, the performance improves with MAF level.}
 \label{fig:ROC}
@@
 Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
-They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples with much success.
+They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples.
 GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
@@
 
-The statistics for RVD is an average statistics as RVD uses six control replicates to estimate the model but returns statistics separately for three set of pair-end case replicates.
+The statistics for RVD is an average statistics as RVD uses six control replicates to estimate the model but returns statistics separately for three sets of pair-end case replicates.
 RVD performed the best among all algorithms when the read depth is as high as 40000x.
-RVD correctly called all the mutated positions across all MAF levels with no false positives when MAF level is 0.3\% or lower.
-RVD tends to be a little bit over sensitive when the MAF level is 1.0\% or higher, which causes specificity slightly lower than 1.00.
-However, RVD program fails when the depth is unable to measure the baseline error rate and is not able to call any mutations.
+RVD called all the mutated positions across all MAF levels with no false positives when MAF level is 0.3\% or lower.
+RVD calls some false positives when the MAF level is 1.0\% or higher, which causes specificity slightly lower than 1.00.
+RVD program fails and can not call any mutations when the depth is unable to measure the baseline error rate.
 
@@
 When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-\begin{figure*}[!bpth]
-    \centering
-    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.65\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf} }}%
-    %
-    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.3\textwidth]{pdf_figs/RVD2calltable.pdf} }}%
-    \caption{Germline and Somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-HG19 Genomic locations are chr7:154,700,000+X.
-(a) Estimated minor allele fraction for called variants in PAXIP1 gene.
-Star (*) indicates $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$.
-The vertical lines represent 95\% credible interval around mean MAF.
-(b) Mutation type classified by RVD2.
-RVD2 identified twelve variants in the 44kbp PAXIP1 gene.
-There were eight germline variants and eight somatic variants.
-Especially, three heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
-The dbSNPv138 column shows the identities for positions found to be common populations SNPs according to dbSNPv138.}%
-    \label{fig:brca_MAF}%
+\begin{figure*}[!hbpt]
+\begin{center}
+\includegraphics[width=0.85\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\caption{Estimated minor allele fraction for Germline and Somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence.
+Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$.
+The  vertical lines represent 95\% credible interval around posterior mean MAF.
+Five positions are common populations SNPs according to dbSNPv138, and the identities are shown below the positions.}
+\label{fig:brca_MAF}
+\end{center}
 \end{figure*}
+
 \subsubsection*{False Discovery Rate Comparison}
@@
 The smaller variance yields improvements in sensitivity at the expense of more false positives.
-Since the FDR only considers positive calls, the performance degrades.
+Since the FDR only considers positive calls, the performance by that measure degrades.
 
@@
 There were eight germline variants and eight somatic mutations in the twelve variants.
-Especially, three heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
-
+%Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
+%TODO :check with pos 154749704G>G in somatic
 Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899 C$>$T, chr7:154753635 T$>$C and chr7:154780960 C$>$T were called germline homozygous mutations.
-Positions chr7:154749704, chr7:154754371 and chr7:154758813, chr7:154766700, chr7:154781769 were called heterozygous in the normal sample.
-Positions chr7:154749704, chr7:154753635 T$>$C, chr7:154754371 T$>$C, chr7:154758813 G$>$A, chr7:154760439 A$>$C, chr7:154766732,  chr7:154766832, chr7:154777118 were called significantly different in tumor and normal sample MAF.
-Positions chr7:154749704, chr7:154754371 T$>$C and chr7:154758813 G$>$A were called loss-of-heterozygosity events.
+Positions\footnote{The notation format is Chromosome:PositionReference\textgreater ControlMutation.} chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations.
+Positions\footnote{The notation format is Chromosome:PositionReference\textgreater TumorMutation.} chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
+Positions chr7:154754371 and chr7:154758813 were considered loss-of-heterozygosity events.
 Some of these mutations are also found to be common population SNPs according to dbSNPv138.
 The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}.
-The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1 for specificity validation.
+The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1 for validation.
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
@@
 According to the technical report, Strelka identified position 22 (chr7:154760439) as variant, but did not call any other variants.
-In particular Strelka missed the three LOH events called by RVD2.
+In particular Strelka missed the two LOH events called by RVD2.
 On the contrary, VarScan2-somatic called most positions among all algorithms, sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
@@
 As shown in Figure~\ref{tbl:comparison_fdr}, the False Discovery Rate for VarScan2-somatic at read depth 53 MAF level 1.0\% is as high as 1.00.
-\citet{spencer2013performance} also mentioned that VarScan2 has tendency to be over-sensitive at high read depth.
+\citet{spencer2013performance} also mentioned that VarScan2 has tendency to call many false positives at high read depth.
 % To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
@@
 However, RVD2 is able to use multiple cores in parallel ( no more than the number of Metropolis-Hasting samples), which can significantly improve time efficiency.
-on a computational mode with 64 2.34GHz processor and 256Gb of RAM, RVd2 takes around 10 minutes to analyze the 400bp long synthetic data for one dilution rate and one downsampling rate.
+on a computational mode with 64 2.34GHz processor and 256Gb of RAM, RVD2 takes around 10 minutes to analyze the 400bp long synthetic data for one dilution rate and one downsampling rate.
 The memory requirement is no higher than the size of the gene sequence times the number of Gibbs samples.

### 2014-03-30 doc/bioinformatics2014/document.tex
edits to manuscript and supplementary info.

@@
 Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
-However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples.
+Therefore, a variant calling algorithm for calling low-frequency polymorphisms in heterogeneous samples is needed.
 
@@
 
-Recently, researchers have developed algorithms to call low-frequency or rare variants in heterogeneous samples.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP.
+Recently, some have developed algorithms to call low-frequency or rare variants in heterogeneous samples.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP.
 VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
@@
 RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}.
-However, that algorithm requires a very high read depth to estimate the sequencing error rate and call variants.
+That algorithm requires a very high read depth to estimate the sequencing error rate and call variants.
 
@@
 	\item For each location $j$:
-	\begin{enumerate}
+	\begin{enumerate}[noitemsep]
 		\item Draw an error rate $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$
@@
 
-The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision;  $ \mu_j $, a local error rate.
-$M_j$, a local precision.
+The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision;  $ \mu_j $, a local error rate; and $M_j$, a local precision.
 The global error rate, $\mu_0$, estimates the expected error rate across all locations.
@@
 The prior distribution allows $ \mu_j $ to borrow information from adjacent positions and allows RVD2 to handle low read depths.
-second, this method smoothly handles multiple replicates in case samples.
+Second, RVD2 handles multiple replicates in case samples.
 Third, RVD2 has a more accurate Bayesian hypothesis testing method compared to a frequentist normal z-test in RVD.
@@
 \subsection{Initialization}
-The initial values for the model parameters and latent variables is obtained by a method-of-moments (MoM) procedure.
+The initial values for the model parameters and latent variables are obtained by a method-of-moments (MoM) procedure.
 MoM works by setting the population moment equal to the sample moment.
@@
 We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples.
-Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.
+Since, each position $j$ is exchangeable given the global hyperparameters, $\mu_0$ and $M_0$, this sampling step can be distributed across up to $J$ processors.
 
@@
 
-Posterior Somatic Test is a two-sided posterior difference test using paired control-case sample.
-To identify somatic mutations, we consider scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
+We use a two-sided posterior difference test using control and case samples to test for somatic mutations.
+We consider scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
 loss-of-heterozygosity) as well as scenarios when the case(tumor) error rate is higher than the control(germline) error rate (e.g.
@@
 The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < \tau ) > 1-\alpha$.
-Threshold $ \tau $ is set at zero.
+We typically set the hreshold $ \tau $ to zero.
 
@@
 
-Posterior Germline Test is a one-sided posterior distribution test using a single control sample. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,
+We use a one-sided posterior distribution test using a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,
 
 \begin{equation}\label{eqn:germ_test}
-	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha,
+	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha.
 \end{equation}
@@
 where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
-\citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available; we choose that value.
+\citet{cressie1984multinomial} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.
 
@@
 If the sample average depth is higher than 500, we use the Benjamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the posterior distribution test \citep{benjamini1995controlling, efron2010large}.
-The average depth threshold is set because Benjamini-Hochberg method is a highly conservative method and will remove many true calls when the read depth is not high enough.
+The average depth threshold is set because Benjamini-Hochberg method is a highly conservative method and will reject many true calls when the read depth is not high enough.
 
@@
 We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms.
-Synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
-HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
+The synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions.
+The HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.
 
 \subsection{Synthetic DNA Sequence Data}
+
 \subsubsection*{Experimental process.}
 Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined MAFs of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele fractons (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
 More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+
 \subsubsection*{Computational process.}
@@
 We used samtools mpileup to generate pileup files using hg19 as reference sequence~\citep{Navin:2010gu}.
-% TODO:cite the technical report. %Not sure the author is correct
+%Not sure the author is correct
 % HCC1187BL origin: http://www.atcc.org/products/all/CRL-2323.aspx
@@
 
-We performed posterior disparity test on synthetic data to identify mutations given it is a haploid DNA sequence.
+We performed the posterior distribution test for somatic mutations on the synthetic data set to identify mutations in the haploid DNA sequence.
 We set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
-We used RVD2 to identify somatic and germline mutations in the diploid HCC1187 sample.
+
+For the HCC1187 data set, we identified both somatic and germline mutations.
 In the posterior somatic test, we set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
 In the posterior germline test, We set  the threshold $ \tau =0.05 $ considering the low average coverage (40x).
-The size of the test is set at $ \alpha = 0.15$, higher than the size of somatic test.
-We are less confident on the germline test because we only use control sample in the germline test compared to normal-tumor paired sample in somatic test.
-We performed $ \chi^2 $ non-uniformity test along all the posterior density test.
+The size of the test is set at $ \alpha = 0.15$.
+We performed $ \chi^2 $ non-uniformity test after the posterior density tests.
 
-\subsection{Performance with read depth}\label{sec:read_depth}
+\subsection{Performance by read depth}\label{sec:read_depth}
 
@@
 Each subfigure shows ROC curves across four different read depths for one MAF level.
-Within one subfigure，the performance improves monotonically with read depth.
-Across different subfigures, the performance improves with MAF level.}
+The performance improves monotonically with read depth.
+The performance improves for all read depths with increasing MAF level.}
 \label{fig:ROC}
@@
 
-We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity.
-Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2 Somatic, Strelka and muTect are designed to call variants in heterogeneous samples, which serve as better comparison to RVD2.
-In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants.
+We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data set by the false discovery rate and sensitivity/specificity.
+Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2 Somatic, Strelka and muTect are have good performance to call variants in heterogeneous samples.
+In a research applications, the false discovery rate is a more relevant performance metric because the aim is generally to identify interesting variants.
 The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
-GATK, Varscan2, Strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.
+GATK, Varscan2, Strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for comparison.
 
@@
 \subsubsection*{Sensitivity/Specificity Comparison}
+
 \begin{figure*}[!htpb]
@@
 \includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
-\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms using synthetic sequence data.}
+\caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms using synthetic DNA data.}
 \label{tbl:comparison_ss}
@@
 \includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
-\caption{False discovery rate comparison of RVD2 with other variant calling algorithms using synthetic sequence data. Blank cells indicate no locations were called variant.}
+\caption{False discovery rate comparison of RVD2 with other variant calling algorithms using synthetic DNA data. Blank cells indicate no locations were called variant.}
 \label{tbl:comparison_fdr}
@@
 \end{figure*}
+
 Figure~\ref{tbl:comparison_ss} shows that samtools, GATK and VarScan2-mpileup all have similar performance.
 They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples.
-GATK, samtools and VarScan2-mpileup are optimized to call genotypes on pure samples.
-Therefore, those algorithms are expected to perform well on the 100\% dilution (pure mutant) sample and poorly on heterogeneous samples.
 VarScan2-somatic is able to call more mixed samples.
@@
 
-The statistics for RVD is an average statistics as RVD uses six control replicates to estimate the model but returns statistics separately for three sets of pair-end case replicates.
-RVD performed the best among all algorithms when the read depth is as high as 40000x.
+The performance statistics for RVD are an average three sets of pair-end case replicates.
+RVD performed the best among all algorithms when the read depth is near 40,000.
 RVD called all the mutated positions across all MAF levels with no false positives when MAF level is 0.3\% or lower.
-RVD calls some false positives when the MAF level is 1.0\% or higher, which causes specificity slightly lower than 1.00.
-RVD program fails and can not call any mutations when the depth is unable to measure the baseline error rate.
+However, RVD can not call any mutations when the depth is too low to measure the baseline error rate and therefore is not useful for low-depth data.
 
-The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample.
+RVD2 has a high sensitivity and specificity for a broad range of read depths and minor-allele frequencies.
 The sensitivity increases considerably with read depth at a slight expense to specificity.
+For the most difficult test with a low read depth and low MAF, performs on-par with muTect.
 With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs.
@@
 When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
+
 \begin{figure*}[!hbpt]
@@
 \includegraphics[width=0.85\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for Germline and Somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
 Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence.
@@
 The  vertical lines represent 95\% credible interval around posterior mean MAF.
-Five positions are common populations SNPs according to dbSNPv138, and the identities are shown below the positions.}
+Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
 \label{fig:brca_MAF}
@@
 
-RVD2 has a lower FDR than other algorithms when the read depth is greater than $10\times$ the MAF with $N=1$ and $\tau$ set to the default value of zero or to the optimal value.
+RVD2 has a lower FDR than other algorithms when the read depth is greater than $10\times$ the inverse MAF with $N=1$ and $\tau$ set to the default value of zero or to the optimal value.
 The FDR is higher when $N=6$ because the variance of the control error rate distribution $P(\mu_j^{\text{control}} | r^{\text{control}})$ is smaller.
@@
 \subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
-\subsubsection{Performance of RVD2.}
 
@@
     \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
-    \caption{(a)Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-VarScan2-somatic reported 60 positions, muTect 11 positions, RVD2 12 positions and Strelka only 1 position.
-This figure uses position index to show the correspondence of positions called by different algorithms for notation simplicity.
-Complete actual positions and depth distribution are provided in Supplementary Table 1 for validation.
-(b) Depth distribution for positions called by RVD2 and muTect.}%
+    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+VarScan2-somatic reported 60 positions, muTect called 11 positions, RVD2 called 12 positions and Strelka called 1 position.
+The unique called positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%
     \label{fig:brca_dot}%
@@
 RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-There were eight germline variants and eight somatic mutations in the twelve variants.
+There were eight germline variants and eight somatic mutations.
 %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
-%TODO :check with pos 154749704G>G in somatic
 Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions\footnote{The notation format is Chromosome:PositionReference\textgreater ControlMutation.} chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations.
-Positions\footnote{The notation format is Chromosome:PositionReference\textgreater TumorMutation.} chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
+Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations.
+Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
 Positions chr7:154754371 and chr7:154758813 were considered loss-of-heterozygosity events.
@@
 The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}.
-The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1 for validation.
+The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1.
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
@@
 Figure~\ref{fig:brca_dot:a} shows mutation detection result from Strelka, RVD2, muTect, and VarScan2-somatic, the state-of-art algorithms able to call mutation from heterogeneous samples.
-For notation simplicity, we use position index to present actual positions in Figure~\ref{fig:brca_dot}, while the correspondence is provided in Supplementary Table 1.
+For notation simplicity, we use position index to present actual positions in Figure~\ref{fig:brca_dot} (the correspondence to genomic positions is provided in Supplementary Table 1).
 
 The mutations called by RVD2 and muTect are the most consistent among all the techniques.
-RVD2 detected twelve germline and somatic mutations, while muTect reported eleven, ten in common.
+RVD2 detected twelve germline and somatic mutations, while muTect reported eleven; ten were called by both.
 In the disagreements, RVD2 did not call position 50 while muTect did not call position 37 and 58.
@@
 
-Strelka was the least sensitive algorithms among all the algorithms.
+Strelka was the least sensitive algorithm among all the algorithms.
 According to the technical report, Strelka identified position 22 (chr7:154760439) as variant, but did not call any other variants.
 In particular Strelka missed the two LOH events called by RVD2.
-On the contrary, VarScan2-somatic called most positions among all algorithms, sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
+VarScan2-somatic called most positions among all algorithms, sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
 VarScan2-somatic  detected all the positions called by RVD2 except position 35, which turns out to be a very likely mutation given the depth distribution in Figure~\ref{fig:brca_dot:b}.
-On the other side, VarScan2-somatic reported fifty positions which were not called by any other three algorithms.
+VarScan2-somatic reported fifty positions which were not called by any other three algorithms.
 The read depth in Supplementary Table 1 suggestions that these positions are very likely to be false positives.
@@
 Another factor that can affect the speed of RVD2 is the number of Metropolis-Hasting sample within one Gibbs sampling run.
-However, RVD2 is able to use multiple cores in parallel ( no more than the number of Metropolis-Hasting samples), which can significantly improve time efficiency.
-on a computational mode with 64 2.34GHz processor and 256Gb of RAM, RVD2 takes around 10 minutes to analyze the 400bp long synthetic data for one dilution rate and one downsampling rate.
+However, RVD2 is able to use multiple cores in parallel, which can significantly improve time efficiency.
+On a computational node with 64 2.34GHz processor and 256Gb of RAM, RVD2 takes around 10 minutes to analyze the one 400bp long synthetic data set.
 The memory requirement is no higher than the size of the gene sequence times the number of Gibbs samples.
-We are currently developing a variational method to estimate RVD2, aiming at decreasing time complexity.
+We are currently developing a variational method to estimate RVD2, aiming at decreasing time cost.
 

### 2014-03-30 doc/bioinformatics2014/supplementary_info.tex
edits to manuscript and supplementary info.

--- 2014-03-17 doc/bioinformatics2014/supplementary_info.tex
@@
 \maketitle
+
+%%%%%%%%%%%%%%%%%%%
+% Appendix A: An simulation example to illustate the idea of proposal graphical model and v.
+%%%%%%%%%%%%%%%%%%%
+\section{Example to illustrate the proposal graphical model and generative process.}\label{sec:appendix_example}
+
+\begin{figure}[!bpth]
+    \centering
+    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/RVD2_model.pdf} }}%
+    \quad
+    \quad
+    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/GMexample.pdf} }}%
+    \caption{(A) RVD2 Graphical Model.
+(B) Example to illustrate the proposal graphical model and generative process.
+Here we want to generate a simulation sample with 3 replicates, and each replicate is a 5bp long sequence.
+Therefore, we have $ J = 5 $, $N = 3$.
+The hyperparameters in orange, $\mu_0 \in \RR^{1 \times 1}$, $M_0 \in \RR^{1 \times 1}$, $M \in \RR^{5 \times 1}$, $n \in \RR^{5 \times 3}$ are set.
+(1) Sample the local error rate $\mu \in \RR^{5 \times 1}$ by drawing random numbers $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
+(2) Use $\mu $ and $ M $ to sample $\theta \in \RR^{5 \times 3}$ by drawing random samples $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
+(3) Use $\theta \in \RR^{5 \times 3}$ and $n \in \RR^{5 \times 3}$ to sample $r \in \RR^{5 \times 3}$ by drawing random numbers $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.}
+    \label{fig:brca_MAF}%
+\end{figure}
 

### 2014-03-31 doc/bioinformatics2014/document.tex
Update the manuscript

@@
 
-We use a two-sided posterior difference test using control and case samples to test for somatic mutations.
+We use a two-sided posterior difference test with control and case paired samples to identify somatic mutations.
 We consider scenarios when the case(tumor) error rate is lower than the control(germline) error rate (e.g.
@@
 
-We use a one-sided posterior distribution test using a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,
+We use a one-sided posterior distribution test with a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,
 
@@
 Matched normal cells were derived from lymphoblastoid cells from peripheral blood.
+Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
 \subsubsection*{Computational process.}
-Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
  The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.
@@
 
-We performed the posterior distribution test for somatic mutations on the synthetic data set to identify mutations in the haploid DNA sequence.
-We set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
+We performed the posterior difference test to identify mutations in the haploid synthetic data. We set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
 
@@
 % rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
+% from folder 'rvd2\results\2013-09-19_operating_characteristics'
 
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
-%\afterpage{\clearpage}
 \subsubsection*{Sensitivity/Specificity Comparison}
@@
 \begin{center}
-\includegraphics[width=0.85\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
 \caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
@@
 Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
-Positions chr7:154754371 and chr7:154758813 were considered loss-of-heterozygosity events.
+Positions chr7:154754371 and chr7:154758813 appears to be loss-of-heterozygosity events.
 Some of these mutations are also found to be common population SNPs according to dbSNPv138.

### 2014-03-31 doc/bioinformatics2014/supplementary_info.tex
Update the manuscript

@@
 Therefore, we have $ J = 5 $, $N = 3$.
-The hyperparameters in orange, $\mu_0 \in \RR^{1 \times 1}$, $M_0 \in \RR^{1 \times 1}$, $M \in \RR^{5 \times 1}$, $n \in \RR^{5 \times 3}$ are set.
-(1) Sample the local error rate $\mu \in \RR^{5 \times 1}$ by drawing random numbers $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
+The hyperparameters in orange, $\mu_0 \in \RR^{1 \times 1}$, $M_0 \in \RR^{1 \times 1}$, $M \in \RR^{5 \times 1}$ are set.
+$n \in \RR^{5 \times 3}$ in the generative process is given as well.
+The generative process is: (1) Sample the local error rate $\mu \in \RR^{5 \times 1}$ by drawing random numbers $\mu_j \thicksim \text{Beta}(\mu_0, M_0)$.
 (2) Use $\mu $ and $ M $ to sample $\theta \in \RR^{5 \times 3}$ by drawing random samples $\theta_{ji} \thicksim \text{Beta}(\mu_j, M_j)$.
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
+\includegraphics[width=1\textwidth]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
 \caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
@@
 
-Samtools mpileup and GATK can accept multiple "tumor'' replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
-VarScan2-mpileup takes multiple ``tumor-normal'' pair replicates so we passed six pair replicates to each algorithm.
-Varscan2-somatic, strelka and muTect do not accept replicate data for the ``normal" or ``tumor" bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
+Samtools mpileup and GATK can accept multiple tumor replicates for variant calling, so we fed six bam files from each case replicate group to mpileup.
+VarScan2-mpileup takes multiple tumor-normal pair replicates so we passed six pair replicates to each algorithm.
+Varscan2-somatic, strelka and muTect do not accept replicate data for the normal or tumor bam files so we used a single bam file from each replicate group with a read depth that most closely matched the overall median depth of the replicates.
 
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_pos.pdf}
+\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_histogram_pos.pdf}
 \caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
@@
 \begin{center}
-\includegraphics[width=120mm]{pdf_figs/HCC1187_histogram_neg.pdf}
+\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_histogram_neg.pdf}
 \caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}

### 2014-03-31 doc/bioinformatics2014/document.tex
Update the manuscript

@@
 Finally, we show that RVD2 is able to call variants on a heterogeneous clinical sample and identify two novel loss-of-heterozygosity events.
-%TODO: Fixed this part
+
 %%%%%%%%%%%%%%%%%%
@@
 Then, it tests for a significant difference between key model parameters in the case and control samples and returns called variant positions.
+
+Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
+In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{jordan2004graphical}.
+A rounded box or ``plate" represents replication of the nodes within the plate.
+The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
+
+\begin{figure}[!bpth]
+\begin{center}
+\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
+\caption{RVD2 Graphical Model.}
+\label{fig:graphical_model}
+\end{center}
+\end{figure}
 
@@
 Importantly, increasing the read depth $n_{ji}$ only reduces the sampling error, but does nothing to reduce experimental variation or variation due to sequence context.
-
-\begin{figure}[!bpth]
-\begin{center}
-\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
-\caption{RVD2 Graphical Model.}
-\label{fig:graphical_model}
-\end{center}
-\end{figure}
-
-Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model.
-In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{jordan2004graphical}.
-A rounded box or ``plate" represents replication of the nodes within the plate.
-The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.
 
@@
 
-\subsubsection*{Experimental process.}
+\subsubsection*{Experimental Methods.}
 Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
@@
 
-\subsubsection*{Computational process.}
+\subsubsection*{Pre-processing Methods.}
  We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
@@
 \subsection{HCC1187 Sequence Data}
-\subsubsection*{Experimental process.}
+\subsubsection*{Experimental Methods.}
 The HCC1187 dataset is a well-recognized baseline dataset from Illumina for evaluating sequence analysis algorithms~\citep{newman2013relative,howarth2011large,howarth2007array}.
@@
 Sequencing libraries were prepared according to the protocol described in the original technical report~\citep{hcc1187techreport}.
-\subsubsection*{Computational process.}
+\subsubsection*{Pre-processing Methods.}
  The raw FASTQ read files were aligned to hg19 using the Isaac aligner to generate BAM files~\citep{raczy2013isaac}.

### 2014-04-02 doc/bioinformatics2014/document.tex
correct a minor error in the manuscript

@@
 %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
-Figure~\ref{fig:brca_MAF:a} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
+Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
 Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations.
@@
 Some of these mutations are also found to be common population SNPs according to dbSNPv138.
-The corresponding identities are shown in the Figure~\ref{fig:brca_MAF:b}.
+The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}.
 The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1.

### 2014-04-09 doc/bioinformatics2014/revision0/document.tex
edits to rvd27 command line. rearranged docs folder.

--- 2014-04-02 doc/bioinformatics2014/document.tex
@@
 DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
+Importantly, accounting for technical errors can drastically improve performance~\citep{Zagordi2010}.
 
@@
 Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations.
-Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
+Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, \\ chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
 Positions chr7:154754371 and chr7:154758813 appears to be loss-of-heterozygosity events.

### 2014-04-09 src/python/rvd27/rvd27.py

@@ -55,11 +55,11 @@ def main():
     argpGibbs.set_defaults(func=gibbs)
 
     # create subparser to compare two model files
-    argpTest = subparsers.add_parser('test_main', 
-                        help='RVD2 algorithm to find mutations in tumor-normal-paired sample. A posteriror difference test and a somatic test will be done in this program.')
+    argpTest = subparsers.add_parser('test', 
+                        help='Posterior density testto find mutations in paired samples.')
 
     argpTest.add_argument('alpha', type=float, default=0.95,
-                help='poseterior difference test probability threshold')
+                help='posterior difference test probability threshold')
     argpTest.add_argument('controlHDF5Name', default=None,
                 help='control model file (HDF5)')
     argpTest.add_argument('caseHDF5Name', default=None,

### 2014-05-02 results/2013-07-18 Test on Gibbs Step Size with Synthetic DNA/Test_Synthetic_DNA_Notebook_T_Test.py
Edit some figure features in previous codes.

@@ -50,15 +50,17 @@ for n in xrange(len(nsample_opt)):
             muCase_s = f['mu'][...]
 
         ## plot histogram
-        position1=49
+        position1=72
         fig=plt.figure()
         ax1=fig.add_subplot(2,1,1)
         ax1.hist(muCase_s[position1,:],20)
         ax1.hist(muControl_s[position1,:],20)
-        title='Histogram of mu when nsample='+str(nsample_opt[n])+' dilution='+str(dilution)+' position='+str(position1+1)
+        title='position='+str(position1+1)
         ax1.set_title(title)
-        ax1.set_ylabel('t')
+        ax1.set_xlabel('$\mu_j$')
+        ax1.set_ylabel('Frequency')
         ax1.legend( ['Case','Control'])
+        ax1.ticklabel_format(axis='y', style='sci', scilimits=(-2,2))
 
         position2=244
         ax2=fig.add_subplot(2,1,2)
@@ -66,9 +68,11 @@ for n in xrange(len(nsample_opt)):
         ax2.hist(muControl_s[position2,:],20)
         title='position='+str(position2+1)
         ax2.set_title(title)
-        ax2.set_ylabel('t')
-        ax2.set_xlabel('mu')
-        ax2.legend(['Case','Control'])
+        ax2.set_ylabel('Frequency')
+        ax2.set_xlabel('$\mu_j$')
+        ax2.ticklabel_format(axis='y', style='sci', scilimits=(-2,2))
+        # ax2.legend(['Case','Control']
+        plt.tight_layout()
         plt.savefig('Histogram of mu_s when nsample='+str(nsample_opt[n])+' Dilution='+str(dilution)+'.jpg')
 
         ##    scipy.stats.ttest_ind(a, b, axis=0, equal_var=True)[source] returns
@@ -95,5 +99,5 @@ for n in xrange(len(nsample_opt)):
         title='tstat '+'with assumed equal variace when nsample='+str(nsample_opt[n])+'.jpg'
     else:
         title='tstat '+'with assumed unequal variace when nsample='+str(nsample_opt[n])+'.jpg'
-        
+    plt.tight_layout()    
     plt.savefig(title)

### 2014-05-02 results/2013-07-20 Test on MH Sample Size with Synthetic DNA/Test_Synthetic_DNA_Notebook_T_Test.py

@@ -21,7 +21,7 @@ import rvd27
 
 dilution_opt=(0.1,0.3,1.0,10.0,100.0)
 gibbs_nsample_opt=[400]
-mh_nsample_opt=[100]
+mh_nsample_opt=[1,10]
 
 
 var=False # whether the variance of muCase_s and muControl_s is assumed equal or not
@@ -57,10 +57,12 @@ for n in xrange(len(gibbs_nsample_opt)):
             ax1=fig.add_subplot(2,1,1)
             ax1.hist(muCase_s[position1,:],20)
             ax1.hist(muControl_s[position1,:],20)
-            title='Histogram of mu_s when ngibbs='+str(gibbs_nsample_opt[n])+'_nmh='+str(mh_nsample_opt[m])+' position='+str(position1+1)
+            title='position='+str(position1+1)
             ax1.set_title(title)
-            ax1.set_ylabel('t')
+            ax1.set_xlabel('$\mu_j$')
+            ax1.set_ylabel('Frequency')
             ax1.legend( ['Case','Control'])
+            ax1.ticklabel_format(axis='y', style='sci', scilimits=(-2,2))
 
             position2=244
             ax2=fig.add_subplot(2,1,2)
@@ -68,9 +70,11 @@ for n in xrange(len(gibbs_nsample_opt)):
             ax2.hist(muControl_s[position2,:],20)
             title='position='+str(position2+1)
             ax2.set_title(title)
-            ax2.set_ylabel('t')
-            ax2.set_xlabel('mu')
-            ax2.legend(['Case','Control'])
+            ax2.set_ylabel('Frequency')
+            ax2.set_xlabel('$\mu_j$')
+            ax2.ticklabel_format(axis='y', style='sci', scilimits=(-2,2))
+            # ax2.legend(['Case','Control'])
+            plt.tight_layout()
             plt.savefig('Histogram of mu_s when ngibbs='+str(gibbs_nsample_opt[n])+'_nmh='+str(mh_nsample_opt[m])+' Dilution='+str(dilution)+'.jpg')
 
             ##    scipy.stats.ttest_ind(a, b, axis=0, equal_var=True)[source] returns

### 2014-05-02 results/2013-08-02 MCMC traces and autocorrelation analysis/MCMC_traces.py

@@ -60,57 +60,75 @@ def MCMCtraces(h5FileName,position,replicate,dilution=None,title=None,figform='.
         mu_s = f['mu'][...]
         theta_s = f['theta'][...]
         
-    fig=plt.figure(figsize=(18,9))
+    fig=plt.figure(figsize=(9.5,11))
     
-    if dilution is not None:
-        plt.suptitle('Case dilution='+str(dilution)+' position='+str(position)+' replicate='+str(replicate))
-    else:
-        plt.suptitle('Control position='+str(position)+' replicate='+str(replicate))
+    # if dilution is not None:
+    #     plt.suptitle('Case dilution='+str(dilution)+' position='+str(position)+' replicate='+str(replicate))
+    # else:
+    #     plt.suptitle('Control position='+str(position)+' replicate='+str(replicate))
     
     '''subplot for theta'''
     (N,J,G)=np.shape(theta_s)
-    ax1=fig.add_subplot(2,3,1)
-    
+    ax1=fig.add_subplot(3,2,1)
+    drotation = 19
     ax1.plot(np.arange(G),theta_s[replicate-1,position-1,:])
-    ax1.set_title('MCMC trace for theta')
+    ax1.set_title(r'MCMC trace for $\theta$')
     ax1.set_xlabel('Gibbs steps')
-    ax1.set_ylabel('theta')
+    # plt.xticks(rotation=drotation) 
+    ax1.set_ylabel(r'$\theta$')
+    ax1.ticklabel_format(axis='x', style='sci', scilimits=(1,2))
+    ax1.ticklabel_format(axis='y', style='sci', scilimits=(-2,2))
 
-    ax2=fig.add_subplot(2,3,2)
+    ax2=fig.add_subplot(3,2,3)
     ax2.acorr(theta_s[replicate-1,position-1,:], usevlines=True, normed=True, detrend=matplotlib.mlab.detrend_mean, maxlags=1599, lw=1)
-    ax2.set_title('MCMC autocorr for theta')
-    ax2.set_xlabel('lags')
-    ax2.set_ylabel('autocorr')
+    ax2.set_title(r'MCMC autocorrelation for $\theta$')
+    ax2.set_xlabel('Lag')
+    # plt.xticks(rotation=drotation) 
[…]

### 2014-05-02 doc/bioinformatics2014/revision0/supplementary_info.tex

--- 2014-03-31 doc/bioinformatics2014/supplementary_info.tex
@@
     \centering
-    \subfloat[\label{fig:brca_MAF:a}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/RVD2_model.pdf} }}%
+    \subfloat[\label{fig:example:a}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/RVD2_model.pdf} }}%
     \quad
     \quad
-    \subfloat[ \label{fig:brca_MAF:b}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/GMexample.pdf} }}%
+    \subfloat[ \label{fig:example:b}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/GMexample.pdf} }}%
     \caption{(A) RVD2 Graphical Model.
@@
 (3) Use $\theta \in \RR^{5 \times 3}$ and $n \in \RR^{5 \times 3}$ to sample $r \in \RR^{5 \times 3}$ by drawing random numbers $r_{ji} \thicksim \text{Binomial}(\theta_{ji}, n_{ji})$.}
-    \label{fig:brca_MAF}%
+    \label{fig:example}
 \end{figure}

### 2014-05-29 results/2014-02-03_variational_inference/rvd_var.py
Update the variational rvd27 codes. The program can run through now, but the variational distribution parameters are not yet properly optimized.

@@ -27,20 +27,20 @@ def main():
     logging.basicConfig(level=log_level, format='%(levelname)s:%(module)s:%(message)s')
 
     ## Generate simulation data
-    J = 5 # number of positions
+    J = 7 # number of positions
     N = 3 # number of replicates
     
     n = np.empty([N,J], dtype=np.int64) # read depth (rep, location)
-    n.fill(100)
+    n.fill(1000)
     
     phi = {'M0':100, 'mu0':0.1, 'M':[1000]*J}
-    (r, theta, mu) = generate_sample(phi, N, J, n, seedint=10241978)
-    
+    (r, theta, mu) = generate_sample(phi, N, J, n, seedint=19891129)
+
     ## model optimization  
-    (phiHat, qHat) = ELBO_opt(r, n, seed = 10241978, pool = None)
+    (phiHat, qHat) = ELBO_opt(r, n, seed = 199096, pool = None)
 
     ## save the parameters.
-    #save_model('model.hdf5', r, n, phiHat, qHat)
+    save_model('model.hdf5', r, n, phiHat, qHat)
     
 def generate_sample(phi, N=3, J=100, n=100, seedint=None):
     """Returns a sample with n reads, N replicates, and
@@ -70,7 +70,11 @@ def generate_sample(phi, N=3, J=100, n=100, seedint=None):
 
 ## compute sufficient statistics
 def EqlogTheta(delta):
-	return psi(delta[0]) - psi(np.sum(delta))
+    try:
+        return psi(delta[0]) - psi(np.sum(delta))
+    except RuntimeWarning:
+        pdb.set_trace()
+	
 
 def Eqlog1_Theta(delta):
 	return psi(delta[1]) - psi(np.sum(delta))
@@ -79,21 +83,18 @@ def EqMu(gam):
 	return gam[0] / (np.sum(gam)) # eps?
 
 def EqlogMu(gam):
-    return psi(gam[0]) - psi(np.sum(gam))
+    try:
+        return psi(gam[0]) - psi(np.sum(gam))
+    except ValueError:
+        pdb.set_trace()
+    
 
 def Eqlog1_Mu(gam):
 	return psi(gam[1]) - psi(np.sum(gam))
 
-# def EqlogGamma(gam, M):
-# 	# Eqlog\gam(muM)
-# 	N = 7
-# 	mu = range(0, 1 , 2^N+1)
-# 	y = [kernel(x, gam , M) for x in mu]
-
[…]

### 2014-06-02 doc/bioinformatics2014/revision1/document.tex
revision 1: reduce the manuscript page number to 8

--- 2014-04-09 doc/bioinformatics2014/revision0/document.tex
@@
 \usepackage{subfig}
-%\usepackage[font={small}]{caption}
+\usepackage{wrapfig}
+\usepackage[font={small}]{caption}
 \usepackage[skip=2pt,font=scriptsize]{caption}
@@
 \title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
-\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{to whom correspondence should be addressed}}
+\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{Patrick Flaherty: pjflaherty@wpi.edu}}
 \address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}
@@
 \section{Availability:}
-\href{http://genomics.wpi.edu/rvd2/rvd2.html}{http://genomics.wpi.edu/rvd2/rvd2.html}
+\href{http://genomics.wpi.edu/rvd2/}{http://genomics.wpi.edu/rvd2/}
 
@@
 DNA sequencing of individual regions of a solid tumor has revealed genetic heterogeneous within an individual sample~\citep{Navin:2010gu}.
-Importantly, accounting for technical errors can drastically improve performance~\citep{Zagordi2010}.
+Importantly, accounting for technical errors can drastically improve performance~\citep{zagordi2010error}.
 
@@
 RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}.
-That algorithm requires a very high read depth to estimate the sequencing error rate and call variants.
 
@@
 \begin{figure}[!bpth]
-\begin{center}
-\includegraphics[width=40mm]{pdf_figs/RVD2_model.pdf}
+\centering
+\vspace{-10pt}
+\includegraphics[width=35mm]{pdf_figs/RVD2_model.pdf}
 \caption{RVD2 Graphical Model.}
+\vspace{-20pt}
 \label{fig:graphical_model}
-\end{center}
 \end{figure}
@@
 For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$.
-$ J $ is the region of interest length and $ N $ is the number of technical replicates in the sample.\label{replicate definition} Technical replicates are used to establish experimental variability in next-generation sequencing procedure~\citep{quackenbush2002microarray,robasky2013role}, though multiple replicates are not necessary for RVD2.
+$ J $ is the region of interest length and $ N $ is the number of technical replicates in the sample.\label{replicate definition} Technical replicates are used to establish experimental variability in next-generation sequencing procedure~\citep{robasky2013role}, though multiple replicates are not necessary for RVD2.
+%~\citep{quackenbush2002microarray,robasky2013role}
 
@@
 The joint distribution over the latent and observed variables for data at location $j$ in replicate $i$ given the parameters can be factorized as
-
 \begin{multline}\label{eqn:jointpdf}
@@
 \end{multline}
-
-where
-
-\begin{align}
-  \begin{split}
-  p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \cdot \\
-  &\quad \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
-  \end{split}\\
+where,
+\begin{align}\nonumber
+  p\left( \mu_j; \mu_0, M_0 \right)  &= \dfrac{\mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}}{\textbf{B}(M_0 \mu_0,M_0 (1-\mu_0) )}, \nonumber \\
  %
-  \begin{split}
-  p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \cdot \\
-  &\quad \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
-  \end{split}\\
+  p\left( \theta_{ji} | \mu_j; M_j \right) &= \dfrac{ \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}}{\textbf{B}(M_j \mu_j, M_j (1-\mu_j))}, \nonumber \\
  %
-  \begin{split}
-  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \\
-  &\quad \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber\\
-  \end{split}\\
+  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
 \end{align}
 
-Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
-\begin{multline}
-p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
-\end{multline}
+%Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
+%\begin{multline}
+%p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) \\ p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
+%\end{multline}
 
-Finally, the log-likelihood of the data set is
-
+The log-likelihood of the data set is
 \begin{multline}
@@
 %%%%%%%%%%%%%
+\vspace{-20pt}
 \subsection{Initialization}
@@
 \begin{figure}[htbp]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{pdf_figs/ROC_without_chi2.pdf}
-\caption{ROC curve varying read depth showing detection performance.
-Each subfigure shows ROC curves across four different read depths for one MAF level.
-The performance improves monotonically with read depth.
-The performance improves for all read depths with increasing MAF level.}
+\centering
+\vspace{-15pt}
+\includegraphics[width=0.45\textwidth]{pdf_figs/ROC_without_chi2.pdf}
+\caption{ROC curve varying read depth showing detection performance. }
+\vspace{-15pt}
 \label{fig:ROC}
-\end{center}
 \end{figure}
-
-% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
 We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs).
@@
 Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.
+% from folder '2013-11-24_six_replicates_Compute_ROC_Synthetic_Qsd_0_1'
 
@@
 \begin{figure*}[!htpb]
-\begin{center}
-\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_ss.pdf}
+\centering
+\vspace{-10pt}
+\includegraphics[width=0.84\textwidth]{pdf_figs/comparison_table_ss.pdf}
 \caption{Sensitivity/Specificity comparison of RVD2 with other variant calling algorithms using synthetic DNA data.}
+\vspace{-10pt}
 \label{tbl:comparison_ss}
-\end{center}
 \end{figure*}
@@
 \begin{figure*}[!tbhp]
-\begin{center}
-\includegraphics[width=1\textwidth]{pdf_figs/comparison_table_fdr.pdf}
+\centering
+\includegraphics[width=0.84\textwidth]{pdf_figs/comparison_table_fdr.pdf}
 \caption{False discovery rate comparison of RVD2 with other variant calling algorithms using synthetic DNA data. Blank cells indicate no locations were called variant.}
+\vspace{-15pt}
 \label{tbl:comparison_fdr}
-\end{center}
 \end{figure*}
@@
 When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases.
-
-\begin{figure*}[!hbpt]
-\begin{center}
-\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence.
-Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$.
-The  vertical lines represent 95\% credible interval around posterior mean MAF.
-Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
-\label{fig:brca_MAF}
-\end{center}
-\end{figure*}
 
@@
 
+\vspace{-10pt}
 \subsection{HCC1187 primary ductal carcinoma sample}\label{sec:brca}
+\begin{figure*}[!hbpt]
+\centering
+\vspace{-10pt}
+\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene.
+Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence.
+Red stars (*) indicates somatic mutations, where $\mu^{\text{case}}$ is significantly different from $ \mu^{\text{control}}$.
+The  vertical lines represent 95\% credible interval around posterior mean MAF.
+Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
+\vspace{-10pt}
+\label{fig:brca_MAF}
+\end{figure*}
 
 \begin{figure*}[!pthb]
+\vspace{-10pt}
     \centering
-    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
+    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
     %
-    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.2\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
+    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.18\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
     \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-VarScan2-somatic reported 60 positions, muTect called 11 positions, RVD2 called 12 positions and Strelka called 1 position.
-The unique called positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%
+The positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%
     \label{fig:brca_dot}%
+\vspace{-10pt}
 \end{figure*}
@@
 % To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
-
+\vspace{-5pt}
 \section{Discussion}
@@
 Such accurate algorithms are a step towards greater access to genomics for clinical diagnostics.
-
+\vspace{-5pt}
 \section*{Acknowledgement}
-We would like to thank Fan Zhang for careful review of this manuscript and helpful comments.
-
-\paragraph{Funding\textcolon}  P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute.
+\paragraph{Funding\textcolon}  P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute. Y.H. was supported by PhRMA Foundation Informatics Grant 2013080079.
 

### 2014-06-02 doc/bioinformatics2014/revision1/document.tex
revision1: Minor modification to the manuscript.

@@
 However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
-Samtools/bcftools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}.
+Samtools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}.
 GATK involves more sophisticate pre- and post-processing steps wherein the genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
@@
 RVD uses a hierarchical Bayesian model to capture the error structure of the data and call variants~\citep{Flaherty:2011ja, cushing2013rvd}.
+That algorithm requires a very high read depth to estimate the sequencing error rate and call variants.
 
@@
 \vspace{-10pt}
-\includegraphics[width=35mm]{pdf_figs/RVD2_model.pdf}
+\includegraphics[width=33mm]{pdf_figs/RVD2_model.pdf}
 \caption{RVD2 Graphical Model.}
@@
 \end{multline}
-where,
+
 \begin{align}\nonumber
-  p\left( \mu_j; \mu_0, M_0 \right)  &= \dfrac{\mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}}{\textbf{B}(M_0 \mu_0,M_0 (1-\mu_0) )}, \nonumber \\
+  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}, \nonumber \\
  %
@@
  %  
-  p\left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \cdot \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
+  p\left( \mu_j; \mu_0, M_0 \right)  &= \dfrac{\mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}}{\textbf{B}(M_0 \mu_0,M_0 (1-\mu_0) )}, \nonumber 
 \end{align}
 
+where \textbf{B}() denotes the beta function.
 %Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,

### 2015-02-02 doc/bioinformatics2014/revision2/document.tex
Added pdf figures in docs folder and removed *.pdf from .gitignore.

--- 2014-06-02 doc/bioinformatics2014/revision1/document.tex
@@
 \title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
-\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{Patrick Flaherty: pjflaherty@wpi.edu}}
+\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{Corresponding Author: pjflaherty@wpi.edu}}
 \address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}
@@
 However, the primary statistical tools for calling variants from NGS data are optimized for homogeneous samples.
-Samtools and GATK uses naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}.
+Samtools and GATK use a naive Bayesian decision rule to call variants~\citep{li2011statistical, depristo2011framework}.
 GATK involves more sophisticate pre- and post-processing steps wherein the genotype prior is fixed and constant across all loci and the likelihood of an allele at a locus is a function of the phred score~\citep{McKenna:2010bv}.
@@
 \end{align}
-
-where \textbf{B}() denotes the beta function.
+where \textbf{B}(x,y) denotes the beta function.
 %Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
@@
 Samples from $\theta_{ji}$ can be drawn from the posterior distribution directly  because the prior and likelihood form a conjugate pair.
-This sampling procedure is repeated until the chain converges to a stationary distribution then we draw samples from the posterior distribution over latent variables.
+This sampling procedure is repeated until the chain converges to a stationary distribution and then we draw samples from the posterior distribution over latent variables.
 
@@
 The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < \tau ) > 1-\alpha$.
-We typically set the hreshold $ \tau $ to zero.
+We typically set the threshold $ \tau $ to zero.
 
@@
 \end{equation}
-
 where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis.
@@
 Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
-Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele fractons (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
+Aliquots of the case and control DNA were mixed at defined fractions to yield defined minor allele fractions (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%.
 More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
@@
 We tested RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
-The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ tinning rate for a final total of $1,600$ samples.
+The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ thinning rate for a final total of $1,600$ samples.
 We drew $1,000$ samples from $ \tilde{\mu}^{\Delta} =  \tilde{\mu}_j^{\text{case}} - \tilde{\mu}_j^{\text{control}}$ to estimate the posterior probability of a variant.
@@
 In the posterior somatic test, we set the threshold $ \tau =0 $ and the size of the test $ \alpha = 0.05$.
-In the posterior germline test, We set  the threshold $ \tau =0.05 $ considering the low average coverage (40x).
+In the posterior germline test, We set  the threshold $ \tau =0.05 $ considering the low average coverage (40$\times$).
 The size of the test is set at $ \alpha = 0.15$.
-We performed $ \chi^2 $ non-uniformity test after the posterior density tests.
+We performed the $\chi^2$ non-uniformity test after the posterior density tests.
 
@@
 We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data set by the false discovery rate and sensitivity/specificity.
-Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2 Somatic, Strelka and muTect are have good performance to call variants in heterogeneous samples.
-In a research applications, the false discovery rate is a more relevant performance metric because the aim is generally to identify interesting variants.
+Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2-somatic, Strelka and muTect are have good performance to call variants in heterogeneous samples.
+In research applications, the false discovery rate is a more relevant performance metric because the aim is generally to identify interesting variants for followup.
 The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives.
-GATK, Varscan2, Strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for comparison.
+GATK, Varscan2, Strelka and muTect are only able to make use of one case and one control sample, so we provide results of RVD2 with the same data set ($N=1$) for comparison.
 
@@
 
-The performance statistics for RVD are an average three sets of pair-end case replicates.
+The performance statistics for RVD are an average of three sets of pair-end case replicates.
 RVD performed the best among all algorithms when the read depth is near 40,000.
@@
 The sensitivity increases considerably with read depth at a slight expense to specificity.
-For the most difficult test with a low read depth and low MAF, performs on-par with muTect.
+For the most difficult test with a low read depth and low MAF, RVD2 performs on-par with muTect.
 With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs.
@@
 
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate for RVD2 compared to samtools, GATK, varscan, Strelka and muTect. Blank cells indicate no positive calls were made.
+Figure~\ref{tbl:comparison_fdr} shows the false discovery rate (FDR) for RVD2 compared to samtools, GATK, varscan, Strelka and muTect. Blank cells indicate no positive calls were made.
 
@@
 \vspace{-10pt}
-\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\includegraphics[width=0.68\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
 \caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene.
@@
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations.
-Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G \textgreater T were called germline mutations.
+Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G\textgreater T were called germline mutations.
 Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, \\ chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
-Positions chr7:154754371 and chr7:154758813 appears to be loss-of-heterozygosity events.
+Positions chr7:154754371 and chr7:154758813 appear to be loss-of-heterozygosity events.
 Some of these mutations are also found to be common population SNPs according to dbSNPv138.
@@
 We also compared to the result shown in original research report where Strelka was used to identify mutations in the same sample ~\citep{hcc1187techreport}.
-Figure~\ref{fig:brca_dot:a} shows mutation detection result from Strelka, RVD2, muTect, and VarScan2-somatic, the state-of-art algorithms able to call mutation from heterogeneous samples.
-For notation simplicity, we use position index to present actual positions in Figure~\ref{fig:brca_dot} (the correspondence to genomic positions is provided in Supplementary Table 1).
+Figure~\ref{fig:brca_dot:a} shows mutation detection result from Strelka, RVD2, muTect, and VarScan2-somatic.
+For notation simplicity, we use position index to present actual positions in Figure~\ref{fig:brca_dot} (the corresponding genomic positions are provided in Supplementary Table 1).
 
@@
 The fact that VarScan2-somatic can be over-sensitive has appeared in synthetic dataset analysis.
-As shown in Figure~\ref{tbl:comparison_fdr}, the False Discovery Rate for VarScan2-somatic at read depth 53 MAF level 1.0\% is as high as 1.00.
+As shown in Figure~\ref{tbl:comparison_fdr}, the false discovery rate for VarScan2-somatic at read depth 53 MAF level 1.0\% is as high as 1.00.
 \citet{spencer2013performance} also mentioned that VarScan2 has tendency to call many false positives at high read depth.
@@
 
-Our inference algorithm uses Gibbs sampling to estimate the RVD2 hierarchical empirical Bayes model.
+Our inference algorithm uses Gibbs sampling to do inference in the RVD2 hierarchical empirical Bayes model.
 This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically.
@@
 Another factor that can affect the speed of RVD2 is the number of Metropolis-Hasting sample within one Gibbs sampling run.
-However, RVD2 is able to use multiple cores in parallel, which can significantly improve time efficiency.
+RVD2 is able to use multiple cores in parallel, which can significantly improve time efficiency.
 On a computational node with 64 2.34GHz processor and 256Gb of RAM, RVD2 takes around 10 minutes to analyze the one 400bp long synthetic data set.
-The memory requirement is no higher than the size of the gene sequence times the number of Gibbs samples.
-We are currently developing a variational method to estimate RVD2, aiming at decreasing time cost.
+The memory requirement is roughly the size of the gene sequence times the number of Gibbs samples.
+We are currently developing a variational method to estimate RVD2, aiming at decreasing computational time.
 
@@
 Such accurate algorithms are a step towards greater access to genomics for clinical diagnostics.
+
 \vspace{-5pt}
-\section*{Acknowledgement}
-\paragraph{Funding\textcolon}  P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute. Y.H. was supported by PhRMA Foundation Informatics Grant 2013080079.
+\section*{Acknowledgements}
+P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute. Y.H. was supported by PhRMA Foundation Informatics Grant 2013080079.
 

### 2015-02-23 doc/bioinformatics2014/revision2/response_to_reviews.txt
Address reviewer comment #1.

**When using synthetic DNA to evaluate the sensitivity and sensibility of RVD2, much is dedicated to (currently) unrealistic coverages. For example, in the Figure1 and Table1 (Figure 3) the authors presents a sequence coverage of 5.000x and even 55.000x. Despite being very useful for the own authors as high hypothetical control, I think ~30x - 100x coverage is more informative, and presenting these extremely high coverages can be misleading.**

The reviewer correctly points out that much next-generation DNA sequencing data has a median depth in the 30x - 100x range. We have added a paragraph to our discussion of the sensitivity and specificity results (second paragraph in Section 5.2). In that paragraph, we discuss three primary median depth ranges and typical applications for those areas. Briefly, whole genome sequencing typically has depths in the range of 10x - 100x as noted. In addition, targeted cancer gene sequencing typically has median read depths in the range 100x - 1,000x because the region of interest is smaller and there is interest in measuring tumor heterogeneity. Microbial and viral sequencing can have median depths from 1,000x - 100,000x because one is interested in identifying rare subpopulations that may prove resistant to antibiotics or antivirals.

We have chosen a broad range of median depths for our results so that researchers in whole genome sequencing, targeted sequencing and microbial sequencing can see how the methodology would perform in their application area.

### 2015-02-23 doc/bioinformatics2014/revision2/document.tex
Address reviewer comment #1.

@@
 
+We compare the empirical performance across a wide range of median read depth ($\sim$40$\times$ -- $\sim$40,000$\times$).
+In typical whole genome applications, the read depth is between 10$\times$ and 100$\times$.
+For targeted cancer sequencing, the median read depth is higher at 100$\times$ to 1,000$\times$.
+And, for microbial or viral sequencing for rare variants, the median read depth is even higher at 1,000$\times$ to 100,000$\times$.
+
 % from folder 'rvd2\results\2013-09-19_operating_characteristics'
@@
 They call the 100\% MAF experiment well even at low depth, but are unable to identify true variants in mixed samples. 
-VarScan2-somatic is able to call more mixed samples.
-However, as the read depth increases the specificity degrades.
+VarScan2-somatic is able to call more mixed samples. However, as the read depth increases the specificity degrades. 
 Strelka is able to call 10\% MAF variants with good performance, but is limited at 1\% MAF and below. 
@@
 
-Figure~\ref{tbl:comparison_fdr} shows the false discovery rate (FDR) for RVD2 compared to samtools, GATK, varscan, Strelka and muTect. Blank cells indicate no positive calls were made.
+Figure~\ref{tbl:comparison_fdr} shows the false discovery rate (FDR) for RVD2 compared to samtools, GATK, varscan, Strelka and muTect. 
+Blank cells indicate no positive calls were made.
 
@@
 VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level. 
-However, the FDR is high due to many false positives.
-Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth.
+However, the FDR is high due to many false positives. Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth. 
 Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF. 

### 2015-02-23 doc/bioinformatics2014/revision2/response_to_reviews.txt
Add discussion of sequencing protocol and results.

@@
-**When using synthetic DNA to evaluate the sensitivity and sensibility of RVD2, much is dedicated to (currently) unrealistic coverages.
+Reviewer 1
+
+1.
+When using synthetic DNA to evaluate the sensitivity and sensibility of RVD2, much is dedicated to (currently) unrealistic coverages.
 For example, in the Figure1 and Table1 (Figure 3) the authors presents a sequence coverage of 5.000x and even 55.000x.
-Despite being very useful for the own authors as high hypothetical control, I think ~30x - 100x coverage is more informative, and presenting these extremely high coverages can be misleading.**
+Despite being very useful for the own authors as high hypothetical control, I think ~30x - 100x coverage is more informative, and presenting these extremely high coverages can be misleading.
 
@@
 We have chosen a broad range of median depths for our results so that researchers in whole genome sequencing, targeted sequencing and microbial sequencing can see how the methodology would perform in their application area.
+
+2.
+Despite having the whole genome sequencing for HCC1187 and HCC1187BL cell lines, the authors focus all presented analysis one a single gene.
+It's not clear why the authors used the coordinates chr7:154738059154782774 (~45Kb), excluding the three first exons (four introns) and the last exon and most of the last intron, when, in fact, PAXIP1 has ~59Kb.
+No further explanation is given regarding why this specific region was chosen.
+
+Thank you for identifying this important omission.
+The region of interest for the PAXIP1 gene should have covered the entire 59Kb coding region.
+We ran RVD2 on the current NCBI start/stop codons (http://www.ncbi.nlm.nih.gov/gene/22976) to give the expanded region of chr7:154,735,400 - 154,794,682.
+This expanded region yields much better results.
+In particular, using the shortened region, we identified SNPs at XXX, XXX, XXX, and XXX.
+An inspection of Fig4 shows that these loci have very similar point estimates of the allele frequency in the case and control.
+In the analysis with the expanded region, RVD2 estimates a lower across-locus precision parameter (M0) and thus does not call these loci.
+However, it does call 2 mutations in the full-length PAXIP1 region of interest.
+
+We have updated our results to reflect the accurate start/stop locations for PAXIP1.
+
+3.
+Instead of using DNA sequencing of primary tumors genomes, which are highly heterogeneous and present many rare (single nucleotide) variants, the authors used the DNA from a commercially available cell line, which is colonially expanded (genetically homogenous).
+Interestingly, chromosome 7, where the investigated gene is located, is tetraploid on HCC1187[1], therefore, one would expect to find only allele frequencies 0%, 25%, 50%, 75 and 100% on variants located in this chromosome.
+Figure 5 seems to be in agreement to this hypothesis, except for variants 49704 and 81769, which has marginally lower estimations of allele frequency.Therefore, the variants detected on HCC1187 are not actually rare variants (MAF=25%) and, as shown on Figure 3 (Table1), may be detected by most of currently available methods.
+Despite the limitations, the analysis is complete, but it would be more informative to apply RVD2 to publicly available (and validated) DNAseq from primary tumors.
+
+4.
+The authors cite a paper from 2011 to refer for the Synthetic DNA sequencing.
+Explicit details on the technology, read length and number of reads generated would greatly ease readers understanding of this section.
+
+Section 4.1 describing the details of the experimental protocol for the synthetic sequencing data is indeed brief.
+We have added the details the reviewer suggested and as expected, the narrative is much clearer.
+Briefly, Supplementary Table S1 in the Nucleic Acid Research publication shows that each sample has roughly 1,000,000 35bp paired-end reads with no errors in the multiplexing barcodes to give an average read depth of ~100,000 across the synthetic amplicon.
+We describe this and more detail as suggested in the protocol section.

### 2015-02-23 doc/bioinformatics2014/revision2/document.tex
Add discussion of sequencing protocol and results.

@@
 \subsubsection*{Experimental Methods.}
-Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control.
+Two 400bp DNA sequences (including linkers) that are identical except at 14 loci with variant bases were synthesized and clonally isolated.
+The samples with the mutations are taken as the case sample and the sample without the mutations is taken as the control. 
 Aliquots of the case and control DNA were mixed at defined fractions to yield defined minor allele fractions (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. 
+Paired-end sequencing was performed on an Illumina GAIIx sequencer (Illumina SCS 2.8) with real-time image analysis and base calling (Illumina RTA 2.8). 
+Eland II (from Illumina pipeline version 1.6) was used with the default parameters to perform sequence alignment to the 300-bp synthetic DNA construct.
 More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}.
+As shown in Supplementary Table S1, each sample has approximately 1,000,000 35bp paired end reads.
 

### 2015-02-23 doc/bioinformatics2014/revision2/response_to_reviews.txt
Start response to reviewer comments about computational scalabilty.

@@
 We describe this and more detail as suggested in the protocol section.
+
+5. The authors greatly expose the comparison of sensibility, specificity, true positive and true negative rates. It seems that the caveat of the method is on memory consumption and
+processing time. It would be very informative to explicitly have a table with execution time
+and memory consumption of all utilized methods, preferably in a more realistic scenario,
+more specifically, a whole genome analysis.
+
+The memory consumption and computational time for the algorithm is an important consideration.
+We intentionally focused on the accuracy of the methodology here and so chose Metropolis-Hastings sampling as our inference algorithm.
+Metropolis-Hastings is known to converge to sampling from the true posterior distribution in the limit.
+As such, presently the algorithm is not as computationally fast as it could be.
+To improve the speed of the algorithm, we have implemented a multi-threaded version to take advantage of a multi-core computational configuration.
+We provide computational speed results in Section X.X and discuss avenues for improvement that we are currently pursuing.

### 2015-02-25 doc/bioinformatics2014/revision2/response_to_reviews.txt
Updated narrative for PAXIP1 region.

@@
 We ran RVD2 on the current NCBI start/stop codons (http://www.ncbi.nlm.nih.gov/gene/22976) to give the expanded region of chr7:154,735,400 - 154,794,682.
-This expanded region yields much better results.
-In particular, using the shortened region, we identified SNPs at XXX, XXX, XXX, and XXX.
-An inspection of Fig4 shows that these loci have very similar point estimates of the allele frequency in the case and control.
+This expanded region yields additional mutations.
+In particular, using the expanded region, we identified 5 additional germline variants and 2 additional somatic variants.
+In expanding the region of interest the previously identified germline variants at chr7:154,766,700 and chr7:154,781,769 are not called.
+A close inspection shows that these loci have very similar point estimates of the allele frequency in the case and control in the shortened region of interest.
 In the analysis with the expanded region, RVD2 estimates a lower across-locus precision parameter (M0) and thus does not call these loci.
-However, it does call 2 mutations in the full-length PAXIP1 region of interest.
+The control M0 parameter is 5.7 in the previous analysis and 4.2 in the 59kbp region.
 

### 2015-02-25 doc/bioinformatics2014/revision2/document.tex
Updated narrative for PAXIP1 region.

@@
 \end{figure*}
+% figure 5 generated from 2014-03-24_germline_and_somatic_test_correction_no_subclassification
 
@@
     \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.18\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
-    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
+    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682.
 The positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%
@@
 
-RVD2 identified twelve variants in the 44kbp PAXIP1 gene from chr7:154738059 to chr7:154782774.
-There were eight germline variants and eight somatic mutations.
+% figure 6 generated from 2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset
+
+RVD2 identified twelve variants in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682.
+There were 11 germline variants and 10 somatic mutations.
 %Especially, two heterozygous germline variants had a loss-of-heterozygosity (LOH) in the tumor sample.
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. 
-Positions chr7:154743899C\textgreater T, chr7:154749704G\textgreater A, chr7:154753635T\textgreater C, chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154766700C\textgreater A, chr7:154780960C\textgreater T, and chr7:154781769G\textgreater T were called germline mutations.
-Positions chr7:154749704G\textgreater G, chr7:154753635T\textgreater C, \\ chr7:154754371T\textgreater C, chr7:154758813G\textgreater A, chr7:154760439A\textgreater C, chr7:154766732T\textgreater G,  chr7:154766832A\textgreater C, chr7:154777118A\textgreater C were identified as significantly different in tumor and normal sample MAF and called somatic mutation.
-Positions chr7:154754371 and chr7:154758813 appear to be loss-of-heterozygosity events.
+Positions 
+chr7:154,743,899C\textgreater T, 
+chr7:154,749,704G\textgreater A, 
+chr7:154,753,635T\textgreater C, 
+chr7:154,754,371T\textgreater C, 
+chr7:154,758,813G\textgreater A, 
+chr7:154,780,960C\textgreater T, 
+chr7:154,784,369A\textgreater G, 
+chr7:154,787,357C\textgreater T, 
+chr7:154,787,367C\textgreater T, 
+chr7:154,787,612A\textgreater G, and 
+chr7:154,788,798A\textgreater G 
+were called germline mutations. 
+%
+Positions 
+chr7:154,749,704G\textgreater A, 
+chr7:154,753,635T\textgreater C, 
+chr7:154,754,371T\textgreater C, 
+chr7:154,758,813G\textgreater A, 
+chr7:154,760,439A\textgreater C, 
+chr7:154,766,732T\textgreater G,  
+chr7:154,766,832A\textgreater C,
+chr7:154,777,118A\textgreater C,
+chr7:154,787,612A\textgreater C,
+chr7:154,788,798A\textgreater C, and
+were identified as significantly different in tumor and normal sample and called somatic mutations.
+% 
+Positions chr7:154,754,371 and chr7:154,758,813 appear to be loss-of-heterozygosity events. 
 Some of these mutations are also found to be common population SNPs according to dbSNPv138. 
-The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}.
-The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1.
+The corresponding identities are shown in the Figure~\ref{fig:brca_MAF}. The read depth distribution for positions called by RVD2 are provided in Supplementary Table 1. 
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

### 2015-03-09 doc/bioinformatics2014/revision2/document.tex
Updated PAXIP1 results

@@
 \vspace{-10pt}
-\includegraphics[width=0.68\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
-\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 44kbp PAXIP1 gene.
+\includegraphics[width=0.9\textwidth]{pdf_figs/HCC1187_MuBarPlot.pdf}
+\caption{Estimated minor allele fraction for germline and somatic mutations called by RVD2 in the 59kbp PAXIP1 gene.
 Blue diamonds ($\diamond$) indicates germline mutations, where $\mu^{\text{control}}$ is significantly different from the reference sequence.
@@
 The  vertical lines represent 95\% credible interval around posterior mean MAF.
-Five positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
+Ten positions are common population SNPs according to dbSNPv138, and the identities are shown below the positions.}
 \vspace{-10pt}
@@
     \centering
-    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.7\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
+    \subfloat[\label{fig:brca_dot:a}]{{\includegraphics[width=0.8\textwidth]{pdf_figs/HCC1187_dotplot.pdf} }}%
     %
-    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.18\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
-    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 44kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682.
+    \subfloat[\label{fig:brca_dot:b}]{{\includegraphics[width=0.21\textwidth]{pdf_figs/position_lookup_chart.pdf} }}%
+    \caption{(a) Positions called by VarScan2-somatic, muTect, RVD2 and Strelka in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682.
 The positions are sorted by index (correspondance to genomic positions shown in Supplementary Table 1) (b) Read counts for each base for positions called by RVD2 and muTect from raw pileup data.}%

### 2015-03-09 doc/bioinformatics2014/revision2/response_to_reviews.txt
Added response to reviewer 2 comment 3.

@@
 We provide computational speed results in Section X.X and discuss avenues for improvement that we are currently pursuing.
+
+Reviewer 2:
+
+3. On figure 6 panel A; at position index 35, 36, 37 and position 50; please check that the symbols shown for miscall versus mutant calls are as you intend them.. looks like two are reversed.
+
+We are unable to identify what symbols are reversed.
+The RVD2 germline and somatic calls in Figure 6 do correspond to the same calls in Figure 5.
+Figure 5 and Figure 6 do use different symbols to denote their data, but the data is consistent.

### 2015-03-09 doc/bioinformatics2014/revision2/response_to_reviews.txt
Respond to reviewer 2 comment 2 regarding -C50 option and correct text in manuscript.

@@
 
+2. on preprocessing methods (4.1); please check what you mean by the C50 option of BWA.. That should be something else.
+
+Thank you for catching this important detail in the description of our methodology. We 
+used the aligned bam files described previously, and ran samtools mmpileup with the -C50 flag to filter for high quality reads.
+The -C50 options is recommended in the samtools manual: http://samtools.sourceforge.net/mpileup.shtml.
+We have corrected the manuscript to reflect this.
+
 3. On figure 6 panel A; at position index 35, 36, 37 and position 50; please check that the symbols shown for miscall versus mutant calls are as you intend them.. looks like two are reversed.

### 2015-03-09 doc/bioinformatics2014/revision2/document.tex
Respond to reviewer 2 comment 2 regarding -C50 option and correct text in manuscript.

@@
 \subsubsection*{Pre-processing Methods.}
- We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.
+The reads were aligned with Eland as described previously.
+We then ran samtools mpileup with the -C50 option to filter for high mapping quality reads.
 To simulate lower coverage data while retaining the error structure of real NGS data, BAM files for the synthetic DNA data were downsampled $10\times$, $100\times$, $1,000\times$, and $10,000\times$ using Picard v1.96.

### 2015-03-09 doc/bioinformatics2014/revision2/response_to_reviews.txt
Responded to reviewer 2 comment 1 regarding -tau.

@@
 
+1. page 4 line 22; check the signs of the threshold boundary conditions in your hypothesis tests.. looks like one should be less than minus Tau..
+
+Thank you for this observant correction.
+Indeed, the equation on page 4 line 22 should be minus tau.
+We have corrected the equation in the main text as well as the same equation in the supplementary information in the caption for Figure S6.
+
 2. on preprocessing methods (4.1); please check what you mean by the C50 option of BWA.. That should be something else.

### 2015-03-09 doc/bioinformatics2014/revision2/document.tex
Responded to reviewer 2 comment 1 regarding -tau.

@@
 homozygous somatic mutation).
-The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < \tau ) > 1-\alpha$.
+The two hypothesis tests are then $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} > \tau ) > 1-\alpha$ and $\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}} < -\tau ) > 1-\alpha$.
 We typically set the threshold $ \tau $ to zero.

### 2015-03-09 doc/bioinformatics2014/revision2/supplementary_info.tex
Added pdf figures in docs folder and removed *.pdf from .gitignore.
Responded to reviewer 2 comment 1 regarding -tau.

--- 2014-05-02 doc/bioinformatics2014/revision0/supplementary_info.tex
@@
 There is little variability across positions indicating that the replication variance does not change greatly across position.
-Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
+
+%Furthermore, we see that $M_j$ does not change with read depth (except where the depth goes to zero) indicating that $M_j$ because $M_j$ is capturing a different process than the read depth.
 
@@
 \includegraphics[width=1\textwidth]{pdf_figs/HCC1187_histogram_neg.pdf}
-\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
+\caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly lower than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j< -\tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
 \label{fig:hist_neg}

### 2015-03-09 doc/bioinformatics2014/revision2/response_to_reviews.txt
Responded to reviewer 1 comment 3. Removed list of PAXIP1 mutation calls from narrative.

@@
 
-We have updated our results to reflect the accurate start/stop locations for PAXIP1.
-
 3.
@@
 Despite the limitations, the analysis is complete, but it would be more informative to apply RVD2 to publicly available (and validated) DNAseq from primary tumors.
+
+Thank you for pointing out the evidence that HCC1887 is tetraploid for chromosome 7.
+We have incorporated this evidence into our discussion of the data with the reference your provided.
+We agree that it would be informaive to apply RVD2 to publicly available and validated DNAseq from primary tumors.
+We have searched extensively for such a data set and we have experienced three limitations.
+First, RVD2 requires sequencing data from a control sample.
+The control sample does not need to be matched, but should be of the same sequence as the case.
+This control sample allows RVD2 to measure the locus-specific error rate and call significant deviations from that error rate in the case.
+This requirement limits the pool of possible data sets.
+Second, we require a reasonable read depth (as shown by the synthetic DNA experiments) to identify rare variants.
+Many DNAseq data sets have a read depth of 10-15x because the investigators are seeking to identify clear variants.
+Few data sets have a read depth higher than 15x.
+Finally, we wanted a data set for which the originators provide a set of validated variants.
+Again, this requirement limits the number of available data sets.
+Taken together, these requirements are met by the HCC1187 data set.
+We expect that with the increased interest in tumor heterogeneity, investigators will be generating more data sets that meet these criteria.
+It will be interesting to explore those data sets with RVD2.
 

### 2015-03-09 doc/bioinformatics2014/revision2/document.tex
Responded to reviewer 1 comment 3. Removed list of PAXIP1 mutation calls from narrative.

@@
 Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. 
-Positions 
-chr7:154,743,899C\textgreater T, 
-chr7:154,749,704G\textgreater A, 
-chr7:154,753,635T\textgreater C, 
-chr7:154,754,371T\textgreater C, 
-chr7:154,758,813G\textgreater A, 
-chr7:154,780,960C\textgreater T, 
-chr7:154,784,369A\textgreater G, 
-chr7:154,787,357C\textgreater T, 
-chr7:154,787,367C\textgreater T, 
-chr7:154,787,612A\textgreater G, and 
-chr7:154,788,798A\textgreater G 
-were called germline mutations. 
 %
-Positions 
-chr7:154,749,704G\textgreater A, 
-chr7:154,753,635T\textgreater C, 
-chr7:154,754,371T\textgreater C, 
-chr7:154,758,813G\textgreater A, 
-chr7:154,760,439A\textgreater C, 
-chr7:154,766,732T\textgreater G,  
-chr7:154,766,832A\textgreater C,
-chr7:154,777,118A\textgreater C,
-chr7:154,787,612A\textgreater C,
-chr7:154,788,798A\textgreater C, and
-were identified as significantly different in tumor and normal sample and called somatic mutations.
+%Positions 
+%chr7:154,743,899C\textgreater T, 
+%chr7:154,749,704G\textgreater A, 
+%chr7:154,753,635T\textgreater C, 
+%chr7:154,754,371T\textgreater C, 
+%chr7:154,758,813G\textgreater A, 
+%chr7:154,780,960C\textgreater T, 
+%chr7:154,784,369A\textgreater G, 
+%chr7:154,787,357C\textgreater T, 
+%chr7:154,787,367C\textgreater T, 
+%chr7:154,787,612A\textgreater G, and 
+%chr7:154,788,798A\textgreater G 
+%were called germline mutations. 
+%%
+%Positions 
+%chr7:154,749,704G\textgreater A, 
+%chr7:154,753,635T\textgreater C, 
+%chr7:154,754,371T\textgreater C, 
+%chr7:154,758,813G\textgreater A, 
+%chr7:154,760,439A\textgreater C, 
+%chr7:154,766,732T\textgreater G,  
+%chr7:154,766,832A\textgreater C,
+%chr7:154,777,118A\textgreater C,
+%chr7:154,787,612A\textgreater C,
+%chr7:154,788,798A\textgreater C, and
+%were identified as significantly different in tumor and normal sample and called somatic mutations.
 % 
-Positions chr7:154,754,371 and chr7:154,758,813 appear to be loss-of-heterozygosity events. 
+Interestingly, positions chr7:154,754,371 and chr7:154,758,813 appear to be loss-of-heterozygosity events. 
 Some of these mutations are also found to be common population SNPs according to dbSNPv138. 
@@
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
+
+Interestingly, karyotyping indicates that chromosome 7 in HCC1187 is tetraploid (http://www.path.cam.ac.uk/~pawefish/BreastCellLineDescriptions/HCC1187.html).
 

### 2015-03-14 doc/bioinformatics2014/revision2/document.tex
Completed careful review #1 of main manuscript.

@@
 \title[RVD2]{RVD2: An ultra-sensitive variant detection model for low-depth heterogeneous next-generation sequencing data}
-\author[He \textit{et~al}]{Yuting He\,$^{1}$, Patrick Flaherty\,$^{1}$\footnote{Corresponding Author: pjflaherty@wpi.edu}}
-\address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA}
+\author[He \textit{et~al}]{Yuting He\,$^{1}$, Fan Zhang\,$^{1}$, Patrick Flaherty\,$^{1,2}$\footnote{Corresponding Author: pjflaherty@wpi.edu}}
+\address{$^{1}$Department of Biomedical Engineering, Worcester Polytechnic Institute, Worcester, MA, USA\\
+$^{2}$Department of Mathematics and Statistics, University of Massachusetts, Amherst, MA, USA}
 
@@
 \section{Motivation:}Next-generation sequencing technology is increasingly being used for clinical diagnostic tests.
-Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
+Clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations.
 Therefore, a variant calling algorithm for calling low-frequency polymorphisms in heterogeneous samples is needed.
@@
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We apply our model and identify twelve mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+We apply our model and identify fifteen mutated loci in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
 
@@
 As a result, NGS is increasingly being used as a general platform for research assays for methylation state~\citep{Laird:2010ab}, DNA mutations~\citep{Consortium:2013co}, copy number variation~\citep{Alkan:2009cr}, promoter occupancy~\citep{Ouyang:2009hc} and others~\citep{Rivera:2013ee}.
-NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbial analysis~\citep{Consortium:2013iz}.
+NGS diagnostics are being translated to clinical applications including noninvasive fetal diagnostics~\citep{Kitzman:2012hea}, infectious disease diagnostics~\citep{Capobianchi:2012em}, cancer diagnostics~\citep{Navin:2010gu}, and human microbiome analysis~\citep{Consortium:2013iz}.
 
@@
 
-Recently, some have developed algorithms to call low-frequency or rare variants in heterogeneous samples.~\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP.
+Recently, some have developed algorithms to call low-frequency or rare variants in heterogeneous samples.
+\citet{yau2010statistical} developed a Bayesian framework which can model the normal DNA contamination and intra-tumor heterogeneity by parameterizing the normal genotype cell proportion at each SNP. 
 VarScan2 combines algorithmic heuristics to call genotypes in the tumor and normal sample pileup data and then applies a Fisher's exact test on the read count data to detect a significant difference in the genotype calls~\citep{Koboldt:2012cg}.
@@
 
-The model generative process given hyperparameters $ \mu_0, M_0$ and $ M $is as follows:
+The model generative process given hyperparameters $ \mu_0, M_0$ and $ M $ is as follows:
 
@@
 \end{align}
-where \textbf{B}(x,y) denotes the beta function.
+where \textbf{B}(\cdot,\cdot) denotes the beta function.
 %Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data,
@@
 First, RVD2 has a $ \text{Beta}(\mu_0, M_0) $ prior on local error rate $ \mu_j $, which captures the global across-position error rate.
-The prior distribution allows $ \mu_j $ to borrow information from adjacent positions and allows RVD2 to handle low read depths.
+The prior distribution allows $ \mu_j $ to share information across adjacent positions and allows RVD2 to handle low read depths.
 Second, RVD2 handles multiple replicates in case samples.
@@
 
-Instead, we have developed a Metropolis-within-Gibbs  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
+Instead, we have developed a Metropolis-within-Gibbs (MwG)  approximate inference algorithm shown in Algorithm~\ref{alg:metro_gibbs}.
 First, the hyperparameters are initialized using method-of-moments (MoM).
-Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting rejection sampling rule.
+Given those hyperparameter estimates, we sample from the marginal posterior distribution for $\mu_j$ given its Markov blanket using a Metropolis-Hasting (M--H) rejection sampling rule.
 Finally, we sample from the marginal posterior distribution for $\theta_{ji}$ given its Markov blanket.
@@
 $\hat{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$.
-The estimates for the position-level parameters are
+The estimates for the local parameters are
 $\hat{\mu}_j = \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}$
@@
 $\hat{M_j} = \frac{ \hat{\mu}_j (1 - \hat{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \hat{\theta}_{ji}^2 } -1$.
-The estimates for the genome-level parameters are
+The estimates for the global parameters are
 $\hat{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \hat{\mu}_j$
@@
 \begin{equation}
-	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
+	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}.
 \end{equation}
@@
 We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j = 0.1 \cdot \hat{\mu}_j\cdot (1-\hat{\mu}_j)$ if $\hat{\mu}_j \in (10^{-3},1-10^{-3})$ and $\sigma_j = 10^{-4}$ otherwise, where $\hat{\mu}_j$ is the MoM estimator of $\mu_j$.
-Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.
+Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples in single Gibbs step for each position.
 
@@
 \begin{equation}\label{eqn:bayes_test}
-	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}}  > \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{case}} - \tilde{\mu}_{jk}^{\text{control}} > \tau } > 1-\alpha,
+	\Pr( \mu_j^{\text{case}} - \mu_j^{\text{control}}  > \tau ) \approx \frac{1}{N_{\text{MwG}}} \sum_{k=1}^{N_{\text{MwG}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{case}} - \tilde{\mu}_{jk}^{\text{control}} > \tau } > 1-\alpha,
 \end{equation}
@@
 
-We use a one-sided posterior distribution test with a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{control} \geq \tau$ with high confidence,
+We use a one-sided posterior distribution test with a single control sample to identify germline mutations. We call a germline mutation if $ \mu_j^{\text{control}} \geq \tau$ with high confidence,
 
@@
 
-We tested RVD2 using synthetic DNA and data from a primary ductal carcinoma sample.
+We tested RVD2 using synthetic DNA and data from the HCC1187 primary ductal carcinoma sample.
 The inference algorithm parameters were set to yield $4,000$ Gibbs samples with a 20\% burn-in and $2\times$ thinning rate for a final total of $1,600$ samples.
@@
 
+We measured the computational time for RVD2 varying the number of Gibbs sampling steps and the median read depth for the 400bp synthetic data data set. 
+Briefly, on a 2.4GHz processor it took $\sim$13 min per 1000 Gibbs samples to fit the model. 
+The computational time is independent of the median read depth due to the model structure; the same performance was observed for a median read depth of 130 and 40,000. 
+As stated previously, due to the independence structure of the model, we are able to perform some of the sampling steps in parallel greatly decreasing the computational time.
+The memory requirement is roughly the size of the gene sequence times the number of Gibbs samples.
+Complete timing results are shown in Supplementary Section 8.
+
 \subsection{Empirical performance compared with other algorithms}\label{sec:comparison}
@@
 For targeted cancer sequencing, the median read depth is higher at 100$\times$ to 1,000$\times$.
-And, for microbial or viral sequencing for rare variants, the median read depth is even higher at 1,000$\times$ to 100,000$\times$.
+For microbial or viral sequencing for rare variants, the median read depth is even higher at 1,000$\times$ to 100,000$\times$.
 
@@
 
-RVD2 identified twelve variants in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682.
+RVD2 identified fifteen variant locations in the 59kbp PAXIP1 gene from chr7:154,735,400 to chr7:154,794,682.
 There were 11 germline variants and 10 somatic mutations.
@@
 %According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
-
-Interestingly, karyotyping indicates that chromosome 7 in HCC1187 is tetraploid (http://www.path.cam.ac.uk/~pawefish/BreastCellLineDescriptions/HCC1187.html).
+Karyotyping indicates that chromosome 7 in HCC1187 is tetraploid~\citep{tetraploid}.
 
@@
 The mutations called by RVD2 and muTect are the most consistent among all the techniques. 
-RVD2 detected twelve germline and somatic mutations, while muTect reported eleven; ten were called by both.
-In the disagreements, RVD2 did not call position 50 while muTect did not call position 37 and 58.
-Referring to the depth distribution shown in Figure~\ref{fig:brca_dot:b}, it can be seen that position 37 and 58 are more likely mutated while position 50 is less likely mutated.
+RVD2 detected 15 germline mutations and 10 somatic mutations, while muTect reported 16 mutations; 11 were called by both. 
+In the disagreements, RVD2 did not call positions 1,39,54, and 84 while muTect did not call positions 41,75, and 77. 
+Referring to the depth distribution shown in Figure~\ref{fig:brca_dot:b}, it can be seen that positions 41, 75 and 77 are more likely mutated while positions 1, 39, 54, and 84 are less likely mutated.
 
-Strelka was the least sensitive algorithm among all the algorithms.
-According to the technical report, Strelka identified position 22 (chr7:154760439) as variant, but did not call any other variants.
+Strelka was the least sensitive algorithm among all the algorithms. According to the technical report, Strelka identified position 26 (chr7:154760439) as variant, but did not call any other variants. 
 In particular Strelka missed the two LOH events called by RVD2. 
-VarScan2-somatic called most positions among all algorithms, sixty positions as shown in Figure~\ref{fig:brca_dot:a}.
-VarScan2-somatic  detected all the positions called by RVD2 except position 35, which turns out to be a very likely mutation given the depth distribution in Figure~\ref{fig:brca_dot:b}.
-VarScan2-somatic reported fifty positions which were not called by any other three algorithms.
+VarScan2-somatic called most positions among all algorithms, 84 positions as shown in Figure~\ref{fig:brca_dot:a}. 
+VarScan2-somatic  detected all the positions called by RVD2 except position 39, which turns out to be a very likely mutation given the depth distribution in Figure~\ref{fig:brca_dot:b}. 
+VarScan2-somatic reported 65 positions which were not called by any other three algorithms. 
 The read depth in Supplementary Table 1 suggestions that these positions are very likely to be false positives. 
-The fact that VarScan2-somatic can be over-sensitive has appeared in synthetic dataset analysis.
 As shown in Figure~\ref{tbl:comparison_fdr}, the false discovery rate for VarScan2-somatic at read depth 53 MAF level 1.0\% is as high as 1.00.
@@
 % To compare the performance of RVD2 with other algorithms, we used Strelka to identify mutationin the same sample. Strelka found three positions: chr7:154754371, chr7:154758813 and chr7:154760439.
+
 \vspace{-5pt}
@@
 RVD2 is able to use multiple cores in parallel, which can significantly improve time efficiency. 
-On a computational node with 64 2.34GHz processor and 256Gb of RAM, RVD2 takes around 10 minutes to analyze the one 400bp long synthetic data set.
-The memory requirement is roughly the size of the gene sequence times the number of Gibbs samples.
-We are currently developing a variational method to estimate RVD2, aiming at decreasing computational time.
+In future studies, we plan to reduce the computational cost by using more sophisticated MCMC sampling methods or deterministic approximation methods such as variational EM or stochastic variational EM.
 

### 2015-03-14 doc/bioinformatics2014/revision2/supplementary_info.tex
Update RVD2 supplementary information. Added RVD2 oo3 document.
Completed careful review #1 of main manuscript.

@@
 
-\section{histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
+\section{Histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
 \begin{figure}[H]
 \begin{center}
-\includegraphics[width=1\textwidth]{pdf_figs/HCC1187_histogram_pos.pdf}
+\includegraphics[height=1\textheight]{pdf_figs/HCC1187_histogram_pos.pdf}
 \caption{Histogram of $ \hat{\mu}_j $ for positions where $\mu^{\text{case}}$ is significantly higher than $ \mu^{\text{control}}$, namely Pr$ (\Delta \hat{\mu}_j> \tau) > 1- \alpha $, where $ \tau=0, \alpha=0.05 $.}
@@
 
+\section{Timing Results}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width=0.7\textwidth]{pdf_figs/timing.pdf}
+\caption{Computational timing measurements for RVD2 on 400bp synthetic DNA data.}
+\end{center}
+\end{figure}
+
 \bibliographystyle{natbib}

### 2015-03-16 doc/bioinformatics2014/revision2/document.tex
Replace/add folder names in comments for each figure in main text and supplement with updated folder.
Submitted revision of bioinformatics RVD2 paper

@@
 We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency.
-We apply our model and identify fifteen mutated loci in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.
+We apply our model and identify fifteen mutated loci in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are likely loss-of-heterozygosity events.
 
@@
 \begin{equation}\label{eqn:germ_test}
-	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MH}}} \sum_{k=1}^{N_{\text{MH}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha.
+	\Pr(\mu_j^{\text{control}}  \geq \tau ) \approx \frac{1}{N_{\text{MwG}}} \sum_{k=1}^{N_{\text{MwG}}} \mathbbm{1}_{\tilde{\mu}_{jk}^{\text{control}} \geq \tau } > 1-\alpha.
 \end{equation}
@@
 \begin{equation}\label{eqn:fisher_combined}
-	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
+	\chi_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
 \end{equation}
@@
 The computational time is independent of the median read depth due to the model structure; the same performance was observed for a median read depth of 130 and 40,000. 
-As stated previously, due to the independence structure of the model, we are able to perform some of the sampling steps in parallel greatly decreasing the computational time.
+As stated previously, due to the independence structure of the model, we are able to perform the sampling step for each location in parallel greatly decreasing the computational time.
 The memory requirement is roughly the size of the gene sequence times the number of Gibbs samples.
-Complete timing results are shown in Supplementary Section 8.
+Complete timing results without parallelization are shown in Supplementary Section 8.
 
@@
 We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data set by the false discovery rate and sensitivity/specificity. 
-Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2-somatic, Strelka and muTect are have good performance to call variants in heterogeneous samples. 
+Among these algorithms, Samtools and GATK are optimized for homogeneous samples, while RVD, VarScan2-somatic, Strelka and muTect have good performance to call variants in heterogeneous samples. 
 In research applications, the false discovery rate is a more relevant performance metric because the aim is generally to identify interesting variants for followup. 
@@
 % rvd2 data are from folder '2013-11-12_one_replicate_optT_with_chi','2013-11-12_one_replicate_T_zero_with_chi2','2013-11-15_six_replicates_optT_with_chi2','2013-11-15_six_replicates_T_zero_with_chi2'
-% from folder 'rvd2\results\2013-09-19_operating_characteristics'
+% For revisions, Figure 3 and Figure 4 are generated from:
+% from folder 'fzhang\Research\rvd\results\2015-03-09_run_RVD_replicates_thresh_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-03-03_run_Samtools_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-03-03_run_GATK_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-03-03_run_MuTect_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-03-03_run_Strelka_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-03-09_run_VarScan2_somatic_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-03-03_run_VarScan2_SS_FDR_downsample_depth_100'
+% from folder 'fzhang\Research\rvd\results\2015-02-23_run_rvd2_SS_FDR_downsample_depth_100'
 
@@
 \end{figure*}
-% figure 5 generated from 2014-03-24_germline_and_somatic_test_correction_no_subclassification
+% figure 5 generated from '2014-03-24_germline_and_somatic_test_correction_no_subclassification'
+% figure 5 generated from 'fzhang\Research\rvd\results\2015-02-25_MubarPlot_expanded_PAXIP1' for revisions.
 
@@
 
-% figure 6 generated from 2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset
+% figure 6 generated from '2014-03-12_dot_plot_mutation_detection_result_HCC1187_dataset'
+% figure 6 generated from 'fzhang\Research\rvd\results\2015-02-25_dotPlot_expanded_PAXIP1' for revisions.
 
@@
 \section*{Acknowledgements}
-P.F. and Y.H. were supported by seed funding from Worcester Polytechnic Institute. Y.H. was supported by PhRMA Foundation Informatics Grant 2013080079.
+P.F. was supported by seed funding from Worcester Polytechnic Institute. Y.H. and F.Z. were supported by PhRMA Foundation Informatics Grant 2013080079.
 

### 2015-03-16 doc/bioinformatics2014/revision2/supplementary_info.tex
Replace/add folder names in comments for each figure in main text and supplement with updated folder.
Submitted revision of bioinformatics RVD2 paper

@@
 
-\section{Histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 44kbp paxip1 gene from chr7:154738059 to chr7:154782774 in hcc1187 dataset}
+\section{Histogram of $ \hat{\mu}_j $ for somatic mutations called by rvd2 in the 59kbp PAXIP1 gene from chr7:154735400 to chr7:154794682 in HCC1187 dataset}
 \begin{figure}[H]
@@
 
+% Figure 5 and Figure 6 in supplementary are from folder '\fzhang\Research\rvd\results\2015-03-04_plot_posterior_mu_supp_PAXIP1'
+
 \bigskip
@@
 
+% Supplementary Table 1 is from folder '\fzhang\Research\rvd\results\2015-02-25_dotPlot_expanded_PAXIP1'
 \includepdf[pages={1-},scale=1]{pdf_figs/position_lookup_chart_all.pdf}
@@
 \section{Timing Results}
-
+% Figure 7 in supplementary is from folder '\fzhang\Research\rvd\results\2015-03-09_timing_rvd2_different_ngibbs'
 \begin{figure}[H]
